### 2016-01-26 inst/doc/futureheatwaves.Rmd
Add current vignette

---
title: "`futureheatwaves` package vignette"
author: "G. Brooke Anderson and Colin Eason"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo = FALSE, message = FALSE}
library(futureheatwaves)
```

## What the package does

The `futureheatwaves` package can process climate data files that you have stored locally to generate a list of heatwaves in each projection as well as characteristics of each heatwave (e.g., length, intensity, timing in the year). The package generates community-specific dataframes based on an input file listing each community and its latitude and longitude. Files with the heatwaves identified and characterized under each ensemble member of each climate model are output as files in a directory specified by the user.

## A basic example of using the package

For these functions to work correctly, you must have the climate projections saved in a specific structure locally on your computer. An example of a directory with the required set-up is given as the `cmip5` directory in the `inst` directory of this package.

You must have a directory that has within it separately subdirectories names "rcp85" and "historical" that contain, respectively, all files drawn from projections for future dates ("rcp85") and all for dates up to 2005 ("historical"). Within each of these subdirectories, you need a subdirectory for each climate model. Within each climate model directory, you need subdirectories for each ensemble member of the model (even if the model only has one ensemble member). Within each ensemble-member directory, you will need three comma-separated files: one file with the climate model output, with columns corresponding to climate grid point and rows corresponding to time, one file that gives the time values corresponding to each row in the climate model output file, and one file giving the latitude and longitude corresponding to the grid locations used in the columns of the climate model output file.

Throughout the directory, subdirectories should be named using names reflecting the climate model or ensemble members they contain. The output files will be written using the same type of directory structure and the same names, so this output will be easier to interpret and use if the directory uses meaningful names for subdirectories.

Once you have your files set up in this structure, define the following variables:

- `dataFolder`: Character string with pathway to directory that contains climate projections. 
- `citycsv`: Character string giving the filepath to a .csv file with latitude and longitude values for each city for which you would like to generate
heatwave projections.
- `coordinateFilenames`: Character string with name of the files containing the latitude and longitude coordinates corresponding to the columns of the time series data.
- `tasFilenames`: Character sting with name of files containing the time series data.
- `timeFilenames`: Character string with name of the files containing the date information corresponding to the rows of the time series data.

For example, I can specify `dataFolder` and `citycsv` as:

```{r}
dataFolder <- "inst/cmip5/" 
citycsv <- "inst/cities.csv"
```

Within the ensemble subdirectories in my files structure, all of the files with climate model data are named "tas_NorthAmerica_12mo.csv", all the files with the corresponding time for rows in that data are in files named "time_NorthAmerica_12mo.csv", and all the files with locations of the grid points are named "latitude_longitude_NorthAmerica_12mo.csv". Therefore, I would specify:

```{r}
coordinateFilenames <- "latitude_longitude_NorthAmerica_12mo.csv"
tasFilenames <- "tas_NorthAmerica_12mo.csv"
timeFilenames <- "time_NorthAmerica_12mo.csv"
```

You will also need to specify where the output files should be locally saved. Save this as the object `out`. For example, if I wanted to write the output heatwave files to a directory called "example_results" in the `tmp` subdirectory of my home directory, I could specify:

```{r}
out <- "~/tmp/example_results/"
```

These selections will be sufficient to run the main function of this package, `gen_hw_set`, using a basic example with default values for things like the heatwave definition, years for which to generate the heatwave datasets (default is 2061--2080), and years to use as a reference for the relative temperature characteristics measured for the heatwaves (default is ...).

Once you have specified these values, you can create files with future heatwaves and their characteristics using the following code:

```{r eval = FALSE}
gen_hw_set(out = out,
           dataFolder = dataFolder,
           citycsv = citycsv,
           coordinateFilenames = coordinateFilenames,
           tasFilenames = tasFilenames,
           timeFilenames = timeFilenames,
           RorCPP = 0)
```

You will first be asked:

```
Warning: This function will write new files to your computer in the  
 ~/tmp/results directory of your computer. 
 Do you want to continue (y / n): 
```

This warning reminds you that this function will create subdirectories and write out files to the directory you specified when defining `out`. If you agree that it is okay to write files and subdirectories to this directory, enter "y" at the prompt, and the function will continue running.

If your file structure is set up correctly, you should get output that looks like this:

```
Processing historical ensemble for bcc1 
Reading ---> r1i1p1 
Read operation complete 
Processing rcp 8.5 ensemble bcc1 
Reading ---> r1i1p1 
Read operation complete 
Creating heatwave dataframe ~~ City:  balt  ~~ City Number:  1  ~~ Cutoff:  84.632 
Creating heatwave dataframe ~~ City:  ny  ~~ City Number:  2  ~~ Cutoff:  78.77876 
Creating heatwave dataframe ~~ City:  nwk  ~~ City Number:  3  ~~ Cutoff:  78.77876 
Creating heatwave dataframe ~~ City:  phil  ~~ City Number:  4  ~~ Cutoff:  84.632 
Creating heatwave dataframe ~~ City:  prov  ~~ City Number:  5  ~~ Cutoff:  75.074 
Writing bcc1: r1i1p1
Writing accumulator 
All operations completed. Exiting.

  modelName length.histens. length.rcpens.
1      bcc1               1              1
```

The function provides status reports on its progress in generating the heatwave projections, as this process can take a while if you include many cities and / or many climate model projections. If you do not want to generate this output when running the function, choose the option `printWarning = FALSE` when calling `gen_hw_set`.

Once the function has completed running, you will have two new subdirectories in the directory you specified with the `out` variable. These will be called `Heatwaves` and `Thresholds`. The `Heatwaves` directory will contain dataframes with all identified heatwaves and their characteristics for each city, while the `Thresholds` folder will contain files with the temperature thresholds used for each city given. The `out` directory specified will also have a file called `hwModelInfo.csv` that contains some basic information about the climate models and the number of ensemble members included for each.

## Using custom options

## How it works

### Parsing the directory structure

After some preliminary checks, the function reads in the structure of the directory specified by `dataFolder` and uses the names of subdirectories to create a list of all of the climate models and their ensemble members within the directory.

The function that does this parsing, `acquireDirectoryStructure`, returns a list object outlining the file structure of the directory containing the climate projections.This list has an element for each climate model (e.g. ,"bcc1"). The first element within each of these elements is the name of the model. The second element within the first-level element gives the file paths for location grids, climate projections, and projection times for each ensemble run of the model.

For example:

```{r}
dataFolder <- system.file("cmip5", package = "futureheatwaves")
finalList <- acquireDirectoryStructure(dataFolder = dataFolder,
   coordinateFilenames = coordinateFilenames,
   tasFilenames = tasFilenames,
   timeFilenames = timeFilenames)

str(finalList[[1]][[1]])
str(finalList[[1]][[2]][1])
```

### 2016-01-26 inst/doc/futureheatwaves.R

@@ -1,0 +1,34 @@
+## ----echo = FALSE, message = FALSE---------------------------------------
+library(futureheatwaves)
+
+## ------------------------------------------------------------------------
+dataFolder <- "inst/cmip5/" 
+citycsv <- "inst/cities.csv"
+
+## ------------------------------------------------------------------------
+coordinateFilenames <- "latitude_longitude_NorthAmerica_12mo.csv"
+tasFilenames <- "tas_NorthAmerica_12mo.csv"
+timeFilenames <- "time_NorthAmerica_12mo.csv"
+
+## ------------------------------------------------------------------------
+out <- "~/tmp/example_results/"
+
+## ----eval = FALSE--------------------------------------------------------
+#  gen_hw_set(out = out,
+#             dataFolder = dataFolder,
+#             citycsv = citycsv,
+#             coordinateFilenames = coordinateFilenames,
+#             tasFilenames = tasFilenames,
+#             timeFilenames = timeFilenames,
+#             RorCPP = 0)
+
+## ------------------------------------------------------------------------
+dataFolder <- system.file("cmip5", package = "futureheatwaves")
+finalList <- acquireDirectoryStructure(dataFolder = dataFolder,
+   coordinateFilenames = coordinateFilenames,
+   tasFilenames = tasFilenames,
+   timeFilenames = timeFilenames)
+
+str(finalList[[1]][[1]])
+str(finalList[[1]][[2]][1])
+

### 2016-01-26 R/preparation.R
Fix example in `acquireDirectoryStructure`

@@ -39,8 +39,8 @@
 #'    tasFilenames = tasFilenames,
 #'    timeFilenames = timeFilenames)
 #'
-#' str(finalList[[1]][[1]])
-#' str(finalList[[1]][[2]][1])
+#' finalList[[1]][[1]]
+#' finalList[[1]][[2]][1]
 #'
 #' @export
 #'

### 2016-01-26 R/preparation.R
Fix example in documentation for `buildStructureModels`

@@ -93,7 +93,13 @@ acquireDirectoryStructure <- function(dataFolder, coordinateFilenames,
         return(finalList)
 }
 
-#' Generate list of file structure from model, experiment, and path
+#' Generate list of file structure
+#'
+#' This function takes input from \code{acquireDirectoryStructure} and
+#'    uses it to generate a list object with the projection directory
+#'    file structure. This parsed file structure is later used to lead other
+#'    code through all climate models and ensemble members in the input
+#'    projection directory.
 #'
 #' @param model Character string of climate model name
 #' @param experiments Character vector of the experiment(s) of interest.
@@ -104,18 +110,33 @@ acquireDirectoryStructure <- function(dataFolder, coordinateFilenames,
 #' @param dataPath Character string of the file path to the directory
 #'    containing the climate projections. Must include the final `/`.
 #'
-#' @return A list of length 3. The first element is the name of the model whose structure was being built.
-#' The second element is the historical experiment hierarchy. The third element is the rcp85 experiment hierarchy. The
-#' second and third elements are return values of buildStructureExperiments.
+#' @return A list of length 3. The first element is the name of the model
+#'    whose structure was being built. The second element is the historical
+#'    experiment hierarchy. The third element is the hierarchy of the future
+#'    projection directory for the model. The second and third elements are
+#'    return values of \code{buildStructureExperiments}.
 #'
 #' @examples
 #' model <- "bcc1"
-#' experiments <- c("rcp85")
+#' experiments <- c("historical", "rcp85")
 #' all <- c("rcp85/bcc1/r1i1p1/latitude_longitude_NorthAmerica_12mo.csv",
 #'          "rcp85/bcc1/r1i1p1/tas_NorthAmerica_12mo.csv",
 #'          "rcp85/bcc1/r1i1p1/time_NorthAmerica_12mo.csv")
-#' dataPath <- "~/Downloads/sample/cmip5/"
-#' buildStructureModels(model, experiments, all, dataPath)
+#'
+#' dataFolder <- system.file("cmip5", package = "futureheatwaves")
+#' dataFolder <- paste0(dataFolder, "/")
+#'
+#' coordinateFilenames <- "latitude_longitude_NorthAmerica_12mo.csv"
+#' tasFilenames <- "tas_NorthAmerica_12mo.csv"
+#' timeFilenames <- "time_NorthAmerica_12mo.csv"
+#'
+#' buildStructureModels(model = model,
+#'                      experiments = experiments,
+#'                      all = all,
[…]

### 2016-01-26 R/preparation.R
Get rid of `all` in `preparation.R`. Closes #51.

@@ -81,13 +81,12 @@ acquireDirectoryStructure <- function(dataFolder, coordinateFilenames,
                                            timeFilenames)) %>%
                 dplyr::select(exp, model, ens, type)
 
-        all <- apply(df_all, 1, paste, collapse = "/")
         models <- as.character(unique(df_all$model))
         experiments <- as.character(unique(df_all$exp))
 
         # Generate the nested lists that will be used for the processing step
         # Structure: model -> experiment -> ensemble
-        finalList <- lapply(models, buildStructureModels, experiments, all,
+        finalList <- lapply(models, buildStructureModels, experiments,
                             dataFolder, coordinateFilenames, tasFilenames,
                             timeFilenames)
         return(finalList)
@@ -101,14 +100,21 @@ acquireDirectoryStructure <- function(dataFolder, coordinateFilenames,
 #'    code through all climate models and ensemble members in the input
 #'    projection directory.
 #'
-#' @param model Character string of climate model name
+#' @param model Character string of climate model name (e.g., "bcc1"). This
+#'    name is typically generated for use in this function from the subdirectory
+#'    names for the climate model within the directory of projection data
+#'    specified by the user in \code{gen_hw_set}.
 #' @param experiments Character vector of the experiment(s) of interest.
 #'    Possible variables are "historical", "rcp85", or both.
-#' @param all Character vectors with the relative pathnames of all
-#'    climate projection files of interest from the directory
-#'    specified in dataPath
 #' @param dataPath Character string of the file path to the directory
 #'    containing the climate projections. Must include the final `/`.
+#' @param coordinateFilenames Character string with name of the files
+#'    containing the latitude and longitude coordinates
+#'    corresponding to the columns of the time series data.
+#' @param tasFilenames Character sting with name of files containing the time
+#'    series data.
+#' @param timeFilenames Character string with name of the files containing the
+#'    date information corresponding to the rows of the time series data.
 #'
 #' @return A list of length 3. The first element is the name of the model
 #'    whose structure was being built. The second element is the historical
@@ -119,9 +125,6 @@ acquireDirectoryStructure <- function(dataFolder, coordinateFilenames,
 #' @examples
 #' model <- "bcc1"
 #' experiments <- c("historical", "rcp85")
[…]

### 2016-01-26 R/main.R
Remove a random copy of `models` to `gar` in `gen_hw_set`
@ColinAE: I'm couldn't find anything this `gar` copy of `models` was being used for. I think it might have just been a relic from some de-bugging.

@@ -99,7 +99,6 @@ gen_hw_set <- function(out,
         # Put the directories into nested list form
         models <- acquireDirectoryStructure(dataFolder, coordinateFilenames,
                                             tasFilenames, timeFilenames)
-        gar <- models
 
         # Read the cities data file
         cities <- read.csv(citycsv)

### 2016-01-26 R/main.R
Edit `createAccumulators` documentation

@@ -339,21 +339,38 @@ checkCustomBounds <- function(boundList, expected_length = 4){
 
 #' Create accumulator closure
 #'
-#' This closure holds data structures that the user wishes to grow at various
-#'    points in the execution of the package. It exists to couple these
-#'    structures together in order to lower the number of parameters of this
-#'    nature the user would have to pass down the the program otherwise. It is
-#'    a closure instead of a list as a pre-emptive measure
+#' This closure holds, adds to, and returns data structures that the user
+#'    wishes to grow at various points in the execution of the package.
+#'
+#' As an example, when the generated
+#'    closure is used with the command "append location list", it will add
+#'    information on the communities and closest grid point locations based
+#'    on the climate model it has just completed analyzing to a growing
+#'    dataframe with this information for all climate models. After the function
+#'    run to generate the heatwave projections is completed, this closure can
+#'    be used with the command "return locations" to output the completed
+#'    dataframe of this location information.The closure can be used in a
+#'    similar manner to aggregate and then return meta-data on the models
+#'    analyzed based on their inclusion in the user-specified projections
+#'    directory.
+#'
+#' This function exists to couple these structures together in order to lower
+#'    the number of parameters of this nature the user would have to pass down
+#'    the the program otherwise. It is a closure instead of a list as a
+#'    pre-emptive measure.
 #'
 #' @return A closure that accepts commands to access and append new data onto
-#'    data structures as the program executes. The utility of this function may
-#'    be expanded by
+#'    data structures as the program executes. The closure created by this
+#'    function accepts two arguments: (1) the command and (2) an element to
+#'    be appended to the end of the data structure of the command. These two
+#'    arguments must be entered in this exact order. The first argument (the
+#'    command) can be one of the following options: "return model information",
+#'    "append model information", "return locations", and "append location
+#'    list". The second argument for the created closure should only be used
+#'    in conjunction with the two "append" commands for the closure.
 #'
-#' @note First argument: The command
-#' Second argument: An element to be appended to the end of the data structure
-#' of the command
-#' This function contains no error checking for the types of elements input
-#' into the data structures it contains
+#' @note This function contains no error checking for the types of elements
+#'    input into the data structures it contains.
 createAccumulators <- function(){
         modelInfoAccumulator <- data.frame(c(), c(), c())
         locationList <- list()

### 2016-01-26 R/preparation.R
Fix spacing

@@ -48,7 +48,6 @@
 acquireDirectoryStructure <- function(dataFolder, coordinateFilenames,
                                       tasFilenames, timeFilenames){
 
-
         # If `dataFolder` does not end in "/", add it.
         # (Only need to repeat here to use with in-package examples)
         split_dataFolder <- unlist(strsplit(dataFolder, split = ""))

### 2016-01-26 R/process.R
Edit documentation for `processModel`

@@ -2,13 +2,35 @@
 
 #' Process valid models
 #'
-#' Process models that are valid and store information about the models
-#' that have been written.
+#' This function takes any models that are valid, identifies and characterizes
+#'    heatwaves within projections for each of its ensemble members, writes
+#'    those dataframes of heatwave projections out to the user-specified output
+#'    directory, and stores information about the models that have been written
+#'    to the output directory using the closure, \code{accumulators}, created by
+#'    \code{createAccumulators}.
 #'
-#' @param model Character string of the climate model (e.g., "bcc1").
-#' @param global The 'global' variable list.
-#' @param custom The user-specified custom data list.
-#' @param accumulators Closures ... generated by createAccumulators
+#' @param model Character string of climate model name (e.g., "bcc1"). This
+#'    name is typically generated for use in this function from the subdirectory
+#'    names for the climate model within the directory of projection data
+#'    specified by the user in \code{gen_hw_set}.
+#' @param global An object name for the 'global' variable list. This
+#'    parameter is passed through from \code{gen_hw_set} and includes user
+#'    specifications for the path to the output directory, the path to the
+#'    input climate projections, the dataframe with community locations, the
+#'    format for the filenames of the coordinate, tas, and time files in the
+#'    ensemble-specific subdirectories of the projection directory input by
+#'    the user, and whether the user prefers to run the function using the R
+#'    or C++ version of the function to identify heatwaves.
+#' @param custom An object name for the 'custom` data list. This parameter is
+#'    passed through from \code{gen_hw_set} and includes user specifications
+#'    for, if specified as other than the default, an alternative function to
+#'    use to identify heatwaves, alternative upper and lower year boundaries
+#'    for the projection period of the heatwave datasets being generated,
+#'    and alternative upper and lower year boundaries for the reference period
+#'    to be used when measuring all heatwave characteristics related to
+#'    relative, rather than absolute, temperature.
+#' @param accumulators An object name for the closure generated by
+#'    \code{createAccumulators}.
 #'
 #' @return A list of rcp 8.5 ensembles for the given model
 processModel <- function(model, global, custom, accumulators){

### 2016-01-26 R/process.R
Edit documentation for `processHistorical`

@@ -65,15 +65,41 @@ processModel <- function(model, global, custom, accumulators){
         return(rcpEnsembles)
 }
 
-#' Calculate threshold temperatures from historical projections
+#' Calculate threshold temperatures
 #'
-#' Calculate the threshold temperatures for the model using the
-#' model's r1i1p1 historical ensemble and saves these to file.
+#' This function calculates the threshold temperatures required to identify
+#'    heatwaves in the climate projection data using each climate model's
+#'    first listed historical ensemble. These values are both used in later
+#'    functions to identify heatwaves in the projections and are also saved
+#'    to file through this function to have for future reference.
 #'
-#' @param model Character string of the climate model (e.g., "bcc1").
-#' @param global The 'global' variable list.
-#' @param custom The user-specified custom data list.
-#' @param reference [What is this parameter?]
+#' @param model Character string of climate model name (e.g., "bcc1"). This
+#'    name is typically generated for use in this function from the subdirectory
+#'    names for the climate model within the directory of projection data
+#'    specified by the user in \code{gen_hw_set}.
+#' @param global An object name for the 'global' variable list. This
+#'    parameter is passed through from \code{gen_hw_set} and includes user
+#'    specifications for the path to the output directory, the path to the
+#'    input climate projections, the dataframe with community locations, the
+#'    format for the filenames of the coordinate, tas, and time files in the
+#'    ensemble-specific subdirectories of the projection directory input by
+#'    the user, and whether the user prefers to run the function using the R
+#'    or C++ version of the function to identify heatwaves.
+#' @param custom An object name for the 'custom` data list. This parameter is
+#'    passed through from \code{gen_hw_set} and includes user specifications
+#'    for, if specified as other than the default, an alternative function to
+#'    use to identify heatwaves, alternative upper and lower year boundaries
+#'    for the projection period of the heatwave datasets being generated,
+#'    and alternative upper and lower year boundaries for the reference period
+#'    to be used when measuring all heatwave characteristics related to
+#'    relative, rather than absolute, temperature.
+#' @param reference FALSE, if the user has not specified custom reference
+#'    boundaries through the \code{referenceBoundaries} argument in
+#'    \code{gen_hw_set}, otherwise a vector with four elements
+#'    (earliest historical year, latest historical year,
+#'    earliest future projection year, latest future projection year)
+#'    giving the user-specified custom date boundaries for the
+#'    heatwave projections.
 #'
[…]

### 2016-01-30 R/process.R
Work more on `process` functions

@@ -52,13 +52,15 @@ processModel <- function(model, global, custom, accumulators){
         rcpens <- model[[3]]
         rcpEnsembles <- lapply(rcpens,
                                processRCP,
-                               modelName,
-                               createEnsembleWriter(modelName, global, custom),
-                               thresholds,
-                               global,
-                               custom,
-                               accumulators,
-                               reference)
+                               modelName = modelName,
+                               ensembleWriter = createEnsembleWriter(modelName,
+                                                                     global,
+                                                                     custom),
+                               thresholds = thresholds,
+                               global = global,
+                               custom = custom,
+                               accumulators = accumulators,
+                               reference = reference)
 
         # Add entry to the modelInfoAccumulator
         accumulators("append model information",
@@ -151,14 +153,47 @@ processHistorical <- function(model, global, custom, reference = FALSE){
 #' For each reference ensemble, create the heatwave dataframe and
 #' write it to file. Also store the locations  vector for each ensemble.
 #'
-#' @param ensemble
-#' @param modelName
-#' @param ensembleWriter
-#' @param thresholds
-#' @param global The 'global' variable list.
-#' @param custom The user-specified custom data list.
-#' @param accumulators Closures ... generated by createAccumulators
-#' @param reference [What is this parameter?]
+#' @param ensemble List with parsed information about the directory structure for
+#'    a specific ensemble member from the user-specified projections directory.
+#'    This list is a subset of the list generated by
+#'    \code{acquireDirectoryStructure}.
+#' @param modelName Character string of climate model name (e.g., "bcc1"). This
+#'    name is typically generated for use in this function from the subdirectory
+#'    names for the climate model within the directory of projection data
+#'    specified by the user in \code{gen_hw_set}.
+#' @param ensembleWriter A closure created by \code{createEnsembleWriter} to
+#'    write output from this process to a file within the output directory
+#'    specified by the user.
+#' @param thresholds A vector with the thresholds to use within each community
+#'    in that community's heatwave definition.These are typically automatically
+#'    determine during the run of the \code{gen_hw_set} function based on
+#'    historical climate projections within the community.
+#' @param global An object name for the 'global' variable list. This
+#'    parameter is passed through from \code{gen_hw_set} and includes user
+#'    specifications for the path to the output directory, the path to the
[…]

### 2016-01-30 R/main.R
Closes #49.

@@ -319,7 +319,7 @@ checkCustomBounds <- function(boundList, expected_length = 4){
 
     # Check if bounds are in the correct range
     if(histLow != FALSE){
-      if(histLow < 1981){
+      if(histLow < 1980){
         stop(paste("Custom boundaries for threshold calculation fall out",
                    "of acceptable range. Stopping."))
       }
@@ -333,14 +333,14 @@ checkCustomBounds <- function(boundList, expected_length = 4){
     }
 
     if(rcpLow != FALSE){
-      if(rcpLow < 2061){
+      if(rcpLow < 2006){
         stop(paste("Custom boundaries for threshold calculation fall out of",
                    "acceptable range. Stopping."))
       }
     }
 
     if(rcpHigh != FALSE){
-      if(rcpHigh > 2080){
+      if(rcpHigh > 2099){
         stop(paste("Custom boundaries for threshold calculation fall out of",
                    "acceptable range. Stopping."))
       }

### 2016-01-30 R/process.R
Work on `process` functions

@@ -202,8 +202,12 @@ processRCP <- function(ensemble, modelName, ensembleWriter, thresholds,
         cat("Processing rcp 8.5 ensemble", modelName, "\n")
 
         # Acquire desired characteristics of the RCP 8.5 ensembles
-        ensemble <- processEnsemble(ensemble, 'rcp', modelName, global,
-                                    custom, reference)
+        ensemble <- processEnsemble(ensemble = ensemble,
+                                    experiment = 'rcp',
+                                    modelName = modelName,
+                                    global = global,
+                                    custom = custom,
+                                    reference = reference)
 
         # Append locations vector to the locations vector accumulator
         accumulators("append location list", ensemble$locations)
@@ -336,7 +340,9 @@ processEnsemble <- function(ensemble, experiment, modelName, global,
         # CUSTOM BLOCK
         # Extract and store the reference period data
         if(reference != FALSE && length(reference) != 1){
-                rbounds <- getBounds(times, experiment, subCustom)
+                rbounds <- getBounds(times = times,
+                                     experiment = experiment,
+                                     custom = subCustom)
                 reference <- data.frame(tas[rbounds[1]:rbounds[2], locations])
                 reference <- apply(series, 1:2, function(element){
                         return((element * 9/5) - 459.67)

### 2016-01-30 R/process.R
Work on functions in `process.R`

@@ -213,7 +213,10 @@ processRCP <- function(ensemble, modelName, ensembleWriter, thresholds,
         accumulators("append location list", ensemble$locations)
 
         # Acquire the heatwave dataframes for every rcp 8.5 ensemble in this model
-        hwFrame <- formHwFrame(ensemble, thresholds, global, custom)
+        hwFrame <- formHwFrame(ensemble = ensemble,
+                               thresholds = thresholds,
+                               global = global,
+                               custom = custom)
 
         # Write every heatwave frame to a .csv
         ensembleWriter(hwFrame)

### 2016-01-30 R/processHeatwaves.R
Edit documentation for `formHwFrame`

@@ -3,16 +3,36 @@
 #' @param ensemble Character string that gives the absolute file path
 #'    for the directory with a particular ensemble member of a climate
 #'    model projection.
-#' @param thresholds [What is this?]
-#' @param global List of global parameters generated by listGlobal.
-#' @param custom List of custom parameters generated by listCustom.
+#' @param thresholds Vector of community-specific thresholds to use in
+#'    identifying heatwaves.
+#' @param global An object name for the 'global' variable list. This
+#'    parameter is passed through from \code{gen_hw_set} and includes user
+#'    specifications for the path to the output directory, the path to the
+#'    input climate projections, the dataframe with community locations, the
+#'    format for the filenames of the coordinate, tas, and time files in the
+#'    ensemble-specific subdirectories of the projection directory input by
+#'    the user, and whether the user prefers to run the function using the R
+#'    or C++ version of the function to identify heatwaves.
+#' @param custom An object name for the 'custom` data list. This parameter is
+#'    passed through from \code{gen_hw_set} and includes user specifications
+#'    for, if specified as other than the default, an alternative function to
+#'    use to identify heatwaves, alternative upper and lower year boundaries
+#'    for the projection period of the heatwave datasets being generated,
+#'    alternative upper and lower year boundaries for the reference period
+#'    to be used when measuring all heatwave characteristics related to
+#'    relative, rather than absolute, temperature, and the percentile
+#'    threshold to use when defining heatwaves.
 #'
-#' @return The combined dataframe of identified heatwaves for all cities of the given ensemble.
+#' @return The combined dataframe of identified heatwaves for all cities of
+#'    the given ensemble.
 formHwFrame <- function(ensemble, thresholds, global, custom){
         # Acquire list of heatwave dataframes for each city
-        hwDataframeList <- apply(data.frame(thresholds), 1, createCityProcessor(global), ensemble, custom)
+        hwDataframeList <- apply(data.frame(thresholds), 1,
+                                 createCityProcessor(global),
+                                 ensemble, custom)
 
-        # Combine the heatwave dataframes contained in hwDataframeList into a single dataframe
+        # Combine the heatwave dataframes contained in hwDataframeList into
+        # a single dataframe
         hwFrame <- consolidate(hwDataframeList)
 
         return(hwFrame)

### 2016-01-30 R/processHeatwaves.R
Work on documentation

@@ -23,8 +23,8 @@
 #'    relative, rather than absolute, temperature, and the percentile
 #'    threshold to use when defining heatwaves.
 #'
-#' @return The combined dataframe of identified heatwaves for all cities of
-#'    the given ensemble.
+#' @return The combined dataframe of identified and characterized heatwaves for
+#'    selected projection date range for all communities specified by the user.
 formHwFrame <- function(ensemble, thresholds, global, custom){
         # Acquire list of heatwave dataframes for each city
         hwDataframeList <- apply(data.frame(thresholds), 1,
@@ -40,8 +40,23 @@ formHwFrame <- function(ensemble, thresholds, global, custom){
 
 #' Identify and aggregate heatwaves for a particular city
 #'
-#' @param global List of global parameters generated by listGlobal.
-#' @param custom List of custom parameters generated by listCustom.
+#' @param global An object name for the 'global' variable list. This
+#'    parameter is passed through from \code{gen_hw_set} and includes user
+#'    specifications for the path to the output directory, the path to the
+#'    input climate projections, the dataframe with community locations, the
+#'    format for the filenames of the coordinate, tas, and time files in the
+#'    ensemble-specific subdirectories of the projection directory input by
+#'    the user, and whether the user prefers to run the function using the R
+#'    or C++ version of the function to identify heatwaves.
+#' @param custom An object name for the 'custom` data list. This parameter is
+#'    passed through from \code{gen_hw_set} and includes user specifications
+#'    for, if specified as other than the default, an alternative function to
+#'    use to identify heatwaves, alternative upper and lower year boundaries
+#'    for the projection period of the heatwave datasets being generated,
+#'    alternative upper and lower year boundaries for the reference period
+#'    to be used when measuring all heatwave characteristics related to
+#'    relative, rather than absolute, temperature, and the percentile
+#'    threshold to use when defining heatwaves.
 #'
 #' @return A closure that will find all heatwaves for a given ensemble. Each passed threshold corresponds to a particular
 #' city. The closure advances an incrementer every time it is called to get this effect.
@@ -78,9 +93,11 @@ createCityProcessor <- function(global, custom){
 
 #' Combine all identified heatwave dataframes together into one.
 #'
-#' @param A list of identified heatwave dataframes created by the closure of createCityProcessor
+#' @param hwDataframeList A list of identified heatwave dataframes created by
+#'    the closure of createCityProcessor
 #'
[…]

### 2016-01-30 R/main.R
Make `gen_hw_set` more robust, remove positional call

@@ -110,8 +110,10 @@ gen_hw_set <- function(out,
         }
 
         # Put the directories into nested list form
-        models <- acquireDirectoryStructure(dataFolder, coordinateFilenames,
-                                            tasFilenames, timeFilenames)
+        models <- acquireDirectoryStructure(dataFolder = dataFolder,
+                                            coordinateFilenames = coordinateFilenames,
+                                            tasFilenames = tasFilenames,
+                                            timeFilenames = timeFilenames)
 
         # Read the cities data file
         cities <- read.csv(citycsv)
@@ -138,8 +140,9 @@ gen_hw_set <- function(out,
         accumulators <- createAccumulators()
 
         # Process the entire dataset
-        referenceEnsembles <- sapply(models, processModel, global, custom,
-                                     accumulators)
+        referenceEnsembles <- sapply(models, processModel,
+                                     global = global,custom = custom,
+                                     accumulators = accumulators)
 
         # Write the model information from the model information accumulator
         out <- accumulators("return model information")

### 2016-02-02 R/IO.R
Pass global to `readLatLong`, etc.

@@ -15,7 +15,7 @@
 #'    column) of each grid point in the climate model. This dataframe
 #'    uses a format where the location of New York City, for example, is
 #'    (40.7127, 285.9941).
-readLatLong <- function(ensemble){
+readLatLong <- function(ensemble, global){
         loc_file <- grep(global$coordinateFilenames, ensemble)
         return(read.csv(ensemble[loc_file], header = FALSE))
 }
@@ -32,7 +32,7 @@ readLatLong <- function(ensemble){
 #' @return A dataframe of climate projection data where each column
 #'    corresponds to a climate model grid point and each row corresponds
 #'    to a date of observation.
-readtas <- function(ensemble){
+readtas <- function(ensemble, global){
         loc_file <- grep(global$tasFilenames, ensemble)
         return(read.csv(ensemble[loc_file], header = FALSE))
 }
@@ -49,7 +49,7 @@ readtas <- function(ensemble){
 #' @return A dataframe of dates corresponding to climate projection data,
 #'    where each row gives a date, split into columns of day in the
 #'    dataframe, four-digit year, numeric month, and numeric day.
-readTimes <- function(ensemble){
+readTimes <- function(ensemble, global){
         loc_file <- grep(global$timeFilenames, ensemble)
         return(read.csv(ensemble[loc_file], header = FALSE))
 }

### 2016-02-02 R/process.R

@@ -283,9 +283,9 @@ processEnsemble <- function(ensemble, modelName, global, custom, type){
 
         # Read the ensemble data
         cat("Reading --->", ensemble[1], "\n")
-        latlong <- readLatLong(ensemble)
-        tas <- readtas(ensemble)
-        times <- readTimes(ensemble)
+        latlong <- readLatLong(ensemble = ensemble, global = global)
+        tas <- readtas(ensemble = ensemble, global = global)
+        times <- readTimes(ensemble = ensemble, global = global)
         cat("Read operation complete", "\n")
 
         # Find indices of the closest points of measurement
@@ -299,6 +299,9 @@ processEnsemble <- function(ensemble, modelName, global, custom, type){
                             type = type)
         start <- bounds[1]
         end <- bounds[2]
+
+        # Get a vector containing the dates of days we are dealing with
+        dates <- formDates(times, bounds)
 
         # Acquire time series for every city
         series <- data.frame(tas[start:end, locations])

### 2016-02-02 R/process.R
Close #59

@@ -337,7 +337,7 @@ processEnsemble <- function(ensemble, modelName, global, custom, type){
 #'    coordinates.
 closest_point <- function(city, latlong){
         latitude <- as.double(city[2])
-        longitude <- 360 - as.double(city[3])
+        longitude <- as.double(city[3])
         aSQ <- abs(latlong[,1] - latitude) ^ 2
         bSQ <- abs(latlong[,2] - longitude) ^ 2
         cSQ <- aSQ + bSQ

### 2016-02-02 R/processHeatwaves.R
Close #58.
The problem was that `dates` weren't being passed through with `reference`. Now it's working fine without the `na.rm` option, as it should.

@@ -279,7 +279,7 @@ createHwDataframe <- function(city, threshold, heatwaves,
         hw.frame$mean.temp.1 <- mean(ref_temps)
         # Summertime is months May through September
         summertime <- as.POSIXlt(ref_dates)$mon %in% c(4:8)
-        hw.frame$mean.summer.temp <- mean(ref_temps[summertime], na.rm = TRUE)
+        hw.frame$mean.summer.temp <- mean(ref_temps[summertime])
 
         hw.frame$city <- city
 

### 2016-02-03 R/exampleApplyFunctions.R
Work on vignette

@@ -19,16 +19,19 @@ number_of_heatwaves <- function(hw_datafr){
         return(out)
 }
 
+#' @export
 heatwave_days <- function(hw_datafr){
         out <- sum(hw_datafr$length)
         return(out)
 }
 
+#' @export
 average_length <- function(hw_datafr){
         out <- mean(hw_datafr$length)
         return(out)
 }
 
+#' @export
 average_mean_temp <- function(hw_datafr){
         out <- mean(hw_datafr$mean.temp)
         return(out)

### 2016-02-03 R/IDHeatwaves.R
Take `hw.number` out of dataframe returned by the ID heatwaves function

@@ -75,9 +75,9 @@ IDheatwaves <- function(threshold, datafr, global, custom){
 #'    or when exploring exposures that, unlike heatwaves, may occur very
 #'    early or late in the calendar year.
 #' @export
-IDHeatwavesR <- function(threshold = stop("Error: unspecified threshold"),
-                         days = 2,
-                         datafr =  stop("Error: 'datafr' unspecified")){
+IDHeatwavesR <- function(threshold, datafr){
+
+        days <- 2
 
         # Add names to the dataframe
         colnames(datafr) <- c("date", "tmpd")
@@ -103,8 +103,7 @@ IDHeatwavesR <- function(threshold = stop("Error: unspecified threshold"),
         # Initialize dataframe containing the columns that will be added to
         # datafr
         hwInfo <- data.frame(hw = c(9),
-                            hw.number = c(9),
-                            first.hw.day = c(9))
+                            hw.number = c(9))
 
         # Current Index
         i <- 1
@@ -120,9 +119,7 @@ IDHeatwavesR <- function(threshold = stop("Error: unspecified threshold"),
                         # Store all desired information about this heatwave
                         hwInfo <- data.frame(hw = c(hwInfo[,1], rep(1, size)),
                                              hw.number = c(hwInfo[,2],
-                                                           rep(counter, size)),
-                                             first.hw.day = c(hwInfo[,3], 1,
-                                                              rep(0, size - 1)))
+                                                           rep(counter, size)))
 
                         # Increment and advance
                         counter <- counter + 1
@@ -130,7 +127,7 @@ IDHeatwavesR <- function(threshold = stop("Error: unspecified threshold"),
                 } else {
                         # If no heatwave at i, then add zeros to the end of
                         # the dataframe
-                        hwInfo <- rbind(hwInfo, c(0, 0, 0))
+                        hwInfo <- rbind(hwInfo, c(0, 0))
                         i <- i + 1
                 }
         }
@@ -169,9 +166,9 @@ IDHeatwavesR <- function(threshold = stop("Error: unspecified threshold"),
 #'    or when exploring exposures that, unlike heatwaves, may occur very
 #'    early or late in the calendar year.
 #' @export
-IDHeatwavesAlternative <- function(threshold = stop("Error: unspecified threshold"),
-                           days = 5,
-                           datafr =  stop("Error: 'datafr' unspecified")){
+IDHeatwavesAlternative <- function(threshold, datafr){
+
[…]

### 2016-02-03 src/IdentifyHeatwaves.cpp
Altered IdentifyHeatwaves.cpp

@@ -86,18 +86,18 @@ void storeHeatwaveEntry(int index, int hwSize, int hwCounter, std::vector<int>& 
 
         for(int i = 0; i < hwSize; i++){
                 // Push back 1 hwSize number of times onto hw.
-                hw[index] = 1;
+                hw[index + i] = 1;
 
                 // Push back hwCounter value of hwSize number of times onto hwNumber.
-                hwNumber[index] = hwCounter;
+                hwNumber[index + i] = hwCounter;
 
                 if(i == 0){
                         // Put the first heatwave day in place.
-                        firstHwDay[index] = 1;
+                        firstHwDay[index + i] = 1;
                 }
                 else{
                         // Put zeroes after the first heatwave day has been placed.
-                        firstHwDay[index] = 0;
+                        firstHwDay[index + i] = 0;
                 }
         }
 }
@@ -121,8 +121,8 @@ void storeZeroes(int index, int potentialHeatwave, std::vector<int>& hw, std::ve
 
         //Store the zeroes.
         for(int i = 0; i < zercount; i++){
-                hw[index] = 0;
-                hwNumber[index] = 0;
-                firstHwDay[index] = 0;
+                hw[index + i] = 0;
+                hwNumber[index + i] = 0;
+                firstHwDay[index + i] = 0;
         }
 }

### 2016-02-03 R/IO.R
Close #56.

@@ -79,7 +79,7 @@ writeThresholds <- function(modelName, threshOut, thresholds, global, custom){
 
         # Create the directory that the file will be written to
         dir.create(paste(writePath, modelName, sep = ""),
-                   showWarnings = FALSE, recursive = TRUE, mode = "0777")
+                   showWarnings = FALSE, recursive = TRUE)
 
         # Write the file
         write.csv(writeThis, file = paste(writePath, modelName,
@@ -130,8 +130,7 @@ createEnsembleWriter <- function(modelName, global, custom){
 
                 # Create the directory that the file will be written to
                 dir.create(paste(writePath, modelName, sep = ""),
-                           showWarnings = FALSE, recursive = TRUE,
-                           mode = "0777")
+                           showWarnings = FALSE, recursive = TRUE)
 
                 # Write the file
                 write.csv(hwFrame,

### 2016-02-03 inst/doc/futureheatwaves.R
Compile latest version of vignette

@@ -1,34 +1,70 @@
 ## ----echo = FALSE, message = FALSE---------------------------------------
 library(futureheatwaves)
 
-## ------------------------------------------------------------------------
-dataFolder <- "inst/cmip5/" 
-citycsv <- "inst/cities.csv"
+## ----eval = FALSE--------------------------------------------------------
+#  gen_hw_set(out = "example_results",
+#             dataFolder = "cmip5" ,
+#             citycsv = "cities.csv",
+#             coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
+#             tasFilenames = "tas_NorthAmerica_12mo.csv",
+#             timeFilenames = "time_NorthAmerica_12mo.csv")
 
-## ------------------------------------------------------------------------
-coordinateFilenames <- "latitude_longitude_NorthAmerica_12mo.csv"
-tasFilenames <- "tas_NorthAmerica_12mo.csv"
-timeFilenames <- "time_NorthAmerica_12mo.csv"
+## ---- eval = FALSE-------------------------------------------------------
+#  # Define the function
+#  average_length <- function(hw_datafr){
+#          ave_length <- mean(hw_datafr$length)
+#          return(ave_length)
+#  }
 
-## ------------------------------------------------------------------------
-out <- "~/tmp/example_results/"
+## ---- eval = FALSE-------------------------------------------------------
+#  # Apply across all heatwave dataframes from all ensemble members
+#  apply_all_models(out = "results",
+#                   FUN = average_length)
+
+## ---- eval = FALSE-------------------------------------------------------
+#  apply_all_models(out = "results",
+#                   FUN = average_length,
+#                   city_specific = TRUE)
+
+## ---- eval = FALSE-------------------------------------------------------
+#  average_mean_temp
 
 ## ----eval = FALSE--------------------------------------------------------
-#  gen_hw_set(out = out,
-#             dataFolder = dataFolder,
-#             citycsv = citycsv,
-#             coordinateFilenames = coordinateFilenames,
-#             tasFilenames = tasFilenames,
-#             timeFilenames = timeFilenames,
-#             RorCPP = 0)
+#  data(hw_datafr)
 
-## ------------------------------------------------------------------------
-dataFolder <- system.file("cmip5", package = "futureheatwaves")
[…]

### 2016-02-03 R/IDHeatwaves.R
Make some tweaks to R wrapper for C++ function

@@ -260,7 +260,8 @@ IDHeatwavesCPPwrapper <- function(datafr, threshold){
         tempsExceedingthreshold <- c(tempsExceedingthreshold, 0)
 
         # Identify heatwaves using the C++ functions.
-        heatwaves <- IDHeatwavesCPP(days, tempsExceedingthreshold)
+        heatwaves <- IDHeatwavesCPP(heatwaveLength = 2,
+                                    tempsExceedingCutoff = tempsExceedingthreshold)
 
         # Attach heatwaves columns onto the data in the datafr
         # variable. Note that the final row, which contains zeroes as

### 2016-02-03 R/IO.R
Close #63. Close #28.

@@ -54,38 +54,6 @@ readTimes <- function(ensemble, global){
         return(read.csv(ensemble[loc_file], header = FALSE))
 }
 
-#' Write temperature thresholds to file
-#'
-#' This function creates a dataframe with a column with names of each of the
-#' user-specified communities and a column with each community's threshold
-#' temperature, as determine from the climate projections for the reference
-#' period and the user-specified percentile threshold. These files are written
-#' out to a directory called "Thresholds" within the user-specified output
-#' directory (as specified in \code{\link{gen_hw_set}}). Within "Thresholds",
-#' there will be a subdirectory for each of the climate models.
-#'
-#' @param threshOut Character string with the file name to use for the file
-#'    the function writes out with threshold temperatures for each community
-#'    for the model.
-#' @inheritParams buildStructureModels
-#' @inheritParams processProjections
-#' @inheritParams processModel
-writeThresholds <- function(modelName, threshOut, thresholds, global, custom){
-
-        writePath <- paste0(global$output, "Thresholds/")
-
-        # Create datastructure that will be written
-        writeThis <- data.frame(global$cities[,1], thresholds)
-
-        # Create the directory that the file will be written to
-        dir.create(paste(writePath, modelName, sep = ""),
-                   showWarnings = FALSE, recursive = TRUE)
-
-        # Write the file
-        write.csv(writeThis, file = paste(writePath, modelName,
-                                          "/", threshOut, ".csv", sep = ''))
-}
-
 #' Ensemble writer factory function
 #'
 #' This function creates a closure that writes a single heatwave list to a
@@ -118,15 +86,6 @@ createEnsembleWriter <- function(modelName, global, custom){
 
                 cat("Writing ", modelName, ": r", i, "i", i, "p", i,
                     "\n", sep = "")
-
-                # TODO: Contradiction detected. writeThresholds is called each time the closure is called
-                # even though the documentation for writeThresholds states it is only supposed to be called
-                # for the r1i1p1 ensemble. Make adjustments.
-                # Write minimum threshold temperatures of each city to a file
-                thresholds <- dplyr::group_by(hwFrame, city) %>%
-                        dplyr::summarize(min = min(min.temp))
-                writeThresholds(modelName, 'minimums',
-                                thresholds, global, custom)
 
                 # Create the directory that the file will be written to
                 dir.create(paste(writePath, modelName, sep = ""),

### 2016-02-03 R/process.R

@@ -62,7 +62,9 @@ processModel <- function(model, global, custom, accumulators){
 
         # Acquire vector of threshold temperatures using the historical
         # ensemble for this model which possesses the name r1i1p1
-        thresholds <- processThresholds(model, global, custom)
+        thresholds <- processThresholds(model = model,
+                                        global = global,
+                                        custom = custom)
 
         # If reference period differs from projection period, get data
         # for reference period
@@ -153,9 +155,6 @@ processThresholds <- function(model, global, custom){
         # threshold percentile (default: 0.98)
         thresholds <- apply(thresholdEnsemble$series, 2, quantile,
                             probs = custom$probThreshold)
-
-        # Save thresholds to file
-        writeThresholds(name, 'thresholds', thresholds, global, custom)
 
         return(thresholds)
 }

### 2016-02-03 R/IO.R
Removed permissions parameter from dir.create() call

@@ -130,8 +130,7 @@ createEnsembleWriter <- function(modelName, global, custom){
 
                 # Create the directory that the file will be written to
                 dir.create(paste(writePath, modelName, sep = ""),
-                           showWarnings = FALSE, recursive = TRUE,
-                           mode = "0777")
+                           showWarnings = FALSE, recursive = TRUE)
 
                 # Write the file
                 write.csv(hwFrame,

### 2016-02-03 R/IDHeatwaves.R
Updated C++ functions

@@ -266,7 +266,7 @@ IDHeatwavesCPPwrapper <- function(datafr, threshold){
         tempsExceedingthreshold <- c(tempsExceedingthreshold, 0)
 
         # Identify heatwaves using the C++ functions.
-        heatwaves <- IDHeatwavesCPP(days, tempsExceedingthreshold)
+        heatwaves <- IDHeatwavesCPP(2, tempsExceedingthreshold)
 
         # Attach heatwaves columns onto the data in the datafr
         # variable. Note that the final row, which contains zeroes as

### 2016-02-03 R/IO.R
Close #25.

@@ -112,13 +112,19 @@ createEnsembleWriter <- function(modelName, global, custom){
 #' @inheritParams processModel
 #'
 #' @return ...
-writeAccumulator <- function(modelInfoAccumulator, global){
+writeAccumulators <- function(modelInfoAccumulator,
+                             locationList,
+                             global){
         colnames(modelInfoAccumulator)[1] <- "Models"
         colnames(modelInfoAccumulator)[2] <- "# of historical ensembles"
         colnames(modelInfoAccumulator)[3] <- "# of future projection ensembles"
-        cat("Writing accumulator", "\n")
+
+        cat("Writing accumulators", "\n")
         writePath <- global$output
         write.csv(modelInfoAccumulator,
                   file = paste0(writePath, "hwModelInfo", ".csv"),
+                  row.names = FALSE)
+        write.csv(locationList,
+                  file = paste0(writePath, "locationList", ".csv"),
                   row.names = FALSE)
 }

### 2016-02-03 R/main.R

@@ -164,7 +164,9 @@ gen_hw_set <- function(out,
 
         # Write the model information from the model information accumulator
         out <- accumulators("return model information")
-        writeAccumulator(accumulators("return model information"), global)
+        writeAccumulators(modelInfoAccumulator = accumulators("return model information"),
+                          locationList = accumulators("return locations"),
+                          global = global)
 
         cat("All operations completed. Exiting.", "\n\n")
         return(out)
@@ -293,9 +295,6 @@ checkCustomBounds <- function(boundList){
         }
 }
 
-# TODO: It may be possible to make the accumulator system extremely robust.
-# Consider this possibility after all else is finished.
-
 #' Create accumulator closure
 #'
 #' This closure holds, adds to, and returns data structures that the user
@@ -332,7 +331,7 @@ checkCustomBounds <- function(boundList){
 #'    input into the data structures it contains.
 createAccumulators <- function(){
         modelInfoAccumulator <- data.frame(c(), c(), c())
-        locationList <- list()
+        locationList <- data.frame(c(), c(), c(), c(), c(), c())
 
         function(command, newElement = FALSE){
 
@@ -349,7 +348,7 @@ createAccumulators <- function(){
                         return(locationList)
 
                 } else if(command == "append location list"){
-                        locationList <<- list(locationList, newElement)
+                        locationList <<- rbind(locationList, newElement)
                 }
 
                 # If user passes an invalid command, halt the program.

### 2016-02-03 R/mapCities.R

@@ -1,62 +1,32 @@
-#' Map cities of interest
+#' Create a map of model grid
 #'
-#' @param locationsList
-#' @param cities
-#' @param sampleEnsemble
+#' @param plot_model A character string with the name of the model to plot
+#' @inheritParams gen_hw_set
 #'
-#' @return Checks locations in locationList for ... and, if ...,
-#'    plots a map of ... .
-makeMap <- function(locationsList, cities, sampleEnsemble){
-        identical <- checkLocations(locationsList)
-        if(identical){
-                mapCities(locationsList[1], cities, sampleEnsemble)
-        }
-}
+#' @export
+#'
+#' @importFrom dplyr %>%
+map_grid <- function(plot_model, out){
+        cities <- read.csv(paste(out, "locationList.csv", sep = "/")) %>%
+                mutate(long = long - 360,
+                       long_grid = long_grid - 360) %>%
+                filter(model == plot_model)
 
-#' [Short title for function]
-#'
-#' @param locationList [Whis is this?]
-#'
-#' @return [What does this return?]
-checkLocations <- function(locationsList){
-        identical <- sapply(locationsList, function(locations, locationsList){
-                identical2 <- sapply(locationsList, function(locations2, locations){
-                        if(identical(locations, locations2)){
-                                return(TRUE)
-                        }
-                },
-                locations)
-                return(all(identical2))
-        },
-        locationsList)
-        return(all(identical))
-}
-
-#' Map locations and cities
-#'
-#' @param locations [What is this?]
-#' @param cities Dataframe with a row for every city and columns for
-#'    city ID, latitude, and longitude.
-#' @param sampleEnsemble [What is this?]
-#'
-#' @return Plots  a map of all cities in the \code{cities} dataframe,
-#'     as well as ...
-#TODO: SWAP LONGITUDE AND LATITUDE
-mapCities <- function(locations, cities, sampleEnsemble){
-        latlong <- readLatLong(sampleEnsemble)
-        dataCoordinates <- latlong[locations]
-        cities <- cities[ , 1:3]
-        cities <- -cities$long
+        latlong <- unique(cities[ , c("long_grid", "lat_grid")])
         states <- map_data("state")
+
         map <- ggplot()
         map <- map + geom_polygon(data = states,
                                   aes(x = long, y = lat, group = group),
-                                  colour = "white", fill = "grey10")
-        map <- map + geom_text(data = cities, hjust = .5, vjust = -.5,
-                               aes( x = long, y = lat, label = cities$city),
[…]

### 2016-02-03 R/process.R

@@ -60,9 +60,16 @@ processModel <- function(model, global, custom, accumulators){
 
         # Acquire vector of threshold temperatures using the historical
         # ensemble for this model which possesses the name r1i1p1
-        thresholds <- processThresholds(model = model,
+        thresholdList <- processThresholds(model = model,
                                         global = global,
                                         custom = custom)
+        thresholds <- thresholdList$thresholds
+        out_locations <- thresholdList$out_locations
+        out_locations$model <- modelName
+
+        # Add locations to the locationList accumulator
+        accumulators(command = "append location list",
+                     newElement = out_locations)
 
         # If reference period differs from projection period, get data
         # for reference period
@@ -154,7 +161,10 @@ processThresholds <- function(model, global, custom){
         thresholds <- apply(thresholdEnsemble$series, 2, quantile,
                             probs = custom$probThreshold)
 
-        return(thresholds)
+        out_locations <- thresholdEnsemble$out_locations
+
+        return(list(thresholds = thresholds,
+                    out_locations = out_locations))
 }
 
 processReference <- function(model, global, custom){
@@ -218,8 +228,8 @@ processProjections <- function(ensemble, modelName, ensembleWriter,
         ensembleSeries$reference <- reference
         ensembleSeries$reference_dates <- reference_dates
 
-        # Append locations vector to the locations vector accumulator
-        accumulators("append location list", ensembleSeries$locations)
+#         # Append locations vector to the locations vector accumulator
+#         accumulators("append location list", ensembleSeries$locations)
 
         # Acquire the heatwave dataframes for every ensemble in the
         # model

### 2016-02-03 R/mapCities.R
Work on mapping function

@@ -3,6 +3,10 @@
 #' @param plot_model A character string with the name of the model to plot
 #' @inheritParams gen_hw_set
 #'
+#' @examples
+#' out <- system.file(package = "futureheatwaves")
+#' map_grid(plot_model = "bcc1", out = out)
+#'
 #' @export
 #'
 #' @importFrom dplyr %>%

### 2016-02-03 inst/doc/futureheatwaves.Rmd
Work on vignette

@@
 title: "`futureheatwaves` package vignette"
-author: "G. Brooke Anderson and Colin Eason"
+author: "G. Brooke Anderson, Colin Eason, and Elizabeth Barnes"
 date: "`r Sys.Date()`"
@@
 
+This function requires you to have data on community locations and climate projection data set up in a specific way on your local computer.
+You then use arguments in the main package function, `gen_hw_set`, to direct the function to these files so it can process them and output heatwave projections.
+The final projection files will be written out to your local computer, in a directory that you select and specify through `gen_hw_set`.
+You can then apply user-created R functions to all heatwaves in the output using the `apply_all_models` function.
+
 ## A basic example of using the package
+
+### Required files and directory set-up
+
+#### File with community locations
+
+First, you must have a comma-separated (csv) file the gives unique identifiers for each community for which you wish to make projections, as well as their latitude and longitude.
+The latitude and longitude should be expressed in the same way that latitude and longitude are expressed in the climate projection files to be processed.
+The community location file should have three columns, with the following column names:
+
+- `city`: A unique identifier for each community you will be analyzing.
+- `lat`: The communities' latitudes
+- `long`: The communities' longitudes
+
+Here is an example of the correct format for a community locations file for five US communities (Baltimore, MD; New York, NY; Newark, NJ; Philadelphia, PA; and Providence RI):
+
+```
+"city","lat","long"
+"balt",39.3008,283.3894
+"ny",40.6698,286.0562
+"nwk",40.7241,285.8268
+"phil",40.006817,284.8653
+"prov",41.82195,288.5803
+```
+
+The location of this community locations file will be specified to the main function, `gen_hw_set`, using the argument `citycsv`.
+For example, if I had this file saved as "my_cities.csv" in my working directory, I would specify `citycsv = "my_cities.csv"` in the `gen_hw_set` function call.
+
+#### Files with climate projections
+
+For each ensemble member of each climate model, you must have three files: one giving the gridded climate model output by date, one giving the location of each of the climate model grids, and one giving the date of each of the climate model projections.
+
+These files must be in a certain structure to run correctly through the `gen_hw_set` function.
+Therefore, the function currently requires a certain amount of preprocessing of climate projection files from a format like .`netCDF` to prepare them to go through this function.
+In the future, we plan to extend this package to allow the `gen_hw_set` function to directly process `.netCDF` files.
+In the meantime, you must preprocess the file for the projection from each ensemble member to conform to the formats listed below.
+
+**Climate projection file**: The climate projection file should be a comma-separated file that looks something like this:
+
+```
+"275.74","282.32","285.03"
+"275.94","283.14","285.47"
+"272.9","280.26","284.6"
+"273.61","280.82","284.67"
+"273.97","281.18","284.83"
+```
+
+This is a projection file covering three grid points of a climate model, giving projections for five dates.
+Each column corresponds to one grid point in the climate model.
+Each row corresponds to one date.
+The file does not have a header row; rather, the observations begin immediately on the first row of the file.
+Similarly, the file does not have a column of row numbers.
+Most projection files will include many more columns and rows than this small example, since they will usually cover all the grid points in a climate model and a long time series of observations.
+
+Each climate model ensemble member will need to have a climate projection file like this one.
+They will all be saved in separate, ensemble-specific directories and should all be given the same file name.
+In our example data, these climate projection files are always saved with the filename "tas_NorthAmerica_12mo.csv".
+You will need to specify this filename in the `tasFilenames` argument when running `gen_hw_set`.
+
+**Grid point location file**: The grid point location file should be a comma-separated file that looks something like this:
+
+```
+"40.464","284.06"
+"40.464","286.88"
+"40.464","289.69"
+```
+
+For this file, each row gives the location for the grid point for the corresponding column in the climate projection files.
+For example, the first row gives the latitude and longitude for the projections given in the first column of the climate projection file, the second row of this file gives the latitude and longitude for the projection given in the second column of the climate projection file, and so on.
+
+The first column of this file should give latitude and the second column should give longitude. The file should not have a header row or a separate column of row names.
+
+Each climate model ensemble member will need to have a grid-point location file like this one.
+They will all be saved in separate, ensemble-specific directories and should all be given the same file name.
+In our example data, these grid-point location files are always saved with the filename "latitude_longitude_NorthAmerica_12mo.csv".
+You will need to specify this filename in the `coordinateFilenames` argument when running `gen_hw_set`.
+
+**Projection times data file**: The projection times data file should be a comma-separated file that looks something like this:
+
+```
+"0","2006","1","1"
+"1","2006","1","2"
+"2","2006","1","3"
+"3","2006","1","4"
+"4","2006","1","5"
+```
+
+This file gives a date that corresponds to each row of the climate projection file.
+The file should have four columns: one with row numbers, and then one each with year, month, and day.
+Year should be given with four digits.
+The file should not have a header row.
+
+Each climate model ensemble member will need to have a separate projection times data file like this one.
+They will all be saved in separate, ensemble-specific directories and should all be given the same file name.
+In our example data, these grid-point location files are always saved with the filename "time_NorthAmerica_12mo.csv".
+You will need to specify this filename in the `timeFilenames` argument when running `gen_hw_set`.
+
+#### Directory structure for climate projection files
 
@@
 
+For example, the following shows the file structure for the files in the example data included with the package.
+This example directory includes separate historical (with years 1980--2004) and RCP8.5 (with years 2006--2099) directories, each of which has the required climate projection, grid point locations, and projection time files for two separate models (`bcc1` and `bcc1-a`) with one and two separate ensemble members, respectively(`r1i1p1` for `bcc1`; `r1i1p1` and `r1i1p1-a` for `bcc1-a`).
+Notice that each subdirectory has all required levels-- for example, even though the `bcc1` model only has one ensemble member (in this example), we've still included a directory level for ensembles in the directory structure.
+
+- `cmip5`
+    - `historical`
+        - `bcc1`
+            - `r1i1p1`
+                - `tas_NorthAmerica_12mo.csv`
+                - `latitude_longitude_NorthAmerica_12mo.csv`
+                - `time_NorthAmerica_12mo.csv`
+        - `bcc1-a`
+            - `r1i1p1`
+                - `tas_NorthAmerica_12mo.csv`
+                - `latitude_longitude_NorthAmerica_12mo.csv`
+                - `time_NorthAmerica_12mo.csv`
+            - `r1i1p1-a`
+                - `tas_NorthAmerica_12mo.csv`
+                - `latitude_longitude_NorthAmerica_12mo.csv`
+                - `time_NorthAmerica_12mo.csv`
+    - `rcp85`
+        - `bcc1`
+            - `r1i1p1`
+                - `tas_NorthAmerica_12mo.csv`
+                - `latitude_longitude_NorthAmerica_12mo.csv`
+                - `time_NorthAmerica_12mo.csv`
+        - `bcc1-a`
+            - `r1i1p1`
+                - `tas_NorthAmerica_12mo.csv`
+                - `latitude_longitude_NorthAmerica_12mo.csv`
+                - `time_NorthAmerica_12mo.csv`
+            - `r1i1p1-a`
+                - `tas_NorthAmerica_12mo.csv`
+                - `latitude_longitude_NorthAmerica_12mo.csv`
+                - `time_NorthAmerica_12mo.csv`
+
 Throughout the directory, subdirectories should be named using names reflecting the climate model or ensemble members they contain.
@@
 
-Once you have your files set up in this structure, define the following variables:
+Once you have your files set up in this structure, define the following variables, you will specify the location of the start of the directory using the `dataFolder` argument in the `gen_hw_set` function.
+In the example directory structure above, if the `cmip5` directory was in your current working directory, you would specify `dataFolder = "cmip5"` when calling `gen_hw_set`.
 
-- `dataFolder`: Character string with pathway to directory that contains climate projections. 
-- `citycsv`: Character string giving the filepath to a .csv file with latitude and longitude values for each city for which you would like to generate
-heatwave projections.
-- `coordinateFilenames`: Character string with name of the files containing the latitude and longitude coordinates corresponding to the columns of the time series data.
-- `tasFilenames`: Character sting with name of files containing the time series data.
-- `timeFilenames`: Character string with name of the files containing the date information corresponding to the rows of the time series data.
+#### Specifying where to output results
 
-For example, I can specify `dataFolder` and `citycsv` as:
+You will need to select a directory pathname where all the heatwave projection files generated by the function will be saved.
+If you use a pathname to a directory that does not yet exist, the function will create that directory and then write the results to it.
+If you specify a directory that does already exist, the function will overwrite it with the results from the function run.
+You therefore should be careful when specifying this directory path and be sure to confirm that the function will not overwrite a directory you do not want to lose.
+You will specify the output directory using the `out` argument in the `gen_hw_set` function.
+For example, if you wanted to create a new directory called "example_results" in the current working directory and output files there, you could specify `out = "example_results"` in `gen_hw_set`.
 
-```{r}
-dataFolder <- "inst/cmip5/" 
-citycsv <- "inst/cities.csv"
-```
-
-Within the ensemble subdirectories in my files structure, all of the files with climate model data are named "tas_NorthAmerica_12mo.csv", all the files with the corresponding time for rows in that data are in files named "time_NorthAmerica_12mo.csv", and all the files with locations of the grid points are named "latitude_longitude_NorthAmerica_12mo.csv".
-Therefore, I would specify:
-
-```{r}
-coordinateFilenames <- "latitude_longitude_NorthAmerica_12mo.csv"
-tasFilenames <- "tas_NorthAmerica_12mo.csv"
-timeFilenames <- "time_NorthAmerica_12mo.csv"
-```
-
-You will also need to specify where the output files should be locally saved.
-Save this as the object `out`.
-For example, if I wanted to write the output heatwave files to a directory called "example_results" in the `tmp` subdirectory of my home directory, I could specify:
-
-```{r}
-out <- "~/tmp/example_results/"
-```
+### Running the code to generate heatwave datasets
 
@@
 ```{r eval = FALSE}
-gen_hw_set(out = out,
-           dataFolder = dataFolder,
-           citycsv = citycsv,
-           coordinateFilenames = coordinateFilenames,
-           tasFilenames = tasFilenames,
-           timeFilenames = timeFilenames,
-           RorCPP = 0)
+gen_hw_set(out = "example_results",
+           dataFolder = "cmip5" ,
+           citycsv = "cities.csv",
+           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
+           tasFilenames = "tas_NorthAmerica_12mo.csv",
+           timeFilenames = "time_NorthAmerica_12mo.csv")
 ```
@@
 Warning: This function will write new files to your computer in the 
- ~/tmp/results directory of your computer. 
- Do you want to continue (y / n): 
+ ~/tmp/results/ directory of your computer. If that directory already exists,
+running this function will write over it. 
+ Do you want to continue? (y / n): 
 ```
@@
 If you agree that it is okay to write files and subdirectories to this directory, enter "y" at the prompt, and the function will continue running.
+(If you do not get this warning when running the function, choose the option `printWarning = FALSE` when calling `gen_hw_set`.)
 
@@
 ```
-Processing historical ensemble for bcc1 
+Processing thresholds for bcc1 
 Reading ---> r1i1p1 
 Read operation complete 
-Processing rcp 8.5 ensemble bcc1 
+Processing references for bcc1 
 Reading ---> r1i1p1 
 Read operation complete 
-Creating heatwave dataframe ~~ City:  balt  ~~ City Number:  1  ~~ Cutoff:  84.632 
-Creating heatwave dataframe ~~ City:  ny  ~~ City Number:  2  ~~ Cutoff:  78.77876 
-Creating heatwave dataframe ~~ City:  nwk  ~~ City Number:  3  ~~ Cutoff:  78.77876 
-Creating heatwave dataframe ~~ City:  phil  ~~ City Number:  4  ~~ Cutoff:  84.632 
-Creating heatwave dataframe ~~ City:  prov  ~~ City Number:  5  ~~ Cutoff:  75.074 
+Processing projections for  bcc1 
+Reading ---> r1i1p1 
+Read operation complete 
+Creating heatwave dataframe ~~ City:  balt  ~~ City Number:  1  ~~ Cutoff:  93.59672 
+Creating heatwave dataframe ~~ City:  ny  ~~ City Number:  2  ~~ Cutoff:  86.07236 
+Creating heatwave dataframe ~~ City:  nwk  ~~ City Number:  3  ~~ Cutoff:  86.07236 
+Creating heatwave dataframe ~~ City:  phil  ~~ City Number:  4  ~~ Cutoff:  93.59672 
+Creating heatwave dataframe ~~ City:  prov  ~~ City Number:  5  ~~ Cutoff:  81.06836 
 Writing bcc1: r1i1p1
@@
 All operations completed. Exiting. 
-
-  modelName length.histens. length.rcpens.
-1      bcc1               1              1
 ```
@@
 The function provides status reports on its progress in generating the heatwave projections, as this process can take a while if you include many cities and / or many climate model projections.
-If you do not want to generate this output when running the function, choose the option `printWarning = FALSE` when calling `gen_hw_set`.
 
-Once the function has completed running, you will have two new subdirectories in the directory you specified with the `out` variable.
-These will be called `Heatwaves` and `Thresholds`.
-The `Heatwaves` directory will contain dataframes with all identified heatwaves and their characteristics for each city, while the `Thresholds` folder will contain files with the temperature thresholds used for each city given.
+Once the function has completed running, you will have one new subdirectory in the directory you specified with the `out` variable called `Heatwaves`.
+This directory will contain dataframes with all identified heatwaves and their characteristics for each community.
 The `out` directory specified will also have a file called `hwModelInfo.csv` that contains some basic information about the climate models and the number of ensemble members included for each.
+
+The full output directory structure will look something like this:
+
+- `out`
+    - `Heatwaves`
+        - `Projections`
+            - `bcc1`
+                - `1.csv`
+            - `bcc1-a`
+                - `1.csv`
+                - `2.csv`
+    - `hwModelInfo.csv`
+
+Each of the heatwave files will have one heatwave per row, covering all the heatwaves identified for that particular ensemble member.
+The dataframe has the following columns with characteristics of each heatwave:
+
+- `hw.number`: A sequential number identifying each heatwave in each community;
+- `mean.temp`: Average of mean daily temperature across all days in the heatwave; 
+- `max.temp`: Highest value of mean daily temperature across days in the heatwave; 
+- `min.temp`: Lowest value of mean daily temperature across days in the heatwave;
+- `length`: Number of days in the heatwave; 
+- `start.date`: Date of the first day of the heatwave; \code{end.date}: Date of the last day of the heatwave; 
+- `start.doy`: Numeric day of the year of the first day of the heatwave (1 = Jan. 1, etc.); 
+- `start.month`: Numeric value indicating the month in which the heatwave started (1 = January); 
+- `days.above.80`: Number of days in the heatwave above 80 degrees Fahrenheit; 
+- `days.above.85`: Number of days in the heatwave above 85 degrees Fahrenheit; 
+- `days.above.90`: Number of days in the heatwave above 90 degrees Fahrenheit;
+- `days.above.95`: Number of days in the heatwave above 90 degrees Fahrenheit;
+- `days.above.99th`:Number of days in the heatwave above the 99th percentile temperature for the community, using the period specified by the user with the `referenceBoundaries` argument in `gen_hw_set` as a reference for determining these percentiles;
+- `days.above.99.5th`:Number of days in the heatwave above the 99.5th percentile temperature for the community; 
+- `first.in.season`: Whether the heatwave was the first to occur in its season (as determined by whether the heatwave was the first in its calendar year); 
+- `threshold.temp`: The temperature used as the threshold for the heatwave definition in the selected community; 
+- `mean.temp.quantile`: The percentile of the average daily mean temperature during the heatwave compared to the community's year-round temperature distribution, based on the temperatures for the community during the period specified by the `referenceBoundaries` argument in `gen_hw_set`;
+- `max.temp.quantile`: The percentile of the highest daily mean temperature during the heatwave compared to the community's year-round temperature distribution; 
+- `min.temp.quantile`: The percentile of the lowest daily mean temperature during the heatwave compared to the community's year-round temperature distribution; 
+- `mean.temp.1`: The community's average year-round temperature, based on the temperatures for the community during the period specified by the `referenceBoundaries` argument in `gen_hw_set`;
+- `mean.summer.temp`: The community's average May--September temperature, based on the temperatures for the community during the period specified by the `referenceBoundaries` argument in `gen_hw_set`; and
+- `city`: The identifier for the community, as given in the file specified in the `citycsv` argument of `gen_hw_set`.
+
+### Applying functions to the heatwave datasets
+
+Once you have created a directory of files with characterized heatwaves for each ensemble member, you can explore the results by using the `apply_all_models` function to apply custom R functions across all heatwave dataframes created by the `gen_hw_sets` call.
+You specify the location of the directory with the heatwave dataframes using the `out` argument when calling `apply_all_models`.
+Typically, this will be the same value as that for the `out` argument in `gen_hw_set`.
+
+For example, if you wanted to to get the average length of the heatwaves identified within each ensemble member, you could write a simple function to calculate average length and then use `apply_all_models` to apply it across the dataframes for heatwaves for all ensemble members in all climate models:
+
+```{r, eval = FALSE}
+# Define the function
+average_length <- function(hw_datafr){
+        ave_length <- mean(hw_datafr$length)
+        return(ave_length)
+}
+```
+
+```{r, eval = FALSE}
+# Apply across all heatwave dataframes from all ensemble members
+apply_all_models(out = "results", 
+                 FUN = average_length)
+```
+
+```
+  model ensemble value
+ bcc1-a        1   3.5
+ bcc1-a        2   3.5
+   bcc1        1   3.5
+```
+
+You can also use the `city_specific` argument in `apply_all_models` to apply the function separately by community to all heatwaves within the dataframe of heatwaves for each ensemble member.
+For example:
+
+```{r, eval = FALSE}
+apply_all_models(out = "results", 
+                 FUN = average_length,
+                 city_specific = TRUE)
+```
+
+```
+   model ensemble city value
+  bcc1-a        1 balt   3.5
+  bcc1-a        1  nwk   3.5
+  bcc1-a        1   ny   3.5
+  bcc1-a        1 phil   3.5
+  bcc1-a        1 prov   3.5
+  bcc1-a        2 balt   3.5
+  bcc1-a        2  nwk   3.5
+  bcc1-a        2   ny   3.5
+  bcc1-a        2 phil   3.5
+  bcc1-a        2 prov   3.5
+    bcc1        1 balt   3.5
+    bcc1        1  nwk   3.5
+    bcc1        1   ny   3.5
+    bcc1        1 phil   3.5
+    bcc1        1 prov   3.5
+```
+
+To work, the functions you apply using `apply_all_models` must follow a certain structure.
+They must require only a dataframe of heatwaves, in the format of those output by `gen_hw_set`, as the input, and they must output a single-value vector (i.e., a vector of length one).
+
+We have included several functions as simple examples of the type of functions that can be used with `apply_all_models`:
+
+- `number_of_heatwaves`: Determines the number of heatwaves ; 
+- `heatwave_days`: Sums up the total number of heatwave days by adding the lengths of all heatwaves; 
+- `average_length`: Calculates the average length of heatwaves; and 
+- `average_mean_temp`: Calculates the average mean temperature across all heatwaves
+
+You can see the code of any of these functions, to use to develop your own, by typing just the function name (with no parentheses) in your R console.
+
+```{r, eval = FALSE}
+average_mean_temp
+```
+
+```
+function(hw_datafr){
+        out <- mean(hw_datafr$mean.temp)
+        return(out)
+}
+<environment: namespace:futureheatwaves>
+```
+
+You can use this functionality to either explore characteristics of the identified heatwaves or to estimate the health impacts of the identified heatwaves.
+
+To help users create their own functions, we have included an example dataframe representative of the heatwave dataframes that `gen_hw_set` outputs. This data can be loaded using the call:
+
+```{r eval = FALSE}
+data(hw_datafr)
+```
+
+### Mapping grid points
+
+The package also has a function that allows you to plot the locations of grid points for each climate model that match study communities, as well as show the links between climate grid points and associated communities.
+
+```{r fig.width = 5, message = FALSE, warning = FALSE}
+out <- system.file(package = "futureheatwaves")
+map_grid(plot_model = "bcc1", out = out)
+```
 
@@
 
-## How it works
+### Using a custom heatwave definition
+
+The default is for this package to use the following definition to identify heatwaves:
+
+> A heatwave is two or more days at or above a community-specific threshold temperature, with the threshold determined as the $98^{th}$ percentile of year-round temperature in the community during some reference period (by default, 1981--2004).
+
+Two components of this definition can be easily customized, without creating a new R function to use to identify heatwaves.
+
+First, the percentile used to identify a heatwave can be changed using the `probThreshold` option when calling the `gen_hw_set` function.
+This option can take values between 0 and 1.
+The default value is 0.98, or a definition with a threshold of the $98^{th}$ percentile of the community's temperature.
+
+Second, it is possible to specify the range of years of data for the community that should be used when determining this threshold.
+For example, if you wanted to base the threshold for each community on current community climates, you could leave this option as its default, which uses climate projections from the years 1981 to 2004 to determine the thresholds.
+If you wanted to use a different set of years, you could set the start and end year
+bounds for this reference period within the first two values of the option `dataBoundaries` in the function `gen_hw_set`.
+
+It is also possible to use a completely customized heatwave definition.
+To do this, you will need to load a function, implementing the definition to identify heatwaves in a time series of temperature data, into your R session.
+You can then reference the custom function using the `IDheatwavesFunction` option in `gen_hw_set`.
+
+To work correctly, this custom function for identifying heatwaves must allow specific inputs and generate specific outputs.
+The function must allow the following inputs (even if it does not use them within the function code):
+
+- `datafr`: A dataframe with columns for date and temperature for a community.
+The first column should be the date of each observation, in the Date class.
+The second column should be the temperature.
+At this point in the normal processing of the `gen_hw_set` function, temperatures have been converted from Kelvin (the expected units for the climate projection files) to Fahrenheit.
+Therefore, if you are writing a custom function that uses an absolute temperature threshold, it should be included in the function in Fahrenheit units.
+- `threshold`: A single-value numerical vector; in the default definition, this is the threshold temperature for each community, as determined earlier in the function call, calculated as a percentile of the community's year-round temperature.
+
+The function must return the input dataframe (`datafr`), with the following columns added:
+
+- `hw`: A binary variable indicating whether a day was part of a heatwave (1: day in a heatwave; 0: not in a heatwave) and
+- `hw.number`: A non-negative integer that is 0 for all days that are not in heatwaves and, for days in heatwaves, gives a unique number for each separate heatwave.
+For example, all days in the second heatwaves in the time series would get the value 2 for this column, while all days in the fifth heatwave would get the value 5, and all days not in heatwaves would get the value 0.
+
+The custom function can then be referenced using the `IDheatwavesFunction` option in `gen_hw_set`, and it will be used to identify heatwaves in the dataset.
+For example, we have included an alternative ID definition function in this package, called `IDHeatwavesAlternative`, which identifies heatwaves as five or more days (compared to two or more days required by the default definition) above a community's $98^{th}$ percentile temperature.
+To use this as the heatwave definition, you would run:
+
+```{r, eval = FALSE}
+gen_hw_set(out = "example_results",
+           dataFolder = "cmip5" ,
+           citycsv = "cities.csv",
+           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
+           tasFilenames = "tas_NorthAmerica_12mo.csv",
+           timeFilenames = "time_NorthAmerica_12mo.csv",
+           IDheatwavesFunction = "IDHeatwavesAlternative")
+```
+
+### Customizing projection dates
+
+By default, the function will generate heatwave projections for the selected communities for the years 2061 to 2080.
+This projection date range can be changed by specifying alternative starting and ending years as the third and fourth values of the `projectionBoundaries` option in the `gen_hw_set` function.
+The starting year cannot be earlier than 1980 and the ending year cannot be later than 2099.
+Further, the custom date boundaries cannot span 2005 (i.e., they must end in 2004 or earlier or must start in 2006 or later).
+
+For example, to create projections for 2071 to 2090, you could call:
+
+```{r, eval = FALSE}
+gen_hw_set(out = "example_results",
+           dataFolder = "cmip5" ,
+           citycsv = "cities.csv",
+           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
+           tasFilenames = "tas_NorthAmerica_12mo.csv",
+           timeFilenames = "time_NorthAmerica_12mo.csv",
+           projectionBoundaries = c(2071, 2090))
+```
+
+### Customizing dates used to determine thresholds or reference data
+
+Similarly, while the default is to use data from 1981 to 2004 to determine the $98^{th}$ percentile temperature for each community to use in their heatwave definitions, the user can customize which date range to use when pulling the temperature data to use to calculate these thresholds.
+This option is specified using the `thresholdBoundaries` option in `gen_hw_set`.
+For example, to use the years 2061 to 2080 to calculate these threshold temperatures, you could run:
+
+```{r, eval = FALSE}
+gen_hw_set(out = "example_results",
+           dataFolder = "cmip5" ,
+           citycsv = "cities.csv",
+           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
+           tasFilenames = "tas_NorthAmerica_12mo.csv",
+           timeFilenames = "time_NorthAmerica_12mo.csv",
+           thresholdBoundaries = c(2061, 2080))
+```
+
+You can customize one other set of time boundaries.
+The heatwave datasets include several heatwave characteristics that incorporate relative temperature.
+These take the absolute temperature measures of heatwave and compare them to the community's typical temperature distributions to generate relative measures of the intensity of the heatwave.
+The default is to use temperatures from a community for the period 2061 to 2080 for these reference temperatures to calculate relative temperatures of each heatwave.
+However, you can use the `referenceBoundaries` option of `gen_hw_set` to specify which years you would like to use for these reference temperatures.
+This can be useful in exploring the role of adaptation in future heatwaves.
+For example, if you wanted to use temperature projections for 2031 to 2040 when calculating relative characteristics of heatwaves, you could run:
+
+```{r, eval = FALSE}
+gen_hw_set(out = "example_results",
+           dataFolder = "cmip5" ,
+           citycsv = "cities.csv",
+           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
+           tasFilenames = "tas_NorthAmerica_12mo.csv",
+           timeFilenames = "time_NorthAmerica_12mo.csv",
+           referenceBoundaries = c(2031, 2040))
+```
+
+As with the `projectionBoundaries` option, there are some restrictions on which year ranges can be selected.
+The starting year cannot be earlier than 1980 and the ending year cannot be later than 2099.
+Further, the custom date boundaries cannot span 2005 (i.e., they must end in 2004 or earlier or must start in 2006 or later).
+
+## Example data
+
+We used, as an example, data from the Beijing Climate Center, China Meteorological Administration model projections included in CMIP5.
+See: [this link](http://cmip-pcmdi.llnl.gov/cmip5/availability.html).
+)
+
+The data for this example of includes climate projections for grid points near several major East Coast cities: New York, NY; Philadelphia, PA; Newark, NJ; Baltimore, MD, Providence, RI.
+For the example CMIP5 data we used, all of these communities could be linked to three grid points on the climate model.
+
+The historical projections range over the years 1980 to 2005, while the future projections range over the years 2006 to 2099.
+
+## How the package works
 
@@
 
-For example:
+This parsed version of the directory structure helps the following functions walk through all of the climate projections for different models and ensembles within the specified projections directory.
 
-```{r}
-dataFolder <- system.file("cmip5", package = "futureheatwaves")
-finalList <- acquireDirectoryStructure(dataFolder = dataFolder,
-   coordinateFilenames = coordinateFilenames,
-   tasFilenames = tasFilenames,
-   timeFilenames = timeFilenames)
+### Aggregating heatwave projections
 
-str(finalList[[1]][[1]])
-str(finalList[[1]][[2]][1])
-```
+Once the function has generated a list describing the climate models and ensemble members included in the user-specified `dataFolder` directory of climate projections, the function will run through all ensemble members of any climate model within this directory, identify and characterize the heatwaves for that ensemble member, and save the resulting dataframe to output.
+
+The package uses closures to aggregate data over all the different ensemble members of all climate models included in the directory of projection data ...
+
+### Determining thresholds for heatwave definition
+
+To determine the threshold value for each community, the function pulls climate projection data from either the default historical years (1981--2004) or from a customized range of dates, as specified by the first two values in the `dataBoundaries` option for the `gen_hw_set` function.
+This option allows the user to select years in the recent past to use when determining these threshold temperatures.
+[Can we use future years for this, too?]
+
+All climate projection data for the grid point closest to each community is pulled and use to estimate the community-specific thresholds.
+The percentile value calculated for each community is either a default of 0.98 or is specified by the user using the `probThreshold` option in the `gen_hw_set` function.
+
+These community-specific thresholds are then passed through to later calls in the function, where they are used to identify heatwaves in a time series of climate projection data.
+
+### Identifying heatwaves in an ensemble member's projection data
+
+...
+
+### Characterizing heatwaves
+
+...
+
+### Writing out results to the user's local computer
+
+...
+
+### Applying user-specified functions across all heatwave datasets
+
+...
+
+## Caveats
+
+- Heatwaves near the start / end of the dataset
+- Downscaling
+
+## Possible future developments
+
+- Allowing netCDF for input files
+- Allowing to specify exact date for boundaries, not year (would allow easier use for southern hemisphere-- reduce problems with "edge-case" heatwaves)

### 2016-02-03 R/mapCities.R

@@ -12,25 +12,27 @@
 #' @importFrom dplyr %>%
 map_grid <- function(plot_model, out){
         cities <- read.csv(paste(out, "locationList.csv", sep = "/")) %>%
-                mutate(long = long - 360,
-                       long_grid = long_grid - 360) %>%
-                filter(model == plot_model)
+                dplyr::mutate(long = long - 360,
+                             long_grid = long_grid - 360) %>%
+                dplyr::filter(model == plot_model)
 
         latlong <- unique(cities[ , c("long_grid", "lat_grid")])
-        states <- map_data("state")
+        states <- ggplot2::map_data("state")
 
-        map <- ggplot()
-        map <- map + geom_polygon(data = states,
-                                  aes(x = long, y = lat, group = group),
+        map <- ggplot2::ggplot()
+        map <- map + ggplot2::geom_polygon(data = states,
+                                  ggplot2::aes(x = long, y = lat, group = group),
                                   colour = "lightgray", fill = "white")
-        map <- map + geom_point(data = latlong,
-                                aes(x = long_grid, y = lat_grid),
-                                color = 132, alpha = 0.6)
-        map <- map + geom_segment(data = cities, aes(x = long, y = lat,
+        map <- map + ggplot2::geom_point(data = latlong,
+                                         ggplot2::aes(x = long_grid, y = lat_grid),
+                                         color = 132, alpha = 0.6)
+        map <- map + ggplot2::geom_segment(data = cities,
+                                           ggplot2::aes(x = long, y = lat,
                                                      xend = long_grid,
                                                      yend = lat_grid),
                                   size = 0.9, alpha = 0.6, color = 132)
-        map <- map + coord_map("albers", lat0=30, lat1=40) + theme_map()
-        map <- map + ggtitle(plot_model)
+        map <- map + ggplot2::coord_map("albers", lat0=30, lat1=40) +
+                ggthemes::theme_map()
+        map <- map + ggplot2::ggtitle(plot_model)
         return(map)
 }

### 2016-02-03 inst/doc/futureheatwaves.R

@@ -32,6 +32,10 @@ library(futureheatwaves)
 ## ----eval = FALSE--------------------------------------------------------
 #  data(hw_datafr)
 
+## ----fig.width = 5, message = FALSE, warning = FALSE---------------------
+out <- system.file(package = "futureheatwaves")
+map_grid(plot_model = "bcc1", out = out)
+
 ## ---- eval = FALSE-------------------------------------------------------
 #  gen_hw_set(out = "example_results",
 #             dataFolder = "cmip5" ,

### 2016-02-16 inst/doc/futureheatwaves.Rmd
Work on vignette

@@
 
-The `futureheatwaves` package can process climate data files that you have stored locally to generate a list of heatwaves in each projection as well as characteristics of each heatwave (e.g., length, intensity, timing in the year).
-The package generates community-specific dataframes based on an input file listing each community and its latitude and longitude.
-Files with the heatwaves identified and characterized under each ensemble member of each climate model are output as files in a directory specified by the user.
+Often, it is of interest in climate impact research to explore impact estimates for climate projections from many different climate models, some of which have multiple ensemble members.
+The scope of this data means that researchers often must spend a lot of time writing code scripts to process the data from different climate models or ensemble members.
+This is the case, for example, in studying the climate impacts of heatwaves, since multi-day heatwave events must be indentified and characterized for each cliamte model and ensemble member before impacts can be estimated.
+We have created this package to automate this process in a way that still allows extensive customization by the user in choices like how to define a heatwave.
 
-This function requires you to have data on community locations and climate projection data set up in a specific way on your local computer.
+The `futureheatwaves` package can process climate files, which are stored locally on your computer, to generate a list of the heatwaves in each projection as well as characteristics of each heatwave (e.g., length, intensity, timing in the year).
+You can identify heatwaves based either on a default definition (two or more days with temperatures at or above the community's $98^{th}$ percentile of year-round temperature), or you can write your own R function to use any heatwave definition of your choice, or to explore several heatwave definitions.
+The package identifies community-specific heatwaves, based on an input file listing each community and its latitude and longitude.
+of each climate model are output as files in a directory specified by the user.
+
+Once you create these dataframes of community-specific heatwaves, another function in the package allows you to apply custom functions across all the heatwave dataframes that were generated.
+This functionality can be used to generate summary statistics (e.g., determine average heatwave length or total heatwave days), or can be used to apply more complex functions (e.g., apply epidemiologic effect estimates across the heatwaves to generate health impact estimates).
+
+This package does require some pre-processing, in terms of getting climate projection data into a specific format and storing files in a specific directory structure.
+However, once the user has completed this pre-processing and directory set-up, the package allows the user to quickly process through data from many projections, generate heatwave datasets for each, and apply custom functions across each heatwave definition.
+
+This package requires you to have data on community locations and climate projection data set up in a specific way on your local computer.
 You then use arguments in the main package function, `gen_hw_set`, to direct the function to these files so it can process them and output heatwave projections.
@@
 
+### Example data
+
+We have included data files in the package to serve as example files so users can try out this package before applying it to their own directories.
+The files are all included as comma-separated files, rather than as saved R objects, to allow users to see how to create the appropriate directory set-up for this package.
+
+We used, as an example, data from two climate models, the Beijing Climate Center (BCC) and the National Center for Atmospheric Research's (NCAR's) Community Climate System Model (CCSM).
+We include one ensemble member from BCC (r1i1p1) and two from CCSM (r1i1p1 and r2i1p1).
+
+... the Beijing Climate Center, China Meteorological Administration model projections included in CMIP5. See: [this link](http://cmip-pcmdi.llnl.gov/cmip5/availability.html). 
+)
+
+[More on NCAR model]
+
+To ensure that the size of this example data is reasonably small, we have only included projection data for model grid points near several major East Coast cities: New York, NY; Philadelphia, PA; Newark, NJ; Baltimore, MD, Providence, RI.
+Further, to keep the file sizes reasonably small, the historical projections range over the years ...
+to ..., while the future projections range over the years ...
+to ...
+.
+
+These location of these files within the downloaded package be determined using  `system.file`:
+
+```{r}
+system.file("cities.csv", package = "futureheatwaves")
+```
+
+In the later sections of this vignette, we show how to use the package functions with these example files as inputs.
+
 ### Required files and directory set-up
@@
 First, you must have a comma-separated (csv) file the gives unique identifiers for each community for which you wish to make projections, as well as their latitude and longitude.
-The latitude and longitude should be expressed in the same way that latitude and longitude are expressed in the climate projection files to be processed.
+The latitude and longitude should be expressed in the same way that latitude and longitude are expressed in the climate projection files to be processed (e.g., if latitude for New York City is expressed as 286 in the climate projection files, it cannot be expressed as -74 or 74 W in the community file).
 The community location file should have three columns, with the following column names:
@@
 - `lat`: The communities' latitudes
-- `long`: The communities' longitudes
+- `lon`: The communities' longitudes
 
@@
 ```
-"city","lat","long"
+"city","lat","lon"
 "balt",39.3008,283.3894
@@
 
-The location of this community locations file will be specified to the main function, `gen_hw_set`, using the argument `citycsv`.
+You will specify the location of this community locations file within your computer file directory when running the main function (`gen_hw_set`) by using the argument `citycsv`.
 For example, if I had this file saved as "my_cities.csv" in my working directory, I would specify `citycsv = "my_cities.csv"` in the `gen_hw_set` function call.
+You can use either absolute or relative filepaths in this argument.
+
+If your community locations file has different column names for the latitude and longitude columns, you can specify this using the `lat_lon_colnames` argument in `gen_hw_set`.
+For example, if these columns were named "latitude" and "longitude", you could specify `lat_lon_colnames = c("latitude", "longitude")` when calling `gen_hw_set`.
+
+#### Directory structure for climate projection files
+
+For these functions to work correctly, you must have climate projection files saved in a specific structure locally on your computer.
+An example of a directory with the required set-up is given as the `cmip5` directory that comes installed with this package.
+To identify the location on your own computer of this directory, you can run:
+
+```{r}
+system.file("cmip5", package = "futureheatwaves")
+```
+
+You must have a directory that has within it separately subdirectories names "rcp85" and "historical" that contain, respectively, all files drawn from projections for future dates ("rcp85") and all for dates up to 2005 ("historical").
+Within each of these subdirectories, you need a subdirectory for each climate model.
+Within each climate model directory, you need subdirectories for each ensemble member of the model (even if the model only has one ensemble member).
+Within each ensemble-member directory, you will need three comma-separated files: one file with the climate model output, with columns corresponding to climate grid point and rows corresponding to time, one file that gives the time values corresponding to each row in the climate model output file, and one file giving the latitude and longitude corresponding to the grid locations used in the columns of the climate model output file.
+
+For example, the following shows the file structure for the files in the example data included with the package.
+This example directory includes separate historical (with years 1980--2004) and RCP8.5 (with years 2006--2099) directories, each of which has the required climate projection, grid point locations, and projection time files for two separate models (`bcc1` and `ccsm`) with one and two separate ensemble members, respectively(`r1i1p1` for `bcc1`; `r1i1p1` and `r2i1p1` for `ccsm`).
+The user specifies that name and location of the directory (through the `dataFolder` option in `gen_hw_set`) as well as the names and year ranges of the two subdirectories (through the `dataDirectories` option in `gen_hw_set`; e.g., `dataDirectories = list("historical" = c(1980, 2004), "rcp85" = c(2006, 2099))` if the two subdirectories have the names "historical" and "rcp85").
+
+- `cmip5`
+    - `historical`
+        - `bcc1`
+            - `r1i1p1`
+                - `tas_NorthAmerica_12mo.csv`
+                - `latitude_longitude_NorthAmerica_12mo.csv`
+                - `time_NorthAmerica_12mo.csv`
+        - `ccsm`
+            - `r1i1p1`
+                - `tas_NorthAmerica_12mo.csv`
+                - `latitude_longitude_NorthAmerica_12mo.csv`
+                - `time_NorthAmerica_12mo.csv`
+    - `rcp85`
+        - `bcc1`
+            - `r1i1p1`
+                - `tas_NorthAmerica_12mo.csv`
+                - `latitude_longitude_NorthAmerica_12mo.csv`
+                - `time_NorthAmerica_12mo.csv`
+        - `ccsm`
+            - `r1i1p1`
+                - `tas_NorthAmerica_12mo.csv`
+                - `latitude_longitude_NorthAmerica_12mo.csv`
+                - `time_NorthAmerica_12mo.csv`
+            - `r2i1p1`
+                - `tas_NorthAmerica_12mo.csv`
+                - `latitude_longitude_NorthAmerica_12mo.csv`
+                - `time_NorthAmerica_12mo.csv`
+
+Notice that each subdirectory has all required levels-- for example, even though the `bcc1` model only has one ensemble member (in this example), we've still included a directory level for ensembles in the directory structure.
+Also notice that the final subdirectory always includes three files (with climate projections, grid point locations, and times; these files are explained in the next section), even if multiple ensemble members of the same climate model share the same grid point locations or times.
+
+Throughout the directory, subdirectories should be named using names reflecting the climate model they contain.
+The output files will be written using the same type of directory structure and the same names, so this output will be easier to interpret and use if the directory uses meaningful names for subdirectories.
+
+Once you have your files set up in this structure, you will specify the location of the start of the directory using the `dataFolder` argument in the `gen_hw_set` function.
+In the example directory structure above, if the `cmip5` directory was in your current working directory, you would specify `dataFolder = "cmip5"` when calling `gen_hw_set`.
 
@@
 You will need to specify this filename in the `timeFilenames` argument when running `gen_hw_set`.
-
-#### Directory structure for climate projection files
-
-For these functions to work correctly, you must have the climate projections saved in a specific structure locally on your computer.
-An example of a directory with the required set-up is given as the `cmip5` directory in the `inst` directory of this package.
-
-You must have a directory that has within it separately subdirectories names "rcp85" and "historical" that contain, respectively, all files drawn from projections for future dates ("rcp85") and all for dates up to 2005 ("historical").
-Within each of these subdirectories, you need a subdirectory for each climate model.
-Within each climate model directory, you need subdirectories for each ensemble member of the model (even if the model only has one ensemble member).
-Within each ensemble-member directory, you will need three comma-separated files: one file with the climate model output, with columns corresponding to climate grid point and rows corresponding to time, one file that gives the time values corresponding to each row in the climate model output file, and one file giving the latitude and longitude corresponding to the grid locations used in the columns of the climate model output file.
-
-For example, the following shows the file structure for the files in the example data included with the package.
-This example directory includes separate historical (with years 1980--2004) and RCP8.5 (with years 2006--2099) directories, each of which has the required climate projection, grid point locations, and projection time files for two separate models (`bcc1` and `bcc1-a`) with one and two separate ensemble members, respectively(`r1i1p1` for `bcc1`; `r1i1p1` and `r1i1p1-a` for `bcc1-a`).
-Notice that each subdirectory has all required levels-- for example, even though the `bcc1` model only has one ensemble member (in this example), we've still included a directory level for ensembles in the directory structure.
-
-- `cmip5`
-    - `historical`
-        - `bcc1`
-            - `r1i1p1`
-                - `tas_NorthAmerica_12mo.csv`
-                - `latitude_longitude_NorthAmerica_12mo.csv`
-                - `time_NorthAmerica_12mo.csv`
-        - `bcc1-a`
-            - `r1i1p1`
-                - `tas_NorthAmerica_12mo.csv`
-                - `latitude_longitude_NorthAmerica_12mo.csv`
-                - `time_NorthAmerica_12mo.csv`
-            - `r1i1p1-a`
-                - `tas_NorthAmerica_12mo.csv`
-                - `latitude_longitude_NorthAmerica_12mo.csv`
-                - `time_NorthAmerica_12mo.csv`
-    - `rcp85`
-        - `bcc1`
-            - `r1i1p1`
-                - `tas_NorthAmerica_12mo.csv`
-                - `latitude_longitude_NorthAmerica_12mo.csv`
-                - `time_NorthAmerica_12mo.csv`
-        - `bcc1-a`
-            - `r1i1p1`
-                - `tas_NorthAmerica_12mo.csv`
-                - `latitude_longitude_NorthAmerica_12mo.csv`
-                - `time_NorthAmerica_12mo.csv`
-            - `r1i1p1-a`
-                - `tas_NorthAmerica_12mo.csv`
-                - `latitude_longitude_NorthAmerica_12mo.csv`
-                - `time_NorthAmerica_12mo.csv`
-
-Throughout the directory, subdirectories should be named using names reflecting the climate model or ensemble members they contain.
-The output files will be written using the same type of directory structure and the same names, so this output will be easier to interpret and use if the directory uses meaningful names for subdirectories.
-
-Once you have your files set up in this structure, define the following variables, you will specify the location of the start of the directory using the `dataFolder` argument in the `gen_hw_set` function.
-In the example directory structure above, if the `cmip5` directory was in your current working directory, you would specify `dataFolder = "cmip5"` when calling `gen_hw_set`.
 
@@
 
-## Example data
-
-We used, as an example, data from the Beijing Climate Center, China Meteorological Administration model projections included in CMIP5.
-See: [this link](http://cmip-pcmdi.llnl.gov/cmip5/availability.html).
-)
-
-The data for this example of includes climate projections for grid points near several major East Coast cities: New York, NY; Philadelphia, PA; Newark, NJ; Baltimore, MD, Providence, RI.
-For the example CMIP5 data we used, all of these communities could be linked to three grid points on the climate model.
-
-The historical projections range over the years 1980 to 2005, while the future projections range over the years 2006 to 2099.
-
 ## How the package works

### 2016-02-16 inst/doc/futureheatwaves.R

@@ -1,6 +1,12 @@
 ## ----echo = FALSE, message = FALSE---------------------------------------
 library(futureheatwaves)
 
+## ------------------------------------------------------------------------
+system.file("cities.csv", package = "futureheatwaves")
+
+## ------------------------------------------------------------------------
+system.file("cmip5", package = "futureheatwaves")
+
 ## ----eval = FALSE--------------------------------------------------------
 #  gen_hw_set(out = "example_results",
 #             dataFolder = "cmip5" ,

### 2016-02-19 R/mapCities.R
Make `mapCities` more robust to different longitude column names

@@ -11,12 +11,14 @@
 #'
 #' @importFrom dplyr %>%
 map_grid <- function(plot_model, out){
-        cities <- read.csv(paste(out, "locationList.csv", sep = "/")) %>%
-                dplyr::mutate(long = long - 360,
-                             long_grid = long_grid - 360) %>%
+        cities <- read.csv(paste(out, "locationList.csv", sep = "/"),
+                           col.names = c("city","lat", "lon",
+                                         "lat_grid", "lon_grid", "model")) %>%
+                dplyr::mutate(lon = lon - 360,
+                             lon_grid = lon_grid - 360) %>%
                 dplyr::filter(model == plot_model)
 
-        latlong <- unique(cities[ , c("long_grid", "lat_grid")])
+        latlong <- unique(cities[ , c("lon_grid", "lat_grid")])
         states <- ggplot2::map_data("state")
 
         map <- ggplot2::ggplot()
@@ -24,14 +26,14 @@ map_grid <- function(plot_model, out){
                                   ggplot2::aes(x = long, y = lat, group = group),
                                   colour = "lightgray", fill = "white")
         map <- map + ggplot2::geom_point(data = latlong,
-                                         ggplot2::aes(x = long_grid, y = lat_grid),
+                                         ggplot2::aes(x = lon_grid, y = lat_grid),
                                          color = 132, alpha = 0.6)
         map <- map + ggplot2::geom_segment(data = cities,
-                                           ggplot2::aes(x = long, y = lat,
-                                                     xend = long_grid,
+                                           ggplot2::aes(x = lon, y = lat,
+                                                     xend = lon_grid,
                                                      yend = lat_grid),
                                   size = 0.9, alpha = 0.6, color = 132)
-        map <- map + ggplot2::coord_map("albers", lat0=30, lat1=40) +
+        map <- map + ggplot2::coord_map("albers", lat0 = 30, lat1 = 40) +
                 ggthemes::theme_map()
         map <- map + ggplot2::ggtitle(plot_model)
         return(map)

### 2016-02-19 R/main.R
Close #88

@@ -213,22 +213,6 @@ gen_hw_set <- function(out,
 #'    column and times by row; 2. A file with the longitude and latitude
 #'    of each grid point in the projection file and; 3. A file with the
 #'    date of each of the rows in the projection file.)
-#'
-#' @examples
-#' out <- "~/tmp/results"
-#' dataFolder <- "inst/cmip5/"
-#' citycsv <- "inst/cities.csv"
-#' coordinateFilenames <- "latitude_longitude_NorthAmerica_12mo.csv"
-#' tasFilenames <- "tas_NorthAmerica_12mo.csv"
-#' timeFilenames <- "time_NorthAmerica_12mo.csv"
-#' IDheatwavesFunction <- FALSE
-#' dataBoundaries <- FALSE
-#' referenceBoundaries <- FALSE
-#'
-#' check_params(out, dataFolder, citycsv,
-#'    coordinateFilenames, tasFilenames, timeFilenames,
-#'    IDheatwavesFunction,
-#'    dataBoundaries, referenceBoundaries)
 check_params <- function(out,
                          dataFolder,
                          dataDirectories,

### 2016-02-19 R/preparation.R

@@ -20,22 +20,6 @@
 #'    of this function. All other files will be removed when creating the
 #'    directory structure.
 #'
-#' @examples
-#' dataFolder <- system.file("cmip5", package = "futureheatwaves")
-#' coordinateFilenames <- "latitude_longitude_NorthAmerica_12mo.csv"
-#' tasFilenames <- "tas_NorthAmerica_12mo.csv"
-#' timeFilenames <- "time_NorthAmerica_12mo.csv"
-#'
-#' finalList <- acquireDirectoryStructure(dataFolder = dataFolder,
-#'    coordinateFilenames = coordinateFilenames,
-#'    tasFilenames = tasFilenames,
-#'    timeFilenames = timeFilenames)
-#'
-#' finalList[[1]][[1]]
-#' finalList[[1]][[2]][1]
-#'
-#' @export
-#'
 #' @importFrom dplyr %>%
 acquireDirectoryStructure <- function(dataFolder, coordinateFilenames,
                                       tasFilenames, timeFilenames,
@@ -121,24 +105,6 @@ acquireDirectoryStructure <- function(dataFolder, coordinateFilenames,
 #'    experiment hierarchy. The third element is the hierarchy of the future
 #'    projection directory for the model. The second and third elements are
 #'    return values of \code{\link{buildStructureExperiments}}.
-#'
-#' @examples
-#' modelName <- "bcc1"
-#' experiments <- c("historical", "rcp85")
-#'
-#' dataFolder <- system.file("cmip5", package = "futureheatwaves")
-#' dataFolder <- paste0(dataFolder, "/")
-#'
-#' coordinateFilenames <- "latitude_longitude_NorthAmerica_12mo.csv"
-#' tasFilenames <- "tas_NorthAmerica_12mo.csv"
-#' timeFilenames <- "time_NorthAmerica_12mo.csv"
-#'
-#' buildStructureModels(model = modelName,
-#'                      experiments = experiments,
-#'                      dataFolder,
-#'                      coordinateFilenames = coordinateFilenames,
-#'                      tasFilenames = tasFilenames,
-#'                      timeFilenames = timeFilenames)
 buildStructureModels <- function(modelName, experiments,
                                  dataFolder,
                                  coordinateFilenames, tasFilenames,
@@ -171,25 +137,6 @@ buildStructureModels <- function(modelName, experiments,
 #'    is a return value of the \code{\link{buildStructureEnsembles}} function for
 #'    one of the ensemble members in that experiment of that climate model
[…]

### 2016-02-19 R/IDHeatwaves.R
Close #84

@@ -244,7 +244,7 @@ IDHeatwavesAlternative <- function(threshold, datafr){
 #'    (\code{hw.number}), and whether the day was the first day in a heatwave
 #'    (\code{first.hw.day}, 0 /1).
 #'
-#' @note This function gives identical results to the \link{\code{IDHeatwavesR}}
+#' @note This function gives identical results to the \code{\link{IDHeatwavesR}}
 #'    function, but should take less time to run.
 #'
 #' @export

### 2016-02-19 R/IO.R
Close #86

@@ -40,6 +40,7 @@ process_cities_file <- function(cities, lat_lon_colnames){
 #'    the file paths to (1) the latitude and longitude file; (2) the
 #'    climate projection file; and (3) the projection dates file
 #'    for the selected climate model.
+#' @inheritParams processModel
 #'
 #' @return A dataframe of the latitude and longitude data of the ensemble
 #'    with columns for the latitude (first column) and longitude (second
@@ -60,6 +61,7 @@ readLatLong <- function(ensemble, global){
 #' \code{\link{gen_hw_set}}.
 #'
 #' @inheritParams readLatLong
+#' @inheritParams processModel
 #'
 #' @return A dataframe of climate projection data where each column
 #'    corresponds to a climate model grid point and each row corresponds
@@ -77,6 +79,7 @@ readtas <- function(ensemble, global){
 #' \code{\link{gen_hw_set}}.
 #'
 #' @inheritParams readLatLong
+#' @inheritParams processModel
 #'
 #' @return A dataframe of dates corresponding to climate projection data,
 #'    where each row gives a date, split into columns of day in the
@@ -139,8 +142,10 @@ createEnsembleWriter <- function(modelName, global, custom){
 #' climate models included in the analysis, including the number of
 #' historical (1980--2004) and future (2006--2099) ensemble members.
 #'
-#' @param modelInfoAccumulator The dataframe that accumulates the number of
-#'    ensembles for each model.
+#' @param modelInfoAccumulator A dataframe with information about the number of
+#'    ensembles for each climate model.
+#' @param locationList A dataframe with information about community locations
+#'    and the closest grid point from each climate model to each community.
 #' @inheritParams processModel
 #'
 #' @return ...

### 2016-02-19 R/RcppExports.R

@@ -36,6 +36,7 @@ storeHeatwaveEntry <- function(index, hwSize, hwCounter, hw, hwNumber) {
 #' @param potentialHeatwave Size of the potential heatwave that turned out not to be a heatwave.
 #' @param hw A reference of the vector that contains the heatwaves
 #' @param hwNumber A reference of the vector that contains the heatwave numbers.
+#' @inheritParams storeHeatwaveEntry
 #'
 #' @export
 storeZeroes <- function(index, potentialHeatwave, hw, hwNumber) {

### 2016-02-19 R/applyHeatwaves.R

@@ -17,6 +17,7 @@
 #'    applied to the full dataframe.
 #' @param city_specific TRUE / FALSE: Should the function be calculated
 #'    separately for each community within the dataframe?
+#' @inheritParams gen_hw_set
 #'
 #' @return A dataframe with the value output by the user-specified
 #'    function, as applied to every dataframe of heatwaves and characteristics

### 2016-02-19 R/main.R

@@ -63,6 +63,22 @@
 #' @param lat_lon_colnames A character vector of length two with the column names
 #'    in the \code{cities} dataframe for latitude (first vector element) and
 #'    longitude (second vector element)
+#' @param models_to_run A character vector with either "all" (the default),
+#'    in which case the function runs through all models in the specified data
+#'    directory, or the names of the models to run, using the names of each
+#'    model's subdirectory within the data directory (e.g.,
+#'    \code{c("bcc1", "ccsm")})
+#' @param dataDirectories A list object, with two elements for each of the
+#'    two subdirectories included in the main directory. Typically, these will
+#'    be a historical directory and a climate projection directory. Each element
+#'    of the list should be named with the names of the subdirectories and
+#'    should provide a numeric vector with the starting year and ending year of
+#'    the data within each of the two subdirectories (e.g.,
+#'    \code{list("historical" = c(1980, 2004), "rcp85" = c(2006, 2099))})
+#' @param threshold_ensemble A character vector giving the name of the ensemble
+#'    member that should be used when determining the threshold temperature for
+#'    the heatwave definition for each community for each climate model (e.g.,
+#'    \code{"r1i1p1"})
 #'
 #' @return This function returns a dataframe listing the name of each climate
 #'    model used, as well as the number of historical and future projection
@@ -276,6 +292,7 @@ check_params <- function(out,
 #'    c(historical period earliest year, historical period latest year,
 #'    future projection period earliest year, future projection period
 #'    latest year).
+#' @inheritParams gen_hw_set
 checkCustomBounds <- function(boundList, dataDirectories){
 
         if(class(boundList) != "numeric"){

### 2016-02-19 R/preparation.R

@@ -96,8 +96,6 @@ acquireDirectoryStructure <- function(dataFolder, coordinateFilenames,
 #'    specified by the user in \code{\link{gen_hw_set}}.
 #' @param experiments Character vector of the experiment(s) of interest.
 #'    Possible variables are "historical", "rcp85", or both.
-#' @param dataPath Character string of the file path to the directory
-#'    containing the climate projections. Must include the final `/`.
 #' @inheritParams gen_hw_set
 #'
 #' @return A list of length 3. The first element is the name of the model
@@ -130,6 +128,8 @@ buildStructureModels <- function(modelName, experiments,
 #'
 #' @param experiment Character string of the experiment of interest.
 #'    Possible variables are "historical" or "rcp85".
+#' @param dataPath Character string of the file path to the directory
+#'    containing the climate projections. Must include the final `/`.
 #' @inheritParams buildStructureModels
 #' @inheritParams gen_hw_set
 #'

### 2016-02-19 R/process.R

@@ -117,13 +117,6 @@ processModel <- function(model, global, custom, accumulators){
 #' functions to identify heatwaves in the projections and are also saved
 #' to file through this function to have for future reference.
 #'
-#' @param reference FALSE, if the user has not specified custom reference
-#'    boundaries through the \code{referenceBoundaries} argument in
-#'    \code{\link{gen_hw_set}}, otherwise a vector with four elements
-#'    (earliest historical year, latest historical year,
-#'    earliest future projection year, latest future projection year)
-#'    giving the user-specified custom date boundaries for the
-#'    heatwave projections.
 #' @inheritParams processModel
 #'
 #' @return A list with two elements: (1) a vector with one element for
@@ -212,6 +205,17 @@ processReference <- function(model, global, custom){
 #'    in that community's heatwave definition.These are typically automatically
 #'    determine during the run of the \code{gen_hw_set} function based on
 #'    historical climate projections within the community.
+#' @param reference FALSE, if the user has not specified custom reference
+#'    boundaries through the \code{referenceBoundaries} argument in
+#'    \code{\link{gen_hw_set}}, otherwise a vector with four elements
+#'    (earliest historical year, latest historical year,
+#'    earliest future projection year, latest future projection year)
+#'    giving the user-specified custom date boundaries for the
+#'    heatwave projections.
+#' @param reference_dates Either FALSE, if the same dates are being used for the
+#'    projection period and for the period to use as a reference for
+#'    relative heatwave characteristics, or a numeric vector with the start and
+#'    end years to use for the reference period
 #' @inheritParams buildStructureModels
 #' @inheritParams processModel
 #' @inheritParams processThresholds
@@ -269,6 +273,7 @@ processProjections <- function(ensemble, modelName, ensembleWriter,
 #' @inheritParams buildStructureModels
 #' @inheritParams processThresholds
 #' @inheritParams processProjections
+#' @inheritParams getBounds
 #'
 #' @return Returns a list with the following:
 #'    \code{locations}: A numberical vector with the indices of the columns
@@ -371,6 +376,8 @@ closest_point <- function(city, latlong){
 #'    corresponds to the row with the same row number in the climate
 #'    projection dataframe. Read in by
 #'    \code{\link{processEnsemble}}.
+#' @param type A character vector giving the type of boundaries that are being
+#'   checked. Possible values are "threshold", "projections", and "reference".
[…]

### 2016-02-19 R/processHeatwaves.R

@@ -150,7 +150,9 @@ consolidate <- function(hwDataframeList){
 #' This function takes a dataframe with identified heatwaves and returns
 #' a dataframe that lists and characterizes all of the heatwaves.
 #'
-#' @param A dataframe with the following columns: (1) \code{date}: Date
+#' @param city A character vector with the identification of the community
+#'    being processed
+#' @param heatwaves A dataframe with the following columns: (1) \code{date}: Date
 #'    of each observation, in class "Date"; (2) \code{tmpd}: Temperature
 #'    (in typical runs of this function, this will be in degrees Fahrenheit);
 #'    (3) \code{hw}: A binary variable designating whether a day is in a

### 2016-02-19 src/IdentifyHeatwaves.cpp

@@ -98,6 +98,7 @@ void storeHeatwaveEntry(int index, int hwSize, int hwCounter, std::vector<int>& 
 //' @param potentialHeatwave Size of the potential heatwave that turned out not to be a heatwave.
 //' @param hw A reference of the vector that contains the heatwaves
 //' @param hwNumber A reference of the vector that contains the heatwave numbers.
+//' @inheritParams storeHeatwaveEntry
 //'
 //' @export
 // [[Rcpp::export]]

### 2016-02-28 R/IDHeatwaves.R
Close #97.

@@ -165,6 +165,9 @@ IDHeatwavesR <- function(threshold, datafr, numDays){
 IDHeatwavesAlternative <- function(threshold, datafr, numDays){
 
         days <- numDays
+        if(threshold <= 80){
+                threshold <- 80
+        }
 
         # Add names to the dataframe
         colnames(datafr) <- c("date", "tmpd")

### 2016-02-28 R/IDHeatwaves.R
Update to allow `numDays` option throughout

@@ -52,6 +52,7 @@ IDheatwaves <- function(threshold, datafr, global, custom){
 #'
 #' @inheritParams closest_point
 #' @inheritParams IDheatwaves
+#' @inheritParams gen_hw_set
 #'
 #' @return Returns the dataframe entered as \code{datafr}, but with new
 #'    columns providing heat wave identifiers. The returned dataframe will
@@ -142,6 +143,7 @@ IDHeatwavesR <- function(threshold, datafr, numDays){
 #'
 #' @inheritParams closest_point
 #' @inheritParams IDheatwaves
+#' @inheritParams gen_hw_set
 #'
 #' @return Returns the dataframe entered as \code{datafr}, but with new
 #'    columns providing heat wave identifiers. The returned dataframe will
@@ -238,6 +240,7 @@ IDHeatwavesAlternative <- function(threshold, datafr, numDays){
 #'
 #' @inheritParams closest_point
 #' @inheritParams IDheatwaves
+#' @inheritParams gen_hw_set
 #'
 #' @return Returns the dataframe entered as \code{datafr}, but with new
 #'    columns providing heat wave identifiers. The returned dataframe will

### 2016-02-28 tests/testthat/test_IDHeatwaves.R

@@ -2,8 +2,9 @@ library(futureheatwaves)
 context("Identifying heatwaves")
 
 data("datafr", package = "futureheatwaves")
-hws_r <- IDHeatwavesR(datafr = datafr, threshold = 84.632)
-hws_cpp <- IDHeatwavesCPPwrapper(datafr = datafr, threshold = 84.632)
+hws_r <- IDHeatwavesR(datafr = datafr, threshold = 84.632, numDays = 2)
+hws_cpp <- IDHeatwavesCPPwrapper(datafr = datafr, threshold = 84.632,
+                                 numDays)
 
 test_that("C++ and R ID heatwaves identify same heatwaves", {
         expect_equal(hws_r$hw, hws_cpp$hw)

### 2016-02-28 tests/testthat/test_IDHeatwaves.R
Fix test to allow `numDays`

@@ -4,7 +4,7 @@ context("Identifying heatwaves")
 data("datafr", package = "futureheatwaves")
 hws_r <- IDHeatwavesR(datafr = datafr, threshold = 84.632, numDays = 2)
 hws_cpp <- IDHeatwavesCPPwrapper(datafr = datafr, threshold = 84.632,
-                                 numDays)
+                                 numDays = 2)
 
 test_that("C++ and R ID heatwaves identify same heatwaves", {
         expect_equal(hws_r$hw, hws_cpp$hw)

### 2016-02-29 R/process.R
Close #100

@@ -11,7 +11,7 @@
 #'    \code{acquireDirectoryStructure}.
 #' @param global An list object created by \code{\link{gen_hw_set}} that
 #'    includes user specifications (e.g., the path to the output directory, the
-#'    path to the input climate projections, the dataframe withcity
+#'    path to the input climate projections, the dataframe with city
 #'    locations).
 #' @param custom An list object created by \code{\link{gen_hw_set}} that
 #'    includes user specifications (e.g., the name of the R function to
@@ -20,11 +20,15 @@
 #'    definition, alternative upper and lower year boundaries
 #'    for the projection period of the heat wave datasets being generated).
 #' @param accumulators The closure generated by
-#'    \code{\link{createAccumulators}}.
+#'    \code{\link{createAccumulators}} that allows you to append model
+#'    information and grid location data as you process through models to
+#'    a growing list.
+#' @inheritParams gen_hw_set
 #'
 #' @return A list object with processed projections, as well as various other
 #'    elements needed for further processing to identify heat waves.
-processModel <- function(model, global, custom, accumulators){
+processModel <- function(model, global, custom, accumulators,
+                         dataDirectories){
 
         # Acquire characteristics of each experiment of the model
         modelName <- model[[1]]
@@ -63,9 +67,9 @@ processModel <- function(model, global, custom, accumulators){
 
         # Process the projection data
 
-        # First, check whether this should process files from the "historical"
-        # or the "rcp85" directory for the model
-        if(custom$getBounds[3] < 2005){
+        # First, check which of the two experiment subdirectories should be
+        # used for the projections
+        if(custom$getBounds[3] <= dataDirectories[[1]][2]){
                 proj_ens <- histens
         } else {
                 proj_ens <- rcpens
@@ -98,15 +102,12 @@ processModel <- function(model, global, custom, accumulators){
 #' @inheritParams processModel
 #'
 #' @return A list with two elements: (1) a vector with one element for
-#'    eachcity included in the user-specifiedcity location
+#'    each city included in the user-specified city location
 #'    file, with each element value giving the threshold temperature for the
-#'    heat wave definition for acity, in the same order thatcities
-#'    are listed in the user-specificed \code{citycsv} file and (2) a
-#'    \code{processModel} object, which is FALSE if the user did not
-#'    specify custom year boundaries with \code{dataBoundaries} in
-#'    \code{\link{gen_hw_set}} and otherwise is ... .
-#'    This function also writes threshold temperatures to files within
[…]

### 2016-02-29 R/main.R
Close #101

@@ -187,7 +187,8 @@ gen_hw_set <- function(out,
                                             timeFilenames = timeFilenames,
                                             models_to_run = models_to_run,
                                             dataDirectories = dataDirectories,
-                                            threshold_ensemble = threshold_ensemble)
+                                            threshold_ensemble = threshold_ensemble,
+                                            thresholdBoundaries = thresholdBoundaries)
 
         # Read the cities data file
         cities <- utils::read.csv(citycsv) %>%
@@ -223,7 +224,8 @@ gen_hw_set <- function(out,
         referenceEnsembles <- sapply(models, processModel,
                                      global = global,
                                      custom = custom,
-                                     accumulators = accumulators)
+                                     accumulators = accumulators,
+                                     dataDirectories = dataDirectories)
 
         # Write the model information from the model information accumulator
         out <- accumulators("return model information")

### 2016-02-29 R/mapCities.R
Update documentation with new example file location

@@ -11,7 +11,7 @@
 #'    edited using \code{ggplot2} functions.
 #'
 #' @examples
-#' out <- system.file(package = "futureheatwaves")
+#' out <- system.file("example_results", package = "futureheatwaves")
 #' map_grid(plot_model = "bcc1", out = out)
 #'
 #' @export

### 2016-03-09 R/mapCities.R
Close #107

@@ -11,7 +11,7 @@
 #'    edited using \code{ggplot2} functions.
 #'
 #' @examples
-#' out <- system.file("example_results", package = "futureheatwaves")
+#' out <- system.file("extdata/example_results", package = "futureheatwaves")
 #' map_grid(plot_model = "bcc1", out = out)
 #'
 #' @export

### 2016-03-09 tests/testthat/test_preparation.R

@@ -1,7 +1,7 @@
 library(futureheatwaves)
 context("Acquiring directory structure")
 
-dataFolder <- system.file("cmip5", package = "futureheatwaves")
+dataFolder <- system.file("extdata/cmip5", package = "futureheatwaves")
 coordinateFilenames <- "latitude_longitude_NorthAmerica_12mo.csv"
 tasFilenames <- "tas_NorthAmerica_12mo.csv"
 timeFilenames <- "time_NorthAmerica_12mo.csv"

### 2016-03-09 R/zzz.R
Delete `zzz.R`
I checked-- Hadley doesn't seem to be using this `loadRcppModules`, etc., in his `.onLoad` for `devtools`, and I was getting errors when I tried to use it here. I had been commenting out the entire file, but based on this, I'll remove it.

@@ -1,6 +1,0 @@
-# .onLoad <- function(libname, pkgname){
-#         Rcpp::loadRcppModules()
-# }
-# .onUnload <- function(libpath){
-#         library.dynam.unload("futureheatwaves", libpath)
-# }

### 2016-03-09 R/exampleApplyFunctions.R
Add examples for example functions for `apply_all_models`

@@ -18,6 +18,10 @@
 #'    used to explore heat waves using the \code{\link{apply_all_models}}
 #'    function.
 #'
+#' @examples
+#' data(hw_datafr)
+#' number_of_heatwaves(hw_datafr)
+#'
 #' @export
 number_of_heatwaves <- function(hw_datafr){
         out <- nrow(hw_datafr)
@@ -41,6 +45,10 @@ number_of_heatwaves <- function(hw_datafr){
 #'    used to explore heat waves using the \code{\link{apply_all_models}}
 #'    function.
 #'
+#' @examples
+#' data(hw_datafr)
+#' heatwave_days(hw_datafr)
+#'
 #' @export
 heatwave_days <- function(hw_datafr){
         out <- sum(hw_datafr$length)
@@ -62,6 +70,10 @@ heatwave_days <- function(hw_datafr){
 #' @note This function is an example of a function that can be created and
 #'    used to explore heat waves using the \code{\link{apply_all_models}}
 #'    function.
+#'
+#' @examples
+#' data(hw_datafr)
+#' average_length(hw_datafr)
 #'
 #' @export
 average_length <- function(hw_datafr){
@@ -88,6 +100,10 @@ average_length <- function(hw_datafr){
 #' @note This function is an example of a function that can be created and
 #'    used to explore heat waves using the \code{\link{apply_all_models}}
 #'    function.
+#'
+#' @examples
+#' data(hw_datafr)
+#' average_mean_temp(hw_datafr)
 #'
 #' @export
 average_mean_temp <- function(hw_datafr){

### 2016-03-09 R/IDHeatwaves.R
Add examples for HW id functions

@@ -72,6 +72,14 @@ IDheatwaves <- function(threshold, datafr, global, custom){
 #'    when using this function either with Southern Hemisphere cities
 #'    or when exploring exposures that, unlike heat waves, may occur very
 #'    early or late in the calendar year.
+#'
+#' @examples
+#' \donttest{
+#' data(datafr)
+#' hw_ids <- IDHeatwavesR(threshold = 80, datafr,
+#'                        numDays = 3)
+#' }
+#'
 #' @export
 IDHeatwavesR <- function(threshold, datafr, numDays){
 
@@ -138,8 +146,9 @@ IDHeatwavesR <- function(threshold, datafr, numDays){
 #' Identify heat waves in a time series
 #'
 #' This function identifies heat waves in a time series of temperature
-#' data using a heat wave definition that a heat wave must be five or more
-#' days with temperatures equal to or above some threshold temperature.
+#' data using a heat wave definition that a heat wave must be a certain
+#' number of days above the greater of either a given threshold or
+#' 80 degrees Fahrenheit.
 #'
 #' @inheritParams closest_point
 #' @inheritParams IDheatwaves
@@ -163,6 +172,14 @@ IDHeatwavesR <- function(threshold, datafr, numDays){
 #'    when using this function either with Southern Hemisphere cities
 #'    or when exploring exposures that, unlike heat waves, may occur very
 #'    early or late in the calendar year.
+#'
+#' @examples
+#' \donttest{
+#' data(datafr)
+#' hw_ids <- IDHeatwavesAlternative(threshold = 80, datafr,
+#'                                  numDays = 3)
+#' }
+#'
 #' @export
 IDHeatwavesAlternative <- function(threshold, datafr, numDays){
 
@@ -260,6 +277,11 @@ IDHeatwavesAlternative <- function(threshold, datafr, numDays){
 #'    when using this function either with Southern Hemisphere cities
 #'    or when exploring exposures that, unlike heat waves, may occur very
 #'    early or late in the calendar year.
+#'
+#' @examples
+#' data(datafr)
+#' hw_ids <- IDHeatwavesCPPwrapper(threshold = 80, datafr,
+#'                                 numDays = 3)
 #'
 #' @export
 IDHeatwavesCPPwrapper <- function(datafr, threshold, numDays){

### 2016-03-09 R/main.R
Add some examples to additional functions

@@ -114,6 +114,22 @@
 #'    projections and dates. For more information on customizing this function,
 #'    see the \code{futureheatwaves} vignette.
 #'
+#' @examples
+#' \donttest{
+#' projection_dir_location <- system.file("extdata/cmip5",
+#'                                       package = "futureheatwaves")
+#' city_file_location <- system.file("extdata/cities.csv",
+#'                                  package = "futureheatwaves")
+#' gen_hw_set(out = "example_results",
+#'            dataFolder = projection_dir_location ,
+#'            dataDirectories = list("historical" = c(1990, 1999),
+#'                                   "rcp85" = c(2060, 2079)),
+#'            citycsv = city_file_location,
+#'            coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
+#'            tasFilenames = "tas_NorthAmerica_12mo.csv",
+#'            timeFilenames = "time_NorthAmerica_12mo.csv")
+#' }
+#'
 #' @export
 #'
 #' @importFrom dplyr %>%

### 2016-03-09 README.Rmd
Close #105

---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# Overview

The `futureheatwaves` package takes a directory of climate projection files and, for each, identifies and characterizes all heat waves. The definition used to identify heat waves can be customized. Characterizations include several metrics of heat wave length, intensity, and timing in the year. Heat waves can be explored by applying custom functions across all generated heat wave files.

# Installing

The development version of this package can be installed directly from GitHub using `install_github` from the `devtools` package:

```{r eval = FALSE}
library(devtools)
install_github("geanders/futureheatwaves", build_vignettes = TRUE)
library(futureheatwaves)
```

Because this package includes some C++ functions, this installation from GitHub requires the compiling of some code. If you get an error when trying to install this package from GitHub, it may be because your computer lacks the required tools to compile that part of the code. In this case, you can either download and install the required tools or email me to get a source version of the code for your operating system (where the C++ code has already been compiled).

# Using the package

There are extensive details on using this package in the vignette that comes with the package and can be accessed using:

```{r eval = FALSE}
vignette("futureheatwaves")
```

### 2016-03-23 README.Rmd
Update README.Rmd

@@
 
-There are extensive details on using this package in the vignette that comes with the package and can be accessed using:
+There are extensive details on using this package in the package vignette, which can be accessed once you install the package using:
 
@@
 ```
+
+An uncompiled version of the vignette can be found, without having to install the package, [here](https://github.com/geanders/futureheatwaves/blob/master/vignettes/futureheatwaves.Rmd).

### 2016-03-23 inst/doc/futureheatwaves.Rmd
Work on vignette

--- 2016-01-26 inst/doc/futureheatwaves.Rmd
@@
   %\VignetteEncoding{UTF-8}
+references:
+- id: gent2011 
+  title: The Community Climate System Model Version 4
+  author: 
+  - family: Gent 
+    given: Peter R.
+  - family: Danabasoglu 
+    given: Gokhan
+  - family: Donner
+    given: Leo J. 
+  - family: others
+  container-title: Journal of Climate
+  volume: 24
+  DOI: 10.1175/2011JCLI4083.1
+  page: 4973-4991
+  type: article-journal
+  issued: 
+    year: 2011
+- id: smith2013
+  title: "Heat waves in the United States: definitions, patterns and trends"
+  author: 
+  - family: Smith
+    given: Tiffany T 
+  - family: Zaitchik
+    given: Benjamin F
+  - family: Gohlke
+    given: Julia M
+  container-title: Climatic Change
+  volume: 118
+  DOI: 10.1007/s10584-012-0659-2
+  page: 811-825
+  type: article-journal
+  issued: 
+    year: 2013
+- id: taylor2012
+  title: An Overview of CMIP5 and the Experiment Design
+  author: 
+  - family: Taylor
+    given: Karl E.
+  - family: Stouffer
+    given: Ronald J.
+  - family: Meehl
+    given: Gerald A.
+  container-title: Bulletin of the American Meteorological Society
+  volume: 93
+  DOI: 10.1175/BAMS-D-11-00094.1
+  page: 485–498
+  type: article-journal
+  issued: 
+    year: 2012
+- id: xin2013
+  title: Introduction of CMIP5 experiments carried out with the climate system models of Beijing Climate Center
+  author:
+  - family: Xin
+    given: Xiao-Ge
+  - family: Wu
+    given: Tong-Wen
+  - family: Zhang
+    given: Jie
+  container-title: Advances in Climate Change Research
+  volume: 4
+  DOI: 10.3724/SP.J.1248.2013.041
+  issue: 1
+  page: 41-49
+  type: article-journal
+  issued:
+    year: 2013
 ---
 
-```{r echo = FALSE, message = FALSE}
+```{r echo = FALSE, message = FALSE, warning = FALSE}
 library(futureheatwaves)
+library(ggplot2)
 ```
@@
 
-The `futureheatwaves` package can process climate data files that you have stored locally to generate a list of heatwaves in each projection as well as characteristics of each heatwave (e.g., length, intensity, timing in the year).
-The package generates community-specific dataframes based on an input file listing each community and its latitude and longitude.
-Files with the heatwaves identified and characterized under each ensemble member of each climate model are output as files in a directory specified by the user.
+Often, it is of interest in climate impact research to explore impact estimates for climate projections from many different climate models, some of which have multiple ensemble members.
+The scope of this data means that researchers often must spend a lot of time writing code scripts to process the data from different climate models or ensemble members.
+This is the case, for example, in studying the impacts of heat waves under climate change, since multi-day heatwave events must be indentified and characterized for each climate model and ensemble member before impacts can be estimated.
+We have created this package to automate this process in a way that still allows extensive customization by the user in choices like how to define a heatwave.
+
+The `futureheatwaves` package can process climate files, which are stored locally on your computer, to generate a list of the heatwaves in each projection as well as characteristics of each heatwave (e.g., length, intensity, timing in the year).
+You can identify heatwaves based either on a default definition (two or more days with temperatures at or above the community's $98^{th}$ percentile of year-round temperature), or you can write your own R function to use any heat wave definition of your choice, or to explore several heat wave definitions.
+The package identifies community-specific heat waves, based on an input file listing each community and its latitude and longitude.
+of each climate model are output as files in a directory specified by the user.
+
+Once you create these dataframes of community-specific heat waves, another function in the package allows you to apply custom functions across all the heat wave dataframes that were generated.
+This functionality can be used to generate summary statistics (e.g., determine average heat wave length or total heat wave days), or can be used to apply more complex functions (e.g., apply epidemiologic effect estimates across the heat waves to generate health impact estimates).
+
+This package does require some pre-processing, in terms of getting climate projection data into a specific format and storing files in a specific directory structure.
+However, once the user has completed this pre-processing and directory set-up, the package allows the user to quickly process through data from many projections, generate heat wave datasets for each, and apply custom functions across each heat wave definition.
+
+This package requires you to have data on community locations and climate projection data set up in a specific way on your local computer.
+You then use arguments in the main package function, `gen_hw_set`, to direct the function to these files so it can process them and output heat wave projections.
+The final projection files will be written out to your local computer, in a directory that you select and specify through `gen_hw_set`.
+You can then apply user-created R functions to all heat waves in the output using the `apply_all_models` function.
 
@@
 
-For these functions to work correctly, you must have the climate projections saved in a specific structure locally on your computer.
-An example of a directory with the required set-up is given as the `cmip5` directory in the `inst` directory of this package.
+### Example data
 
-You must have a directory that has within it separately subdirectories names "rcp85" and "historical" that contain, respectively, all files drawn from projections for future dates ("rcp85") and all for dates up to 2005 ("historical").
-Within each of these subdirectories, you need a subdirectory for each climate model.
-Within each climate model directory, you need subdirectories for each ensemble member of the model (even if the model only has one ensemble member).
-Within each ensemble-member directory, you will need three comma-separated files: one file with the climate model output, with columns corresponding to climate grid point and rows corresponding to time, one file that gives the time values corresponding to each row in the climate model output file, and one file giving the latitude and longitude corresponding to the grid locations used in the columns of the climate model output file.
+We have included data files in the package to serve as example files so users can try out this package before applying it to their own directories.
+The files are all included as comma-separated files, rather than as saved R objects, to allow users to see how to create the appropriate directory set-up for this package.
 
-Throughout the directory, subdirectories should be named using names reflecting the climate model or ensemble members they contain.
-The output files will be written using the same type of directory structure and the same names, so this output will be easier to interpret and use if the directory uses meaningful names for subdirectories.
+We used, as an example, data from two climate models used for the fifth Coupled Model Intercomparison Project (CMIP5) [@taylor2012]: (1) the model of the Beijing Climate Center, China Meteorological Administration (BCC) [@xin2013] and (2) the National Center for Atmospheric Research's (NCAR's) Community Climate System Model (CCSM4) [@gent2011].
+We include one ensemble member from BCC (r1i1p1) and two from CCSM (r1i1p1 and r2i1p1).
 
-Once you have your files set up in this structure, define the following variables:
+To ensure that the size of this example data is reasonably small, we have only included projection data for model grid points near several major East Coast cities: New York, NY; Philadelphia, PA; Newark, NJ; Baltimore, MD, Providence, RI.
+Further, to keep the file sizes reasonably small, the historical projections range over the years 1990 to 1999, while the future projections range over the years 2060 to 2079.
 
-- `dataFolder`: Character string with pathway to directory that contains climate projections. 
-- `citycsv`: Character string giving the filepath to a .csv file with latitude and longitude values for each city for which you would like to generate
-heatwave projections.
-- `coordinateFilenames`: Character string with name of the files containing the latitude and longitude coordinates corresponding to the columns of the time series data.
-- `tasFilenames`: Character sting with name of files containing the time series data.
-- `timeFilenames`: Character string with name of the files containing the date information corresponding to the rows of the time series data.
-
-For example, I can specify `dataFolder` and `citycsv` as:
+These files are all stored in the package's `extdata` directory. These location of these files within the downloaded package can be determined using `system.file`:
 
 ```{r}
-dataFolder <- "inst/cmip5/" 
-citycsv <- "inst/cities.csv"
+system.file("extdata/cities.csv", package = "futureheatwaves")
 ```
 
-Within the ensemble subdirectories in my files structure, all of the files with climate model data are named "tas_NorthAmerica_12mo.csv", all the files with the corresponding time for rows in that data are in files named "time_NorthAmerica_12mo.csv", and all the files with locations of the grid points are named "latitude_longitude_NorthAmerica_12mo.csv".
-Therefore, I would specify:
+In the later sections of this vignette, we show how to use these example files as inputs in the package functions.
 
-```{r}
-coordinateFilenames <- "latitude_longitude_NorthAmerica_12mo.csv"
-tasFilenames <- "tas_NorthAmerica_12mo.csv"
-timeFilenames <- "time_NorthAmerica_12mo.csv"
+### Processing climate projections with the example data
+
+The `gen_hw_set` function processes a directory of climate projections and creates files of identified and characterized heat waves.
+For example, if you wanted to use the example city location and climate projection data, and save results to a directory called "example_results" in your current working directory, you could run:
+
+```{r eval = FALSE}
+projection_dir_location <- system.file("extdata/cmip5",
+                                       package = "futureheatwaves")
+city_file_location <- system.file("extdata/cities.csv",
+                                  package = "futureheatwaves")
+
+gen_hw_set(out = "example_results",
+           dataFolder = projection_dir_location ,
+           dataDirectories = list("historical" = c(1990, 1999),
+                                       "rcp85" = c(2060, 2079)),
+           citycsv = city_file_location,
+           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
+           tasFilenames = "tas_NorthAmerica_12mo.csv",
+           timeFilenames = "time_NorthAmerica_12mo.csv")
 ```
 
-You will also need to specify where the output files should be locally saved.
-Save this as the object `out`.
-For example, if I wanted to write the output heatwave files to a directory called "example_results" in the `tmp` subdirectory of my home directory, I could specify:
+Notice that, in this call, you must specify the directory where you would like results written (`out`), the location of the directory of climate projections (`dataFolder`), the names of the two main subdirectories of that climate projection directory, as well as their year boundaries (`dataDirectories`), the file name of the community location file (`citycsv`), and the names used for the final grid coordinate, climate projection, and projection date files (`coordinateFilenames`, `tasFilenames`, and `timeFilenames`).
+There are more details later in this vignette describing how to set-up the input directory and location files for your own projection data.
 
-```{r}
-out <- "~/tmp/example_results/"
-```
-
-These selections will be sufficient to run the main function of this package, `gen_hw_set`, using a basic example with default values for things like the heatwave definition, years for which to generate the heatwave datasets (default is 2061--2080), and years to use as a reference for the relative temperature characteristics measured for the heatwaves (default is ...).
-
-Once you have specified these values, you can create files with future heatwaves and their characteristics using the following code:
-
-```{r eval = FALSE}
-gen_hw_set(out = out,
-           dataFolder = dataFolder,
-           citycsv = citycsv,
-           coordinateFilenames = coordinateFilenames,
-           tasFilenames = tasFilenames,
-           timeFilenames = timeFilenames,
-           RorCPP = 0)
-```
-
-You will first be asked:
+When you run this call, you will first be advised that the function will write files to your computer, and you will need to confirm that you want to proceed:
 
@@
  Warning: This function will write new files to your computer in the 
- ~/tmp/results directory of your computer. 
- Do you want to continue (y / n): 
+ ~/tmp/results/ directory of your computer. If that directory already exists,
+running this function will write over it. 
+ Do you want to continue? (y / n): 
 ```
@@
 If you agree that it is okay to write files and subdirectories to this directory, enter "y" at the prompt, and the function will continue running.
+(If you do not want to get this warning when running the function-- for example, when looping and calling this function repeatedly-- choose the option `printWarning = FALSE` when calling `gen_hw_set`.)
 
-If your file structure is set up correctly, you should get output that looks like this:
+If your file structure is set up correctly, you should get output that looks something like this:
 
 ```
-Processing historical ensemble for bcc1 
+Processing thresholds for bcc1 
 Reading ---> r1i1p1 
 Read operation complete 
-Processing rcp 8.5 ensemble bcc1 
+Processing projections for  bcc1 
 Reading ---> r1i1p1 
 Read operation complete 
-Creating heatwave dataframe ~~ City:  balt  ~~ City Number:  1  ~~ Cutoff:  84.632 
-Creating heatwave dataframe ~~ City:  ny  ~~ City Number:  2  ~~ Cutoff:  78.77876 
-Creating heatwave dataframe ~~ City:  nwk  ~~ City Number:  3  ~~ Cutoff:  78.77876 
-Creating heatwave dataframe ~~ City:  phil  ~~ City Number:  4  ~~ Cutoff:  84.632 
-Creating heatwave dataframe ~~ City:  prov  ~~ City Number:  5  ~~ Cutoff:  75.074 
+Creating heatwave dataframe ~~ City:  balt  ~~ City Number:  1  ~~ Cutoff:  84.25436 
+Creating heatwave dataframe ~~ City:  ny  ~~ City Number:  2  ~~ Cutoff:  78.11672 
+Creating heatwave dataframe ~~ City:  nwk  ~~ City Number:  3  ~~ Cutoff:  78.11672 
+Creating heatwave dataframe ~~ City:  phil  ~~ City Number:  4  ~~ Cutoff:  84.25436 
+Creating heatwave dataframe ~~ City:  prov  ~~ City Number:  5  ~~ Cutoff:  74.498 
 Writing bcc1: r1i1p1
-Writing accumulator 
+Processing thresholds for ccsm 
+Reading ---> r1i1p1 
+Read operation complete 
+Processing projections for  ccsm 
+Reading ---> r1i1p1 
+Read operation complete 
+Creating heatwave dataframe ~~ City:  balt  ~~ City Number:  1  ~~ Cutoff:  82.92344 
+Creating heatwave dataframe ~~ City:  ny  ~~ City Number:  2  ~~ Cutoff:  81.212 
+Creating heatwave dataframe ~~ City:  nwk  ~~ City Number:  3  ~~ Cutoff:  81.212 
+Creating heatwave dataframe ~~ City:  phil  ~~ City Number:  4  ~~ Cutoff:  81.44672 
+Creating heatwave dataframe ~~ City:  prov  ~~ City Number:  5  ~~ Cutoff:  80.88872 
+Writing ccsm: r1i1p1
+Processing projections for  ccsm 
+Reading ---> r2i1p1 
+Read operation complete 
+Creating heatwave dataframe ~~ City:  balt  ~~ City Number:  1  ~~ Cutoff:  82.92344 
+Creating heatwave dataframe ~~ City:  ny  ~~ City Number:  2  ~~ Cutoff:  81.212 
+Creating heatwave dataframe ~~ City:  nwk  ~~ City Number:  3  ~~ Cutoff:  81.212 
+Creating heatwave dataframe ~~ City:  phil  ~~ City Number:  4  ~~ Cutoff:  81.44672 
+Creating heatwave dataframe ~~ City:  prov  ~~ City Number:  5  ~~ Cutoff:  80.88872 
+Writing ccsm: r2i2p2
+Writing accumulators 
 All operations completed. Exiting. 
-
-  modelName length.histens. length.rcpens.
-1      bcc1               1              1
 ```
 
-The function provides status reports on its progress in generating the heatwave projections, as this process can take a while if you include many cities and / or many climate model projections.
-If you do not want to generate this output when running the function, choose the option `printWarning = FALSE` when calling `gen_hw_set`.
+The `gen_hw_set` function provides status reports on its progress in generating the heat wave projections.
+This helps you see that the function call is progressing, as this process can take a while if you include many cities and / or many climate model projections.
 
-Once the function has completed running, you will have two new subdirectories in the directory you specified with the `out` variable.
-These will be called `Heatwaves` and `Thresholds`.
-The `Heatwaves` directory will contain dataframes with all identified heatwaves and their characteristics for each city, while the `Thresholds` folder will contain files with the temperature thresholds used for each city given.
-The `out` directory specified will also have a file called `hwModelInfo.csv` that contains some basic information about the climate models and the number of ensemble members included for each.
+Once the function has completed running, you will have one new subdirectory in the `out` directory you specified called "Heatwaves".
+This directory will contain dataframes with all identified heat waves and their characteristics for each ensemble member of each climate model in your original directory.
+The `out` directory specified will also have a file called `hwModelInfo.csv` that contains some basic information about the climate models and the number of ensemble members included for each as well as `locationList.csv`, a file that gives the closest model gridpoint to each community for each climate model, and `hwModelInfo.csv`, a file that lists each of the climate models and the number of ensemble members for each within each subdirectory (e.g., "historical" and "rcp85" for the example).
+
+The full output directory structure will look something like this:
+
+- `out`
+    - `Heatwaves`
+        - `Projections`
+            - `bcc1`
+                - `1.csv`
+            - `ccsm`
+                - `1.csv`
+                - `2.csv`
+    - `hwModelInfo.csv`
+    - `locationList.csv`
+
+Each of the heat wave files will have one heat wave per row, covering all the heat waves identified for that particular ensemble member.
+The dataframe has the following columns with characteristics of each heat wave:
+
+- `hw.number`: The number of the heat wave out of all heat waves identified in the location;
+- `mean.temp`: Average daily temperature across days in the heat wave, in degrees Fahrenheit;
+- `max.temp`: Highest daily temperature across days in the heat wave, in degrees Fahrenheit;
+- `min.temp`: Lowest daily temperature across days in the heat wave, in degrees Fahrenheit
+- `length`: Number of days in the heat wave;
+- `start.date`: Date of the first day of the heat wave;
+- `end.date`: Date of the last day of the heat wave;
+- `start.doy`: Day of the year of the first day of the heat wave (1 = Jan. 1, etc.);
+- `start.month`: Month in which the heat wave started (1 = January, etc.);
+- `days.above.80`: Number of days in the heat wave above $80^oF$;
+- `days.above.85`: Number of days in the heat wave above $85^oF$;
+- `days.above.90`: Number of days in the heat wave above $90^oF$;
+- `days.above.95`: Number of days in the heat wave above $95^oF$;
+- `days.above.99th`: Number of days in the heat wave above the $99^{th}$ percentile temperature for the location, using the period specified by the user with the `referenceBoundaries` argument in `gen_hw_set` as a reference for determining these percentiles;
+- `days.above.99.5th`: Number of days in the heat wave above the $99.5^{th}$ percentile temperature for the location, using the period specified by the user with the `referenceBoundaries` argument in `gen_hw_set` as a reference for determining these percentiles;
+- `first.in.season`: Whether the heat wave was the first to occur in its calendar year in its location (Note: this characteristic is likely not useful in southern hemisphere studies.);
+- `threshold.temp`: The temperature used as the threshold for the heat wave definition in the location;
+- `mean.temp.quantile`: The percentile of the average daily mean temperature during the heat wave compared to the location's year-round temperature distribution, based on the temperatures for the city during the period specified by the `referenceBoundaries` argument in `gen_hw_set`;
+- `max.temp.quantile`: The percentile of the highest daily mean temperature during the heat wave compared to the location's year-round temperature distribution;
+- `min.temp.quantile`: The percentile of the lowest daily mean temperature during the heat wave compared to the location's year-round temperature distribution;
+- `mean.temp.1`: The location's average year-round temperature, based on the temperatures for the location during the period specified by the `referenceBoundaries` argument in `gen_hw_set`;
+- `mean.summer.temp`: The location's average May--September temperature, based on the temperatures for the location during the period specified by the `referenceBoundaries` argument in `gen_hw_set`; and
+- `city`: The identifier for the location, as given in the file specified in the `citycsv` argument of `gen_hw_set`.
+
+An example of one of these heat wave dataframes is given in the `hw_datafr` dataset included with the package. You can access this data using:
+
+```{r}
+data(hw_datafr)
+```
+
+This will load the example dataframe into your R session as the object `hw_datafr`:
+
+```{r}
+hw_datafr[1:3, c("hw.number", "mean.temp", "length", "start.date",
+                 "mean.temp.quantile", "city")]
+```
+
+You will need to select a directory pathname where all the heat wave files will be saved.
+For example, if you wanted to create a new directory called "example_results" in the current working directory and output files there, you could specify `out = "example_results"` in `gen_hw_set`.
+If you use a pathname to a directory that does not yet exist, the function will create that directory and then write the results to it.
+If you specify a directory that does already exist, the function will overwrite it with the results from the function run.
+You therefore should be careful when specifying this directory path and be sure to confirm that the function will not overwrite a directory you do not want to lose.
+You will specify the output directory using the `out` argument in the `gen_hw_set` function.
+
+### Required files and directory set-up
+
+To apply `gen_hw_set` to your own data, rather than the example data, you will need to set up your files and directory using a certain structure, to allow the function to process it correctly.
+
+#### File with community locations
+
+First, you must input to `gen_hw_set` a comma-separated (csv) file with the unique identifiers for each community for which you wish to make projections, as well as their latitude and longitude.
+The latitude and longitude should be expressed in the same way that latitude and longitude are expressed in the climate projection files to be processed (e.g., if latitude for New York City is expressed as 286 in the climate projection files, it cannot be expressed as -74 or 74 W in the community file).
+The community location file should have three columns, with the following column names:
+
+- `city`: A unique identifier for each community you will be analyzing.
+- `lat`: The communities' latitudes
+- `lon`: The communities' longitudes
+
+Here is an example of the correct format for a community locations file for five US communities (Baltimore, MD; New York, NY; Newark, NJ; Philadelphia, PA; and Providence RI):
+
+```
+"city","lat","lon"
+"balt",39.3008,283.3894
+"ny",40.6698,286.0562
+"nwk",40.7241,285.8268
+"phil",40.006817,284.8653
+"prov",41.82195,288.5803
+```
+
+You must specify the filepath of thislocations file using the argument `citycsv` when running the main function (`gen_hw_set`).
+For example, if you had this file saved as "my_cities.csv" in your current working directory, you would specify `citycsv = "my_cities.csv"` in the `gen_hw_set` function call.
+You can use either absolute or relative filepaths in this argument.
+
+If your community locations file has different column names for the latitude and longitude columns, you can specify this using the `lat_lon_colnames` argument in `gen_hw_set`.
+For example, if these columns were named "latitude" and "longitude", you could specify `lat_lon_colnames = c("latitude", "longitude")` when calling `gen_hw_set`.
+
+#### Directory structure for climate projection files
+
+For these functions to work correctly, you must have climate projection files saved in a specific structure locally on your computer.
+An example of a directory with the required set-up is given as the `cmip5` directory that comes installed with this package.
+To identify the location on your own computer of this directory, you can run:
+
+```{r}
+system.file("extdata/cmip5", package = "futureheatwaves")
+```
+
+You must have a directory that has within it two separate subdirectories.
+This set-up is required because this package was developed to work with CMIP5 projection data, in which climate projections up to 2004 or 2005 are stored in "historical experiment" files, while projections from 2006 and later are stored in separate files for each climate model ensemble member.
+The `gen_hw_set` function requires two subdirectories within the climate projection directory, to match up with this structure, where one subdirectory includes time series that cover one range of dates (e.g., 1980--2004) and the other subdirectory includes climate projections for the same climate model ensemble members, but for a different range of dates (e.g., 2006--2099).
+
+You can still use `gen_hw_set` if the full time range you are interested in are saved in a single set of files, but the function does require you to set up your climate projections directory to have two subdirectories.
+To work around this requirement in this case, you could copy the single subdirectory to have two copies and leave the second subdirectory in as a "dummy" subdirectory.
+
+You must specify in the `gen_hw_set` function the names of the two subdirectories as well as the date ranges that each of the subdirectories covers.
+For example, for a set-up with climate projection files for 1980--2004 in a subdirectory called "historical" and files for 2006--2099 in a subdirectory called "rcp85", you would include in `gen_hw_set` the argument: `dataDirectories = list("historical" = c(1980, 2004), "rcp85" = c(2006, 2099))`.
+The function uses this information to figure out which directory to search in for temperature data based on the user's choices of custom date ranges for certain arguments (e.g., dates for which to get projections).
+
+Within each climate model's directory, you need subdirectories for each ensemble member of the model (even if the model only has one ensemble member).
+Within each ensemble-member directory, you will need three comma-separated files: one file with the climate model output, with columns corresponding to climate grid point and rows corresponding to time, one file that gives the time values corresponding to each row in the climate model output file, and one file giving the latitude and longitude corresponding to the grid locations used in the columns of the climate model output file (more details on the required set-up of these files are given later in the vignette).
+
+For example, the following shows the file structure for the files in the example data included with the package.
+This example directory includes separate historical (with years 1990--1999) and RCP8.5 (with years 2060--2079) directories, each of which has the required climate projection, grid point locations, and projection time files for two separate models (`bcc1` and `ccsm`) with one and two separate ensemble members, respectively(`r1i1p1` for `bcc1`; `r1i1p1` and `r2i1p1` for `ccsm`).
+
+- `cmip5`
+    - `historical`
+        - `bcc1`
+            - `r1i1p1`
+                - `tas_NorthAmerica_12mo.csv`
+                - `latitude_longitude_NorthAmerica_12mo.csv`
+                - `time_NorthAmerica_12mo.csv`
+        - `ccsm`
+            - `r1i1p1`
+                - `tas_NorthAmerica_12mo.csv`
+                - `latitude_longitude_NorthAmerica_12mo.csv`
+                - `time_NorthAmerica_12mo.csv`
+    - `rcp85`
+        - `bcc1`
+            - `r1i1p1`
+                - `tas_NorthAmerica_12mo.csv`
+                - `latitude_longitude_NorthAmerica_12mo.csv`
+                - `time_NorthAmerica_12mo.csv`
+        - `ccsm`
+            - `r1i1p1`
+                - `tas_NorthAmerica_12mo.csv`
+                - `latitude_longitude_NorthAmerica_12mo.csv`
+                - `time_NorthAmerica_12mo.csv`
+            - `r2i1p1`
+                - `tas_NorthAmerica_12mo.csv`
+                - `latitude_longitude_NorthAmerica_12mo.csv`
+                - `time_NorthAmerica_12mo.csv`
+
+The user specifies the filepath of this directory of climate projections (through the `dataFolder` option in `gen_hw_set`).
+Notice that each subdirectory in this example has all required levels of the directory structure-- for example, even though the `bcc1` model only has one ensemble member (in this example), we've still included a directory level for ensemble members in the directory structure.
+Also notice that the final subdirectory always includes three files (with climate projections, grid point locations, and times; these files are explained in the next section), even if multiple ensemble members of the same climate model share the same grid point locations or times.
+
+#### Files with climate projections
+
+For each ensemble member of each climate model, you must have three files: one giving the gridded climate model output by date, one giving the location of each of the climate model grids, and one giving the date of each of the climate model projections.
+
+These files must be in a certain structure to run correctly through the `gen_hw_set` function.
+Therefore, the function currently requires a certain amount of preprocessing of climate projection files from a format like `.netCDF` to prepare them to be processed by this function.
+You must preprocess the file for the projection from each ensemble member to conform to the formats listed below.
+
+**Climate projection file**: The climate projection file should be a comma-separated file of temperatures that looks something like this:
+
+```
+267.6,281.17,285.69
+269.33,280.48,285.28
+269.29,280.12,285.05
+271.35,280.19,284.97
+272.97,281.57,285.22
+```
+
+This is a projection file covering three grid points of a climate model, giving projections for five dates.
+Each column corresponds to one grid point in the climate model.
+Each row corresponds to one date.
+The file does not have a header row; rather, the observations begin immediately on the first row of the file.
+Similarly, the file does not have a column of row numbers.
+Most projection files will include many more columns and rows than this small example, since they will usually cover all the grid points in a climate model and a long time series of observations.
+These projection files must not have gaps in dates (e.g., they should be year-round and not limited to the warm season) and cannot have any missing observations.
+In our example data, these climate projection files are saved for all ensemble members with the filename "tas_NorthAmerica_12mo.csv".
+You will need to specify this filename in the `tasFilenames` argument when running `gen_hw_set`.
+
+**Grid point location file**: The grid point location file should be a comma-separated file that looks something like this:
+
+```
+40.464,284.06
+40.464,286.88
+40.464,289.69
+```
+
+For this file, each row gives the location for the grid point for the corresponding column in the climate projection files.
+For example, the first row gives the latitude (first column) and longitude (second column) for the projections given in the first column of the climate projection file, the second row of this file gives the latitude and longitude for the projection given in the second column of the climate projection file, and so on.
+The file should not have a header row or a separate column of row names.
+
+In our example data, these grid-point location files are saved for all climate model ensemble members with the filename "latitude_longitude_NorthAmerica_12mo.csv".
+You will need to specify this filename in the `coordinateFilenames` argument when running `gen_hw_set`.
+
+**Projection times data file**: The projection times data file should be a comma-separated file that looks something like this:
+
+```
+51100,1990,1,1
+51101,1990,1,2
+51102,1990,1,3
+51103,1990,1,4
+51104,1990,1,5
+```
+
+This file gives a date that corresponds to each row of the climate projection file.
+The file should have four columns: one with row numbers, and then one each with year, month, and day.
+Year should be given with four digits.
+The file should not have a header row.
+In our example data, these grid-point location files are always saved with the filename "time_NorthAmerica_12mo.csv".
+You will need to specify this filename in the `timeFilenames` argument when running `gen_hw_set`.
+
+### Applying functions across all heat wave datasets
+
+Once you have created a directory of files with characterized heat waves for each ensemble member, you can explore or extend your results by using the `apply_all_models` function.
+This function allows you to apply custom R functions across all heat wave dataframes created by the `gen_hw_sets` call.
+
+You specify the location of the directory with the heat wave dataframes using the `out` argument when calling `apply_all_models`.
+Typically, this will be the same value as that for the `out` argument in `gen_hw_set`.
+You can apply any R function you have created, as long as it follows certain standards in accepting input and returning output.
+
+As an example, if you wanted to to get the average length of the heat waves identified within each ensemble member, you could write a simple function to calculate average length and then use `apply_all_models` to apply it across the dataframes for heat waves for all ensemble members in all climate models:
+
+```{r, eval = FALSE}
+# Define the function
+average_mean_temp <- function(hw_datafr){
+        out <- mean(hw_datafr$mean.temp)
+        return(out)
+        }
+```
+
+This function can then be applied across an example output directory included in the package, `example_results`, using:
+
+```{r}
+out <- system.file("extdata/example_results", package = "futureheatwaves")
+apply_all_models(out = out, 
+                 FUN = average_mean_temp)
+```
+
+You can also use the `city_specific` argument in `apply_all_models` to apply the function separately by location. For example:
+
+```{r}
+apply_all_models(out = out, 
+                 FUN = average_mean_temp,
+                 city_specific = TRUE)
+```
+
+To work, the functions you apply using `apply_all_models` must follow a certain structure.
+They must input a dataframe of heat waves, in the format of those output by `gen_hw_set`, and they must output a single-value vector (i.e., a vector of length one).
+You can use this functionality to either explore characteristics of the identified heat waves or to estimate the health impacts of the identified heat waves.
+
+We have included several functions as simple examples of the type of functions that can be used with `apply_all_models`:
+
+- `number_of_heatwaves`: Determines the number of heat waves ; 
+- `heatwave_days`: Sums up the total number of heat wave days by adding the lengths of all heat waves; 
+- `average_length`: Calculates the average length of heat waves; and 
+- `average_mean_temp`: Calculates the average mean temperature across all heat waves.
+
+You can see the code of any of these functions, to use as examples when developing your own, by typing just the function name (with no parentheses) in your R console.
+
+```{r, eval = FALSE}
+average_mean_temp
+```
+
+```
+function(hw_datafr){
+        out <- mean(hw_datafr$mean.temp)
+        return(out)
+}
+<environment: namespace:futureheatwaves>
+```
+
+To help users create their own functions, we have included an example dataframe representative of the heat wave dataframes that `gen_hw_set` outputs. This data can be loaded using the call:
+
+```{r eval = FALSE}
+data(hw_datafr)
+```
 
@@
 
-## How it works
+### Using a custom heat wave definition
 
-### Parsing the directory structure
+A variety of different heat wave definitions have been used to identify heat waves in a time series of temperature data [@smith2013].
+Researchers might want to use a specific definition, for example, because it matches the definition used by local health officials to declare heat wave warnings or, in the case of health impact assessments, to match with a definition used in an epidemiologic study.
 
-After some preliminary checks, the function reads in the structure of the directory specified by `dataFolder` and uses the names of subdirectories to create a list of all of the climate models and their ensemble members within the directory.
+The default is for this package to use the following definition to identify heat waves:
 
-The function that does this parsing, `acquireDirectoryStructure`, returns a list object outlining the file structure of the directory containing the climate projections.This list has an element for each climate model (e.g.
-,"bcc1").
-The first element within each of these elements is the name of the model.
-The second element within the first-level element gives the file paths for location grids, climate projections, and projection times for each ensemble run of the model.
+> A heat wave is two or more days at or above a community-specific threshold temperature, with the threshold determined as the $98^{th}$ percentile of year-round temperature in the community during some reference period (by default, 1981--2004).
 
-For example:
+Three components of this definition can be easily customized, without creating a new R function to use to identify heat waves.
+With a bit more work, in terms of writing a custom R function, the customization of the heat wave definition is quite extensive.
+
+First, the percentile used to identify a heat wave can be changed using the `probThreshold` option when calling the `gen_hw_set` function.
+This option can take values between 0 and 1.
+The default value is 0.98, or a definition with a threshold of the $98^{th}$ percentile of the community's temperature.
+Second, the number of days required in the heat wave definition can similarly be specified using the `numDays` argument in `gen_hw_set`.
+
+Third, it is possible to specify the range of years of data for the community that should be used when determining this threshold.
+For example, if you wanted to base the threshold for each community on current community climates, you could leave this option as its default, which uses climate projections from the years 1981 to 2004 to determine the thresholds.
+If you wanted to use a different set of years, you could set the start and end year bounds for this period using the option `thresholdBoundaries` in the function `gen_hw_set`.
+
+It is also possible to use a completely customized heat wave definition.
+To do this, you will need to write and load a function that implements your chosen heat wave definition.
+You can then reference the custom function using the `IDheatwavesFunction` option in `gen_hw_set`.
+To work correctly, this custom function must allow only specific inputs and generate only specific outputs.
+
+The function must allow the following inputs (even if it does not use them within the function code):
+
+- `datafr`: A dataframe with columns for date and temperature for a community.
+The first column should be the date of each observation, in the Date class.
+The second column should be the temperature.
+At this point in the normal processing of the `gen_hw_set` function, temperatures have been converted to Fahrenheit.
+Therefore, if you are writing a custom function that uses an absolute temperature threshold, it should be included in the function in Fahrenheit units.
+- `threshold`: A single-value numerical vector; in the default definition, this is the threshold temperature for each community, as determined earlier in the function call, calculated as a percentile of the community's year-round temperature.
+This should be an absolute temperature value, in degrees Fahrenheit.
+- `numDays`: The number of days above the threshold required for a heat wave.
+
+The function must return only the input dataframe (`datafr`), with the following columns added:
+
+- `hw`: A binary variable indicating whether a day was part of a heat wave (1: day in a heat wave; 0: not in a heat wave) and
+- `hw.number`: A non-negative integer that is 0 for all days that are not in heat waves and, for days in heat waves, gives a unique number for each separate heat wave.
+For example, all days in the second heat waves in the time series would get the value 2 for this column, while all days in the fifth heat wave would get the value 5, and all days not in heat waves would get the value 0.
+
+To help in developing your own heat wave identification functions, the package includes an example of the required input dataframe, `datafr`. You can load this dataframe by calling:
 
 ```{r}
-dataFolder <- system.file("cmip5", package = "futureheatwaves")
-finalList <- acquireDirectoryStructure(dataFolder = dataFolder,
-   coordinateFilenames = coordinateFilenames,
-   tasFilenames = tasFilenames,
-   timeFilenames = timeFilenames)
+data(datafr)
+```
 
-str(finalList[[1]][[1]])
-str(finalList[[1]][[2]][1])
+If your heat wave function is properly set up to use in this package, it can take as input this example dataframe and a threshold value and will return the original dataframe with the columns `hw` and `hw.number`, as defined above, returned:
+
+```{r}
+head(datafr, 3)
+id_of_hws <- IDHeatwavesR(datafr = datafr, threshold = 95, numDays = 2)
+head(id_of_hws, 3)
 ```
+
+The custom function can then be referenced using the `IDheatwavesFunction` option in `gen_hw_set`, and it will be used to identify heat waves in the dataset.
+You should specify the function name in quotation marks for this argument.
+For example, we have included an alternative ID definition function in this package, called `IDHeatwavesAlternative`, which identifies heat waves two or more days above the higher of either a community's $98^{th}$ percentile temperature or $80^oF$.
+To use this as the heat wave definition, you would run:
+
+```{r, eval = FALSE}
+# Set up path names to example data
+projection_dir_location <- system.file("extdata/cmip5",
+                                       package = "futureheatwaves")
+city_file_location <- system.file("extdata/cities.csv",
+                                  package = "futureheatwaves")
+
+gen_hw_set(out = "example_results",
+           dataFolder = projection_dir_location ,
+           dataDirectories = list("historical" = c(1990, 1999),
+                                       "rcp85" = c(2060, 2079)),
+           citycsv = city_file_location,
+           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
+           tasFilenames = "tas_NorthAmerica_12mo.csv",
+           timeFilenames = "time_NorthAmerica_12mo.csv",
+           IDheatwavesFunction = "IDHeatwavesAlternative")
+```
+
+### Customizing projection dates
+
+By default, the function will generate heat wave projections for the selected communities for the years 2070 to 2079.
+This projection date range can be changed by specifying alternative starting and ending years in the `projectionBoundaries` option in the `gen_hw_set` function.
+The starting year cannot be earlier than the starting date specified for the earlier subdirectory (e.g., "historical") or later than the last date of the later subdirectory (e.g., "rcp85").
+Further, the custom date boundaries cannot span between the two subdirectories.
+
+For example, to create projections for 2060 to 2069, you could call:
+
+```{r, eval = FALSE}
+gen_hw_set(out = "example_results",
+           dataFolder = projection_dir_location ,
+           dataDirectories = list("historical" = c(1990, 1999),
+                                  "rcp85" = c(2060, 2079)),
+           citycsv = city_file_location,
+           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
+           tasFilenames = "tas_NorthAmerica_12mo.csv",
+           timeFilenames = "time_NorthAmerica_12mo.csv",
+           projectionBoundaries = c(2060, 2069))
+```
+
+### Customizing dates for reference data
+
+The heat wave datasets characterize heat wave in several ways that are based on relative temperature.
+These characteristics are measured by taking the absolute temperature measures of the heat wave (e.g., average temperature during the heat wave is $90^oF$) and comparing them to the location’s typical temperature distributions.
+This process generates relative measures of how intense the heat wave is compared to what is normal in that loaction (e.g., $90^oF$ is in the $99^{th}$ percentile of year-round
+temperatures in the location).
+The default is to use temperatures for the period 2070 to 2079 for these reference temperatures, to align with the example data.
+However, the user can change this specification using the `referenceBoundaries` option of `gen_hw_set` to specify which years to use for these reference temperatures.
+This functionality can be useful in exploring the role of adaptation in future heat waves.
+
+For example, to use temperature projections for 1990 to 1999 when calculating relative characteristics of heat waves, to explore the assumption that cities remain adapted to their present-day climate, rather than changing in adaptation as climate change increases temperatures, you could run:
+
+```{r, eval = FALSE}
+gen_hw_set(out = "example_results",
+           dataFolder = projection_dir_location ,
+           dataDirectories = list("historical" = c(1990, 1999),
+                                       "rcp85" = c(2060, 2079)),
+           citycsv = city_file_location,
+           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
+           tasFilenames = "tas_NorthAmerica_12mo.csv",
+           timeFilenames = "time_NorthAmerica_12mo.csv",
+           referenceBoundaries = c(1990, 1999))
+```
+
+As with the `projectionBoundaries` option, there are some restrictions on which year ranges can be selected.
+The starting year cannot be earlier than the first year in the earlier subdirectory and the ending year cannot be later than the last year of the later subdirectory.
+Further, the custom date boundaries cannot span the two directories.
+
+### Mapping grid points
+
+The package also has a function that allows you to plot the locations of grid points for each climate model that match study locations, as well as show the links between climate grid points and associated locations.
+The `map_grid` function allows you to specify a model (`plot_model`), as well as the location of your output directory (usually, this will be the directory specified using the `out` argument in `gen_hw_set`) and will create a plot of the model grid points associated with your community locations, as well as draw lines between each community and its closest model grid point for that climate model:
+
+```{r fig.width = 5, message = FALSE, warning = FALSE}
+out <- system.file("extdata/example_results", package = "futureheatwaves")
+map_grid(plot_model = "bcc1", out = out)
+```
+
+Because the resulting plot is a "ggplot" object, you can use `ggplot2` functions to change the formatting for this output.
+For example, you could change the title and theme with the following code (results not shown):
+
+```{r fig.width = 5, message = FALSE, warning = FALSE, eval = FALSE}
+a <- map_grid(plot_model = "bcc1", out = out)
+a + ggtitle("BCC1 CMIP5 model") + theme_dark()
+```
+
+If you want to plot several of these maps for different climate models and then plot them together, you can use the `grid.arrange` function from the `gridExtra` package (results not shown):
+
+```{r fig.width = 5, fig.height = 7, eval = FALSE}
+library(gridExtra)
+a <- map_grid(plot_model = "bcc1", out = out)
+b <- map_grid(plot_model = "ccsm", out = out)
+grid.arrange(a, b, ncol = 1)
+```
+
+## Extensions
+
+The functionality of this package can be easily expanded by loops.
+For example, to explore the role of the heat wave definition on projections, the user could create a loop to run `gen_hw_set` and `apply_all_models` to the same directory of climate projections but with a variety of different functions used to pick out the heat waves in the projections.
+
+Also, while this package was created to be used for research on heat waves in climate change projections, with some adaptations it can also be used more broadly.
+For example, there are other exposures like wildfires and air pollution where it may be interesting to identify extended periods of high exposures in projection time series.
+This package could be applied to explore these exposures.
+The directory of projection data would need to be set up in the same structure as for exploring heat waves, and the `input_metric` should be set as `fahrenheit`, to pass the exposure values through to the characterized datasets without performing a conversion.
+A user could also use this package to explore events that must be lower than some minimum threshold (e.g., cold waves), but it would take some extra coding and conversions, since the functions in this package are written to identify periods above a threshold (for example, the user could multiple all projected temperatures by -1, and then the coldest temperatures would register as being the highest).
+
+Users doing these kinds of extensions will need to pay attention to a few points.
+First, some of the event characteristics (first in the calendar year, average of May--September temperatures, days above $90^{o}F$) might not be meaningful for studies of other types of events.
+Further, because event periods are usually defined as a string of multiple days exceeding some threshold, the functions in this package may miss the first and last event of the year.
+For example, if the first day of the time series is the last day of an event, this function would not identify that event because it lacks data from the earlier days that make allow this day to meet the event definition.
+This issues would lead to, a most, missing two events out of each projection, but should be considered if studying events that might occur near the start and end of projection data.
+If there is adequate interest from researchers, in the future we may adapt the package to make these secondary uses more straightforward.
+
+## References

### 2016-03-23 inst/doc/futureheatwaves.R

@@ -1,25 +1,24 @@
-## ----echo = FALSE, message = FALSE---------------------------------------
+## ----echo = FALSE, message = FALSE, warning = FALSE----------------------
 library(futureheatwaves)
+library(ggplot2)
 
 ## ------------------------------------------------------------------------
-system.file("cities.csv", package = "futureheatwaves")
-
-## ------------------------------------------------------------------------
-system.file("cmip5", package = "futureheatwaves")
+system.file("extdata/cities.csv", package = "futureheatwaves")
 
 ## ----eval = FALSE--------------------------------------------------------
-#  projection_dir_location <- system.file("cmip5", package = "futureheatwaves")
-#  city_file_location <- system.file("cities.csv", package = "futureheatwaves")
+#  projection_dir_location <- system.file("extdata/cmip5",
+#                                         package = "futureheatwaves")
+#  city_file_location <- system.file("extdata/cities.csv",
+#                                    package = "futureheatwaves")
 #  
 #  gen_hw_set(out = "example_results",
 #             dataFolder = projection_dir_location ,
-#             dataDirectories = list("historical" = c(1980, 2004),
-#                                          "rcp85" = c(2006, 2099)),
+#             dataDirectories = list("historical" = c(1990, 1999),
+#                                         "rcp85" = c(2060, 2079)),
 #             citycsv = city_file_location,
 #             coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
 #             tasFilenames = "tas_NorthAmerica_12mo.csv",
-#             timeFilenames = "time_NorthAmerica_12mo.csv",
-#             lat_lon_colnames = c("lat", "long"))
+#             timeFilenames = "time_NorthAmerica_12mo.csv")
 
 ## ------------------------------------------------------------------------
 data(hw_datafr)
@@ -28,22 +27,25 @@ data(hw_datafr)
 hw_datafr[1:3, c("hw.number", "mean.temp", "length", "start.date",
                  "mean.temp.quantile", "city")]
 
+## ------------------------------------------------------------------------
+system.file("extdata/cmip5", package = "futureheatwaves")
+
 ## ---- eval = FALSE-------------------------------------------------------
 #  # Define the function
-#  average_length <- function(hw_datafr){
-#          ave_length <- mean(hw_datafr$length)
-#          return(ave_length)
-#  }
[…]

### 2016-03-24 R/IDHeatwaves.R
Final edits on code and documentation

@@ -44,11 +44,8 @@ IDheatwaves <- function(threshold, datafr, global, custom){
 #' Identify heat waves in a time series
 #'
 #' This function identifies heat waves in a time series of temperature
-#' data using a heat wave definition that a heat wave must be two or more
-#' days with temperatures equal to or above some threshold temperature.
-#'
-#' This function is the default function used to identify heat waves in
-#' \code{\link{gen_hw_set}}.
+#' data using a heat wave definition that a heat wave must be a certain number
+#' of days with temperatures equal to or above some threshold temperature.
 #'
 #' @inheritParams closest_point
 #' @inheritParams IDheatwaves
@@ -76,7 +73,7 @@ IDheatwaves <- function(threshold, datafr, global, custom){
 #' @examples
 #' \donttest{
 #' data(datafr)
-#' hw_ids <- IDHeatwavesR(threshold = 80, datafr,
+#' hw_ids <- IDHeatwavesR(threshold = 80, datafr = datafr,
 #'                        numDays = 3)
 #' }
 #'
@@ -176,7 +173,7 @@ IDHeatwavesR <- function(threshold, datafr, numDays){
 #' @examples
 #' \donttest{
 #' data(datafr)
-#' hw_ids <- IDHeatwavesAlternative(threshold = 80, datafr,
+#' hw_ids <- IDHeatwavesAlternative(threshold = 80, datafr = datafr,
 #'                                  numDays = 3)
 #' }
 #'
@@ -249,11 +246,14 @@ IDHeatwavesAlternative <- function(threshold, datafr, numDays){
 #' Identify heat waves in a time series
 #'
 #' This function identifies heat waves in a time series of temperature
-#' data using a heat wave definition that a heat wave must be two or more
-#' days with temperatures equal to or above some threshold temperature. This
+#' data using a heat wave definition that a heat wave must be a certain number
+#' of days with temperatures equal to or above some threshold temperature. This
 #' function uses a compiled C++ function for part of the process, making it
 #' faster than the R analogue, \code{\link{IDHeatwavesR}}, although the two
 #' functions give identical results.
+#'
+#' This function is the default function used to identify heat waves in
+#' \code{\link{gen_hw_set}}.
 #'
 #' @inheritParams closest_point
 #' @inheritParams IDheatwaves
[…]

### 2016-03-24 R/applyHeatwaves.R

@@ -8,28 +8,28 @@
 #' \code{city_specific} is set to \code{TRUE}.
 #'
 #' @param out Character string with pathname to which heat wave files were
-#'    written by \code{\link{gen_hw_set}}.Typically, this will be the same
+#'    written by \code{\link{gen_hw_set}}. Typically, this will be the same
 #'    pathname as that specified with \code{out} when running
 #'    \code{\link{gen_hw_set}}.
 #' @param FUN A character string with the name of a function to apply to
 #'    the heat wave dataframes located in the directory specified by
-#'    \code{out}. This function must take only the argument, \code{hw_datafr},
+#'    \code{out}. This function must take the argument, \code{hw_datafr},
 #'    a dataframe of identified and characterized heat waves, as generated by
 #'    \code{\link{gen_hw_set}}. For an example of one of these dataframes, load
 #'    the \code{\link{hw_datafr}} data. The function should output a single
-#'    value when applied to the full dataframe.
-#' @param city_specific TRUE / FALSE: Should the function be applied
-#'    separately for each city?
-#' @param ... Optional arguments to FUN
+#'    value when applied to the full heat wave dataframe. The function can
+#'    also have other arguments, which are passed through the \code{...}
+#'    argument of \code{apply_all_models}.
+#' @param city_specific TRUE or FALSE, specifying whether the function should
+#'    be applied separately for each study location.
+#' @param ... Optional arguments to \code{FUN}.
 #'
-#' @return A dataframe with the value output by the function specified by the
-#'    user with \code{FUN}, as applied to every dataframe of heat waves and
-#'    characteristics for every ensemble member of every climate model.
+#' @return A dataframe with the value output by the \code{FUN} function, as
+#'    applied to all the heat wave dataframes generated by \code{gen_hw_set}.
 #'
 #' @note The function input as \code{FUN} must follow a very specific
-#'    structure. It most have only one argument, and that argument
-#'    must be a dataframe with heat waves and their characteristics, as
-#'    generated by the \code{\link{gen_hw_set}} function. See the
+#'    structure. It most have as an argument a dataframe with characterized heat
+#'    waves, as generated by the \code{\link{gen_hw_set}} function. See the
 #'    \code{futureheatwaves} vignette for more guidance on creating and
 #'    applying a custom function to explore the heat waves identified and
 #'    characterized by \code{\link{gen_hw_set}}.

### 2016-03-24 R/exampleApplyFunctions.R

@@ -84,18 +84,18 @@ average_length <- function(hw_datafr){
 #' Calculate average temperature of heat waves
 #'
 #' This function takes a dataframe of heat waves, as created by
-#' \code{\link{gen_hw_set}}, and calculates the average of the mean daily
+#' \code{\link{gen_hw_set}}, and calculates the average of the daily
 #' temperature during each heat wave across all heat waves in the dataframe.
 #'
 #' @param hw_datafr A dataframe of heat waves and their characteristics, as
 #'    created by \code{\link{gen_hw_set}}
 #'
 #' @return A numeric value with the average temperature (degrees Fahrenheit)
-#'    for a dataset of heat waves, as generated by \code{\link{gen_hw_set}}.
-#'    This value is the mean of the average temperature of each heat wave,
-#'    not the average temperature across all heat wave days (to calculate that,
-#'    you would need to calculate a weighted mean, weighted by the number of
-#'    days in each heat wave).
+#'    during all heat waves for a dataset of heat waves, as generated by
+#'    \code{\link{gen_hw_set}}. This value is the mean of the average
+#'    temperature of each heat wave, not the average temperature across all
+#'    heat wave days (to calculate that, you would need to calculate a weighted
+#'    mean, weighted by the number of days in each heat wave).
 #'
 #' @note This function is an example of a function that can be created and
 #'    used to explore heat waves using the \code{\link{apply_all_models}}

### 2016-03-24 R/main.R

@@ -1,9 +1,10 @@
 #' Create and write heat wave projections
 #'
 #' This function creates datasets of identified and characterized heat waves
-#' for all ensemble members in a directory of climate projections for a
-#' user-specified set of cities.The resulting heat wave projections are written
-#' out to a specified directory on the user's local computer.
+#' for all ensemble members in all climate models in a directory of climate
+#' projections for a user-specified set of locations.The resulting heat wave
+#' projections are written out to a specified directory on the user's local
+#' computer.
 #'
 #' @param out Character string with pathway to directory to which
 #'    heat wave files will be written. This should be a pathname to a directory
@@ -19,62 +20,68 @@
 #' @param citycsv Character string giving the filepath to a
 #'    comma-separated (.csv) file with, for each study city, a unique city
 #'    identifier, latitude, and longitude. These values must be specified with
-#'    the column names \code{city}, \code{lat}, and \code{lon}.
+#'    the column names \code{city}, \code{lat}, and \code{lon}. See the
+#'    \code{futureheatwaves} vignette for guidance on setting up this
+#'    file.
 #' @param coordinateFilenames Character string the with filename of each
-#'    grid point location file. This filename should be identical for all ensemble
-#'    members included in the \code{dataFolder} directory. See the package
-#'    vignette for an example of the required structure for this file.
+#'    grid point location file. This filename should be identical for all
+#'    ensemble member subdirectories included in the \code{dataFolder} directory.
+#'    See the package vignette for an example of the required structure for this
+#'    file.
 #' @param tasFilenames Character string the with filename of each climate
 #'    projection file. This filename should be identical for all ensemble
-#'    members included in the \code{dataFolder} directory. See the package
-#'    vignette for an example of the required structure for this file.
+#'    member subdirectories included in the \code{dataFolder} directory. See the
+#'    package vignette for an example of the required structure for this file.
 #' @param timeFilenames Character string the with filename of each projection
 #'    dates file. This filename should be identical for all ensemble
-#'    members included in the \code{dataFolder} directory. See the package
-#'    vignette for an example of the required structure for this file.
+#'    member  subdirectories included in the \code{dataFolder} directory. See
+#'    the package vignette for an example of the required structure for this
+#'    file.
 #' @param IDheatwavesFunction A character string with the name of the R function
[…]

### 2016-03-24 R/mapCities.R

@@ -1,14 +1,23 @@
 #' Create a map of model grid
 #'
 #' This function creates a map of the grid points of a climate model used for
-#' the study cities and draws lines connecting each study city to its climate
-#' model grid point.
+#' the study locations and draws lines connecting each study city to its climate
+#' model grid point. It currently can only be used for studies within the
+#' United States.
 #'
 #' @param plot_model A character string with the name of the model to plot
-#' @inheritParams gen_hw_set
+#' @inheritParams apply_all_models
+#'
+#' @return A \code{ggplot2} object with a map of grid points for the climate
+#'    model that were used in processing heat waves for the study locations,
+#'    with a line drawn from each study locations to the grid point used for
+#'    it.
 #'
 #' @note This function creates a \code{ggplot2} object, so the output can be
-#'    edited using \code{ggplot2} functions.
+#'    edited using \code{ggplot2} functions. For this function to work
+#'    correctly, longitude must be expressed using non-negative decimal
+#'    degrees when setting up the climate projection files and community
+#'    location file for \code{\link{gen_hw_set}}.
 #'
 #' @examples
 #' out <- system.file("extdata/example_results", package = "futureheatwaves")

### 2016-03-24 R/preparation.R

@@ -23,14 +23,6 @@ acquireDirectoryStructure <- function(dataFolder, coordinateFilenames,
                                       threshold_ensemble,
                                       thresholdBoundaries){
 
-        # If `dataFolder` does not end in "/", add it.
-        # (Only need to repeat here to use with in-package examples)
-        split_dataFolder <- unlist(strsplit(dataFolder, split = ""))
-        last_char <- split_dataFolder[length(split_dataFolder)]
-        if(last_char != "/"){
-                dataFolder <- paste0(dataFolder, "/")
-        }
-
         # Acquire all pathnames to csv files rooted at dataPath
         all <- list.files(dataFolder, recursive = TRUE,
                           pattern = "\\.csv$")
@@ -114,16 +106,23 @@ buildStructureModels <- function(modelName, experiments,
                                  coordinateFilenames, tasFilenames,
                                  timeFilenames, dataDirectories){
         return(list(modelName,
-                    buildStructureExperiments(modelName, experiments[1],
-                                              dataFolder,
-                                              coordinateFilenames,
-                                              tasFilenames, timeFilenames,
-                                              dataDirectories),
-                    buildStructureExperiments(modelName, experiments[2],
-                                              dataFolder,
-                                              coordinateFilenames,
-                                              tasFilenames, timeFilenames,
-                                              dataDirectories)))
+                    buildStructureExperiments(modelName = modelName,
+                                              experiment = experiments[1],
+                                              dataPath = dataFolder,
+                                              coordinateFilenames =
+                                                      coordinateFilenames,
+                                              tasFilenames = tasFilenames,
+                                              timeFilenames = timeFilenames,
+                                              dataDirectories = dataDirectories),
+                    buildStructureExperiments(modelName = modelName,
+                                              experiment = experiments[2],
+                                              dataPath = dataFolder,
+                                              coordinateFilenames =
+                                                      coordinateFilenames,
+                                              tasFilenames = tasFilenames,
+                                              timeFilenames = timeFilenames,
+                                              dataDirectories =
+                                                      dataDirectories)))
 }
 
 #' Generate file structure for an experiment
[…]

### 2016-03-24 R/process.R

@@ -109,7 +109,7 @@ processModel <- function(model, global, custom, accumulators,
 #'    components of the dataframe of climate model grid locations that will
 #'    ultimately be output as the climate model grid locations file.
 processThresholds <- function(model, global, custom){
-        name <- model[[1]]
+        mod_name <- model[[1]]
 
         # To find the threshold, use the first ensemble member within
         # the relevant directory, historical or rcp
@@ -122,11 +122,11 @@ processThresholds <- function(model, global, custom){
                 global$threshold_ensemble
         thresholdDirs <- thresholdDirs_model[r1i1p1_i][[1]]
 
-        cat("Processing thresholds for", name, "\n")
+        cat("Processing thresholds for", mod_name, "\n")
 
         # Acquire characteristics of the first historical ensemble
         thresholdEnsemble <- processEnsemble(ensemble = thresholdDirs,
-                                              modelName = name,
+                                              modelName = mod_name,
                                               global = global,
                                               custom = custom,
                                               type = "threshold")
@@ -213,7 +213,7 @@ processProjections <- function(ensemble, modelName, ensembleWriter,
                                thresholds, global, custom, accumulators,
                                reference, reference_dates){
 
-        cat("Processing projections for ", modelName, "\n")
+        cat("Processing projections for", modelName, "\n")
 
         # Acquire desired characteristics of the projection ensemble
         ensembleSeries <- processEnsemble(ensemble = ensemble,
@@ -283,7 +283,8 @@ processEnsemble <- function(ensemble, modelName, global, custom, type){
         end <- bounds[2]
 
         # Get a vector containing the dates of days we are dealing with
-        dates <- formDates(times, bounds)
+        dates <- formDates(times = times,
+                           bounds = bounds)
 
         # Acquire time series for every city
         series <- data.frame(tas[start:end, locations])
@@ -333,9 +334,8 @@ closest_point <- function(city, latlong){
         aSQ <- abs(latlong[,1] - latitude) ^ 2
         bSQ <- abs(latlong[,2] - longitude) ^ 2
         cSQ <- aSQ + bSQ
-        c <- sqrt(cSQ)
-        smallest <- min(c)
-        index <- match(smallest, c)
+        c_distance <- sqrt(cSQ)
+        index <- which.min(c_distance)
         return(index)
 }
 

### 2016-03-24 inst/doc/futureheatwaves.R
Final edits of vignette

@@ -6,15 +6,17 @@ library(ggplot2)
 system.file("extdata/cities.csv", package = "futureheatwaves")
 
 ## ----eval = FALSE--------------------------------------------------------
+#  # Identify location of example files
 #  projection_dir_location <- system.file("extdata/cmip5",
 #                                         package = "futureheatwaves")
 #  city_file_location <- system.file("extdata/cities.csv",
 #                                    package = "futureheatwaves")
 #  
+#  # Process example files
 #  gen_hw_set(out = "example_results",
 #             dataFolder = projection_dir_location ,
 #             dataDirectories = list("historical" = c(1990, 1999),
-#                                         "rcp85" = c(2060, 2079)),
+#                                          "rcp85" = c(2060, 2079)),
 #             citycsv = city_file_location,
 #             coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
 #             tasFilenames = "tas_NorthAmerica_12mo.csv",
@@ -22,36 +24,64 @@ system.file("extdata/cities.csv", package = "futureheatwaves")
 
 ## ------------------------------------------------------------------------
 data(hw_datafr)
-
-## ------------------------------------------------------------------------
 hw_datafr[1:3, c("hw.number", "mean.temp", "length", "start.date",
                  "mean.temp.quantile", "city")]
 
 ## ------------------------------------------------------------------------
 system.file("extdata/cmip5", package = "futureheatwaves")
 
-## ---- eval = FALSE-------------------------------------------------------
-#  # Define the function
-#  average_mean_temp <- function(hw_datafr){
-#          out <- mean(hw_datafr$mean.temp)
-#          return(out)
-#          }
+## ------------------------------------------------------------------------
+average_mean_temp <- function(hw_datafr){
+        out <- mean(hw_datafr$mean.temp)
+        return(out)
+        }
 
 ## ------------------------------------------------------------------------
 out <- system.file("extdata/example_results", package = "futureheatwaves")
-apply_all_models(out = out, 
-                 FUN = average_mean_temp)
+apply_all_models(out = out, FUN = average_mean_temp)
 
 ## ------------------------------------------------------------------------
-apply_all_models(out = out, 
-                 FUN = average_mean_temp,
-                 city_specific = TRUE)
+apply_all_models(out = out, FUN = average_mean_temp, city_specific = TRUE)
 
[…]

### 2016-12-26 R/mapCities.R
Change the mapsource for ggmap function

@@ -191,24 +191,29 @@ map_grid_ggmap <- function(plot_model, out, lon_transform = FALSE){
                                  f = 0.5)
         map <- suppressWarnings(get_map(location = bbox,
                                         source = "stamen",
-                                        maptype = "watercolor"))
+                                        maptype = "toner-lite"))
 
         latlong <- unique(cities[ , c("lon_grid", "lat_grid")])
 
         map <- ggmap::ggmap(map) +
+                ggplot2::geom_segment(data = cities,
+                                      ggplot2::aes_string(x = "lon",
+                                                          y = "lat",
+                                                          xend = "lon_grid",
+                                                          yend = "lat_grid"),
+                                      size = 1.2, alpha = 0.8,
+                                      color = "darkred") +
+                ggplot2::geom_point(data = cities,
+                                    ggplot2::aes_string(x = "lon",
+                                                        y = "lat"),
+                                    size = 1.5, color = "darkred",
+                                    alpha = 0.8) +
                 ggplot2::geom_point(data = latlong,
                                     ggplot2::aes_string(x = "lon_grid",
                                                         y = "lat_grid"),
-                                    size = 2, color = "black", alpha = 0.8)
-        map <- map + ggplot2::geom_segment(data = cities,
-                                           ggplot2::aes_string(x = "lon",
-                                                               y = "lat",
-                                                               xend = "lon_grid",
-                                                               yend = "lat_grid"),
-                                           size = 1.2, alpha = 0.8,
-                                           color = "black")
-        map <- map + ggthemes::theme_map()
-        map <- map + ggplot2::ggtitle(plot_model)
+                                    size = 2, color = "red", alpha = 0.8) +
+                ggthemes::theme_map() +
+                ggplot2::ggtitle(plot_model)
         return(map)
 }
 

### 2016-12-26 R/mapCities.R
Fix ggmap function so you don't need `lon_transform`

@@ -85,18 +85,12 @@ map_grid <- function(plot_model, out){
 #' @export
 #'
 #' @importFrom dplyr %>%
-map_grid_leaflet <- function(plot_model, out, lon_transform = FALSE){
+map_grid_leaflet <- function(plot_model, out){
         cities <- utils::read.csv(paste(out, "locationList.csv", sep = "/"),
                                   col.names = c("city","lat", "lon",
                                                 "lat_grid", "lon_grid",
                                                 "model")) %>%
                 dplyr::filter_(~ model == plot_model)
-
-        if(lon_transform){
-                cities <- cities %>%
-                        dplyr::mutate_(lon = ~ lon - 360,
-                                       lon_grid = ~ lon_grid - 360)
-        }
 
         cities <- cities %>%
                 dplyr::mutate_(city_popup = ~ paste0("<b>City:</b> ",
@@ -169,22 +163,22 @@ map_grid_leaflet <- function(plot_model, out, lon_transform = FALSE){
 #'
 #' @examples
 #' out <- system.file("extdata/example_results", package = "futureheatwaves")
-#' map_grid_ggmap(plot_model = "bcc1", out = out, lon_transform = TRUE)
+#' map_grid_ggmap(plot_model = "bcc1", out = out)
 #'
 #' @export
 #'
 #' @importFrom dplyr %>%
-map_grid_ggmap <- function(plot_model, out, lon_transform = FALSE){
+map_grid_ggmap <- function(plot_model, out){
         cities <- utils::read.csv(paste(out, "locationList.csv", sep = "/"),
                                   col.names = c("city","lat", "lon",
                                                 "lat_grid", "lon_grid", "model")) %>%
                 dplyr::filter_(~ model == plot_model)
 
-        if(lon_transform){
-                cities <- cities %>%
-                        dplyr::mutate_(lon = ~ lon - 360,
-                                       lon_grid = ~ lon_grid - 360)
-        }
+        cities <- cities %>%
+                dplyr::mutate_(lon = ~ ifelse(lon > 180, -(360 - lon), lon),
+                               lon_grid = ~ ifelse(lon_grid > 180,
+                                                   -(360 - lon_grid),
+                                                   lon_grid))
 
         bbox <- ggmap::make_bbox(lon = c(cities$lon, cities$lon_grid),
                                  lat = c(cities$lat, cities$lat_grid),

### 2016-12-26 R/mapCities.R
Use circle markers in leaflet function

@@ -127,7 +127,7 @@ map_grid_leaflet <- function(plot_model, out){
 
         out_map <- leaflet::leaflet() %>%
                 leaflet::addTiles() %>%
-                leaflet::addMarkers(data = cities, lng = ~lon, lat = ~lat,
+                leaflet::addCircleMarkers(data = cities, lng = ~lon, lat = ~lat,
                                     popup = ~ city_popup) %>%
                 leaflet::addCircleMarkers(data = grid_data,
                                           lng = ~lon_grid, lat = ~lat_grid,
@@ -183,9 +183,9 @@ map_grid_ggmap <- function(plot_model, out){
         bbox <- ggmap::make_bbox(lon = c(cities$lon, cities$lon_grid),
                                  lat = c(cities$lat, cities$lat_grid),
                                  f = 0.5)
-        map <- suppressWarnings(get_map(location = bbox,
-                                        source = "stamen",
-                                        maptype = "toner-lite"))
+        map <- suppressWarnings(ggmap::get_map(location = bbox,
+                                               source = "stamen",
+                                                maptype = "toner-lite"))
 
         latlong <- unique(cities[ , c("lon_grid", "lat_grid")])
 

### 2016-12-26 R/main.R
Rearrange parameter order for `gen_hw_set`

@@ -153,9 +153,9 @@ gen_hw_set <- function(out,
                        printWarning = TRUE,
                        threshold_ensemble = "r1i1p1",
                        lat_lon_colnames = c("lat", "lon"),
-                       seasonal_months = c(5:9),
+                       above_threshold = TRUE,
                        absolute_thresholds = c(299.82, 302.60, 305.37, 308.15),
-                       above_threshold = TRUE){
+                       seasonal_months = c(5:9)){
 
         # Add warning for user that this will write new files
         if(printWarning){

### 2016-12-27 R/data.R
Update example heat wave dataframe

@@ -20,14 +20,14 @@
 #' This example dataset can be used in developing custom functions to use with
 #' \code{\link{apply_all_models}}.
 #'
-#' @format A dataframe with 45 rows and 23 columns. The variables are:
+#' @format A dataframe with 258 rows and 23 columns. The variables are:
 #' \itemize{
 #'    \item hw.number: A sequential number identifying each heat wave in a city;
-#'    \item mean.temp: Average daily temperature across all days in the
+#'    \item mean.var: Average daily temperature across all days in the
 #'       heat wave, in degrees Fahrenheit;
-#'    \item max.temp: Highest daily temperature across days in the
+#'    \item max.var: Highest daily temperature across days in the
 #'       heat wave, in degrees Fahrenheit;
-#'    \item min.temp: Lowest daily temperature across days in the
+#'    \item min.var: Lowest daily temperature across days in the
 #'       heat wave, in degrees Fahrenheit
 #'    \item length: Number of days in the heat wave;
 #'    \item start.date: Date of the first day of the heat wave;
@@ -36,13 +36,13 @@
 #'       (1 = Jan. 1, etc.);
 #'    \item start.month: Month in which the heat wave started (1 = January,
 #'       etc.);
-#'    \item days.above.80: Number of days in the heat wave above 80 degrees
+#'    \item days.above.abs.threshold.1: Number of days in the heat wave above 80 degrees
 #'        Fahrenheit;
-#'    \item days.above.85: Number of days in the heat wave above 85 degrees
+#'    \item days.above.abs.threshold.2: Number of days in the heat wave above 85 degrees
 #'        Fahrenheit;
-#'    \item days.above.90: Number of days in the heat wave above 90 degrees
+#'    \item days.above.abs.threshold.3: Number of days in the heat wave above 90 degrees
 #'        Fahrenheit;
-#'    \item days.above.95: Number of days in the heat wave above 90 degrees
+#'    \item days.above.abs.threshold.4: Number of days in the heat wave above 90 degrees
 #'        Fahrenheit;
 #'    \item days.above.99th: Number of days in the heat wave above the 99th
 #'        percentile temperature for the city, using the period specified
@@ -54,27 +54,27 @@
 #'        by the user with the \code{referenceBoundaries} argument in
 #'        \code{\link{gen_hw_set}} as a reference for determining these
 #'        percentiles;
-#'    \item first.in.season: Whether the heat wave was the first to occur in its
[…]

### 2016-12-27 README.Rmd
Update vignette

@@
 
+You can download the latest stable version of this package directly from CRAN:
+
+```{r eval = FALSE}
+install.packages("futureheatwaves")
+```
+
 The development version of this package can be installed directly from GitHub using `install_github` from the `devtools` package:
@@
 
-An uncompiled version of the vignette can be found, without having to install the package, [here](https://github.com/geanders/futureheatwaves/blob/master/vignettes/futureheatwaves.Rmd).
+A version of this vignette is also available through CRAN [here](https://cran.r-project.org/web/packages/futureheatwaves/vignettes/futureheatwaves.html).
+
+The development version of the package also includes a vignette with details on how to prepare CMIP5 netCDF climate model output files for use with this package.
+If you have downloaded the development version of the package directly from GitHub, you can access that vignette with:
+
+```{r eval = FALSE}
+vignette("starting_from_netcdf")
+```

### 2016-12-27 R/mapCities.R
Move grid map function to development version

@@ -137,74 +137,3 @@ map_grid_leaflet <- function(plot_model, out){
         return(out_map)
 }
 
-#' Create a map of model grid using ggmap
-#'
-#' This function creates a map of the grid points of a climate model used for
-#' the study locations and draws lines connecting each study city to its climate
-#' model grid point, using as a base a map pulled from Google Maps with
-#' \code{ggmap}.
-#'
-#' @inheritParams map_grid
-#' @inheritParams map_grid_leaflet
-#' @inheritParams apply_all_models
-#'
-#' @return A \code{ggplot2} object with a map of grid points for the climate
-#'    model that were used in processing heat waves for the study locations,
-#'    with a line drawn from each study locations to the grid point used for
-#'    it.
-#'
-#' @note This function creates a \code{ggplot2} object, so the output can be
-#'    edited using \code{ggplot2} functions.
-#'
-#' @examples \dontrun{
-#' out <- system.file("extdata/example_results", package = "futureheatwaves")
-#' map_grid_ggmap(plot_model = "bcc1", out = out)
-#' }
-#'
-#' @export
-#'
-#' @importFrom dplyr %>%
-map_grid_ggmap <- function(plot_model, out){
-        cities <- utils::read.csv(paste(out, "locationList.csv", sep = "/"),
-                                  col.names = c("city","lat", "lon",
-                                                "lat_grid", "lon_grid", "model")) %>%
-                dplyr::filter_(~ model == plot_model)
-
-        cities <- cities %>%
-                dplyr::mutate_(lon = ~ ifelse(lon > 180, -(360 - lon), lon),
-                               lon_grid = ~ ifelse(lon_grid > 180,
-                                                   -(360 - lon_grid),
-                                                   lon_grid))
-
-        bbox <- ggmap::make_bbox(lon = c(cities$lon, cities$lon_grid),
-                                 lat = c(cities$lat, cities$lat_grid),
-                                 f = 0.5)
-        map <- suppressWarnings(ggmap::get_map(location = bbox,
-                                               source = "stamen",
-                                                maptype = "toner-lite"))
-
-        latlong <- unique(cities[ , c("lon_grid", "lat_grid")])
-
-        map <- ggmap::ggmap(map) +
-                ggplot2::geom_segment(data = cities,
-                                      ggplot2::aes_string(x = "lon",
-                                                          y = "lat",
-                                                          xend = "lon_grid",
-                                                          yend = "lat_grid"),
-                                      size = 1.2, alpha = 0.8,
[…]

### 2016-12-27 R/IO.R
Get rid of commented out tries for netCDF input

@@ -54,17 +54,6 @@ readLatLong <- function(ensemble, global){
         return(locations)
 }
 
-# readLatLong_ncdf <- function(ensemble, global){
-#         loc_file <- "~/Downloads/tas_day_GFDL-ESM2G_historical_r1i1p1_19810101-19851231.nc"
-#         ncdf_file <- ncdf4::nc_open(loc_file)
-#         ncdf_lon <- ncdf4::ncvar_get(ncdf_file, "lon")
-#         ncdf_lat <- ncdf4::ncvar_get(ncdf_file, "lat")
-#         ncdf4::nc_close(ncdf_file)
-#         locations <- expand.grid(ncdf_lat, ncdf_lon)
-#         colnames(locations) <- c("V1", "V2")
-#         return(locations)
-# }
-
 #' Read climate projection data
 #'
 #' This function reads climate projection data for a climate
@@ -107,19 +96,6 @@ readTimes <- function(ensemble, global){
         loc_file <- grep(global$timeFilenames, ensemble)
         return(utils::read.csv(ensemble[loc_file], header = FALSE))
 }
-
-# readTimes_ncdf <- function(ensemble, global){
-#         library(PCICt)
-#         loc_file <- "~/Downloads/tas_day_GFDL-ESM2G_historical_r1i1p1_19810101-19851231.nc"
-#         ncdf_file <- ncdf4::nc_open(loc_file)
-#         ncdf_time <- ncdf4.helpers::nc.get.time.series(ncdf_file, v = "tas", time.dim.name = "time")
-#         ncdf4::nc_close(ncdf_file)
-#         times <- data.frame(V1 = 1:length(ncdf_time),
-#                                 V2 = as.integer(format(ncdf_time, "%Y")),
-#                                 V3 = as.integer(format(ncdf_time, "%m")),
-#                                 V4 = as.integer(format(ncdf_time, "%d")))
-#         return(times)
-# }
 
 #' Ensemble writer factory function
 #'

### 2016-12-27 R/process.R

@@ -260,19 +260,12 @@ processProjections <- function(ensemble, modelName, ensembleWriter,
 #'    for the specified ensemble member.
 processEnsemble <- function(ensemble, modelName, global, custom, type){
 
-        # loc_file <- "~/Downloads/tas_day_GFDL-ESM2G_historical_r1i1p1_19810101-19851231.nc"
-        # ncdf_file <- ncdf4::nc_open(loc_file)
-
         # Read the ensemble data
         cat("Reading --->", ensemble[1], "\n")
 
         latlong <- readLatLong(ensemble = ensemble, global = global)
         climate_lats <- unique(latlong[ , 1])
         climate_longs <- unique(latlong[ , 2])
-
-        # climate_lats <- ncdf4::ncvar_get(ncdf_file, "lat")
-        # climate_longs <- ncdf4::ncvar_get(ncdf_file, "lon")
-        # latlong <- expand.grid(climate_lats, climate_longs)
 
         times <- readTimes(ensemble = ensemble, global = global)
 
@@ -282,21 +275,6 @@ processEnsemble <- function(ensemble, modelName, global, custom, type){
         locations <- apply(global$cities, 1, closest_point, latlong = latlong)
         out_locations <- cbind(global$cities, latlong[locations, ])
         colnames(out_locations)[4:5] <- c("lat_grid", "long_grid")
-
-        # lat_grid <- sapply(global$cities$lat, FUN = function(x){
-        #         closest_index <- which.min(abs(x - climate_lats))
-        #         return(data.frame(closest_index = closest_index,
-        #                           closest_lat = climate_lats[closest_index]))
-        # })
-        # long_grid <- sapply(global$cities$lon, FUN = function(x){
-        #         closest_index <- which.min(abs(x - climate_longs))
-        #         return(data.frame(closest_index = closest_index,
-        #                           closest_long = climate_longs[closest_index]))
-        # })
-        # out_locations <- cbind(global$cities,
-        #                        lat_grid = lat_grid["closest_lat", ],
-        #                        long_grid = long_grid["closest_long", ])
-
 
         # Acquire the boundaries for the time series
         # Structure: c(start index, end index, # of elements spanning
@@ -314,22 +292,6 @@ processEnsemble <- function(ensemble, modelName, global, custom, type){
         # Acquire time series for every city
         tas <- readtas(ensemble = ensemble, global = global, locations = locations)
         series <- data.frame(tas[start:end, ])
-
[…]

### 2016-12-29 paper_drafts/RJournalRevisions/Example.tex
Add revisions to R Journal to git

\documentclass[margin=10pt]{standalone}
\usepackage{tikz}
\usepackage{tikz-qtree}
\usetikzlibrary{trees,calc,arrows.meta,positioning,decorations.pathreplacing,bending}

\tikzset{
    edge from parent/.style={draw, thick, blue!70!black},
    no edge from this parent/.style={
        every child/.append style={
        edge from parent/.style={draw=none}}},
    level 3/.style={yshift=5cm},
         }

\begin{document}

\begin{tikzpicture}[
    level/.style={sibling distance=40mm/#1},
    text=blue!70!black,
    >=latex,
    font=\sffamily
    ]

\node (z){cn}
  child {node (a) {cn/2}
    child {node  (b) {cn/4}}
    child {node (g) {cn/4}}
  }
    child {node (d) {cn/2}
      child {node  (e) {cn/4}}
      child {node (f) {cn/4}}
};

\node[left=7 of z]  (ln1) {}[no edge from this parent]
    child {node (ln2) {climate models}[no edge from this parent]
        child {node (ln3) {ensemble members}[no edge from this parent]}};

\draw[blue!70!black,dashed,thick]
    ($(a.west)+(-1em,0)$) -- (ln2.east);
\draw[blue!70!black,dashed,thick]
    ($(b.west)+(-1em,0)$) -- (ln3);

\end{tikzpicture}
\end{document}

### 2016-12-29 paper_drafts/RJournalRevisions/anderson-eason-barnes.Rmd
Add revisions to R Journal to git

---
title: >
    Working with daily climate model output data in R and the \pkg{futureheatwaves} package
author:
  - name: G. Brooke Anderson
    affiliation: Colorado State University
    address:
    - Department of Environmental & Radiological Health Sciences
    - 1681 Campus Delivery
    - Fort Collins, Colorado 80523
    email:  brooke.anderson@colostate.edu
  - name: Colin Eason
    affiliation: Colorado State University
    address:
    - Department of Computer Science
    - 1873 Campus Delivery
    - Fort Collins, Colorado 80523
    email:  aimesce@gmail.com
  - name: Elizabeth A. Barnes
    affiliation: Colorado State University
    address:
    - Department of Atmospheric Science
    - 1371 Campus Delivery
    - Fort Collins, CO 80523
    email:  eabarnes@atmos.colostate.edu
abstract: >
  Research on climate change impacts can require extensive processing, to incorporate output from ensembles of different climate models and different simulations of each model. This processing can be particularly extensive when identifying and characterizing extreme events like heat waves and frost day spells, as these must be processed from model output with daily time steps and may use identification threholds based on local distributions (e.g., 98th percentile temperature for that location). Here, we provide an overview of working with daily climate model output data in R. Further, we present the \pkg{futureheatwaves} package, which we developed to ease the process of identifying, characterizing, and exploring multi-day extreme events in climate model output. This package takes as input a directory of projection files from many climate models, identify all heat waves for multiple cities using a user-specified heat wave definition, and write an output directory of files with these heat waves and their characteristics (e.g., length, absolute and relative temperatures). The \pkg{futureheatwaves} package then allows the user to create custom functions and apply them across all heat wave files; this functionality can be used either to explore heat wave patterns (e.g., average length and intensity across different climate models) or to apply epidemiological estimates to project health impacts under different scenarios.
preamble: >
  % Supposedly allows you to add stuff to LaTeX preamble. Not for me.
output: rticles::rjournal_article
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(futureheatwaves)
library(ggplot2)
```

# Introduction

Research on climate change impacts can require extensive processing. This is not only because output files for a single climate model can be large, but also because of the rising popularity of ensemble techniques \citep{IPCCch1}, in which, to better characterize uncertainty in projections, impacts are assessed for multiple climate models, multiple simulations of each climate model, and multiple climate experiments. Projections of regional climate change over the next century are subject to high uncertainty due to three distinct sources: (1) internal climate variability, i.e. climate noise, (2) climate model uncertainty, i.e. the same forcing can produce a different response in different models and (3) scenario uncertainty, i.e. uncertainty in future climate forcings (e.g. \citet{hawkins2009potential}). One approach for ensuring that these sources of uncertainty are characterized is to simulate the future climate many times with multiple models and for multiple future scenarios.

For example, the Coupled Model Intercomparison Project, phase 5 (CMIP5; \citet{taylor2012overview}) brought together dozens of major climate modeling groups around the world to simulate the same future radiative forcing scenarios, but with their own models. This created an ensemble of state-of-the-art climate model projections that allow researchers to study projections and their uncertainties. Most of these modeling groups additionally performed more than one simulation for each scenario and model (i.e. multiple ensemble members), perturbing the initial conditions by a very tiny amount to quantify uncertainties due to internal climate variability.

This processing is particularly intensive for smaller time steps, like daily climate model output. While some climate impacts can be assessed using climate model output at monthly, seasonal, or yearly time steps, the impacts of multi-day extreme events must be assessed using output in daily time steps. Such multi-day extreme events include heat waves, cold spells, frost day spells, and droughts. The assessment is further complicated if extreme events are identified based on conditions that are rare for a certain location (e.g., 98th percentile of local temperature distribution for identifying heat waves) \citep{IPCCch1}, as this event definition must be determined at each study location from climate model output. Further, it is often of interest to create summaries of multiple characteristics of these extreme events. One interest, for example, in interpreting climate change scenarios output is whether frequency of heat waves or warm spells will change, as well as how certain characteristics of these extreme events (e.g., length, intensity) will change under different scenarios \citep{IPCCch1}.

We begin this article with an overview of climate model output data, particularly daily data, for R users. We outline where data from CMIP5 can be obtained as well as how to work with the file format (netCDF) from R. We overview some R packages that can be useful when working with this data, as well as aspects of the data (e.g., non-standard calendars) of which users should be aware when working with daily climate model output in R.

After this overview, we present the \pkg{futureheatwaves} package. which we created to aid in identifying and characterizing any type of multi-day extreme event from daily climate model output (Table \ref{tab:goals}). Further, this package provides some functionality particularly useful in identifying and characterizing heat waves specifically. Quantifying the impacts of heat waves on human health suffer from additional sources of uncertainty beyond those inherent in projections of regional changes in surface temperature. These include: (1) uncertainty in the ability of communities to adapt to changing temperatures (the adaptation scenario) and (2) uncertainty in the definition of a heat wave itself. Thus, identifying and characterizing the impacts of future heat waves in state-of-the-art climate models requires analyzing hundreds of projections based on combinations of anthropogenic activity, climate model and ensemble member, adaptation scenario, and heat wave definition. Such an analysis is non-trivial. For example, to estimate the possible impacts of heat waves using 32 ensemble members of CMIP5, two forcing scenarios, two different heat wave definitions, and five progressive assumptions about adaptation to heat, would require that one identify and characterize heat waves in 640 separate temperature time series for each region of interest.

\begin{table}[t]
\begin{center}
\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
\begin{tabular}{lL{\dimexpr0.8\textwidth-2\tabcolsep\relax}}
\toprule
 & Design goals of the \pkg{futureheatwaves} package \\
\midrule
1 & Make processing of large sets of climate projections more practical for researchers exploring the potential impacts of heat waves and other multi-day extreme events.\\
2 & Speed up processing time by incorporating C++ in event identification.\\
3 & Keep track of the names of climate models, and number of ensemble members processed for each.\\
4 & Not only identify, but also characterize, all extreme events within each climate projection, to allow the exploration of patterns in these characteristics across different projections and also to allow the use of more complex impacts models, including models that include effect modification by event characteristics (e.g., event length, event intensity). For example, this package allows the user to apply a health effects model where risk of mortality is not the same for every heat wave, but rather is modified by heat wave length, intensity, or other measured characteristics.\\
5 & Give users extensive power in customizing the process, including allowing custom event definitions.\\
6 & Allow users to easily explore the extreme events identified within all climate projections by applying custom functions across heat wave data sets from all projections at once.\\
7 & Create output that is in a ``tidy" data format, allowing it to work well with \pkg{ggplot2} for visualization.\\
\bottomrule
\end{tabular}
\end{center}
\caption{Design goals for the \pkg{futureheatwaves} package.}
\label{tab:goals}
\end{table}

The \pkg{futureheatwaves} package automates the process of identifying and characterizing multi-day extreme events across different ensemble members of one or more climate models. A variety of different heat wave definitions have been used to identify heat waves in a time series of temperature data \citep{smith2013heat}, and the choice of heat wave definitions can influence both health effect estimates \citep{chen2015influence, kent2014heat} and projected heat wave trends \citep{smith2013heat}. Further, other types of extreme events will be defined differently than heat waves (for example, frost day spells may be defined as one or more days with temperature at or below $0^oC$). This package therefore allows the user to create and use a custom the extreme event definition used to identify events in the climate model output. Finally, this package allows the user to explore the extreme events identified in the climate model output with a function that can take a user-defined R function and apply it across all extreme event files generated for the separate ensemble members.

# An overview of climate model output for R users

## CMIP5 climate model output data

For climate impact studies, a top source for climate model output files is the Coupled Model Intercomparison Project, which is currently in its fifth phase (CMIP5). Over 20 climate modeling groups have created one or more climate models which, for this project, are run using standardized scenarios \citep{taylor2012overview}. The resulting output is uniform across modeling groups and has a consistent structure, which allows comparison of simulations from different models \citep{IPCCch9}. For CMIP5, each group created simulations under several experiments, with experiments varying in terms of the radiative forcing. This radiative forcing depends on time-varying model inputs (greenhouse gas emissions or concentrations, land use changes, etc.), which are specified for each experiment \citep{taylor2012overview, IPCCch9}. Experiments include historical experiments (run using forcings consistent with observed and reconstructed data for 1850--2005), pre-Industrial control experiments, and experiments of future scenarios of radiative forcing over the 21st century or longer (e.g., RCP4.5, RCP8.5) \citep{taylor2012overview}. Some modeling groups created ensembles of output for a specific model and experiment, in which they ran the experiment multiple times with the model with very small changes to the initial conditions, resulting in multiple ensemble members for a single climate model and experiment. CMIP5 climate model output is created at a number of different time steps (e.g., daily, monthly, seasonal, yearly) \citep{taylor2010cmip5}, and some variables are reported at multiple depths in the ocean or atmosphere (e.g., ocean temperature). Here, we will focus on data with a daily time step for variables reported at a single depth (e.g., near-surface air temperature).

The CMIP5 climate model output data is distributed across data nodes at different climate modeling centers \citep{taylor2012overview}, but can be accessed centrally at the World Climate Research Programme CMIP5 data portal at https://pcmdi.llnl.gov/search/cmip5/. Users must register before downloading data, and some data are restricted to non-commercial use. There is a separate file for each combination of climate model, experiment, modeling realm (e.g., atmosphere, ocean), variable, time step, and ensemble member \citep{taylor2012overview, taylor2010cmip5}. For finer time scales, this output is further split across multiple files for specific year ranges (e.g., 5 years of output for each file) \citep{taylor2010cmip5}. Filenames for CMIP5 files can be parsed to generate information about the output variable, climate model, experiment, and ensemble member for the simulation \citep{taylor2010cmip5}. Files can be searched and downloaded through a point-and-click web interface. The data portal also allows you to generate a specific wget script, which you can use to download many files at once. A wget script can also be generated through Earth System Grid Federation's Search RESTful API. Tips on efficiently searching and downloading the data, including through use of wget scripts and the search API, are available as user tutorials through the website of the University of Colorado Boulder's Earth System CoG (e.g., https://www.earthsystemcog.org/projects/cog/doc/wget for a tutorial on downloading files using wget).

CMIP5 files are saved in Network Common Data Format (netCDF). NetCDF is a binary file format that allows storage of data representing a regular array. These climate model output files can be as large as several gigabytes [?]. For climate model output at a single depth (e.g., near-surface air temperature), the data is a 3-dimensional array, with dimensions representing time and two metrics of location (e.g., latitude and longitude) (Figure \ref{fig:netcdfexample}). Global climate models generate output at regularly-spaced time steps, typically at regularly-spaced grid points around the world. The latitude and longitude spacing of grid points vary by climate model, but are typically 1--2 degrees for atmospheric variables in CMIP5 models \citep{IPCCch9}. Each data point in the netCDF array gives the modeled value of the variable (e.g., surface temperature) for a single time point and location. For CMIP5 climate model output, the location units are in degrees east and degrees north for longitude and latitude, respectively. For daily output files, the time unit is in days since a specified origin date-time (e.g., days since 1850-01-01 00:00:00) \citep{taylor2010cmip5}. All CMIP5 output files are requred to include certain metadata  \citep{taylor2010cmip5}. This required metadata includes the experiment, forcing agents input to the model to create the simulation, time step, institution and institutional contact information, climate model, and modeling realm \citep{taylor2010cmip5}. The metadata also must include units for all of the coordinate variables (e.g., longitude, latitude, time).

\begin{figure}
\begin{center}
\includegraphics[width = 0.9\textwidth]{netcdf_structure_example}
\end{center}
\caption{Example of structure of a NetCDF climate model output file for a variable reported at a single depth, like surface air temperature. Data are stored in a three-dimensional array, with measurements at each time step and grid location. Surface temperature data are typically indexed in climate model output files by longitude, latitude, and time, in that order. For example, if the air surface temperature ("tas") is read into an R object called \code{tas}, you can access the value for the first day at the fourth longitude and third latitude with \code{tas[4, 3, 1]}. In addition to the output variable (temperature in this example), vectors with the ordered values of each dimension (longitude, latitude, and time) can also be read in from the netCDF file, as well as attribute data (e.g., units for variables, the calendar used for time).}
\label{fig:netcdfexample}
\end{figure}

\citet{taylor2012overview} and \citet{meehl2007wcrp} are excellent resources for finding out more about the CMIP climate model output data.

## Working with climate model output in R

A few R packages can be used to work with the netCDF file format used for CMIP5 files. Earlier packages to work with netCDF files included \pkg{ncdf} and \pkg{ncvar}, but these do not work with the newer netCDF version 4 released in 2008 and are no longer available through CRAN. More recent packages, including \pkg{ncdf4} \citep{ncdf4} and \pkg{RNetCDF} \citep{michna2013rnetcdf, RNetCDF}, work with both version 4 and netCDF's older version 3. Climate model output data for CMIP5 is required to conform with the earlier version (version 3) \citep{taylor2010cmip5} and so should work with any of these packages, although it is safer to write code using functions that can be used with version for in case future phases of CMIP do not require files to conform with netCDF version 3.

The netCDF format allows you to access metadata and variables describing the dimensions of the data without reading the full file into memory. The metadata describes the dimensions of the array and the units, etc., of each variable and is printed by the print method of the R object returned by the \code{nc\_open} function of \pkg{ncdf4}. For CMIP5 files, the dimension variables include the times and locations corresponding to each array in the data. Before reading in a variable from a netCDF file, a connection to the file must be opened, for example with the \code{nc\_open} function from \pkg{ncdf4}. Variables can then be read in using the \code{ncvar\_get} function from \pkg{ncdf4}, with the \code{varid} parameter set to "lat", "lon", or "time" (as a caveat, many climate models output to non-Gregorian calendars, in which case the time variable should be read in using a different function, as discussed later in this section). The climate output variable (e.g., near-surface air temperature) can similarly be read in using \code{ncvar\_get}. In this case, the \code{varid} parameter should be set using the appropriate CMIP5 variable name (e.g., "tas" for near-surface air temperature); these variable names can be found in the CMIP Requested Output tables \citep{taylor2010cmip5}. In practice, you can use the dimensional time and location data to identify the location of the variable data you need in the netCDF array and use indexing to read only that data into memory, without needing to read in the full file \citep{michna2013rnetcdf}. Once the user is done reading in data from the file, the connection can be closed (e.g., with the \code{nc\_close} function from \pkg{ncdf4}).

Since the late 1500s, Western dates have been set using the Gregorian calendar, which has 365.2425-day years. Some climate models, however, output to different calendars, including the Julian calendar (365.25-day years), a calendar where there are no leap years (365-day years), a calendar where every year is a leap year (366-day years), and a calendar of twelve 30-day months (360-day years) \citep{cfconventions}. With these non-Gregorian calendars, R's base functions for converting a vector to a Date class based on the number of days since an origin date (\code{as.Date}, \code{as.POSIXct}) do not return the desired values. The \pkg{PCICt} \citep{PCICt} and \pkg{ncdf4.helpers} \citep{ncdf4.helpers} packages provide further functionality with netCDF files that can be useful when working with climate model output data and provide particular help in working with different calendars. CMIP5 netCDF files include information on the calendar used, which the \code{nc.get.time.series} function in \pkg{ncdf4.helpers} use to convert the "time" variable in the file to an object of the PCICt class, which provides Date-like functionality for 360- and 365-day calendars \citep{PCICt}. As a note, while these functions will help with handling most CMIP5 files, the CMIP5 standards allows use of other calendars which may not be succesfully handled by these functions, so it is important to assess whether the time variable range in the PCICt object correctly matches the expected date ranges for a file as you process CMIP5 data in R.

The size of the climate model output files can be large enough that it may make more sense to work with smaller chunks of the data, rather than reading all data into memory and working with the data all at once \citep{RCMIP5}. This problem aggregates when working with multiple climate models and more than one ensemble member for each of those climate models.

The following code gives an example of using a CMIP5 file and some of the R packages and functions discussed. Results include a map of near-surface air temperatures from a single climate model on a specific date (July 1, 2075) for China (Figure [x]) and a time series of daily near-surface air temperature simulations for the climate model grid point closest to Beijing, China (Figure [x]). To try this example, you will need to download the file ... from ... and save it in the "tmp" subdirectory of your home directory.

```{r}

```

In addition to these general packages for working with netCDF files, there are several R packages specifically for working with climate model output data, including \pkg{RCMIP5} \citep{RCMIP5} and \pkg{wux} \citep{wux}. However, these packages are more useful for working with data output at time steps of a month or higher and have limited utility with the daily climate model output data required for studies of multi-day extreme events.

The \pkg{RCMIP5} package includes functions to read in CMIP5 data from netCDF files, scan a directory of CMIP5 files and determine models with continuous available data, create objects of a special "cmip5data" class to work with CMIP5 data within R, and parse the filenames for all files in a directory to extract information within the filename. For this package, most functions only work with monthly or less frequent data (e.g., the functions \code{checkTimePeriod} and \code{cmip5data}, as well as functions that work with \code{cmip5data} objects including \code{filterDimensions} and \code{getProjectionMatrix}) \citep{RCMIP5}. While the \code{loadCMIP5} function does successfully load the daily data as a \code{cmip5data} object, most of the methods for this object type do not do anything meaningful for daily data. The package's \code{getFileInfo} function, however, will work with CMIP5 files of any time step; this function identifies all CMIP5 files in a directory and creates a dataframe that parses the information contained in the file name. As a note, the \code{get.split.filename.cmip5} function in the \pkg{ncdf4.helper} package similarly can be used to parse information contained in CMIP5 file names \citep{ncdf4.helpers}.

The \pkg{wux} package \citep{wux} includes functions that allow the user to download CMIP5 monthly-aggregated output directly from within R with the \code{CMIP5fromESGF} function. However, this function does not allow downloading of climate model output with finer time steps, like daily data. This package uses the \code{models2wux} function to read in climate model output netCDF files and convert to a WUX dataframe that can be used by other functions in the package. While this function can input climate model output with daily time steps, if the element "what.timesteps" of the \code{modelinput} list input is set to "daily", the function aggregates this data to a monthly or less frequent (e.g., seasonal) aggregation when creating the WUX dataframe. Therefore, while this package provides very useful functionality for working with averaged output of daily climate model output data, it cannot easily be used to identify and characterize multi-day extreme events like heat waves.

More tips on working with CMIP5 files in R are provided in the "starting_from_netcdf" vignette of the \pkg{futureheatwaves} package.

# The \pkg{futureheatwaves} package

## How the package works

\begin{widefigure}
\includegraphics[width = 0.9\textwidth]{OverviewFigure}
\caption{Overview of the functionality of the \pkg{futureheatwaves} package. The package takes a directory with climate projection files (left), for one or more climate models, with one or more ensemble members for each climate model (this example figure shows four climate models with one or two ensemble members each). The gen\_hw\_set function processes these files to create a data frame for each ensemble member, identifying and characterizing all multi-day extreme events (e.g., heat waves) in the time series projection for that ensemble member. The apply\_all\_models function allows users to explore these extreme events by applying user-created functions across all the extreme event data frames, creating a summary data frame with results.}
\label{fig:overview}
\end{widefigure}

The \pkg{futureheatwaves} package was facilitated to aid in identifying, characterizing, and exploring multi-day extreme events in daily climate model output data. Figure \ref{fig:overview} gives an overview of the two primary functions of the \pkg{futureheatwaves} package. First, the \code{gen\_hw\_set} function processes a directory of climate projection files that are stored locally on the user's computer (Figure \ref{fig:overview}, "Climate projections"), to generate a list of all extreme events in each projection, as well as over a dozen characteristics of each identified extreme event (Table \ref{tab:hwcharacteristics}). The extreme events are identified and characterized at one or more study locations (e.g., cities), with study locations specified by the user in an input file. The extreme events identified for each ensemble member are output as separate files in a directory specified by the user (Figure \ref{fig:overview}, ``Extreme events datasets"). Second, the \code{apply\_all\_models} function allows the user to apply custom functions across all the extreme event data frames generated by \code{gen\_hw\_set} to summaries of extreme events across all climate models and ensemble members (Figure \ref{fig:overview}, right).

\begin{table}
\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
\begin{tabular}{lL{\dimexpr0.8\textwidth-2\tabcolsep\relax}}
\toprule
Column name & Description of characteristic \\
\midrule
\code{mean.var} & Average daily value of the variable across all days in the extreme event, in the units in which the variable is expressed in input files (e.g., average daily mean temperature during the heat wave in degrees Kelvin) \\
\code{max.var} & Highest daily value of the variable across all days in the extreme event, in the units in which the variable is expressed in input files \\
\code{min.var} & Lowest daily value of the variable across all days in the extreme event, in the units in which the variable is expressed in input files \\
\code{length} & Number of days in the event \\
\code{start.date} & Date of the first day of the event \\
\code{end.date} & Date of the last day of the event \\
\code{start.doy} & Day of the year of the first day of the event (1 = Jan. 1, etc.)\\
\code{start.month} & Month in which the event started (1 = January) \\
\code{days.above.abs.thresh.1} & Number of days in the event above a specified absolute threshold (default is the number of days in the event above $80^oF$ / $26.7^oC$, but this can be changed with the \code{absolute\_thresholds} argument in \code{gen\_hw\_set}) \\
\code{days.above.abs.thresh.2} & Number of days in the event above a specified absolute threshold (default is the number of days in the event above $85^oF$ / $29.4^oC$) \\ 
\code{days.above.abs.thresh.3} & Number of days in the event above a specified absolute threshold (default is the number of days in the event above $90^oF$ / $32.3^oC$) \\
\code{days.above.abs.thresh.4} & Number of days in the event above a specified absolute threshold (default is the number of days in the event above $95^oF$ / $35.0^oC$) \\
\code{days.above.99th} & Number of days in the event above the $99^{th}$ percentile of the variable for the location, using the period specified with the \code{referenceBoundaries} argument in \code{gen\_hw\_set} as a reference for determining these percentiles \\
\code{days.above.99.5th} & Number of days in the event above the $99.5^{th}$ percentile of the variable for the location, using the period specified with the \code{referenceBoundaries} argument in \code{gen\_hw\_set} as a reference for determining these percentiles \\
\code{first.in.year} & Whether the event was the first to occur in its calendar year in the location \\
\code{mean.var.quantile} & The percentile of the average variable value during the event compared to the location's year-round distribution of the variable, based on the variable distribution for the location during the period specified by the \code{referenceBoundaries} argument in \code{gen\_hw\_set} \\
\code{max.var.quantile} & The percentile of the maximum variable value during the event compared to the location's year-round distribution of the variable, based on the variable distribution for the location during the period specified by the \code{referenceBoundaries} argument in \code{gen\_hw\_set} \\
\code{min.var.quantile} & The percentile of the minimum variable value during the event compared to the location's year-round distribution of the variable, based on the variable distribution for the location during the period specified by the \code{referenceBoundaries} argument in \code{gen\_hw\_set} \\
\code{mean.seasonal.var} & The location's average seasonal value of the variable (by default, season is set to May--September, but this can be changed with the \code{seasonal\_months} argument in \code{gen\_hw\_set}), based on the variable values for the location during the years specified by the \code{referenceBoundaries} argument in \code{gen\_hw\_set} \\
\code{mean.yearround.var} & The location's average year-round value of the variable, based on the variable values for the location during the years specified by the \code{referenceBoundaries} argument in \code{gen\_hw\_set} \\
\bottomrule
\end{tabular}
\caption{Extreme event characteristics measured by the \code{gen\_hw\_set} function in the \pkg{futureheatwaves} package. The left column gives the name of each variable's column in the extreme event datasets created by the \code{gen\_hw\_set} function. When characterizing extreme events below a threshold, like cold spells, appropriate alternatives are given for some columns (e.g., \code{days.below.abs.thresh.1}, \code{days.below.1st}).}
\label{tab:hwcharacteristics}
\end{table}

The user can extensively customize the definition used to identify extreme events. The default is a definition for heat waves (two or more days with temperatures at or above the city's $98^{th}$ percentile of year-round temperature), but users can write a custom R function with either a different heat wave definition or with a definition appropriate for a different type of extreme event (e.g., one or more days at or below $0^oC$ for frost day spells). 
Some of these heat wave characteristics are based on relative temperature, which is a measure of how the temperature during a heat wave compares to the typical distribution of temperature at that location \citep{IPCCch1}. These relative temperature metrics will vary depending on whether you calculate them based on a location's present-day temperature distribution or on the location's temperature distribution in the future, since typical temperature distributions in many locations are expected to change with climate change. The package therefore allows the user to specify date ranges of the temperature distributions to be used in calculating these relative temperature metrics in each location.

Second, once the user creates these data frames of location-specific heat waves, the `apply_all_models` function allows the user to apply a custom function across all the heat wave data frames that were generated to create a summary data frame (Figure \ref{fig:overview}, right). This functionality can be used to generate summary statistics (e.g., determine average heat wave length or total heat wave days), or can be used to apply more complex functions (e.g., apply epidemiologic effect estimates across the heat waves to generate health impact estimates).

## Example data

We have included data files in the package to serve as example files, so users can try this package before applying it to their own directory of climate projection files. The example files are all included as comma-separated (.csv) files, rather than as saved R objects, because the `gen_hw_set` function requires a directory of comma-separated files as input and csv is a general file format familiar to most users. This example data therefore also allow users to see how to create the appropriate directory set-up for this package.

This example data comes from two climate models that are a part of CMIP5: (1) the model of the Beijing Climate Center, China Meteorological Administration (BCC) \citep{xin2013introduction} and (2) the National Center for Atmospheric Research's (NCAR's) Community Climate System Model, version 4 (CCSM4) \citep{gent2011community}. We include one ensemble member from BCC (r1i1p1) and two from CCSM (r1i1p1 and r2i1p1). To ensure that the size of this example data is reasonably small, we have only included projection data for grid points from these climate models that are near five U.S. east coast cities: New York, NY; Philadelphia, PA; Newark, NJ; Baltimore, MD, and Providence, RI. Further, to keep the file sizes reasonably small, the historical projections range over the years 1990 to 1999, while the future projections are limited to 2060 to 2079. Users' applications of this package will likely use directories with many more climate model ensemble members and more locations; however, the operation of the package is the same for this smaller example application as it would be for a much larger application.

Once the \pkg{futureheatwaves} package is installed and loaded, the user can find the local location of these files using R's `system.file` function. In the later sections of this article, we show how to use the package functions with these example files as inputs.

When setting up to run climate model output data beyond this example data, any downloaded CMIP5 files will require some pre-processing, in terms of getting the getting the data into a specific format and storing these files in a specific directory structure. A file of study locations must also be saved locally in a specific format. Extensive details on the required format of files and required input directory structure is available in the \pkg{futureheatwaves} vignette "futureheatwaves", while tips on processing CMIP5 files into the required structure are given in the package's "starting_from_netcdf" vignette.

## Basic example of processing projection files

Once these files and directories are set up, \pkg{futureheatwaves} can process them to identify and characterize heat waves in each ensemble member's projection for each  location using the `gen_hw_set` function. For example, to process the example climate projections directory included with the package, the user can run:

```{r eval = FALSE}
projection_dir_location <- system.file("extdata/cmip5",
                                       package = "futureheatwaves")
city_file_location <- system.file("extdata/cities.csv",
                                  package = "futureheatwaves")

gen_hw_set(out = "example_results",
           dataFolder = projection_dir_location ,
           dataDirectories = list("historical" = c(1990, 1999),
                                        "rcp85" = c(2060, 2079)),
           citycsv = city_file_location,
           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
           tasFilenames = "tas_NorthAmerica_12mo.csv",
           timeFilenames = "time_NorthAmerica_12mo.csv")
```

This code first identifies and saves as objects the path names on the user's computer of the example climate projections directory (`projection_dir_location`) and the file of study locations (`city_file_location`). The `gen_hw_set` function processes this example input and creates a new directory, `example_results`, with files of identified and characterized heat waves, in the user's current working directory. In this example code, this processing is done using  default values for the heat wave definition, years for which to generate the heat wave data sets, etc. Ways to customize these choices are fully explained later in the text.

In this call, the user must specify the directory where the results should be written (`out`), the location of the directory of climate projections (`dataFolder`), the names of the two main subdirectories of that climate projection directory, as well as their year boundaries (`dataDirectories`; see, for example, Figure \ref{fig:directories}), the location of the city location file (`citycsv`), and the names used for the grid coordinate, climate projection, and projection date files (`coordinateFilenames`, `tasFilenames`, and `timeFilenames`). When `gen_hw_set` is run, the user is advised that the function will write files to his computer and must agree to proceed:

```
 Warning: This function will write new files to your computer in the 
 ~/tmp/example_results/ directory of your computer. If that directory already exists,
running this function will write over it. 
 Do you want to continue? (y / n): 
```

\noindent The function provides status reports while processing. This helps the user see that the function call is progressing, as `gen_hw_set` can take a while to run if it is processing many locations and / or many climate ensemble members. Once the function has completed running, results will be written locally to the directory specified by the `out` argument of `gen_hw_set`. The structure of this output directory is shown in Figure \ref{fig:directories} (right). This directory will include files with some basic information about the climate models and the closest grid points of each climate model to each location. The output directory will also include a directory with files of identified and classified heat waves for each ensemble member, including all characteristics in Table \ref{tab:hwcharacteristics}.

## Customizing the extreme event definition

The default heat wave definition for this package is:

> A \dfn{heat wave} is two or more days at or above a city-specific threshold temperature, with the threshold determined as the $98^{th}$ percentile of year-round temperature in the city during some reference period (by default, 1990--1999).

Many different definitions of heat waves exist (e.g., \citet{smith2013heat, kent2014heat}), so researchers will often want to use alternative ways to define heat waves. Researchers might want to use a specific definition, for example, because it matches the definition used by local health officials to declare heat wave warnings or, in the case of health impact assessments, to match with a definition used in an epidemiological study. Three components of the heat wave definition can be easily customized in the `gen_hw_set` function call, without creating a new R function to use to identify heat waves. The customization of the heat wave definition is even more extensive as one has the option of writing a custom R function.

First, the percentile used to identify a heat wave can be changed using the `probThreshold` option in `gen_hw_set`. This option can take values between 0 and 1. The default value is 0.98, or a definition with a threshold of the $98^{th}$ percentile of the location's temperature. For example, to identify heat waves as two or more days at or above the city's $99^{th}$ percentile of year-round temperature, the user could run:

```{r eval = FALSE}
gen_hw_set(out = "example_results",
           dataFolder = projection_dir_location ,
           dataDirectories = list("historical" = c(1990, 1999),
                                        "rcp85" = c(2060, 2079)),
           citycsv = city_file_location,
           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
           tasFilenames = "tas_NorthAmerica_12mo.csv",
           timeFilenames = "time_NorthAmerica_12mo.csv",
           probThreshold = 0.99)
```

\noindent Second, the user can change the number of days used in the heat wave definition using the `numDays` argument in the `gen_hw_set` function. Combined, these two customization choices can allow the user to identify heat waves using many of the heat wave definitions used in previous climate and health research-- for example, 9 of the 16 heat wave definitions used in \citet{kent2014heat} could be fit using different combinations of these two options for specifying threshold percentile and number of days.

Third, it is possible to specify the range of years that should be used when determining this threshold. For example, if you wanted to base the threshold for each location on its current climate, you could leave this option as its default, while if you wanted to use a different set of years (i.e. a threshold relative to future projected temperatures), you could set the start and end year bounds for this reference period using the `thresholdBoundaries` argument in the function `gen_hw_set`. For example, to use temperatures in each city from 2070 to 2079 to determine threshold temperatures, you would run:

```{r eval = FALSE}
gen_hw_set(out = "example_results",
           dataFolder = projection_dir_location ,
           dataDirectories = list("historical" = c(1990, 1999),
                                        "rcp85" = c(2060, 2079)),
           citycsv = city_file_location,
           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
           tasFilenames = "tas_NorthAmerica_12mo.csv",
           timeFilenames = "time_NorthAmerica_12mo.csv",
           thresholdBoundaries = c(2070, 2079))
```

It is also possible to use a completely customized heat wave definition. This functionality allows the user to use heat wave definitions that either require a number of days above an absolute threshold (e.g., maximum temperature of $\ge95^oF$ for $\ge1$ day; \citet{kent2014heat, tan2007heat}) or that require a combination of thresholds to be met (e.g., maximum daily temperature above a lower threshold every day of the heat wave and above a higher threshold for a certain number of days; \citet{kent2014heat, peng2011toward}).

To use a customized heat wave definition, the user can write and load an R function that implements the definition. To work correctly, this custom function must allow only specific inputs and generate only specific outputs. Details about this required structure are provided in the \pkg{futureheatwaves} package vignette.

To help in developing custom heat wave identification functions, the package includes an example of the required input data frame, `datafr`, which can be loaded with `data(datafr)`. If a custom heat wave function is properly set up to use in this package, it can take as input this example data frame and a threshold value and will return the original data frame with the columns for whether the day was part of a heat wave (`hw`) and, if so, the sequential number of the heat wave out of all heat waves in that location (`hw.number`).

The custom function can then be referenced using the `IDheatwavesFunction` option in `gen_hw_set`, and it will be used to identify heat waves in the data set. As an example, we have included an alternative ID definition function in this package, called `IDHeatwavesAlternative`, which identifies heat waves as two or more days above the higher of either the location's $98^{th}$ percentile temperature or $80^{o}F$, with the idea that this definition prevents the identification of heat waves at mild temperatures in locations with mild or cool climates. The following code processes the example climate projecting files using this function as the heat wave definition:

```{r, eval = FALSE}
gen_hw_set(out = "example_results",
           dataFolder = projection_dir_location ,
           dataDirectories = list("historical" = c(1990, 1999),
                                        "rcp85" = c(2060, 2079)),
           citycsv = city_file_location,
           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
           tasFilenames = "tas_NorthAmerica_12mo.csv",
           timeFilenames = "time_NorthAmerica_12mo.csv",
           IDheatwavesFunction = "IDHeatwavesAlternative")
```

With this ability to write and use a custom heat wave identification function, most heat wave definitions can be applied with this package. The exception is definitions that require two separate temperature metrics (e.g., maximum temperature above a certain threshold or minimum temperature above a different threshold; \citet{kent2014heat, robinson2001definition}), because the `gen_hw_set` function cannot take as input projections of more than one temperature metric. Definitions that use a heat metric that combines multiple inputs (e.g., heat index, which is a combined measure of temperature and air moisture) can be used, but require that the calculation of the metric is done when creating the input files to include in the ``Climate projections" directory (Figure \ref{fig:overview}, left), rather than as part of the heat wave definition function.

To increase processing speed when identifying heat waves, we coded parts of the default heat wave identification function in C++ and synced it with R using the \CRANpkg{Rcpp} package \citep{Rcpp}. Users should consider a similar strategy for custom heat wave definitions, especially if processing a large number of climate projection files.

## Exploring extreme events

Once you have created a directory of files with characterized heat waves for each ensemble member, the results can be explored using the `apply_all_models` function. This function allows the user to apply custom R functions across all heat wave data frames created by the `gen_hw_sets` call. The user can apply any R function that follows certain standards in accepting input and returning output. Full details on these standards are given in the \pkg{futureheatwaves} package vignette.

As an example, if the user wanted to to get the average temperature of the heat waves identified within each ensemble member, he or she could write a simple function:

\begin{verbatim}
average_mean_temp <- function(hw_datafr){
        out <- mean(hw_datafr$mean.var)
        return(out)
        }
\end{verbatim}

\noindent The `apply_all_models` function can then apply this `average_mean_temp` function across the heat wave data frames for all ensemble members in all climate models:

```{r echo = FALSE}
average_mean_temp <- function(hw_datafr){
        out <- mean(hw_datafr$mean.var)
        return(out)
        }
```

```{r}
out <- system.file("extdata/example_results", package = "futureheatwaves")
apply_all_models(out = out, FUN = average_mean_temp)
```

Note that the location of the directory with the heat wave data frames must be specified using the `out` argument when calling \code{apply\_all\_models}. Typically, this will be the directory path for the directory specified with the `out` argument in \code{gen\_hw\_set}. Location-specific results can be generated using the \code{city\_specific} argument in \code{apply\_all\_models}:

```{r}
apply_all_models(out = out, FUN = average_mean_temp, city_specific = TRUE)
```

This output is structured as ``tidy" data \citep{wickham2014tidy}, allowing it to be used easily with the graphing package \CRANpkg{ggplot2} \citep{ggplot2}.

The `apply_all_models` function can also be used to project the health impacts of heat waves. As a very simplistic example, \citet{anderson2009weather} estimated that heat waves, defined as two or more days at or above a community's $98^{th}$ percentile temperature, were associated with an added relative risk of 1.032 for cardiorespiratory mortality risk in 107 U.S. communities. A simple estimate of excess deaths associated with this added heat wave risk in a community can be calculated as \citep{peng2011toward}:

\begin{equation}
E_c = N_c * (RR - 1) * L_c 
\label{eq:healthimpacts}
\end{equation}

where:

- $E_c$ is the total number of excess deaths in community $c$;
- $N_c$ is the baseline average daily mortality in community $c$; 
- $RR$ is the relative risk of cardiorespiratory mortality per day associated with a heat wave; and 
- $L_c$ is the total number of heat wave days in community $c$ over the study period.

This impact assessment equation can be translated into a function that merges each projection's heat wave data frame with a data frame of community-specific baseline mortality rates ($B_c$), calculates equation \ref{eq:healthimpacts} for each heat wave, and then sums up the total excess deaths across all heat waves:

\begin{verbatim}
excess_deaths <- function(hw_datafr, base_mortality, RR = 1.032){
        hw_datafr <- dplyr::left_join(hw_datafr, base_mortality,
                                      by = "city") %>%
                dplyr::mutate(excess_deaths = base_mort * length * RR)
        out <- sum(hw_datafr$excess_deaths)
        return(out)
}
\end{verbatim}

\noindent Once defined in R, this function can be applied across all heat waves from all climate models' ensemble members, provided that you have a data frame called `city_mortality` with columns with each community's identifier (`city`) and baseline mortality rate (`base_mort`), using the call:

```{r eval = FALSE}
apply_all_models(out = out, FUN = excess_deaths, base_mortality = city_mortality)
```

As examples of the type of functions that can be used with `apply_all_models`, we have included several simple functions for exploring heat waves (Table \ref{tab:applyfunctions}). We included an example data frame representative of the heat wave data frames that `gen_hw_set` outputs, to use to test new functions. This data can be loaded with `data(hw_datafr)`. The process for creating a new function to use to explore heatwaves should be to:

1. Load the example data frame with `load(hw_datafr)`;
2. Build a function that works with `hw_datafr` as an input; and 
3. Apply the function with `apply_all_models` to process all of the heat wave files.

\begin{table}
\centering
\begin{tabular}{ll}
\toprule
Function & Description \\
\midrule
\code{number\_of\_heatwaves} & Determines the number of extreme events\\
\code{heatwave\_days} & Sums up the total number of extreme event days across all heat waves\\
\code{average\_length} & Calculates the average length of extreme events\\
\code{average\_mean\_temp} & Calculates the average mean temperature across all extreme events\\
\bottomrule
\end{tabular}
\caption{Example functions included in the \pkg{futureheatwaves} package that can be used to explore extreme events using the \code{apply\_all\_models} function.}
\label{tab:applyfunctions}
\end{table}

## Customizing dates for projections and reference temperatures

By default, the `gen_hw_set` function will generate heat wave projections for the years 2070 to 2079, to align with the example data. This projection date range can be changed by specifying alternative starting and ending years in the `projectionBoundaries` argument of the `gen_hw_set` function. For example, to create projections for 2060 to 2079, the user would call:

```{r, eval = FALSE}
gen_hw_set(out = "example_results",
           dataFolder = projection_dir_location ,
           dataDirectories = list("historical" = c(1990, 1999),
                                        "rcp85" = c(2060, 2079)),
           citycsv = city_file_location,
           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
           tasFilenames = "tas_NorthAmerica_12mo.csv",
           timeFilenames = "time_NorthAmerica_12mo.csv",
           projectionBoundaries = c(2060, 2079))
```

The heat wave data sets characterize heat wave in several ways that are based on relative temperature (for example, the `mean.temp.quantile`, `max.temp.quantile`, and `min.temp.quantile` columns described in Table \ref{tab:hwcharacteristics}). These characteristics are measured for each of the heat waves identified by the `gen_hw_set` function by taking the absolute temperature of the heat wave (e.g., average temperature during the heat wave is $90^oF$) and comparing it to the heat wave location's typical temperature distribution. This process generates relative measures of how intense the heat wave is compared to what is normal in that location (e.g., $90^oF$ is in the $99^{th}$ percentile of year-round temperatures in the location).

The values of these relative heat wave characteristics will depend on what time period is used to determine the location's typical temperature distribution. For example, the same heat wave would likely have much higher values for these relative characteristics if the temperature during the heat wave is compared to typical temperatures in the present day for that location versus if the temperature is compared to typical temperatures much later in the $21^{st}$ century for the same location.

The default in the `gen_hw_set` function is to use temperatures for the period 2070 to 2079 for these reference temperatures to calculate all relative heat wave characteristics. However, the user can change this specification using the `referenceBoundaries` option of `gen_hw_set`. This functionality can be useful in exploring the role of adaptation in future heat waves. For example, to use temperature projections for 1990 to 1999 when calculating relative characteristics of heat waves, to explore the assumption that cities remain adapted to their present-day climate, rather than changing in adaptation as climate change increases temperatures, a user could run:

```{r, eval = FALSE}
gen_hw_set(out = "example_results",
           dataFolder = projection_dir_location ,
           dataDirectories = list("historical" = c(1990, 1999),
                                        "rcp85" = c(2060, 2079)),
           citycsv = city_file_location,
           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
           tasFilenames = "tas_NorthAmerica_12mo.csv",
           timeFilenames = "time_NorthAmerica_12mo.csv",
           referenceBoundaries = c(1990, 1999))
```

For these date range specifications, there are some restrictions on which year ranges can be selected. The starting year cannot be earlier than the first year in the first subdirectory, the ending year cannot be later than the last year of the second subdirectory, and the custom date boundaries cannot span the two subdirectories.

## Mapping grid points

It can be useful to explore the location of the climate model grid point used to pull climate model output for each study location with a given climate model. Therefore, the package has a function called `map_grid_leaflet` that plots the locations of grid points used for each location from each climate model. This function is built using the htmlWidget \pkg{leaflet} package \citep{leaflet}. The following code illustrates the use of this function with the example data to create Figure \ref{fig:gridmap}, which plots the grid points used in the example data from the BCC climate model in the example data:

```{r eval = FALSE}
out <- system.file("extdata/example_results", package = "futureheatwaves")
map_grid_leaflet(plot_model = "bcc1", out = out)
```

\begin{figure}
\begin{center}
\includegraphics[width = 0.9\textwidth]{ExampleLeaflet}
\end{center}
\caption{Snapshot of an example interactive map created using the \code{map\_grid\_leaflet} showing the locations of study cities and their matching climate model grid points for the BCC climate model example data included with \pkg{futureheatwaves}. The lines on the map connect each climate model grid point to the study location(s) for which that grid point was used. The interactive maps include pop-ups with city identiers; one is shown open in this snapshot as an example. From this map, you can see that the climate model grid point closest to New York City for this climate model is over the Atlantic Ocean.}
\label{fig:gridmap}
\end{figure}

## Extensions

The functionality of this package can be easily expanded by loops. For example, to explore the role of the heat wave definition on projections, the user could create a loop to run `gen_hw_set` and `apply_all_models` to the same directory of climate projections but with a variety of different functions used to identify the heat waves in the projections.

Also, while this package was created to be used for research on heat waves in climate change projections, with some modifications it can be used more broadly. For example, there are other episodes like wildfires and air pollution where it may be interesting to identify extended periods of high exposures in projection time series, and this package could be applied to explore these exposures. The directory of projection data would need to be set up in the same structure as for exploring heat waves, and the `input_metric` should be set as `fahrenheit`, to pass the exposure values through to the characterized data sets without performing a conversion. A user could also use this package to explore events that must be lower than some minimum threshold (e.g., cold waves), but it would take some extra coding and conversions, since the functions in this package are written to identify periods above a threshold (for example, the user could multiple all projected temperatures by -1, and then the coldest temperatures would register as being the highest).

```{r eval = FALSE}
## Cold wave example
gen_hw_set(out = "example_results",
           dataFolder = projection_dir_location ,
           dataDirectories = list("historical" = c(1990, 1999),
                                  "rcp85" = c(2060, 2079)),
           citycsv = city_file_location,
           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
           tasFilenames = "tas_NorthAmerica_12mo.csv",
           timeFilenames = "time_NorthAmerica_12mo.csv", 
           probThreshold = 0.10, 
           above_threshold = FALSE, 
           absolute_thresholds = c(266, 263, 260, 258), 
           seasonal_months = c(12, 1, 2))
```

>  Extreme statistics and spell-lengths: "A 'frost day' is defined as one during which the minimum temperature falls below freezing point (0 degC). This is described as a climatological statistic, in which the minimum temperature is first calculated within each day, and then the number of days or spell lengths meeting the specified condition are evaluated." (from one example in http://cfconventions.org/Data/cf-conventions/cf-conventions-1.6/build/cf-conventions.html#calendar)

Users doing these kinds of extensions will need to pay attention to a few points. First, some of the event characteristics (first in the calendar year, average of May--September temperatures, days above $90^{o}F$) might not be meaningful for studies of other types of events. Further, because event periods are usually defined as a string of multiple days exceeding some threshold, the functions in this package may miss the first and last event of the time period. For example, if the first day of the time series is the last day of an event, this function would not identify that event because it lacks data from the earlier days that allow this day to meet the event definition. This issue would lead to, at most, missing two events out of each projection, but should be considered if studying events that might occur near the start and end of projection data. If there is adequate interest from researchers, in the future we may adapt the package to make these secondary applications part of the package.

# Acknowledgements

This work was supported by grants from the National Institute of Environmental Health Sciences (R00ES022631), the National Science Foundation (1331399), and the Colorado State University Vice President for Research.

\bibliography{Anderson}

### 2016-12-29 paper_drafts/RJournalRevisions/anderson-eason-barnes.tex
Add revisions to R Journal to git

--- 2016-12-29 paper_drafts/RJournalRevisions/anderson-eason-barnes.Rmd
@@
----
-title: >
-    Working with daily climate model output data in R and the \pkg{futureheatwaves} package
-author:
-  - name: G. Brooke Anderson
-    affiliation: Colorado State University
-    address:
-    - Department of Environmental & Radiological Health Sciences
-    - 1681 Campus Delivery
-    - Fort Collins, Colorado 80523
-    email:  brooke.anderson@colostate.edu
-  - name: Colin Eason
-    affiliation: Colorado State University
-    address:
-    - Department of Computer Science
-    - 1873 Campus Delivery
-    - Fort Collins, Colorado 80523
-    email:  aimesce@gmail.com
-  - name: Elizabeth A. Barnes
-    affiliation: Colorado State University
-    address:
-    - Department of Atmospheric Science
-    - 1371 Campus Delivery
-    - Fort Collins, CO 80523
-    email:  eabarnes@atmos.colostate.edu
-abstract: >
-  Research on climate change impacts can require extensive processing, to incorporate output from ensembles of different climate models and different simulations of each model.
-This processing can be particularly extensive when identifying and characterizing extreme events like heat waves and frost day spells, as these must be processed from model output with daily time steps and may use identification threholds based on local distributions (e.g., 98th percentile temperature for that location).
-Here, we provide an overview of working with daily climate model output data in R.
-Further, we present the \pkg{futureheatwaves} package, which we developed to ease the process of identifying, characterizing, and exploring multi-day extreme events in climate model output.
-This package takes as input a directory of projection files from many climate models, identify all heat waves for multiple cities using a user-specified heat wave definition, and write an output directory of files with these heat waves and their characteristics (e.g., length, absolute and relative temperatures).
-The \pkg{futureheatwaves} package then allows the user to create custom functions and apply them across all heat wave files; this functionality can be used either to explore heat wave patterns (e.g., average length and intensity across different climate models) or to apply epidemiological estimates to project health impacts under different scenarios.
-preamble: >
-  % Supposedly allows you to add stuff to LaTeX preamble. Not for me.
-output: rticles::rjournal_article
----
+% !TeX root = RJwrapper.tex
+\title{Working with daily climate model output data in R and the
+\pkg{futureheatwaves} package}
+\author{by G. Brooke Anderson, Colin Eason, Elizabeth A. Barnes}
 
-```{r echo = FALSE, message = FALSE, warning = FALSE}
-library(futureheatwaves)
-library(ggplot2)
-```
+\maketitle
 
-# Introduction
+\abstract{%
+Research on climate change impacts can require extensive processing, to
+incorporate output from ensembles of different climate models and
+different simulations of each model. This processing can be particularly
+extensive when identifying and characterizing extreme events like heat
+waves and frost day spells, as these must be processed from model output
+with daily time steps and may use identification threholds based on
+local distributions (e.g., 98th percentile temperature for that
+location). Here, we provide an overview of working with daily climate
+model output data in R. Further, we present the \pkg{futureheatwaves}
+package, which we developed to ease the process of identifying,
+characterizing, and exploring multi-day extreme events in climate model
+output. This package takes as input a directory of projection files from
+many climate models, identify all heat waves for multiple cities using a
+user-specified heat wave definition, and write an output directory of
+files with these heat waves and their characteristics (e.g., length,
+absolute and relative temperatures). The \pkg{futureheatwaves} package
+then allows the user to create custom functions and apply them across
+all heat wave files; this functionality can be used either to explore
+heat wave patterns (e.g., average length and intensity across different
+climate models) or to apply epidemiological estimates to project health
+impacts under different scenarios.
+}
+
+\section{Introduction}\label{introduction}
 
 Research on climate change impacts can require extensive processing.
-This is not only because output files for a single climate model can be large, but also because of the rising popularity of ensemble techniques \citep{IPCCch1}, in which, to better characterize uncertainty in projections, impacts are assessed for multiple climate models, multiple simulations of each climate model, and multiple climate experiments.
-Projections of regional climate change over the next century are subject to high uncertainty due to three distinct sources: (1) internal climate variability, i.e.
-climate noise, (2) climate model uncertainty, i.e.
-the same forcing can produce a different response in different models and (3) scenario uncertainty, i.e.
-uncertainty in future climate forcings (e.g.
-\citet{hawkins2009potential}).
-One approach for ensuring that these sources of uncertainty are characterized is to simulate the future climate many times with multiple models and for multiple future scenarios.
+This is not only because output files for a single climate model can be
+large, but also because of the rising popularity of ensemble techniques
+\citep{IPCCch1}, in which, to better characterize uncertainty in
+projections, impacts are assessed for multiple climate models, multiple
+simulations of each climate model, and multiple climate experiments.
+Projections of regional climate change over the next century are subject
+to high uncertainty due to three distinct sources: (1) internal climate
+variability, i.e.~climate noise, (2) climate model uncertainty, i.e.~the
+same forcing can produce a different response in different models and
+(3) scenario uncertainty, i.e.~uncertainty in future climate forcings
+(e.g. \citet{hawkins2009potential}). One approach for ensuring that
+these sources of uncertainty are characterized is to simulate the future
+climate many times with multiple models and for multiple future
+scenarios.
 
-For example, the Coupled Model Intercomparison Project, phase 5 (CMIP5; \citet{taylor2012overview}) brought together dozens of major climate modeling groups around the world to simulate the same future radiative forcing scenarios, but with their own models.
-This created an ensemble of state-of-the-art climate model projections that allow researchers to study projections and their uncertainties.
-Most of these modeling groups additionally performed more than one simulation for each scenario and model (i.e.
-multiple ensemble members), perturbing the initial conditions by a very tiny amount to quantify uncertainties due to internal climate variability.
+For example, the Coupled Model Intercomparison Project, phase 5 (CMIP5;
+\citet{taylor2012overview}) brought together dozens of major climate
+modeling groups around the world to simulate the same future radiative
+forcing scenarios, but with their own models. This created an ensemble
+of state-of-the-art climate model projections that allow researchers to
+study projections and their uncertainties. Most of these modeling groups
+additionally performed more than one simulation for each scenario and
+model (i.e.~multiple ensemble members), perturbing the initial
+conditions by a very tiny amount to quantify uncertainties due to
+internal climate variability.
 
-This processing is particularly intensive for smaller time steps, like daily climate model output.
-While some climate impacts can be assessed using climate model output at monthly, seasonal, or yearly time steps, the impacts of multi-day extreme events must be assessed using output in daily time steps.
-Such multi-day extreme events include heat waves, cold spells, frost day spells, and droughts.
-The assessment is further complicated if extreme events are identified based on conditions that are rare for a certain location (e.g., 98th percentile of local temperature distribution for identifying heat waves) \citep{IPCCch1}, as this event definition must be determined at each study location from climate model output.
-Further, it is often of interest to create summaries of multiple characteristics of these extreme events.
-One interest, for example, in interpreting climate change scenarios output is whether frequency of heat waves or warm spells will change, as well as how certain characteristics of these extreme events (e.g., length, intensity) will change under different scenarios \citep{IPCCch1}.
+This processing is particularly intensive for smaller time steps, like
+daily climate model output. While some climate impacts can be assessed
+using climate model output at monthly, seasonal, or yearly time steps,
+the impacts of multi-day extreme events must be assessed using output in
+daily time steps. Such multi-day extreme events include heat waves, cold
+spells, frost day spells, and droughts. The assessment is further
+complicated if extreme events are identified based on conditions that
+are rare for a certain location (e.g., 98th percentile of local
+temperature distribution for identifying heat waves) \citep{IPCCch1}, as
+this event definition must be determined at each study location from
+climate model output. Further, it is often of interest to create
+summaries of multiple characteristics of these extreme events. One
+interest, for example, in interpreting climate change scenarios output
+is whether frequency of heat waves or warm spells will change, as well
+as how certain characteristics of these extreme events (e.g., length,
+intensity) will change under different scenarios \citep{IPCCch1}.
 
-We begin this article with an overview of climate model output data, particularly daily data, for R users.
-We outline where data from CMIP5 can be obtained as well as how to work with the file format (netCDF) from R.
-We overview some R packages that can be useful when working with this data, as well as aspects of the data (e.g., non-standard calendars) of which users should be aware when working with daily climate model output in R.
+We begin this article with an overview of climate model output data,
+particularly daily data, for R users. We outline where data from CMIP5
+can be obtained as well as how to work with the file format (netCDF)
+from R. We overview some R packages that can be useful when working with
+this data, as well as aspects of the data (e.g., non-standard calendars)
+of which users should be aware when working with daily climate model
+output in R.
 
-After this overview, we present the \pkg{futureheatwaves} package.
-which we created to aid in identifying and characterizing any type of multi-day extreme event from daily climate model output (Table \ref{tab:goals}).
-Further, this package provides some functionality particularly useful in identifying and characterizing heat waves specifically.
-Quantifying the impacts of heat waves on human health suffer from additional sources of uncertainty beyond those inherent in projections of regional changes in surface temperature.
-These include: (1) uncertainty in the ability of communities to adapt to changing temperatures (the adaptation scenario) and (2) uncertainty in the definition of a heat wave itself.
-Thus, identifying and characterizing the impacts of future heat waves in state-of-the-art climate models requires analyzing hundreds of projections based on combinations of anthropogenic activity, climate model and ensemble member, adaptation scenario, and heat wave definition.
-Such an analysis is non-trivial.
-For example, to estimate the possible impacts of heat waves using 32 ensemble members of CMIP5, two forcing scenarios, two different heat wave definitions, and five progressive assumptions about adaptation to heat, would require that one identify and characterize heat waves in 640 separate temperature time series for each region of interest.
+After this overview, we present the \pkg{futureheatwaves} package. which
+we created to aid in identifying and characterizing any type of
+multi-day extreme event from daily climate model output (Table
+\ref{tab:goals}). Further, this package provides some functionality
+particularly useful in identifying and characterizing heat waves
+specifically. Quantifying the impacts of heat waves on human health
+suffer from additional sources of uncertainty beyond those inherent in
+projections of regional changes in surface temperature. These include:
+(1) uncertainty in the ability of communities to adapt to changing
+temperatures (the adaptation scenario) and (2) uncertainty in the
+definition of a heat wave itself. Thus, identifying and characterizing
+the impacts of future heat waves in state-of-the-art climate models
+requires analyzing hundreds of projections based on combinations of
+anthropogenic activity, climate model and ensemble member, adaptation
+scenario, and heat wave definition. Such an analysis is non-trivial. For
+example, to estimate the possible impacts of heat waves using 32
+ensemble members of CMIP5, two forcing scenarios, two different heat
+wave definitions, and five progressive assumptions about adaptation to
+heat, would require that one identify and characterize heat waves in 640
+separate temperature time series for each region of interest.
 
@@
 
-The \pkg{futureheatwaves} package automates the process of identifying and characterizing multi-day extreme events across different ensemble members of one or more climate models.
-A variety of different heat wave definitions have been used to identify heat waves in a time series of temperature data \citep{smith2013heat}, and the choice of heat wave definitions can influence both health effect estimates \citep{chen2015influence, kent2014heat} and projected heat wave trends \citep{smith2013heat}.
-Further, other types of extreme events will be defined differently than heat waves (for example, frost day spells may be defined as one or more days with temperature at or below $0^oC$).
-This package therefore allows the user to create and use a custom the extreme event definition used to identify events in the climate model output.
-Finally, this package allows the user to explore the extreme events identified in the climate model output with a function that can take a user-defined R function and apply it across all extreme event files generated for the separate ensemble members.
+The \pkg{futureheatwaves} package automates the process of identifying
+and characterizing multi-day extreme events across different ensemble
+members of one or more climate models. A variety of different heat wave
+definitions have been used to identify heat waves in a time series of
+temperature data \citep{smith2013heat}, and the choice of heat wave
+definitions can influence both health effect estimates
+\citep{chen2015influence, kent2014heat} and projected heat wave trends
+\citep{smith2013heat}. Further, other types of extreme events will be
+defined differently than heat waves (for example, frost day spells may
+be defined as one or more days with temperature at or below \(0^oC\)).
+This package therefore allows the user to create and use a custom the
+extreme event definition used to identify events in the climate model
+output. Finally, this package allows the user to explore the extreme
+events identified in the climate model output with a function that can
+take a user-defined R function and apply it across all extreme event
+files generated for the separate ensemble members.
 
-# An overview of climate model output for R users
+\section{An overview of climate model output for R
+users}\label{an-overview-of-climate-model-output-for-r-users}
 
-## CMIP5 climate model output data
+\subsection{CMIP5 climate model output
+data}\label{cmip5-climate-model-output-data}
 
-For climate impact studies, a top source for climate model output files is the Coupled Model Intercomparison Project, which is currently in its fifth phase (CMIP5).
-Over 20 climate modeling groups have created one or more climate models which, for this project, are run using standardized scenarios \citep{taylor2012overview}.
-The resulting output is uniform across modeling groups and has a consistent structure, which allows comparison of simulations from different models \citep{IPCCch9}.
-For CMIP5, each group created simulations under several experiments, with experiments varying in terms of the radiative forcing.
-This radiative forcing depends on time-varying model inputs (greenhouse gas emissions or concentrations, land use changes, etc.), which are specified for each experiment \citep{taylor2012overview, IPCCch9}.
-Experiments include historical experiments (run using forcings consistent with observed and reconstructed data for 1850--2005), pre-Industrial control experiments, and experiments of future scenarios of radiative forcing over the 21st century or longer (e.g., RCP4.5, RCP8.5) \citep{taylor2012overview}.
-Some modeling groups created ensembles of output for a specific model and experiment, in which they ran the experiment multiple times with the model with very small changes to the initial conditions, resulting in multiple ensemble members for a single climate model and experiment.
-CMIP5 climate model output is created at a number of different time steps (e.g., daily, monthly, seasonal, yearly) \citep{taylor2010cmip5}, and some variables are reported at multiple depths in the ocean or atmosphere (e.g., ocean temperature).
-Here, we will focus on data with a daily time step for variables reported at a single depth (e.g., near-surface air temperature).
+For climate impact studies, a top source for climate model output files
+is the Coupled Model Intercomparison Project, which is currently in its
+fifth phase (CMIP5). Over 20 climate modeling groups have created one or
+more climate models which, for this project, are run using standardized
+scenarios \citep{taylor2012overview}. The resulting output is uniform
+across modeling groups and has a consistent structure, which allows
+comparison of simulations from different models \citep{IPCCch9}. For
+CMIP5, each group created simulations under several experiments, with
+experiments varying in terms of the radiative forcing. This radiative
+forcing depends on time-varying model inputs (greenhouse gas emissions
+or concentrations, land use changes, etc.), which are specified for each
+experiment \citep{taylor2012overview, IPCCch9}. Experiments include
+historical experiments (run using forcings consistent with observed and
+reconstructed data for 1850--2005), pre-Industrial control experiments,
+and experiments of future scenarios of radiative forcing over the 21st
+century or longer (e.g., RCP4.5, RCP8.5) \citep{taylor2012overview}.
+Some modeling groups created ensembles of output for a specific model
+and experiment, in which they ran the experiment multiple times with the
+model with very small changes to the initial conditions, resulting in
+multiple ensemble members for a single climate model and experiment.
+CMIP5 climate model output is created at a number of different time
+steps (e.g., daily, monthly, seasonal, yearly) \citep{taylor2010cmip5},
+and some variables are reported at multiple depths in the ocean or
+atmosphere (e.g., ocean temperature). Here, we will focus on data with a
+daily time step for variables reported at a single depth (e.g.,
+near-surface air temperature).
 
-The CMIP5 climate model output data is distributed across data nodes at different climate modeling centers \citep{taylor2012overview}, but can be accessed centrally at the World Climate Research Programme CMIP5 data portal at https://pcmdi.llnl.gov/search/cmip5/.
-Users must register before downloading data, and some data are restricted to non-commercial use.
-There is a separate file for each combination of climate model, experiment, modeling realm (e.g., atmosphere, ocean), variable, time step, and ensemble member \citep{taylor2012overview, taylor2010cmip5}.
-For finer time scales, this output is further split across multiple files for specific year ranges (e.g., 5 years of output for each file) \citep{taylor2010cmip5}.
-Filenames for CMIP5 files can be parsed to generate information about the output variable, climate model, experiment, and ensemble member for the simulation \citep{taylor2010cmip5}.
-Files can be searched and downloaded through a point-and-click web interface.
-The data portal also allows you to generate a specific wget script, which you can use to download many files at once.
-A wget script can also be generated through Earth System Grid Federation's Search RESTful API.
-Tips on efficiently searching and downloading the data, including through use of wget scripts and the search API, are available as user tutorials through the website of the University of Colorado Boulder's Earth System CoG (e.g., https://www.earthsystemcog.org/projects/cog/doc/wget for a tutorial on downloading files using wget).
+The CMIP5 climate model output data is distributed across data nodes at
+different climate modeling centers \citep{taylor2012overview}, but can
+be accessed centrally at the World Climate Research Programme CMIP5 data
+portal at \url{https://pcmdi.llnl.gov/search/cmip5/}. Users must
+register before downloading data, and some data are restricted to
+non-commercial use. There is a separate file for each combination of
+climate model, experiment, modeling realm (e.g., atmosphere, ocean),
+variable, time step, and ensemble member
+\citep{taylor2012overview, taylor2010cmip5}. For finer time scales, this
+output is further split across multiple files for specific year ranges
+(e.g., 5 years of output for each file) \citep{taylor2010cmip5}.
+Filenames for CMIP5 files can be parsed to generate information about
+the output variable, climate model, experiment, and ensemble member for
+the simulation \citep{taylor2010cmip5}. Files can be searched and
+downloaded through a point-and-click web interface. The data portal also
+allows you to generate a specific wget script, which you can use to
+download many files at once. A wget script can also be generated through
+Earth System Grid Federation's Search RESTful API. Tips on efficiently
+searching and downloading the data, including through use of wget
+scripts and the search API, are available as user tutorials through the
+website of the University of Colorado Boulder's Earth System CoG (e.g.,
+\url{https://www.earthsystemcog.org/projects/cog/doc/wget} for a
+tutorial on downloading files using wget).
 
-CMIP5 files are saved in Network Common Data Format (netCDF).
-NetCDF is a binary file format that allows storage of data representing a regular array.
-These climate model output files can be as large as several gigabytes [?].
-For climate model output at a single depth (e.g., near-surface air temperature), the data is a 3-dimensional array, with dimensions representing time and two metrics of location (e.g., latitude and longitude) (Figure \ref{fig:netcdfexample}).
-Global climate models generate output at regularly-spaced time steps, typically at regularly-spaced grid points around the world.
-The latitude and longitude spacing of grid points vary by climate model, but are typically 1--2 degrees for atmospheric variables in CMIP5 models \citep{IPCCch9}.
-Each data point in the netCDF array gives the modeled value of the variable (e.g., surface temperature) for a single time point and location.
-For CMIP5 climate model output, the location units are in degrees east and degrees north for longitude and latitude, respectively.
-For daily output files, the time unit is in days since a specified origin date-time (e.g., days since 1850-01-01 00:00:00) \citep{taylor2010cmip5}.
-All CMIP5 output files are requred to include certain metadata  \citep{taylor2010cmip5}.
-This required metadata includes the experiment, forcing agents input to the model to create the simulation, time step, institution and institutional contact information, climate model, and modeling realm \citep{taylor2010cmip5}.
-The metadata also must include units for all of the coordinate variables (e.g., longitude, latitude, time).
+CMIP5 files are saved in Network Common Data Format (netCDF). NetCDF is
+a binary file format that allows storage of data representing a regular
+array. These climate model output files can be as large as several
+gigabytes {[}?{]}. For climate model output at a single depth (e.g.,
+near-surface air temperature), the data is a 3-dimensional array, with
+dimensions representing time and two metrics of location (e.g., latitude
+and longitude) (Figure \ref{fig:netcdfexample}). Global climate models
+generate output at regularly-spaced time steps, typically at
+regularly-spaced grid points around the world. The latitude and
+longitude spacing of grid points vary by climate model, but are
+typically 1--2 degrees for atmospheric variables in CMIP5 models
+\citep{IPCCch9}. Each data point in the netCDF array gives the modeled
+value of the variable (e.g., surface temperature) for a single time
+point and location. For CMIP5 climate model output, the location units
+are in degrees east and degrees north for longitude and latitude,
+respectively. For daily output files, the time unit is in days since a
+specified origin date-time (e.g., days since 1850-01-01 00:00:00)
+\citep{taylor2010cmip5}. All CMIP5 output files are requred to include
+certain metadata \citep{taylor2010cmip5}. This required metadata
+includes the experiment, forcing agents input to the model to create the
+simulation, time step, institution and institutional contact
+information, climate model, and modeling realm \citep{taylor2010cmip5}.
+The metadata also must include units for all of the coordinate variables
+(e.g., longitude, latitude, time).
 
@@
 
-\citet{taylor2012overview} and \citet{meehl2007wcrp} are excellent resources for finding out more about the CMIP climate model output data.
+\citet{taylor2012overview} and \citet{meehl2007wcrp} are excellent
+resources for finding out more about the CMIP climate model output data.
 
-## Working with climate model output in R
+\subsection{Working with climate model output in
+R}\label{working-with-climate-model-output-in-r}
 
-A few R packages can be used to work with the netCDF file format used for CMIP5 files.
-Earlier packages to work with netCDF files included \pkg{ncdf} and \pkg{ncvar}, but these do not work with the newer netCDF version 4 released in 2008 and are no longer available through CRAN.
-More recent packages, including \pkg{ncdf4} \citep{ncdf4} and \pkg{RNetCDF} \citep{michna2013rnetcdf, RNetCDF}, work with both version 4 and netCDF's older version 3.
-Climate model output data for CMIP5 is required to conform with the earlier version (version 3) \citep{taylor2010cmip5} and so should work with any of these packages, although it is safer to write code using functions that can be used with version for in case future phases of CMIP do not require files to conform with netCDF version 3.
+A few R packages can be used to work with the netCDF file format used
+for CMIP5 files. Earlier packages to work with netCDF files included
+\pkg{ncdf} and \pkg{ncvar}, but these do not work with the newer netCDF
+version 4 released in 2008 and are no longer available through CRAN.
+More recent packages, including \pkg{ncdf4} \citep{ncdf4} and
+\pkg{RNetCDF} \citep{michna2013rnetcdf, RNetCDF}, work with both version
+4 and netCDF's older version 3. Climate model output data for CMIP5 is
+required to conform with the earlier version (version 3)
+\citep{taylor2010cmip5} and so should work with any of these packages,
+although it is safer to write code using functions that can be used with
+version for in case future phases of CMIP do not require files to
+conform with netCDF version 3.
 
-The netCDF format allows you to access metadata and variables describing the dimensions of the data without reading the full file into memory.
-The metadata describes the dimensions of the array and the units, etc., of each variable and is printed by the print method of the R object returned by the \code{nc\_open} function of \pkg{ncdf4}.
-For CMIP5 files, the dimension variables include the times and locations corresponding to each array in the data.
-Before reading in a variable from a netCDF file, a connection to the file must be opened, for example with the \code{nc\_open} function from \pkg{ncdf4}.
-Variables can then be read in using the \code{ncvar\_get} function from \pkg{ncdf4}, with the \code{varid} parameter set to "lat", "lon", or "time" (as a caveat, many climate models output to non-Gregorian calendars, in which case the time variable should be read in using a different function, as discussed later in this section).
-The climate output variable (e.g., near-surface air temperature) can similarly be read in using \code{ncvar\_get}.
-In this case, the \code{varid} parameter should be set using the appropriate CMIP5 variable name (e.g., "tas" for near-surface air temperature); these variable names can be found in the CMIP Requested Output tables \citep{taylor2010cmip5}.
-In practice, you can use the dimensional time and location data to identify the location of the variable data you need in the netCDF array and use indexing to read only that data into memory, without needing to read in the full file \citep{michna2013rnetcdf}.
-Once the user is done reading in data from the file, the connection can be closed (e.g., with the \code{nc\_close} function from \pkg{ncdf4}).
+The netCDF format allows you to access metadata and variables describing
+the dimensions of the data without reading the full file into memory.
+The metadata describes the dimensions of the array and the units, etc.,
+of each variable and is printed by the print method of the R object
+returned by the \code{nc\_open} function of \pkg{ncdf4}. For CMIP5
+files, the dimension variables include the times and locations
+corresponding to each array in the data. Before reading in a variable
+from a netCDF file, a connection to the file must be opened, for example
+with the \code{nc\_open} function from \pkg{ncdf4}. Variables can then
+be read in using the \code{ncvar\_get} function from \pkg{ncdf4}, with
+the \code{varid} parameter set to ``lat'', ``lon'', or ``time'' (as a
+caveat, many climate models output to non-Gregorian calendars, in which
+case the time variable should be read in using a different function, as
+discussed later in this section). The climate output variable (e.g.,
+near-surface air temperature) can similarly be read in using
+\code{ncvar\_get}. In this case, the \code{varid} parameter should be
+set using the appropriate CMIP5 variable name (e.g., ``tas'' for
+near-surface air temperature); these variable names can be found in the
+CMIP Requested Output tables \citep{taylor2010cmip5}. In practice, you
+can use the dimensional time and location data to identify the location
+of the variable data you need in the netCDF array and use indexing to
+read only that data into memory, without needing to read in the full
+file \citep{michna2013rnetcdf}. Once the user is done reading in data
+from the file, the connection can be closed (e.g., with the
+\code{nc\_close} function from \pkg{ncdf4}).
 
-Since the late 1500s, Western dates have been set using the Gregorian calendar, which has 365.2425-day years.
-Some climate models, however, output to different calendars, including the Julian calendar (365.25-day years), a calendar where there are no leap years (365-day years), a calendar where every year is a leap year (366-day years), and a calendar of twelve 30-day months (360-day years) \citep{cfconventions}.
-With these non-Gregorian calendars, R's base functions for converting a vector to a Date class based on the number of days since an origin date (\code{as.Date}, \code{as.POSIXct}) do not return the desired values.
-The \pkg{PCICt} \citep{PCICt} and \pkg{ncdf4.helpers} \citep{ncdf4.helpers} packages provide further functionality with netCDF files that can be useful when working with climate model output data and provide particular help in working with different calendars.
-CMIP5 netCDF files include information on the calendar used, which the \code{nc.get.time.series} function in \pkg{ncdf4.helpers} use to convert the "time" variable in the file to an object of the PCICt class, which provides Date-like functionality for 360- and 365-day calendars \citep{PCICt}.
-As a note, while these functions will help with handling most CMIP5 files, the CMIP5 standards allows use of other calendars which may not be succesfully handled by these functions, so it is important to assess whether the time variable range in the PCICt object correctly matches the expected date ranges for a file as you process CMIP5 data in R.
+Since the late 1500s, Western dates have been set using the Gregorian
+calendar, which has 365.2425-day years. Some climate models, however,
+output to different calendars, including the Julian calendar (365.25-day
+years), a calendar where there are no leap years (365-day years), a
+calendar where every year is a leap year (366-day years), and a calendar
+of twelve 30-day months (360-day years) \citep{cfconventions}. With
+these non-Gregorian calendars, R's base functions for converting a
+vector to a Date class based on the number of days since an origin date
+(\code{as.Date}, \code{as.POSIXct}) do not return the desired values.
+The \pkg{PCICt} \citep{PCICt} and \pkg{ncdf4.helpers}
+\citep{ncdf4.helpers} packages provide further functionality with netCDF
+files that can be useful when working with climate model output data and
+provide particular help in working with different calendars. CMIP5
+netCDF files include information on the calendar used, which the
+\code{nc.get.time.series} function in \pkg{ncdf4.helpers} use to convert
+the ``time'' variable in the file to an object of the PCICt class, which
+provides Date-like functionality for 360- and 365-day calendars
+\citep{PCICt}. As a note, while these functions will help with handling
+most CMIP5 files, the CMIP5 standards allows use of other calendars
+which may not be succesfully handled by these functions, so it is
+important to assess whether the time variable range in the PCICt object
+correctly matches the expected date ranges for a file as you process
+CMIP5 data in R.
 
-The size of the climate model output files can be large enough that it may make more sense to work with smaller chunks of the data, rather than reading all data into memory and working with the data all at once \citep{RCMIP5}.
-This problem aggregates when working with multiple climate models and more than one ensemble member for each of those climate models.
+The size of the climate model output files can be large enough that it
+may make more sense to work with smaller chunks of the data, rather than
+reading all data into memory and working with the data all at once
+\citep{RCMIP5}. This problem aggregates when working with multiple
+climate models and more than one ensemble member for each of those
+climate models.
 
-The following code gives an example of using a CMIP5 file and some of the R packages and functions discussed.
-Results include a map of near-surface air temperatures from a single climate model on a specific date (July 1, 2075) for China (Figure [x]) and a time series of daily near-surface air temperature simulations for the climate model grid point closest to Beijing, China (Figure [x]).
-To try this example, you will need to download the file ...
-from ...
-and save it in the "tmp" subdirectory of your home directory.
+The following code gives an example of using a CMIP5 file and some of
+the R packages and functions discussed. Results include a map of
+near-surface air temperatures from a single climate model on a specific
+date (July 1, 2075) for China (Figure {[}x{]}) and a time series of
+daily near-surface air temperature simulations for the climate model
+grid point closest to Beijing, China (Figure {[}x{]}). To try this
+example, you will need to download the file \ldots{} from \ldots{} and
+save it in the ``tmp'' subdirectory of your home directory.
 
-```{r}
+In addition to these general packages for working with netCDF files,
+there are several R packages specifically for working with climate model
+output data, including \pkg{RCMIP5} \citep{RCMIP5} and \pkg{wux}
+\citep{wux}. However, these packages are more useful for working with
+data output at time steps of a month or higher and have limited utility
+with the daily climate model output data required for studies of
+multi-day extreme events.
 
-```
+The \pkg{RCMIP5} package includes functions to read in CMIP5 data from
+netCDF files, scan a directory of CMIP5 files and determine models with
+continuous available data, create objects of a special ``cmip5data''
+class to work with CMIP5 data within R, and parse the filenames for all
+files in a directory to extract information within the filename. For
+this package, most functions only work with monthly or less frequent
+data (e.g., the functions \code{checkTimePeriod} and \code{cmip5data},
+as well as functions that work with \code{cmip5data} objects including
+\code{filterDimensions} and \code{getProjectionMatrix}) \citep{RCMIP5}.
+While the \code{loadCMIP5} function does successfully load the daily
+data as a \code{cmip5data} object, most of the methods for this object
+type do not do anything meaningful for daily data. The package's
+\code{getFileInfo} function, however, will work with CMIP5 files of any
+time step; this function identifies all CMIP5 files in a directory and
+creates a dataframe that parses the information contained in the file
+name. As a note, the \code{get.split.filename.cmip5} function in the
+\pkg{ncdf4.helper} package similarly can be used to parse information
+contained in CMIP5 file names \citep{ncdf4.helpers}.
 
-In addition to these general packages for working with netCDF files, there are several R packages specifically for working with climate model output data, including \pkg{RCMIP5} \citep{RCMIP5} and \pkg{wux} \citep{wux}.
-However, these packages are more useful for working with data output at time steps of a month or higher and have limited utility with the daily climate model output data required for studies of multi-day extreme events.
+The \pkg{wux} package \citep{wux} includes functions that allow the user
+to download CMIP5 monthly-aggregated output directly from within R with
+the \code{CMIP5fromESGF} function. However, this function does not allow
+downloading of climate model output with finer time steps, like daily
+data. This package uses the \code{models2wux} function to read in
+climate model output netCDF files and convert to a WUX dataframe that
+can be used by other functions in the package. While this function can
+input climate model output with daily time steps, if the element
+``what.timesteps'' of the \code{modelinput} list input is set to
+``daily'', the function aggregates this data to a monthly or less
+frequent (e.g., seasonal) aggregation when creating the WUX dataframe.
+Therefore, while this package provides very useful functionality for
+working with averaged output of daily climate model output data, it
+cannot easily be used to identify and characterize multi-day extreme
+events like heat waves.
 
-The \pkg{RCMIP5} package includes functions to read in CMIP5 data from netCDF files, scan a directory of CMIP5 files and determine models with continuous available data, create objects of a special "cmip5data" class to work with CMIP5 data within R, and parse the filenames for all files in a directory to extract information within the filename.
-For this package, most functions only work with monthly or less frequent data (e.g., the functions \code{checkTimePeriod} and \code{cmip5data}, as well as functions that work with \code{cmip5data} objects including \code{filterDimensions} and \code{getProjectionMatrix}) \citep{RCMIP5}.
-While the \code{loadCMIP5} function does successfully load the daily data as a \code{cmip5data} object, most of the methods for this object type do not do anything meaningful for daily data.
-The package's \code{getFileInfo} function, however, will work with CMIP5 files of any time step; this function identifies all CMIP5 files in a directory and creates a dataframe that parses the information contained in the file name.
-As a note, the \code{get.split.filename.cmip5} function in the \pkg{ncdf4.helper} package similarly can be used to parse information contained in CMIP5 file names \citep{ncdf4.helpers}.
+More tips on working with CMIP5 files in R are provided in the
+``starting\_from\_netcdf'' vignette of the \pkg{futureheatwaves}
+package.
 
-The \pkg{wux} package \citep{wux} includes functions that allow the user to download CMIP5 monthly-aggregated output directly from within R with the \code{CMIP5fromESGF} function.
-However, this function does not allow downloading of climate model output with finer time steps, like daily data.
-This package uses the \code{models2wux} function to read in climate model output netCDF files and convert to a WUX dataframe that can be used by other functions in the package.
-While this function can input climate model output with daily time steps, if the element "what.timesteps" of the \code{modelinput} list input is set to "daily", the function aggregates this data to a monthly or less frequent (e.g., seasonal) aggregation when creating the WUX dataframe.
-Therefore, while this package provides very useful functionality for working with averaged output of daily climate model output data, it cannot easily be used to identify and characterize multi-day extreme events like heat waves.
+\section{\texorpdfstring{The \pkg{futureheatwaves}
+package}{The  package}}\label{the-package}
 
-More tips on working with CMIP5 files in R are provided in the "starting_from_netcdf" vignette of the \pkg{futureheatwaves} package.
-
-# The \pkg{futureheatwaves} package
-
-## How the package works
+\subsection{How the package works}\label{how-the-package-works}
 
@@
 
-The \pkg{futureheatwaves} package was facilitated to aid in identifying, characterizing, and exploring multi-day extreme events in daily climate model output data.
-Figure \ref{fig:overview} gives an overview of the two primary functions of the \pkg{futureheatwaves} package.
-First, the \code{gen\_hw\_set} function processes a directory of climate projection files that are stored locally on the user's computer (Figure \ref{fig:overview}, "Climate projections"), to generate a list of all extreme events in each projection, as well as over a dozen characteristics of each identified extreme event (Table \ref{tab:hwcharacteristics}).
-The extreme events are identified and characterized at one or more study locations (e.g., cities), with study locations specified by the user in an input file.
-The extreme events identified for each ensemble member are output as separate files in a directory specified by the user (Figure \ref{fig:overview}, ``Extreme events datasets").
-Second, the \code{apply\_all\_models} function allows the user to apply custom functions across all the extreme event data frames generated by \code{gen\_hw\_set} to summaries of extreme events across all climate models and ensemble members (Figure \ref{fig:overview}, right).
+The \pkg{futureheatwaves} package was facilitated to aid in identifying,
+characterizing, and exploring multi-day extreme events in daily climate
+model output data. Figure \ref{fig:overview} gives an overview of the
+two primary functions of the \pkg{futureheatwaves} package. First, the
+\code{gen\_hw\_set} function processes a directory of climate projection
+files that are stored locally on the user's computer (Figure
+\ref{fig:overview}, ``Climate projections''), to generate a list of all
+extreme events in each projection, as well as over a dozen
+characteristics of each identified extreme event (Table
+\ref{tab:hwcharacteristics}). The extreme events are identified and
+characterized at one or more study locations (e.g., cities), with study
+locations specified by the user in an input file. The extreme events
+identified for each ensemble member are output as separate files in a
+directory specified by the user (Figure \ref{fig:overview}, ``Extreme
+events datasets``). Second, the \code{apply\_all\_models} function
+allows the user to apply custom functions across all the extreme event
+data frames generated by \code{gen\_hw\_set} to summaries of extreme
+events across all climate models and ensemble members (Figure
+\ref{fig:overview}, right).
 
@@
 
-The user can extensively customize the definition used to identify extreme events.
-The default is a definition for heat waves (two or more days with temperatures at or above the city's $98^{th}$ percentile of year-round temperature), but users can write a custom R function with either a different heat wave definition or with a definition appropriate for a different type of extreme event (e.g., one or more days at or below $0^oC$ for frost day spells).
-Some of these heat wave characteristics are based on relative temperature, which is a measure of how the temperature during a heat wave compares to the typical distribution of temperature at that location \citep{IPCCch1}.
-These relative temperature metrics will vary depending on whether you calculate them based on a location's present-day temperature distribution or on the location's temperature distribution in the future, since typical temperature distributions in many locations are expected to change with climate change.
-The package therefore allows the user to specify date ranges of the temperature distributions to be used in calculating these relative temperature metrics in each location.
+The user can extensively customize the definition used to identify
+extreme events. The default is a definition for heat waves (two or more
+days with temperatures at or above the city's \(98^{th}\) percentile of
+year-round temperature), but users can write a custom R function with
+either a different heat wave definition or with a definition appropriate
+for a different type of extreme event (e.g., one or more days at or
+below \(0^oC\) for frost day spells). Some of these heat wave
+characteristics are based on relative temperature, which is a measure of
+how the temperature during a heat wave compares to the typical
+distribution of temperature at that location \citep{IPCCch1}. These
+relative temperature metrics will vary depending on whether you
+calculate them based on a location's present-day temperature
+distribution or on the location's temperature distribution in the
+future, since typical temperature distributions in many locations are
+expected to change with climate change. The package therefore allows the
+user to specify date ranges of the temperature distributions to be used
+in calculating these relative temperature metrics in each location.
 
-Second, once the user creates these data frames of location-specific heat waves, the `apply_all_models` function allows the user to apply a custom function across all the heat wave data frames that were generated to create a summary data frame (Figure \ref{fig:overview}, right).
-This functionality can be used to generate summary statistics (e.g., determine average heat wave length or total heat wave days), or can be used to apply more complex functions (e.g., apply epidemiologic effect estimates across the heat waves to generate health impact estimates).
+Second, once the user creates these data frames of location-specific
+heat waves, the \texttt{apply\_all\_models} function allows the user to
+apply a custom function across all the heat wave data frames that were
+generated to create a summary data frame (Figure \ref{fig:overview},
+right). This functionality can be used to generate summary statistics
+(e.g., determine average heat wave length or total heat wave days), or
+can be used to apply more complex functions (e.g., apply epidemiologic
+effect estimates across the heat waves to generate health impact
+estimates).
 
-## Example data
+\subsection{Example data}\label{example-data}
 
-We have included data files in the package to serve as example files, so users can try this package before applying it to their own directory of climate projection files.
-The example files are all included as comma-separated (.csv) files, rather than as saved R objects, because the `gen_hw_set` function requires a directory of comma-separated files as input and csv is a general file format familiar to most users.
-This example data therefore also allow users to see how to create the appropriate directory set-up for this package.
+We have included data files in the package to serve as example files, so
+users can try this package before applying it to their own directory of
+climate projection files. The example files are all included as
+comma-separated (.csv) files, rather than as saved R objects, because
+the \texttt{gen\_hw\_set} function requires a directory of
+comma-separated files as input and csv is a general file format familiar
+to most users. This example data therefore also allow users to see how
+to create the appropriate directory set-up for this package.
 
-This example data comes from two climate models that are a part of CMIP5: (1) the model of the Beijing Climate Center, China Meteorological Administration (BCC) \citep{xin2013introduction} and (2) the National Center for Atmospheric Research's (NCAR's) Community Climate System Model, version 4 (CCSM4) \citep{gent2011community}.
-We include one ensemble member from BCC (r1i1p1) and two from CCSM (r1i1p1 and r2i1p1).
-To ensure that the size of this example data is reasonably small, we have only included projection data for grid points from these climate models that are near five U.S.
-east coast cities: New York, NY; Philadelphia, PA; Newark, NJ; Baltimore, MD, and Providence, RI.
-Further, to keep the file sizes reasonably small, the historical projections range over the years 1990 to 1999, while the future projections are limited to 2060 to 2079.
-Users' applications of this package will likely use directories with many more climate model ensemble members and more locations; however, the operation of the package is the same for this smaller example application as it would be for a much larger application.
+This example data comes from two climate models that are a part of
+CMIP5: (1) the model of the Beijing Climate Center, China Meteorological
+Administration (BCC) \citep{xin2013introduction} and (2) the National
+Center for Atmospheric Research's (NCAR's) Community Climate System
+Model, version 4 (CCSM4) \citep{gent2011community}. We include one
+ensemble member from BCC (r1i1p1) and two from CCSM (r1i1p1 and r2i1p1).
+To ensure that the size of this example data is reasonably small, we
+have only included projection data for grid points from these climate
+models that are near five U.S. east coast cities: New York, NY;
+Philadelphia, PA; Newark, NJ; Baltimore, MD, and Providence, RI.
+Further, to keep the file sizes reasonably small, the historical
+projections range over the years 1990 to 1999, while the future
+projections are limited to 2060 to 2079. Users' applications of this
+package will likely use directories with many more climate model
+ensemble members and more locations; however, the operation of the
+package is the same for this smaller example application as it would be
+for a much larger application.
 
-Once the \pkg{futureheatwaves} package is installed and loaded, the user can find the local location of these files using R's `system.file` function.
-In the later sections of this article, we show how to use the package functions with these example files as inputs.
+Once the \pkg{futureheatwaves} package is installed and loaded, the user
+can find the local location of these files using R's
+\texttt{system.file} function. In the later sections of this article, we
+show how to use the package functions with these example files as
+inputs.
 
-When setting up to run climate model output data beyond this example data, any downloaded CMIP5 files will require some pre-processing, in terms of getting the getting the data into a specific format and storing these files in a specific directory structure.
-A file of study locations must also be saved locally in a specific format.
-Extensive details on the required format of files and required input directory structure is available in the \pkg{futureheatwaves} vignette "futureheatwaves", while tips on processing CMIP5 files into the required structure are given in the package's "starting_from_netcdf" vignette.
+When setting up to run climate model output data beyond this example
+data, any downloaded CMIP5 files will require some pre-processing, in
+terms of getting the getting the data into a specific format and storing
+these files in a specific directory structure. A file of study locations
+must also be saved locally in a specific format. Extensive details on
+the required format of files and required input directory structure is
+available in the \pkg{futureheatwaves} vignette ``futureheatwaves'',
+while tips on processing CMIP5 files into the required structure are
+given in the package's ``starting\_from\_netcdf'' vignette.
 
-## Basic example of processing projection files
+\subsection{Basic example of processing projection
+files}\label{basic-example-of-processing-projection-files}
 
-Once these files and directories are set up, \pkg{futureheatwaves} can process them to identify and characterize heat waves in each ensemble member's projection for each  location using the `gen_hw_set` function.
-For example, to process the example climate projections directory included with the package, the user can run:
+Once these files and directories are set up, \pkg{futureheatwaves} can
+process them to identify and characterize heat waves in each ensemble
+member's projection for each location using the \texttt{gen\_hw\_set}
+function. For example, to process the example climate projections
+directory included with the package, the user can run:
 
-```{r eval = FALSE}
+\begin{Schunk}
+\begin{Sinput}
 projection_dir_location <- system.file("extdata/cmip5",
@@
            timeFilenames = "time_NorthAmerica_12mo.csv")
-```
+\end{Sinput}
+\end{Schunk}
 
-This code first identifies and saves as objects the path names on the user's computer of the example climate projections directory (`projection_dir_location`) and the file of study locations (`city_file_location`).
-The `gen_hw_set` function processes this example input and creates a new directory, `example_results`, with files of identified and characterized heat waves, in the user's current working directory.
-In this example code, this processing is done using  default values for the heat wave definition, years for which to generate the heat wave data sets, etc.
+This code first identifies and saves as objects the path names on the
+user's computer of the example climate projections directory
+(\texttt{projection\_dir\_location}) and the file of study locations
+(\texttt{city\_file\_location}). The \texttt{gen\_hw\_set} function
+processes this example input and creates a new directory,
+\texttt{example\_results}, with files of identified and characterized
+heat waves, in the user's current working directory. In this example
+code, this processing is done using default values for the heat wave
+definition, years for which to generate the heat wave data sets, etc.
 Ways to customize these choices are fully explained later in the text.
 
-In this call, the user must specify the directory where the results should be written (`out`), the location of the directory of climate projections (`dataFolder`), the names of the two main subdirectories of that climate projection directory, as well as their year boundaries (`dataDirectories`; see, for example, Figure \ref{fig:directories}), the location of the city location file (`citycsv`), and the names used for the grid coordinate, climate projection, and projection date files (`coordinateFilenames`, `tasFilenames`, and `timeFilenames`).
-When `gen_hw_set` is run, the user is advised that the function will write files to his computer and must agree to proceed:
+In this call, the user must specify the directory where the results
+should be written (\texttt{out}), the location of the directory of
+climate projections (\texttt{dataFolder}), the names of the two main
+subdirectories of that climate projection directory, as well as their
+year boundaries (\texttt{dataDirectories}; see, for example, Figure
+\ref{fig:directories}), the location of the city location file
+(\texttt{citycsv}), and the names used for the grid coordinate, climate
+projection, and projection date files (\texttt{coordinateFilenames},
+\texttt{tasFilenames}, and \texttt{timeFilenames}). When
+\texttt{gen\_hw\_set} is run, the user is advised that the function will
+write files to his computer and must agree to proceed:
 
-```
+\begin{verbatim}
  Warning: This function will write new files to your computer in the 
@@
  Do you want to continue? (y / n): 
-```
+\end{verbatim}
 
-\noindent The function provides status reports while processing.
-This helps the user see that the function call is progressing, as `gen_hw_set` can take a while to run if it is processing many locations and / or many climate ensemble members.
-Once the function has completed running, results will be written locally to the directory specified by the `out` argument of `gen_hw_set`.
-The structure of this output directory is shown in Figure \ref{fig:directories} (right).
-This directory will include files with some basic information about the climate models and the closest grid points of each climate model to each location.
-The output directory will also include a directory with files of identified and classified heat waves for each ensemble member, including all characteristics in Table \ref{tab:hwcharacteristics}.
+\noindent The function provides status reports while processing. This
+helps the user see that the function call is progressing, as
+\texttt{gen\_hw\_set} can take a while to run if it is processing many
+locations and / or many climate ensemble members. Once the function has
+completed running, results will be written locally to the directory
+specified by the \texttt{out} argument of \texttt{gen\_hw\_set}. The
+structure of this output directory is shown in Figure
+\ref{fig:directories} (right). This directory will include files with
+some basic information about the climate models and the closest grid
+points of each climate model to each location. The output directory will
+also include a directory with files of identified and classified heat
+waves for each ensemble member, including all characteristics in Table
+\ref{tab:hwcharacteristics}.
 
-## Customizing the extreme event definition
+\subsection{Customizing the extreme event
+definition}\label{customizing-the-extreme-event-definition}
 
@@
 
-> A \dfn{heat wave} is two or more days at or above a city-specific threshold temperature, with the threshold determined as the $98^{th}$ percentile of year-round temperature in the city during some reference period (by default, 1990--1999).
+\begin{quote}
+A \dfn{heat wave} is two or more days at or above a city-specific
+threshold temperature, with the threshold determined as the \(98^{th}\)
+percentile of year-round temperature in the city during some reference
+period (by default, 1990--1999).
+\end{quote}
 
-Many different definitions of heat waves exist (e.g., \citet{smith2013heat, kent2014heat}), so researchers will often want to use alternative ways to define heat waves.
-Researchers might want to use a specific definition, for example, because it matches the definition used by local health officials to declare heat wave warnings or, in the case of health impact assessments, to match with a definition used in an epidemiological study.
-Three components of the heat wave definition can be easily customized in the `gen_hw_set` function call, without creating a new R function to use to identify heat waves.
-The customization of the heat wave definition is even more extensive as one has the option of writing a custom R function.
+Many different definitions of heat waves exist (e.g.,
+\citet{smith2013heat, kent2014heat}), so researchers will often want to
+use alternative ways to define heat waves. Researchers might want to use
+a specific definition, for example, because it matches the definition
+used by local health officials to declare heat wave warnings or, in the
+case of health impact assessments, to match with a definition used in an
+epidemiological study. Three components of the heat wave definition can
+be easily customized in the \texttt{gen\_hw\_set} function call, without
+creating a new R function to use to identify heat waves. The
+customization of the heat wave definition is even more extensive as one
+has the option of writing a custom R function.
 
-First, the percentile used to identify a heat wave can be changed using the `probThreshold` option in `gen_hw_set`.
-This option can take values between 0 and 1.
-The default value is 0.98, or a definition with a threshold of the $98^{th}$ percentile of the location's temperature.
-For example, to identify heat waves as two or more days at or above the city's $99^{th}$ percentile of year-round temperature, the user could run:
+First, the percentile used to identify a heat wave can be changed using
+the \texttt{probThreshold} option in \texttt{gen\_hw\_set}. This option
+can take values between 0 and 1. The default value is 0.98, or a
+definition with a threshold of the \(98^{th}\) percentile of the
+location's temperature. For example, to identify heat waves as two or
+more days at or above the city's \(99^{th}\) percentile of year-round
+temperature, the user could run:
 
-```{r eval = FALSE}
+\begin{Schunk}
+\begin{Sinput}
 gen_hw_set(out = "example_results",
@@
            probThreshold = 0.99)
-```
+\end{Sinput}
+\end{Schunk}
 
-\noindent Second, the user can change the number of days used in the heat wave definition using the `numDays` argument in the `gen_hw_set` function.
-Combined, these two customization choices can allow the user to identify heat waves using many of the heat wave definitions used in previous climate and health research-- for example, 9 of the 16 heat wave definitions used in \citet{kent2014heat} could be fit using different combinations of these two options for specifying threshold percentile and number of days.
+\noindent Second, the user can change the number of days used in the
+heat wave definition using the \texttt{numDays} argument in the
+\texttt{gen\_hw\_set} function. Combined, these two customization
+choices can allow the user to identify heat waves using many of the heat
+wave definitions used in previous climate and health research-- for
+example, 9 of the 16 heat wave definitions used in \citet{kent2014heat}
+could be fit using different combinations of these two options for
+specifying threshold percentile and number of days.
 
-Third, it is possible to specify the range of years that should be used when determining this threshold.
-For example, if you wanted to base the threshold for each location on its current climate, you could leave this option as its default, while if you wanted to use a different set of years (i.e.
-a threshold relative to future projected temperatures), you could set the start and end year bounds for this reference period using the `thresholdBoundaries` argument in the function `gen_hw_set`.
-For example, to use temperatures in each city from 2070 to 2079 to determine threshold temperatures, you would run:
+Third, it is possible to specify the range of years that should be used
+when determining this threshold. For example, if you wanted to base the
+threshold for each location on its current climate, you could leave this
+option as its default, while if you wanted to use a different set of
+years (i.e.~a threshold relative to future projected temperatures), you
+could set the start and end year bounds for this reference period using
+the \texttt{thresholdBoundaries} argument in the function
+\texttt{gen\_hw\_set}. For example, to use temperatures in each city
+from 2070 to 2079 to determine threshold temperatures, you would run:
 
-```{r eval = FALSE}
+\begin{Schunk}
+\begin{Sinput}
 gen_hw_set(out = "example_results",
@@
            thresholdBoundaries = c(2070, 2079))
-```
+\end{Sinput}
+\end{Schunk}
 
 It is also possible to use a completely customized heat wave definition.
-This functionality allows the user to use heat wave definitions that either require a number of days above an absolute threshold (e.g., maximum temperature of $\ge95^oF$ for $\ge1$ day; \citet{kent2014heat, tan2007heat}) or that require a combination of thresholds to be met (e.g., maximum daily temperature above a lower threshold every day of the heat wave and above a higher threshold for a certain number of days; \citet{kent2014heat, peng2011toward}).
+This functionality allows the user to use heat wave definitions that
+either require a number of days above an absolute threshold (e.g.,
+maximum temperature of \(\ge95^oF\) for \(\ge1\) day;
+\citet{kent2014heat, tan2007heat}) or that require a combination of
+thresholds to be met (e.g., maximum daily temperature above a lower
+threshold every day of the heat wave and above a higher threshold for a
+certain number of days; \citet{kent2014heat, peng2011toward}).
 
-To use a customized heat wave definition, the user can write and load an R function that implements the definition.
-To work correctly, this custom function must allow only specific inputs and generate only specific outputs.
-Details about this required structure are provided in the \pkg{futureheatwaves} package vignette.
+To use a customized heat wave definition, the user can write and load an
+R function that implements the definition. To work correctly, this
+custom function must allow only specific inputs and generate only
+specific outputs. Details about this required structure are provided in
+the \pkg{futureheatwaves} package vignette.
 
-To help in developing custom heat wave identification functions, the package includes an example of the required input data frame, `datafr`, which can be loaded with `data(datafr)`.
-If a custom heat wave function is properly set up to use in this package, it can take as input this example data frame and a threshold value and will return the original data frame with the columns for whether the day was part of a heat wave (`hw`) and, if so, the sequential number of the heat wave out of all heat waves in that location (`hw.number`).
+To help in developing custom heat wave identification functions, the
+package includes an example of the required input data frame,
+\texttt{datafr}, which can be loaded with \texttt{data(datafr)}. If a
+custom heat wave function is properly set up to use in this package, it
+can take as input this example data frame and a threshold value and will
+return the original data frame with the columns for whether the day was
+part of a heat wave (\texttt{hw}) and, if so, the sequential number of
+the heat wave out of all heat waves in that location
+(\texttt{hw.number}).
 
-The custom function can then be referenced using the `IDheatwavesFunction` option in `gen_hw_set`, and it will be used to identify heat waves in the data set.
-As an example, we have included an alternative ID definition function in this package, called `IDHeatwavesAlternative`, which identifies heat waves as two or more days above the higher of either the location's $98^{th}$ percentile temperature or $80^{o}F$, with the idea that this definition prevents the identification of heat waves at mild temperatures in locations with mild or cool climates.
-The following code processes the example climate projecting files using this function as the heat wave definition:
+The custom function can then be referenced using the
+\texttt{IDheatwavesFunction} option in \texttt{gen\_hw\_set}, and it
+will be used to identify heat waves in the data set. As an example, we
+have included an alternative ID definition function in this package,
+called \texttt{IDHeatwavesAlternative}, which identifies heat waves as
+two or more days above the higher of either the location's \(98^{th}\)
+percentile temperature or \(80^{o}F\), with the idea that this
+definition prevents the identification of heat waves at mild
+temperatures in locations with mild or cool climates. The following code
+processes the example climate projecting files using this function as
+the heat wave definition:
 
-```{r, eval = FALSE}
+\begin{Schunk}
+\begin{Sinput}
 gen_hw_set(out = "example_results",
@@
            IDheatwavesFunction = "IDHeatwavesAlternative")
-```
+\end{Sinput}
+\end{Schunk}
 
-With this ability to write and use a custom heat wave identification function, most heat wave definitions can be applied with this package.
-The exception is definitions that require two separate temperature metrics (e.g., maximum temperature above a certain threshold or minimum temperature above a different threshold; \citet{kent2014heat, robinson2001definition}), because the `gen_hw_set` function cannot take as input projections of more than one temperature metric.
-Definitions that use a heat metric that combines multiple inputs (e.g., heat index, which is a combined measure of temperature and air moisture) can be used, but require that the calculation of the metric is done when creating the input files to include in the ``Climate projections" directory (Figure \ref{fig:overview}, left), rather than as part of the heat wave definition function.
+With this ability to write and use a custom heat wave identification
+function, most heat wave definitions can be applied with this package.
+The exception is definitions that require two separate temperature
+metrics (e.g., maximum temperature above a certain threshold or minimum
+temperature above a different threshold;
+\citet{kent2014heat, robinson2001definition}), because the
+\texttt{gen\_hw\_set} function cannot take as input projections of more
+than one temperature metric. Definitions that use a heat metric that
+combines multiple inputs (e.g., heat index, which is a combined measure
+of temperature and air moisture) can be used, but require that the
+calculation of the metric is done when creating the input files to
+include in the ``Climate projections" directory (Figure
+\ref{fig:overview}, left), rather than as part of the heat wave
+definition function.
 
-To increase processing speed when identifying heat waves, we coded parts of the default heat wave identification function in C++ and synced it with R using the \CRANpkg{Rcpp} package \citep{Rcpp}.
-Users should consider a similar strategy for custom heat wave definitions, especially if processing a large number of climate projection files.
+To increase processing speed when identifying heat waves, we coded parts
+of the default heat wave identification function in C++ and synced it
+with R using the \CRANpkg{Rcpp} package \citep{Rcpp}. Users should
+consider a similar strategy for custom heat wave definitions, especially
+if processing a large number of climate projection files.
 
-## Exploring extreme events
+\subsection{Exploring extreme events}\label{exploring-extreme-events}
 
-Once you have created a directory of files with characterized heat waves for each ensemble member, the results can be explored using the `apply_all_models` function.
-This function allows the user to apply custom R functions across all heat wave data frames created by the `gen_hw_sets` call.
-The user can apply any R function that follows certain standards in accepting input and returning output.
-Full details on these standards are given in the \pkg{futureheatwaves} package vignette.
+Once you have created a directory of files with characterized heat waves
+for each ensemble member, the results can be explored using the
+\texttt{apply\_all\_models} function. This function allows the user to
+apply custom R functions across all heat wave data frames created by the
+\texttt{gen\_hw\_sets} call. The user can apply any R function that
+follows certain standards in accepting input and returning output. Full
+details on these standards are given in the \pkg{futureheatwaves}
+package vignette.
 
-As an example, if the user wanted to to get the average temperature of the heat waves identified within each ensemble member, he or she could write a simple function:
+As an example, if the user wanted to to get the average temperature of
+the heat waves identified within each ensemble member, he or she could
+write a simple function:
 
@@
 
-\noindent The `apply_all_models` function can then apply this `average_mean_temp` function across the heat wave data frames for all ensemble members in all climate models:
+\noindent The \texttt{apply\_all\_models} function can then apply this
+\texttt{average\_mean\_temp} function across the heat wave data frames
+for all ensemble members in all climate models:
 
-```{r echo = FALSE}
-average_mean_temp <- function(hw_datafr){
-        out <- mean(hw_datafr$mean.var)
-        return(out)
-        }
-```
-
-```{r}
+\begin{Schunk}
+\begin{Sinput}
 out <- system.file("extdata/example_results", package = "futureheatwaves")
 apply_all_models(out = out, FUN = average_mean_temp)
-```
+\end{Sinput}
+\begin{Soutput}
+#>   model ensemble    value
+#> 1  bcc1        1 302.3745
+#> 2  ccsm        1 302.4458
+#> 3  ccsm        2 302.3428
+\end{Soutput}
+\end{Schunk}
 
-Note that the location of the directory with the heat wave data frames must be specified using the `out` argument when calling \code{apply\_all\_models}.
-Typically, this will be the directory path for the directory specified with the `out` argument in \code{gen\_hw\_set}.
-Location-specific results can be generated using the \code{city\_specific} argument in \code{apply\_all\_models}:
+Note that the location of the directory with the heat wave data frames
+must be specified using the \texttt{out} argument when calling
+\code{apply\_all\_models}. Typically, this will be the directory path
+for the directory specified with the \texttt{out} argument in
+\code{gen\_hw\_set}. Location-specific results can be generated using
+the \code{city\_specific} argument in \code{apply\_all\_models}:
 
-```{r}
+\begin{Schunk}
+\begin{Sinput}
 apply_all_models(out = out, FUN = average_mean_temp, city_specific = TRUE)
-```
+\end{Sinput}
+\begin{Soutput}
+#>    model ensemble city    value
+#> 1   bcc1        1 balt 305.1816
+#> 2   bcc1        1  nwk 300.3367
+#> 3   bcc1        1   ny 300.3367
+#> 4   bcc1        1 phil 305.1816
+#> 5   bcc1        1 prov 298.0402
+#> 6   ccsm        1 balt 303.1277
+#> 7   ccsm        1  nwk 302.4053
+#> 8   ccsm        1   ny 302.4053
+#> 9   ccsm        1 phil 302.3425
+#> 10  ccsm        1 prov 301.8895
+#> 11  ccsm        2 balt 302.9373
+#> 12  ccsm        2  nwk 302.2748
+#> 13  ccsm        2   ny 302.2748
+#> 14  ccsm        2 phil 302.2858
+#> 15  ccsm        2 prov 301.9520
+\end{Soutput}
+\end{Schunk}
 
-This output is structured as ``tidy" data \citep{wickham2014tidy}, allowing it to be used easily with the graphing package \CRANpkg{ggplot2} \citep{ggplot2}.
+This output is structured as ``tidy" data \citep{wickham2014tidy},
+allowing it to be used easily with the graphing package
+\CRANpkg{ggplot2} \citep{ggplot2}.
 
-The `apply_all_models` function can also be used to project the health impacts of heat waves.
-As a very simplistic example, \citet{anderson2009weather} estimated that heat waves, defined as two or more days at or above a community's $98^{th}$ percentile temperature, were associated with an added relative risk of 1.032 for cardiorespiratory mortality risk in 107 U.S.
-communities.
-A simple estimate of excess deaths associated with this added heat wave risk in a community can be calculated as \citep{peng2011toward}:
+The \texttt{apply\_all\_models} function can also be used to project the
+health impacts of heat waves. As a very simplistic example,
+\citet{anderson2009weather} estimated that heat waves, defined as two or
+more days at or above a community's \(98^{th}\) percentile temperature,
+were associated with an added relative risk of 1.032 for
+cardiorespiratory mortality risk in 107 U.S. communities. A simple
+estimate of excess deaths associated with this added heat wave risk in a
+community can be calculated as \citep{peng2011toward}:
 
@@
 
-- $E_c$ is the total number of excess deaths in community $c$;
-- $N_c$ is the baseline average daily mortality in community $c$; 
-- $RR$ is the relative risk of cardiorespiratory mortality per day associated with a heat wave; and 
-- $L_c$ is the total number of heat wave days in community $c$ over the study period.
+\begin{itemize}
+\tightlist
+\item
+  \(E_c\) is the total number of excess deaths in community \(c\);
+\item
+  \(N_c\) is the baseline average daily mortality in community \(c\);
+\item
+  \(RR\) is the relative risk of cardiorespiratory mortality per day
+  associated with a heat wave; and
+\item
+  \(L_c\) is the total number of heat wave days in community \(c\) over
+  the study period.
+\end{itemize}
 
-This impact assessment equation can be translated into a function that merges each projection's heat wave data frame with a data frame of community-specific baseline mortality rates ($B_c$), calculates equation \ref{eq:healthimpacts} for each heat wave, and then sums up the total excess deaths across all heat waves:
+This impact assessment equation can be translated into a function that
+merges each projection's heat wave data frame with a data frame of
+community-specific baseline mortality rates (\(B_c\)), calculates
+equation \ref{eq:healthimpacts} for each heat wave, and then sums up the
+total excess deaths across all heat waves:
 
@@
 
-\noindent Once defined in R, this function can be applied across all heat waves from all climate models' ensemble members, provided that you have a data frame called `city_mortality` with columns with each community's identifier (`city`) and baseline mortality rate (`base_mort`), using the call:
+\noindent Once defined in R, this function can be applied across all
+heat waves from all climate models' ensemble members, provided that you
+have a data frame called \texttt{city\_mortality} with columns with each
+community's identifier (\texttt{city}) and baseline mortality rate
+(\texttt{base\_mort}), using the call:
 
-```{r eval = FALSE}
+\begin{Schunk}
+\begin{Sinput}
 apply_all_models(out = out, FUN = excess_deaths, base_mortality = city_mortality)
-```
+\end{Sinput}
+\end{Schunk}
 
-As examples of the type of functions that can be used with `apply_all_models`, we have included several simple functions for exploring heat waves (Table \ref{tab:applyfunctions}).
-We included an example data frame representative of the heat wave data frames that `gen_hw_set` outputs, to use to test new functions.
-This data can be loaded with `data(hw_datafr)`.
-The process for creating a new function to use to explore heatwaves should be to:
+As examples of the type of functions that can be used with
+\texttt{apply\_all\_models}, we have included several simple functions
+for exploring heat waves (Table \ref{tab:applyfunctions}). We included
+an example data frame representative of the heat wave data frames that
+\texttt{gen\_hw\_set} outputs, to use to test new functions. This data
+can be loaded with \texttt{data(hw\_datafr)}. The process for creating a
+new function to use to explore heatwaves should be to:
 
-1. Load the example data frame with `load(hw_datafr)`;
-2. Build a function that works with `hw_datafr` as an input; and 
-3. Apply the function with `apply_all_models` to process all of the heat wave files.
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Load the example data frame with \texttt{load(hw\_datafr)};
+\item
+  Build a function that works with \texttt{hw\_datafr} as an input; and
+\item
+  Apply the function with \texttt{apply\_all\_models} to process all of
+  the heat wave files.
+\end{enumerate}
 
@@
 
-## Customizing dates for projections and reference temperatures
+\subsection{Customizing dates for projections and reference
+temperatures}\label{customizing-dates-for-projections-and-reference-temperatures}
 
-By default, the `gen_hw_set` function will generate heat wave projections for the years 2070 to 2079, to align with the example data.
-This projection date range can be changed by specifying alternative starting and ending years in the `projectionBoundaries` argument of the `gen_hw_set` function.
-For example, to create projections for 2060 to 2079, the user would call:
+By default, the \texttt{gen\_hw\_set} function will generate heat wave
+projections for the years 2070 to 2079, to align with the example data.
+This projection date range can be changed by specifying alternative
+starting and ending years in the \texttt{projectionBoundaries} argument
+of the \texttt{gen\_hw\_set} function. For example, to create
+projections for 2060 to 2079, the user would call:
 
-```{r, eval = FALSE}
+\begin{Schunk}
+\begin{Sinput}
 gen_hw_set(out = "example_results",
@@
            projectionBoundaries = c(2060, 2079))
-```
+\end{Sinput}
+\end{Schunk}
 
-The heat wave data sets characterize heat wave in several ways that are based on relative temperature (for example, the `mean.temp.quantile`, `max.temp.quantile`, and `min.temp.quantile` columns described in Table \ref{tab:hwcharacteristics}).
-These characteristics are measured for each of the heat waves identified by the `gen_hw_set` function by taking the absolute temperature of the heat wave (e.g., average temperature during the heat wave is $90^oF$) and comparing it to the heat wave location's typical temperature distribution.
-This process generates relative measures of how intense the heat wave is compared to what is normal in that location (e.g., $90^oF$ is in the $99^{th}$ percentile of year-round temperatures in the location).
+The heat wave data sets characterize heat wave in several ways that are
+based on relative temperature (for example, the
+\texttt{mean.temp.quantile}, \texttt{max.temp.quantile}, and
+\texttt{min.temp.quantile} columns described in Table
+\ref{tab:hwcharacteristics}). These characteristics are measured for
+each of the heat waves identified by the \texttt{gen\_hw\_set} function
+by taking the absolute temperature of the heat wave (e.g., average
+temperature during the heat wave is \(90^oF\)) and comparing it to the
+heat wave location's typical temperature distribution. This process
+generates relative measures of how intense the heat wave is compared to
+what is normal in that location (e.g., \(90^oF\) is in the \(99^{th}\)
+percentile of year-round temperatures in the location).
 
-The values of these relative heat wave characteristics will depend on what time period is used to determine the location's typical temperature distribution.
-For example, the same heat wave would likely have much higher values for these relative characteristics if the temperature during the heat wave is compared to typical temperatures in the present day for that location versus if the temperature is compared to typical temperatures much later in the $21^{st}$ century for the same location.
+The values of these relative heat wave characteristics will depend on
+what time period is used to determine the location's typical temperature
+distribution. For example, the same heat wave would likely have much
+higher values for these relative characteristics if the temperature
+during the heat wave is compared to typical temperatures in the present
+day for that location versus if the temperature is compared to typical
+temperatures much later in the \(21^{st}\) century for the same
+location.
 
-The default in the `gen_hw_set` function is to use temperatures for the period 2070 to 2079 for these reference temperatures to calculate all relative heat wave characteristics.
-However, the user can change this specification using the `referenceBoundaries` option of `gen_hw_set`.
-This functionality can be useful in exploring the role of adaptation in future heat waves.
-For example, to use temperature projections for 1990 to 1999 when calculating relative characteristics of heat waves, to explore the assumption that cities remain adapted to their present-day climate, rather than changing in adaptation as climate change increases temperatures, a user could run:
+The default in the \texttt{gen\_hw\_set} function is to use temperatures
+for the period 2070 to 2079 for these reference temperatures to
+calculate all relative heat wave characteristics. However, the user can
+change this specification using the \texttt{referenceBoundaries} option
+of \texttt{gen\_hw\_set}. This functionality can be useful in exploring
+the role of adaptation in future heat waves. For example, to use
+temperature projections for 1990 to 1999 when calculating relative
+characteristics of heat waves, to explore the assumption that cities
+remain adapted to their present-day climate, rather than changing in
+adaptation as climate change increases temperatures, a user could run:
 
-```{r, eval = FALSE}
+\begin{Schunk}
+\begin{Sinput}
 gen_hw_set(out = "example_results",
@@
            referenceBoundaries = c(1990, 1999))
-```
+\end{Sinput}
+\end{Schunk}
 
-For these date range specifications, there are some restrictions on which year ranges can be selected.
-The starting year cannot be earlier than the first year in the first subdirectory, the ending year cannot be later than the last year of the second subdirectory, and the custom date boundaries cannot span the two subdirectories.
+For these date range specifications, there are some restrictions on
+which year ranges can be selected. The starting year cannot be earlier
+than the first year in the first subdirectory, the ending year cannot be
+later than the last year of the second subdirectory, and the custom date
+boundaries cannot span the two subdirectories.
 
-## Mapping grid points
+\subsection{Mapping grid points}\label{mapping-grid-points}
 
-It can be useful to explore the location of the climate model grid point used to pull climate model output for each study location with a given climate model.
-Therefore, the package has a function called `map_grid_leaflet` that plots the locations of grid points used for each location from each climate model.
-This function is built using the htmlWidget \pkg{leaflet} package \citep{leaflet}.
-The following code illustrates the use of this function with the example data to create Figure \ref{fig:gridmap}, which plots the grid points used in the example data from the BCC climate model in the example data:
+It can be useful to explore the location of the climate model grid point
+used to pull climate model output for each study location with a given
+climate model. Therefore, the package has a function called
+\texttt{map\_grid\_leaflet} that plots the locations of grid points used
+for each location from each climate model. This function is built using
+the htmlWidget \pkg{leaflet} package \citep{leaflet}. The following code
+illustrates the use of this function with the example data to create
+Figure \ref{fig:gridmap}, which plots the grid points used in the
+example data from the BCC climate model in the example data:
 
-```{r eval = FALSE}
+\begin{Schunk}
+\begin{Sinput}
 out <- system.file("extdata/example_results", package = "futureheatwaves")
 map_grid_leaflet(plot_model = "bcc1", out = out)
-```
+\end{Sinput}
+\end{Schunk}
 
@@
 The lines on the map connect each climate model grid point to the study location(s) for which that grid point was used.
-The interactive maps include pop-ups with city identiers; one is shown open in this snapshot as an example.
-From this map, you can see that the climate model grid point closest to New York City for this climate model is over the Atlantic Ocean.}
+The interactive maps include pop-ups with city identiers; one is shown open in this snapshot as an example.}
 \label{fig:gridmap}
@@
 
-## Extensions
+\subsection{Extensions}\label{extensions}
 
-The functionality of this package can be easily expanded by loops.
-For example, to explore the role of the heat wave definition on projections, the user could create a loop to run `gen_hw_set` and `apply_all_models` to the same directory of climate projections but with a variety of different functions used to identify the heat waves in the projections.
+The functionality of this package can be easily expanded by loops. For
+example, to explore the role of the heat wave definition on projections,
+the user could create a loop to run \texttt{gen\_hw\_set} and
+\texttt{apply\_all\_models} to the same directory of climate projections
+but with a variety of different functions used to identify the heat
+waves in the projections.
 
-Also, while this package was created to be used for research on heat waves in climate change projections, with some modifications it can be used more broadly.
-For example, there are other episodes like wildfires and air pollution where it may be interesting to identify extended periods of high exposures in projection time series, and this package could be applied to explore these exposures.
-The directory of projection data would need to be set up in the same structure as for exploring heat waves, and the `input_metric` should be set as `fahrenheit`, to pass the exposure values through to the characterized data sets without performing a conversion.
-A user could also use this package to explore events that must be lower than some minimum threshold (e.g., cold waves), but it would take some extra coding and conversions, since the functions in this package are written to identify periods above a threshold (for example, the user could multiple all projected temperatures by -1, and then the coldest temperatures would register as being the highest).
+Also, while this package was created to be used for research on heat
+waves in climate change projections, with some modifications it can be
+used more broadly. For example, there are other episodes like wildfires
+and air pollution where it may be interesting to identify extended
+periods of high exposures in projection time series, and this package
+could be applied to explore these exposures. The directory of projection
+data would need to be set up in the same structure as for exploring heat
+waves, and the \texttt{input\_metric} should be set as
+\texttt{fahrenheit}, to pass the exposure values through to the
+characterized data sets without performing a conversion. A user could
+also use this package to explore events that must be lower than some
+minimum threshold (e.g., cold waves), but it would take some extra
+coding and conversions, since the functions in this package are written
+to identify periods above a threshold (for example, the user could
+multiple all projected temperatures by -1, and then the coldest
+temperatures would register as being the highest).
 
-```{r eval = FALSE}
+\begin{Schunk}
+\begin{Sinput}
 ## Cold wave example
@@
            seasonal_months = c(12, 1, 2))
-```
+\end{Sinput}
+\end{Schunk}
 
->  Extreme statistics and spell-lengths: "A 'frost day' is defined as one during which the minimum temperature falls below freezing point (0 degC).
-This is described as a climatological statistic, in which the minimum temperature is first calculated within each day, and then the number of days or spell lengths meeting the specified condition are evaluated." (from one example in http://cfconventions.org/Data/cf-conventions/cf-conventions-1.6/build/cf-conventions.html#calendar)
+\begin{quote}
+Extreme statistics and spell-lengths: ``A `frost day' is defined as one
+during which the minimum temperature falls below freezing point (0
+degC). This is described as a climatological statistic, in which the
+minimum temperature is first calculated within each day, and then the
+number of days or spell lengths meeting the specified condition are
+evaluated.'' (from one example in
+\url{http://cfconventions.org/Data/cf-conventions/cf-conventions-1.6/build/cf-conventions.html\#calendar})
+\end{quote}
 
-Users doing these kinds of extensions will need to pay attention to a few points.
-First, some of the event characteristics (first in the calendar year, average of May--September temperatures, days above $90^{o}F$) might not be meaningful for studies of other types of events.
-Further, because event periods are usually defined as a string of multiple days exceeding some threshold, the functions in this package may miss the first and last event of the time period.
-For example, if the first day of the time series is the last day of an event, this function would not identify that event because it lacks data from the earlier days that allow this day to meet the event definition.
-This issue would lead to, at most, missing two events out of each projection, but should be considered if studying events that might occur near the start and end of projection data.
-If there is adequate interest from researchers, in the future we may adapt the package to make these secondary applications part of the package.
+Users doing these kinds of extensions will need to pay attention to a
+few points. First, some of the event characteristics (first in the
+calendar year, average of May--September temperatures, days above
+\(90^{o}F\)) might not be meaningful for studies of other types of
+events. Further, because event periods are usually defined as a string
+of multiple days exceeding some threshold, the functions in this package
+may miss the first and last event of the time period. For example, if
+the first day of the time series is the last day of an event, this
+function would not identify that event because it lacks data from the
+earlier days that allow this day to meet the event definition. This
+issue would lead to, at most, missing two events out of each projection,
+but should be considered if studying events that might occur near the
+start and end of projection data. If there is adequate interest from
+researchers, in the future we may adapt the package to make these
+secondary applications part of the package.
 
-# Acknowledgements
+\section{Acknowledgements}\label{acknowledgements}
 
-This work was supported by grants from the National Institute of Environmental Health Sciences (R00ES022631), the National Science Foundation (1331399), and the Colorado State University Vice President for Research.
+This work was supported by grants from the National Institute of
+Environmental Health Sciences (R00ES022631), the National Science
+Foundation (1331399), and the Colorado State University Vice President
+for Research.
 
 \bibliography{Anderson}
+
+\address{%
+G. Brooke Anderson\\
+Colorado State University\\
+Department of Environmental \& Radiological Health Sciences\\ 1681 Campus Delivery\\ Fort Collins, Colorado 80523\\
+}
+\href{mailto:brooke.anderson@colostate.edu}{\nolinkurl{brooke.anderson@colostate.edu}}
+
+\address{%
+Colin Eason\\
+Colorado State University\\
+Department of Computer Science\\ 1873 Campus Delivery\\ Fort Collins, Colorado 80523\\
+}
+\href{mailto:aimesce@gmail.com}{\nolinkurl{aimesce@gmail.com}}
+
+\address{%
+Elizabeth A. Barnes\\
+Colorado State University\\
+Department of Atmospheric Science\\ 1371 Campus Delivery\\ Fort Collins, CO 80523\\
+}
+\href{mailto:eabarnes@atmos.colostate.edu}{\nolinkurl{eabarnes@atmos.colostate.edu}}

### 2016-12-29 paper_drafts/RJournalRevisions/anderson-eason-barnes.Rmd
Update article draft

@@
   Research on climate change impacts can require extensive processing, to incorporate output from ensembles of different climate models and different simulations of each model.
-This processing can be particularly extensive when identifying and characterizing extreme events like heat waves and frost day spells, as these must be processed from model output with daily time steps and may use identification threholds based on local distributions (e.g., 98th percentile temperature for that location).
+This processing can be particularly extensive when identifying and characterizing extreme events like heat waves and frost day spells, as these must be processed from model output with daily time steps and may use identification threholds based on local distributions (e.g., 98\textsuperscript{th} percentile temperature for that location).
 Here, we provide an overview of working with daily climate model output data in R.
 Further, we present the \pkg{futureheatwaves} package, which we developed to ease the process of identifying, characterizing, and exploring multi-day extreme events in climate model output.
-This package takes as input a directory of projection files from many climate models, identify all heat waves for multiple cities using a user-specified heat wave definition, and write an output directory of files with these heat waves and their characteristics (e.g., length, absolute and relative temperatures).
-The \pkg{futureheatwaves} package then allows the user to create custom functions and apply them across all heat wave files; this functionality can be used either to explore heat wave patterns (e.g., average length and intensity across different climate models) or to apply epidemiological estimates to project health impacts under different scenarios.
+This package takes as input a directory of projection files from one or many climate models, identifies all of a specific type of extreme event (e.g., heat waves) for multiple cities using a user-specified event definition, and writes an output directory of files with these extreme events and their characteristics (e.g., length, absolute and relative temperatures).
+The \pkg{futureheatwaves} package then allows the user to create custom functions and apply them across all event files; this functionality can be used either to explore extreme event patterns (e.g., average length and intensity across different climate models) or to apply epidemiological estimates to project health impacts under different scenarios.
 preamble: >
@@
 
-This processing is particularly intensive for smaller time steps, like daily climate model output.
+Processing climate output data is particularly intensive for data in smaller time steps, like daily output.
 While some climate impacts can be assessed using climate model output at monthly, seasonal, or yearly time steps, the impacts of multi-day extreme events must be assessed using output in daily time steps.
@@
 which we created to aid in identifying and characterizing any type of multi-day extreme event from daily climate model output (Table \ref{tab:goals}).
-Further, this package provides some functionality particularly useful in identifying and characterizing heat waves specifically.
+This package can be used to identify and characterize a variety of multi-day extreme events, but provides some functionality particularly useful in identifying and characterizing heat waves specifically.
 Quantifying the impacts of heat waves on human health suffer from additional sources of uncertainty beyond those inherent in projections of regional changes in surface temperature.
@@
 A variety of different heat wave definitions have been used to identify heat waves in a time series of temperature data \citep{smith2013heat}, and the choice of heat wave definitions can influence both health effect estimates \citep{chen2015influence, kent2014heat} and projected heat wave trends \citep{smith2013heat}.
-Further, other types of extreme events will be defined differently than heat waves (for example, frost day spells may be defined as one or more days with temperature at or below $0^oC$).
+Further, other types of extreme events will be defined differently than heat waves (for example, frost day spells may be defined as one or more days with temperature at or below 32\textsuperscript{o}F / 0\textsuperscript{o}C).
 This package therefore allows the user to create and use a custom the extreme event definition used to identify events in the climate model output.
@@
 
-The \pkg{futureheatwaves} package was facilitated to aid in identifying, characterizing, and exploring multi-day extreme events in daily climate model output data.
+We created the \pkg{futureheatwaves} package to aid in identifying, characterizing, and exploring multi-day extreme events in daily climate model output data.
 Figure \ref{fig:overview} gives an overview of the two primary functions of the \pkg{futureheatwaves} package.
@@
 The extreme events identified for each ensemble member are output as separate files in a directory specified by the user (Figure \ref{fig:overview}, ``Extreme events datasets").
-Second, the \code{apply\_all\_models} function allows the user to apply custom functions across all the extreme event data frames generated by \code{gen\_hw\_set} to summaries of extreme events across all climate models and ensemble members (Figure \ref{fig:overview}, right).
+
+Once the user creates these data frames of location-specific heat waves, the \code{apply\_all\_models} function allows the user to apply custom functions across all the extreme event data frames generated by \code{gen\_hw\_set} to summaries of extreme events across all climate models and ensemble members (Figure \ref{fig:overview}, right).
+This functionality can be used to generate summary statistics (e.g., determine average heat wave length or total frost days), or can be used to apply more complex functions (e.g., apply epidemiologic effect estimates across the heat waves to generate health impact estimates).
 
@@
 \code{start.month} & Month in which the event started (1 = January) \\
-\code{days.above.abs.thresh.1} & Number of days in the event above a specified absolute threshold (default is the number of days in the event above $80^oF$ / $26.7^oC$, but this can be changed with the \code{absolute\_thresholds} argument in \code{gen\_hw\_set}) \\
-\code{days.above.abs.thresh.2} & Number of days in the event above a specified absolute threshold (default is the number of days in the event above $85^oF$ / $29.4^oC$) \\ 
-\code{days.above.abs.thresh.3} & Number of days in the event above a specified absolute threshold (default is the number of days in the event above $90^oF$ / $32.3^oC$) \\
-\code{days.above.abs.thresh.4} & Number of days in the event above a specified absolute threshold (default is the number of days in the event above $95^oF$ / $35.0^oC$) \\
-\code{days.above.99th} & Number of days in the event above the $99^{th}$ percentile of the variable for the location, using the period specified with the \code{referenceBoundaries} argument in \code{gen\_hw\_set} as a reference for determining these percentiles \\
-\code{days.above.99.5th} & Number of days in the event above the $99.5^{th}$ percentile of the variable for the location, using the period specified with the \code{referenceBoundaries} argument in \code{gen\_hw\_set} as a reference for determining these percentiles \\
+\code{days.above.abs.thresh.1} & Number of days in the event above a specified absolute threshold (default is the number of days in the event above 80\textsuperscript{o}F / 26.7\textsuperscript{o}C, but this can be changed with the \code{absolute\_thresholds} argument in \code{gen\_hw\_set}) \\
+\code{days.above.abs.thresh.2} & Number of days in the event above a specified absolute threshold (default is the number of days in the event above 85\textsuperscript{o}F / 29.4\textsuperscript{o}C) \\
+\code{days.above.abs.thresh.3} & Number of days in the event above a specified absolute threshold (default is the number of days in the event above 90\textsuperscript{o}F / 32.3\textsuperscript{o}C) \\
+\code{days.above.abs.thresh.4} & Number of days in the event above a specified absolute threshold (default is the number of days in the event above 95\textsuperscript{o}F / 35.0\textsuperscript{o}C) \\
+\code{days.above.99th} & Number of days in the event above the 99\textsuperscript{th} percentile of the variable for the location, using the period specified with the \code{referenceBoundaries} argument in \code{gen\_hw\_set} as a reference for determining these percentiles \\
+\code{days.above.99.5th} & Number of days in the event above the 99.5\textsuperscript{th} percentile of the variable for the location, using the period specified with the \code{referenceBoundaries} argument in \code{gen\_hw\_set} as a reference for determining these percentiles \\
 \code{first.in.year} & Whether the event was the first to occur in its calendar year in the location \\
@@
 
-The user can extensively customize the definition used to identify extreme events.
-The default is a definition for heat waves (two or more days with temperatures at or above the city's $98^{th}$ percentile of year-round temperature), but users can write a custom R function with either a different heat wave definition or with a definition appropriate for a different type of extreme event (e.g., one or more days at or below $0^oC$ for frost day spells).
-Some of these heat wave characteristics are based on relative temperature, which is a measure of how the temperature during a heat wave compares to the typical distribution of temperature at that location \citep{IPCCch1}.
-These relative temperature metrics will vary depending on whether you calculate them based on a location's present-day temperature distribution or on the location's temperature distribution in the future, since typical temperature distributions in many locations are expected to change with climate change.
-The package therefore allows the user to specify date ranges of the temperature distributions to be used in calculating these relative temperature metrics in each location.
-
-Second, once the user creates these data frames of location-specific heat waves, the `apply_all_models` function allows the user to apply a custom function across all the heat wave data frames that were generated to create a summary data frame (Figure \ref{fig:overview}, right).
-This functionality can be used to generate summary statistics (e.g., determine average heat wave length or total heat wave days), or can be used to apply more complex functions (e.g., apply epidemiologic effect estimates across the heat waves to generate health impact estimates).
-
 ## Example data
@@
 We have included data files in the package to serve as example files, so users can try this package before applying it to their own directory of climate projection files.
-The example files are all included as comma-separated (.csv) files, rather than as saved R objects, because the `gen_hw_set` function requires a directory of comma-separated files as input and csv is a general file format familiar to most users.
-This example data therefore also allow users to see how to create the appropriate directory set-up for this package.
-
 This example data comes from two climate models that are a part of CMIP5: (1) the model of the Beijing Climate Center, China Meteorological Administration (BCC) \citep{xin2013introduction} and (2) the National Center for Atmospheric Research's (NCAR's) Community Climate System Model, version 4 (CCSM4) \citep{gent2011community}.
 We include one ensemble member from BCC (r1i1p1) and two from CCSM (r1i1p1 and r2i1p1).
+
 To ensure that the size of this example data is reasonably small, we have only included projection data for grid points from these climate models that are near five U.S.
@@
 
-Once the \pkg{futureheatwaves} package is installed and loaded, the user can find the local location of these files using R's `system.file` function.
+Once the \pkg{futureheatwaves} package is installed and loaded, the user can find the local location of these files using R's \code{system.file} function.
 In the later sections of this article, we show how to use the package functions with these example files as inputs.
 
-When setting up to run climate model output data beyond this example data, any downloaded CMIP5 files will require some pre-processing, in terms of getting the getting the data into a specific format and storing these files in a specific directory structure.
-A file of study locations must also be saved locally in a specific format.
-Extensive details on the required format of files and required input directory structure is available in the \pkg{futureheatwaves} vignette "futureheatwaves", while tips on processing CMIP5 files into the required structure are given in the package's "starting_from_netcdf" vignette.
+When using this package with data other than this example data, the climate model output data will require some pre-processing.
+Data will need to be saved in a specific format, with files stored in a specific directory structure.
+This preprocessing can be done using some of the R packages described earlier for working with CMIP5 data in R.
+Full details of the required file and directory structure are provided in the package's "futureheatwaves" vignette, while tips on conducting this processing starting from CMIP5 netCDF files are given in the "starting_from_netcdf" vignette.
 
-## Basic example of processing projection files
+## Basic example of identifying and characterizing extreme events
 
-Once these files and directories are set up, \pkg{futureheatwaves} can process them to identify and characterize heat waves in each ensemble member's projection for each  location using the `gen_hw_set` function.
-For example, to process the example climate projections directory included with the package, the user can run:
+Once climate model output files are set up as specified in the "futureheatwaves" package vignette, the package can process them to identify and characterize heat waves in each ensemble member's projection for each  location using the \code{gen\_hw\_set} function.
+For example, to process the example climate model output data included with the package, the user can run:
 
@@
 
-This code first identifies and saves as objects the path names on the user's computer of the example climate projections directory (`projection_dir_location`) and the file of study locations (`city_file_location`).
-The `gen_hw_set` function processes this example input and creates a new directory, `example_results`, with files of identified and characterized heat waves, in the user's current working directory.
-In this example code, this processing is done using  default values for the heat wave definition, years for which to generate the heat wave data sets, etc.
-Ways to customize these choices are fully explained later in the text.
+This code first identifies and saves as objects the path names on the user's computer of the example climate projections directory (\code{projection\_dir\_location}) and the file of study locations (\code{city\_file\_location}).
+The \code{gen\_hw\_set} function processes this example input and creates a new directory, \file{example\_results}, with files of identified and characterized heat waves, in the user's current working directory.
+In this example code, the processing is done using  default values for the event definition, years for which to generate the heat wave data sets, etc.
+How and why to customize these choices are explained later in the text.
+Function arguments (e.g., \code{dataDirectories}, \code{tasFilenames}) are used to specify the format of the data and the directory structure.
 
-In this call, the user must specify the directory where the results should be written (`out`), the location of the directory of climate projections (`dataFolder`), the names of the two main subdirectories of that climate projection directory, as well as their year boundaries (`dataDirectories`; see, for example, Figure \ref{fig:directories}), the location of the city location file (`citycsv`), and the names used for the grid coordinate, climate projection, and projection date files (`coordinateFilenames`, `tasFilenames`, and `timeFilenames`).
-When `gen_hw_set` is run, the user is advised that the function will write files to his computer and must agree to proceed:
-
-```
- Warning: This function will write new files to your computer in the 
- ~/tmp/example_results/ directory of your computer. If that directory already exists,
-running this function will write over it. 
- Do you want to continue? (y / n): 
-```
-
-\noindent The function provides status reports while processing.
-This helps the user see that the function call is progressing, as `gen_hw_set` can take a while to run if it is processing many locations and / or many climate ensemble members.
-Once the function has completed running, results will be written locally to the directory specified by the `out` argument of `gen_hw_set`.
-The structure of this output directory is shown in Figure \ref{fig:directories} (right).
-This directory will include files with some basic information about the climate models and the closest grid points of each climate model to each location.
-The output directory will also include a directory with files of identified and classified heat waves for each ensemble member, including all characteristics in Table \ref{tab:hwcharacteristics}.
+Once the function has completed running, results will be written locally to the directory specified by the \code{out} argument of \code{gen\_hw\_set}.
+This directory will include files with some basic information about the climate models and the closest grid points of each climate model to each location, as well as a directory with files of identified and classified extreme events for each ensemble member, including all characteristics in Table \ref{tab:hwcharacteristics}.
 
@@
 
-The default heat wave definition for this package is:
+By default, the package identifies extreme events in climate model output data using a specific definition for heat waves:
 
-> A \dfn{heat wave} is two or more days at or above a city-specific threshold temperature, with the threshold determined as the $98^{th}$ percentile of year-round temperature in the city during some reference period (by default, 1990--1999).
+> A \dfn{heat wave} is two or more days at or above a city-specific threshold temperature, with the threshold determined as the 98\textsuperscript{th} percentile of year-round temperature in the city during some reference period (by default, 1990--1999).
 
-Many different definitions of heat waves exist (e.g., \citet{smith2013heat, kent2014heat}), so researchers will often want to use alternative ways to define heat waves.
-Researchers might want to use a specific definition, for example, because it matches the definition used by local health officials to declare heat wave warnings or, in the case of health impact assessments, to match with a definition used in an epidemiological study.
-Three components of the heat wave definition can be easily customized in the `gen_hw_set` function call, without creating a new R function to use to identify heat waves.
-The customization of the heat wave definition is even more extensive as one has the option of writing a custom R function.
+\noindent However, the user can extensively customize the definition used to identify extreme events.
+Users can write a custom R function with either a different heat wave definition (see \citet{smith2013heat} and \citet{kent2014heat} for listings of definitions used in scientific studies) or with a definition appropriate for a different type of extreme event (e.g., one or more days at or below 32\textsuperscript{o}F / 0\textsuperscript{o}C for frost day spells).
+For heat wave identification, researchers might want to use a specific definition, for example, because it matches the definition used by local health officials to declare heat wave warnings or, in the case of health impact assessments, to match with a definition used in an epidemiological study.
+For other extreme events, a heat wave definition likely will not be applicable and so a customized definition is necessary.
 
-First, the percentile used to identify a heat wave can be changed using the `probThreshold` option in `gen_hw_set`.
-This option can take values between 0 and 1.
-The default value is 0.98, or a definition with a threshold of the $98^{th}$ percentile of the location's temperature.
-For example, to identify heat waves as two or more days at or above the city's $99^{th}$ percentile of year-round temperature, the user could run:
+Three components of the extreme event definition can be easily customized in the \code{gen\_hw\_set} function call, without creating a new R function to use to identify heat waves.
+First, often users wish to define an extreme event based on conditions that are rare in the study location \citep{IPCCch1}.
+In this case, the percentile of the variable of interest required for an extreme event can be changed using the \code{probThreshold} option in \code{gen\_hw\_set}.
+For example, this option could be used to identify heat waves based on the 99\textsuperscript{th} percentile, rather than the 98\textsuperscript{th} percentile, of a city's year-round temperature.
+Second, the user can change the number of days used in the heat wave definition using the \code{numDays} argument in the \code{gen\_hw\_set} function (the default is \samp{numDays = 2}, corresponding to two or more days).
+Combined, these two customization choices allow the user to identify heat waves using many of the heat wave definitions used in previous climate and health research-- for example, 9 of the 16 heat wave definitions used in \citet{kent2014heat} (e.g., [some examples of the 9 that could be fit]) could be fit using different combinations of these two options for specifying threshold percentile and number of days.
+Third, some extreme events like cold waves and frost day spells are defined as a certain number of days below, rather than above, a threshold.
+While the default is to identify events by searching for days above a threshold, this behavior can be changed with the \code{above\_threshold = FALSE} argument in the \code{gen\_hw\_set function}.
 
-```{r eval = FALSE}
-gen_hw_set(out = "example_results",
-           dataFolder = projection_dir_location ,
-           dataDirectories = list("historical" = c(1990, 1999),
-                                        "rcp85" = c(2060, 2079)),
-           citycsv = city_file_location,
-           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
-           tasFilenames = "tas_NorthAmerica_12mo.csv",
-           timeFilenames = "time_NorthAmerica_12mo.csv",
-           probThreshold = 0.99)
-```
-
-\noindent Second, the user can change the number of days used in the heat wave definition using the `numDays` argument in the `gen_hw_set` function.
-Combined, these two customization choices can allow the user to identify heat waves using many of the heat wave definitions used in previous climate and health research-- for example, 9 of the 16 heat wave definitions used in \citet{kent2014heat} could be fit using different combinations of these two options for specifying threshold percentile and number of days.
-
-Third, it is possible to specify the range of years that should be used when determining this threshold.
-For example, if you wanted to base the threshold for each location on its current climate, you could leave this option as its default, while if you wanted to use a different set of years (i.e.
-a threshold relative to future projected temperatures), you could set the start and end year bounds for this reference period using the `thresholdBoundaries` argument in the function `gen_hw_set`.
-For example, to use temperatures in each city from 2070 to 2079 to determine threshold temperatures, you would run:
-
-```{r eval = FALSE}
-gen_hw_set(out = "example_results",
-           dataFolder = projection_dir_location ,
-           dataDirectories = list("historical" = c(1990, 1999),
-                                        "rcp85" = c(2060, 2079)),
-           citycsv = city_file_location,
-           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
-           tasFilenames = "tas_NorthAmerica_12mo.csv",
-           timeFilenames = "time_NorthAmerica_12mo.csv",
-           thresholdBoundaries = c(2070, 2079))
-```
-
-It is also possible to use a completely customized heat wave definition.
-This functionality allows the user to use heat wave definitions that either require a number of days above an absolute threshold (e.g., maximum temperature of $\ge95^oF$ for $\ge1$ day; \citet{kent2014heat, tan2007heat}) or that require a combination of thresholds to be met (e.g., maximum daily temperature above a lower threshold every day of the heat wave and above a higher threshold for a certain number of days; \citet{kent2014heat, peng2011toward}).
-
-To use a customized heat wave definition, the user can write and load an R function that implements the definition.
+Beyond these simpler options, the customization of the event definition is even more extensive as one has the option of writing a custom R function.
+This functionality allows the user to use heat wave definitions that either require a number of days above an absolute threshold (e.g., maximum temperature of $\ge$ 95\textsuperscript{o}F for $\ge1$ day \citet{kent2014heat, tan2007heat}; minimum temperature $\le$ 0\textsuperscript{o}C for $\ge1$ day for frost day spells) or that require a combination of thresholds to be met (e.g., maximum daily temperature one threshold every day of the heat wave and above a higher threshold for a certain number of days; \citet{kent2014heat, peng2011toward}).
+To use a customized event definition, the user can write and load an R function that implements the definition.
+The custom function is passed to the \code{gen\_hw\_set} function using the \code{IDheatwavesFunction} argument.
 To work correctly, this custom function must allow only specific inputs and generate only specific outputs.
 Details about this required structure are provided in the \pkg{futureheatwaves} package vignette.
+To increase processing speed when identifying extreme events, we coded parts of the default event definition function in C++ and synced it with R using the \CRANpkg{Rcpp} package \citep{Rcpp}.
+Users should consider a similar strategy for custom heat wave definitions, especially if processing a large number of climate projection files.
 
-To help in developing custom heat wave identification functions, the package includes an example of the required input data frame, `datafr`, which can be loaded with `data(datafr)`.
-If a custom heat wave function is properly set up to use in this package, it can take as input this example data frame and a threshold value and will return the original data frame with the columns for whether the day was part of a heat wave (`hw`) and, if so, the sequential number of the heat wave out of all heat waves in that location (`hw.number`).
-
-The custom function can then be referenced using the `IDheatwavesFunction` option in `gen_hw_set`, and it will be used to identify heat waves in the data set.
-As an example, we have included an alternative ID definition function in this package, called `IDHeatwavesAlternative`, which identifies heat waves as two or more days above the higher of either the location's $98^{th}$ percentile temperature or $80^{o}F$, with the idea that this definition prevents the identification of heat waves at mild temperatures in locations with mild or cool climates.
-The following code processes the example climate projecting files using this function as the heat wave definition:
-
-```{r, eval = FALSE}
-gen_hw_set(out = "example_results",
-           dataFolder = projection_dir_location ,
-           dataDirectories = list("historical" = c(1990, 1999),
-                                        "rcp85" = c(2060, 2079)),
-           citycsv = city_file_location,
-           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
-           tasFilenames = "tas_NorthAmerica_12mo.csv",
-           timeFilenames = "time_NorthAmerica_12mo.csv",
-           IDheatwavesFunction = "IDHeatwavesAlternative")
-```
-
-With this ability to write and use a custom heat wave identification function, most heat wave definitions can be applied with this package.
-The exception is definitions that require two separate temperature metrics (e.g., maximum temperature above a certain threshold or minimum temperature above a different threshold; \citet{kent2014heat, robinson2001definition}), because the `gen_hw_set` function cannot take as input projections of more than one temperature metric.
-Definitions that use a heat metric that combines multiple inputs (e.g., heat index, which is a combined measure of temperature and air moisture) can be used, but require that the calculation of the metric is done when creating the input files to include in the ``Climate projections" directory (Figure \ref{fig:overview}, left), rather than as part of the heat wave definition function.
-
-To increase processing speed when identifying heat waves, we coded parts of the default heat wave identification function in C++ and synced it with R using the \CRANpkg{Rcpp} package \citep{Rcpp}.
-Users should consider a similar strategy for custom heat wave definitions, especially if processing a large number of climate projection files.
+Finally, extreme events are often based on conditions that are rare for a specific location using location-specific relative thresholds.
+These thresholds are often defined for climate impact studies based on a variable's distribution at that location in present-day or historical data.
+However, it can be interesting to explore trends in extreme events under climate change if extreme events are identified based on variable distributions during the projection period, or another future period.
+It is possible to specify the time period to use when determining a relative threshold for an event definition using the \code{thresholdBoundaries} argument in the function \code{gen\_hw\_set}.
 
@@
 
-Once you have created a directory of files with characterized heat waves for each ensemble member, the results can be explored using the `apply_all_models` function.
-This function allows the user to apply custom R functions across all heat wave data frames created by the `gen_hw_sets` call.
+Once the user has created a directory of characterized event files for each ensemble member ("Extreme event data sets", Figure \ref{fig:overview}), he or she can explore the results using the \code{apply\_all\_models} function.
+This function allows the user to apply custom R functions across all extreme event data frames created by the \code{gen\_hw\_sets} call.
 The user can apply any R function that follows certain standards in accepting input and returning output.
@@
 
-As an example, if the user wanted to to get the average temperature of the heat waves identified within each ensemble member, he or she could write a simple function:
+As an example, if the user wanted to calculate the average temperature of the heat waves identified for each ensemble member, he or she could write a simple function:
 
-\begin{verbatim}
+```
 average_mean_temp <- function(hw_datafr){
@@
         }
-\end{verbatim}
+```
 
-\noindent The `apply_all_models` function can then apply this `average_mean_temp` function across the heat wave data frames for all ensemble members in all climate models:
+\noindent This function could then be applied across all extreme event data sets output by \code{gen\_hw\_set} using the \code{apply\_all\_models} function.
+For example, to apply this function to all the example output results that come with the package, the user could run:
 
@@
 
-Note that the location of the directory with the heat wave data frames must be specified using the `out` argument when calling \code{apply\_all\_models}.
-Typically, this will be the directory path for the directory specified with the `out` argument in \code{gen\_hw\_set}.
+This output gives the results (\code{value} column) of running the custom function for each ensemble member of each climate model.
+Note that the location of the directory with the heat wave data frames must be specified using the \code{out} argument when calling \code{apply\_all\_models}.
+Typically, this will be the directory path for the directory specified with the \code{out} argument in \code{gen\_hw\_set}.
+
 Location-specific results can be generated using the \code{city\_specific} argument in \code{apply\_all\_models}:
@@
 
-The `apply_all_models` function can also be used to project the health impacts of heat waves.
-As a very simplistic example, \citet{anderson2009weather} estimated that heat waves, defined as two or more days at or above a community's $98^{th}$ percentile temperature, were associated with an added relative risk of 1.032 for cardiorespiratory mortality risk in 107 U.S.
-communities.
-A simple estimate of excess deaths associated with this added heat wave risk in a community can be calculated as \citep{peng2011toward}:
+The \code{apply\_all\_models} function can also be used to project the health impacts of heat waves.
+A simple estimate of excess deaths associated with heat wave risk in a community can be calculated as \citep{peng2011toward}:
 
 \begin{equation}
-E_c = N_c * (RR - 1) * L_c 
+E_c = B_c * (RR - 1) * L_c 
 \label{eq:healthimpacts}
@@
 
-- $E_c$ is the total number of excess deaths in community $c$;
-- $N_c$ is the baseline average daily mortality in community $c$; 
-- $RR$ is the relative risk of cardiorespiratory mortality per day associated with a heat wave; and 
-- $L_c$ is the total number of heat wave days in community $c$ over the study period.
+\begin{itemize}
+\item $E_c$ is the total number of excess deaths in community $c$;
+\item $B_c$ is the baseline average daily mortality in community $c$; 
+\item $RR$ is the relative risk of mortality associated with a heat wave day; and 
+\item $L_c$ is the total number of heat wave days in community $c$ over the study period.
+\end{itemize}
 
-This impact assessment equation can be translated into a function that merges each projection's heat wave data frame with a data frame of community-specific baseline mortality rates ($B_c$), calculates equation \ref{eq:healthimpacts} for each heat wave, and then sums up the total excess deaths across all heat waves:
+This impact assessment equation can be translated into a function that merges each projection's heat wave data frame with a data frame of community-specific baseline mortality rates ($B_c$), calculates equation \ref{eq:healthimpacts} for each heat wave, and then sums up the total excess deaths across all heat waves.
+In this example, we have set the default relative risk for a heat wave as 1.032, based on the estimated added heat wave relative risk in 107 U.S.
+communities for heat waves defined as two or more days at or above a community's 98\textsuperscript{th} percentile temperature \citep{anderson2009weather}.
 
-\begin{verbatim}
+```
 excess_deaths <- function(hw_datafr, base_mortality, RR = 1.032){
@@
                                       by = "city") %>%
-                dplyr::mutate(excess_deaths = base_mort * length * RR)
+                dplyr::mutate(excess_deaths = base_mort * (RR - 1) * length)
         out <- sum(hw_datafr$excess_deaths)
+        out <- round(out)
         return(out)
 }
-\end{verbatim}
+```
 
-\noindent Once defined in R, this function can be applied across all heat waves from all climate models' ensemble members, provided that you have a data frame called `city_mortality` with columns with each community's identifier (`city`) and baseline mortality rate (`base_mort`), using the call:
+```{r echo = FALSE}
+excess_deaths <- function(hw_datafr, base_mortality, RR = 1.032){
+        hw_datafr <- dplyr::left_join(hw_datafr, base_mortality,
+                                      by = "city") %>%
+                dplyr::mutate(excess_deaths = base_mort * (RR - 1) * length)
+        out <- sum(hw_datafr$excess_deaths)
+        out <- round(out)
+        return(out)
+}
+```
 
-```{r eval = FALSE}
+\noindent Once defined in R, this function can be applied across all heat waves from all climate models' ensemble members, provided that you have a data frame called \code{city\_mortality} with columns with each community's identifier (\code{city}) and baseline mortality rate (\code{base\_mort}), using the call (this example uses hypothetical baseline mortality rates for the cities and requires the \CRANpkg{dplyr} package \citep{dplyr}):
+
+```{r message = FALSE}
+library(dplyr)
+city_mortality <- data.frame(city = c("balt", "nwk", "ny", "phil", "prov"),
+                             base_mort = c(3, 2.5, 35, 6, 2.5))
 apply_all_models(out = out, FUN = excess_deaths, base_mortality = city_mortality)
@@
 
-As examples of the type of functions that can be used with `apply_all_models`, we have included several simple functions for exploring heat waves (Table \ref{tab:applyfunctions}).
-We included an example data frame representative of the heat wave data frames that `gen_hw_set` outputs, to use to test new functions.
-This data can be loaded with `data(hw_datafr)`.
-The process for creating a new function to use to explore heatwaves should be to:
-
-1. Load the example data frame with `load(hw_datafr)`;
-2. Build a function that works with `hw_datafr` as an input; and 
-3. Apply the function with `apply_all_models` to process all of the heat wave files.
-
-\begin{table}
-\centering
-\begin{tabular}{ll}
-\toprule
-Function & Description \\
-\midrule
-\code{number\_of\_heatwaves} & Determines the number of extreme events\\
-\code{heatwave\_days} & Sums up the total number of extreme event days across all heat waves\\
-\code{average\_length} & Calculates the average length of extreme events\\
-\code{average\_mean\_temp} & Calculates the average mean temperature across all extreme events\\
-\bottomrule
-\end{tabular}
-\caption{Example functions included in the \pkg{futureheatwaves} package that can be used to explore extreme events using the \code{apply\_all\_models} function.}
-\label{tab:applyfunctions}
-\end{table}
+\noindent While this example impact analysis is simplistic, it provides an example of how the \code{apply\_all\_models} function can be used in conjunction with epidemiologic results to estimate the potential health impacts of extreme events with climate change.
 
@@
 The heat wave data sets characterize heat wave in several ways that are based on relative temperature (for example, the `mean.temp.quantile`, `max.temp.quantile`, and `min.temp.quantile` columns described in Table \ref{tab:hwcharacteristics}).
-These characteristics are measured for each of the heat waves identified by the `gen_hw_set` function by taking the absolute temperature of the heat wave (e.g., average temperature during the heat wave is $90^oF$) and comparing it to the heat wave location's typical temperature distribution.
-This process generates relative measures of how intense the heat wave is compared to what is normal in that location (e.g., $90^oF$ is in the $99^{th}$ percentile of year-round temperatures in the location).
+These characteristics are measured for each of the heat waves identified by the `gen_hw_set` function by taking the absolute temperature of the heat wave (e.g., average temperature during the heat wave is 90\textsuperscript{o}F) and comparing it to the heat wave location's typical temperature distribution.
+This process generates relative measures of how intense the heat wave is compared to what is normal in that location (e.g., 90\textsuperscript{o}F is in the 99\textsuperscript{th} percentile of year-round temperatures in the location).
 
 The values of these relative heat wave characteristics will depend on what time period is used to determine the location's typical temperature distribution.
-For example, the same heat wave would likely have much higher values for these relative characteristics if the temperature during the heat wave is compared to typical temperatures in the present day for that location versus if the temperature is compared to typical temperatures much later in the $21^{st}$ century for the same location.
+For example, the same heat wave would likely have much higher values for these relative characteristics if the temperature during the heat wave is compared to typical temperatures in the present day for that location versus if the temperature is compared to typical temperatures much later in the 21\textsuperscript{st} century for the same location.
 
@@
 Users doing these kinds of extensions will need to pay attention to a few points.
-First, some of the event characteristics (first in the calendar year, average of May--September temperatures, days above $90^{o}F$) might not be meaningful for studies of other types of events.
+First, some of the event characteristics (first in the calendar year, average of May--September temperatures, days above 90\textsuperscript{o}F) might not be meaningful for studies of other types of events.
 Further, because event periods are usually defined as a string of multiple days exceeding some threshold, the functions in this package may miss the first and last event of the time period.

### 2016-12-29 paper_drafts/RJournalRevisions/anderson-eason-barnes.tex
Update article draft

@@
 with daily time steps and may use identification threholds based on
-local distributions (e.g., 98th percentile temperature for that
-location). Here, we provide an overview of working with daily climate
-model output data in R. Further, we present the \pkg{futureheatwaves}
-package, which we developed to ease the process of identifying,
-characterizing, and exploring multi-day extreme events in climate model
-output. This package takes as input a directory of projection files from
-many climate models, identify all heat waves for multiple cities using a
-user-specified heat wave definition, and write an output directory of
-files with these heat waves and their characteristics (e.g., length,
-absolute and relative temperatures). The \pkg{futureheatwaves} package
-then allows the user to create custom functions and apply them across
-all heat wave files; this functionality can be used either to explore
-heat wave patterns (e.g., average length and intensity across different
-climate models) or to apply epidemiological estimates to project health
-impacts under different scenarios.
+local distributions (e.g., 98\textsuperscript{th} percentile temperature
+for that location). Here, we provide an overview of working with daily
+climate model output data in R. Further, we present the
+\pkg{futureheatwaves} package, which we developed to ease the process of
+identifying, characterizing, and exploring multi-day extreme events in
+climate model output. This package takes as input a directory of
+projection files from one or many climate models, identifies all of a
+specific type of extreme event (e.g., heat waves) for multiple cities
+using a user-specified event definition, and writes an output directory
+of files with these extreme events and their characteristics (e.g.,
+length, absolute and relative temperatures). The \pkg{futureheatwaves}
+package then allows the user to create custom functions and apply them
+across all event files; this functionality can be used either to explore
+extreme event patterns (e.g., average length and intensity across
+different climate models) or to apply epidemiological estimates to
+project health impacts under different scenarios.
 }
@@
 
-This processing is particularly intensive for smaller time steps, like
-daily climate model output. While some climate impacts can be assessed
-using climate model output at monthly, seasonal, or yearly time steps,
-the impacts of multi-day extreme events must be assessed using output in
-daily time steps. Such multi-day extreme events include heat waves, cold
-spells, frost day spells, and droughts. The assessment is further
-complicated if extreme events are identified based on conditions that
-are rare for a certain location (e.g., 98th percentile of local
+Processing climate output data is particularly intensive for data in
+smaller time steps, like daily output. While some climate impacts can be
+assessed using climate model output at monthly, seasonal, or yearly time
+steps, the impacts of multi-day extreme events must be assessed using
+output in daily time steps. Such multi-day extreme events include heat
+waves, cold spells, frost day spells, and droughts. The assessment is
+further complicated if extreme events are identified based on conditions
+that are rare for a certain location (e.g., 98th percentile of local
 temperature distribution for identifying heat waves) \citep{IPCCch1}, as
@@
 multi-day extreme event from daily climate model output (Table
-\ref{tab:goals}). Further, this package provides some functionality
+\ref{tab:goals}). This package can be used to identify and characterize
+a variety of multi-day extreme events, but provides some functionality
 particularly useful in identifying and characterizing heat waves
@@
 defined differently than heat waves (for example, frost day spells may
-be defined as one or more days with temperature at or below \(0^oC\)).
-This package therefore allows the user to create and use a custom the
-extreme event definition used to identify events in the climate model
-output. Finally, this package allows the user to explore the extreme
-events identified in the climate model output with a function that can
-take a user-defined R function and apply it across all extreme event
-files generated for the separate ensemble members.
+be defined as one or more days with temperature at or below
+32\textsuperscript{o}F / 0\textsuperscript{o}C). This package therefore
+allows the user to create and use a custom the extreme event definition
+used to identify events in the climate model output. Finally, this
+package allows the user to explore the extreme events identified in the
+climate model output with a function that can take a user-defined R
+function and apply it across all extreme event files generated for the
+separate ensemble members.
 
@@
 
-The \pkg{futureheatwaves} package was facilitated to aid in identifying,
+We created the \pkg{futureheatwaves} package to aid in identifying,
 characterizing, and exploring multi-day extreme events in daily climate
@@
 directory specified by the user (Figure \ref{fig:overview}, ``Extreme
-events datasets``). Second, the \code{apply\_all\_models} function
-allows the user to apply custom functions across all the extreme event
-data frames generated by \code{gen\_hw\_set} to summaries of extreme
-events across all climate models and ensemble members (Figure
-\ref{fig:overview}, right).
+events datasets``).
+
+Once the user creates these data frames of location-specific heat waves,
+the \code{apply\_all\_models} function allows the user to apply custom
+functions across all the extreme event data frames generated by
+\code{gen\_hw\_set} to summaries of extreme events across all climate
+models and ensemble members (Figure \ref{fig:overview}, right). This
+functionality can be used to generate summary statistics (e.g.,
+determine average heat wave length or total frost days), or can be used
+to apply more complex functions (e.g., apply epidemiologic effect
+estimates across the heat waves to generate health impact estimates).
 
@@
 
-The user can extensively customize the definition used to identify
-extreme events. The default is a definition for heat waves (two or more
-days with temperatures at or above the city's \(98^{th}\) percentile of
-year-round temperature), but users can write a custom R function with
-either a different heat wave definition or with a definition appropriate
-for a different type of extreme event (e.g., one or more days at or
-below \(0^oC\) for frost day spells). Some of these heat wave
-characteristics are based on relative temperature, which is a measure of
-how the temperature during a heat wave compares to the typical
-distribution of temperature at that location \citep{IPCCch1}. These
-relative temperature metrics will vary depending on whether you
-calculate them based on a location's present-day temperature
-distribution or on the location's temperature distribution in the
-future, since typical temperature distributions in many locations are
-expected to change with climate change. The package therefore allows the
-user to specify date ranges of the temperature distributions to be used
-in calculating these relative temperature metrics in each location.
-
-Second, once the user creates these data frames of location-specific
-heat waves, the \texttt{apply\_all\_models} function allows the user to
-apply a custom function across all the heat wave data frames that were
-generated to create a summary data frame (Figure \ref{fig:overview},
-right). This functionality can be used to generate summary statistics
-(e.g., determine average heat wave length or total heat wave days), or
-can be used to apply more complex functions (e.g., apply epidemiologic
-effect estimates across the heat waves to generate health impact
-estimates).
-
 \subsection{Example data}\label{example-data}
@@
 users can try this package before applying it to their own directory of
-climate projection files. The example files are all included as
-comma-separated (.csv) files, rather than as saved R objects, because
-the \texttt{gen\_hw\_set} function requires a directory of
-comma-separated files as input and csv is a general file format familiar
-to most users. This example data therefore also allow users to see how
-to create the appropriate directory set-up for this package.
-
-This example data comes from two climate models that are a part of
-CMIP5: (1) the model of the Beijing Climate Center, China Meteorological
-Administration (BCC) \citep{xin2013introduction} and (2) the National
-Center for Atmospheric Research's (NCAR's) Community Climate System
-Model, version 4 (CCSM4) \citep{gent2011community}. We include one
-ensemble member from BCC (r1i1p1) and two from CCSM (r1i1p1 and r2i1p1).
+climate projection files. This example data comes from two climate
+models that are a part of CMIP5: (1) the model of the Beijing Climate
+Center, China Meteorological Administration (BCC)
+\citep{xin2013introduction} and (2) the National Center for Atmospheric
+Research's (NCAR's) Community Climate System Model, version 4 (CCSM4)
+\citep{gent2011community}. We include one ensemble member from BCC
+(r1i1p1) and two from CCSM (r1i1p1 and r2i1p1).
+
 To ensure that the size of this example data is reasonably small, we
@@
 Once the \pkg{futureheatwaves} package is installed and loaded, the user
-can find the local location of these files using R's
-\texttt{system.file} function. In the later sections of this article, we
-show how to use the package functions with these example files as
-inputs.
-
-When setting up to run climate model output data beyond this example
-data, any downloaded CMIP5 files will require some pre-processing, in
-terms of getting the getting the data into a specific format and storing
-these files in a specific directory structure. A file of study locations
-must also be saved locally in a specific format. Extensive details on
-the required format of files and required input directory structure is
-available in the \pkg{futureheatwaves} vignette ``futureheatwaves'',
-while tips on processing CMIP5 files into the required structure are
-given in the package's ``starting\_from\_netcdf'' vignette.
-
-\subsection{Basic example of processing projection
-files}\label{basic-example-of-processing-projection-files}
-
-Once these files and directories are set up, \pkg{futureheatwaves} can
-process them to identify and characterize heat waves in each ensemble
-member's projection for each location using the \texttt{gen\_hw\_set}
-function. For example, to process the example climate projections
-directory included with the package, the user can run:
+can find the local location of these files using R's \code{system.file}
+function. In the later sections of this article, we show how to use the
+package functions with these example files as inputs.
+
+When using this package with data other than this example data, the
+climate model output data will require some pre-processing. Data will
+need to be saved in a specific format, with files stored in a specific
+directory structure. This preprocessing can be done using some of the R
+packages described earlier for working with CMIP5 data in R. Full
+details of the required file and directory structure are provided in the
+package's ``futureheatwaves'' vignette, while tips on conducting this
+processing starting from CMIP5 netCDF files are given in the
+``starting\_from\_netcdf'' vignette.
+
+\subsection{Basic example of identifying and characterizing extreme
+events}\label{basic-example-of-identifying-and-characterizing-extreme-events}
+
+Once climate model output files are set up as specified in the
+``futureheatwaves'' package vignette, the package can process them to
+identify and characterize heat waves in each ensemble member's
+projection for each location using the \code{gen\_hw\_set} function. For
+example, to process the example climate model output data included with
+the package, the user can run:
 
@@
 user's computer of the example climate projections directory
-(\texttt{projection\_dir\_location}) and the file of study locations
-(\texttt{city\_file\_location}). The \texttt{gen\_hw\_set} function
+(\code{projection\_dir\_location}) and the file of study locations
+(\code{city\_file\_location}). The \code{gen\_hw\_set} function
 processes this example input and creates a new directory,
-\texttt{example\_results}, with files of identified and characterized
-heat waves, in the user's current working directory. In this example
-code, this processing is done using default values for the heat wave
-definition, years for which to generate the heat wave data sets, etc.
-Ways to customize these choices are fully explained later in the text.
-
-In this call, the user must specify the directory where the results
-should be written (\texttt{out}), the location of the directory of
-climate projections (\texttt{dataFolder}), the names of the two main
-subdirectories of that climate projection directory, as well as their
-year boundaries (\texttt{dataDirectories}; see, for example, Figure
-\ref{fig:directories}), the location of the city location file
-(\texttt{citycsv}), and the names used for the grid coordinate, climate
-projection, and projection date files (\texttt{coordinateFilenames},
-\texttt{tasFilenames}, and \texttt{timeFilenames}). When
-\texttt{gen\_hw\_set} is run, the user is advised that the function will
-write files to his computer and must agree to proceed:
-
-\begin{verbatim}
- Warning: This function will write new files to your computer in the 
- ~/tmp/example_results/ directory of your computer. If that directory already exists,
-running this function will write over it. 
- Do you want to continue? (y / n): 
-\end{verbatim}
-
-\noindent The function provides status reports while processing. This
-helps the user see that the function call is progressing, as
-\texttt{gen\_hw\_set} can take a while to run if it is processing many
-locations and / or many climate ensemble members. Once the function has
-completed running, results will be written locally to the directory
-specified by the \texttt{out} argument of \texttt{gen\_hw\_set}. The
-structure of this output directory is shown in Figure
-\ref{fig:directories} (right). This directory will include files with
-some basic information about the climate models and the closest grid
-points of each climate model to each location. The output directory will
-also include a directory with files of identified and classified heat
-waves for each ensemble member, including all characteristics in Table
-\ref{tab:hwcharacteristics}.
+\file{example\_results}, with files of identified and characterized heat
+waves, in the user's current working directory. In this example code,
+the processing is done using default values for the event definition,
+years for which to generate the heat wave data sets, etc. How and why to
+customize these choices are explained later in the text. Function
+arguments (e.g., \code{dataDirectories}, \code{tasFilenames}) are used
+to specify the format of the data and the directory structure.
+
+Once the function has completed running, results will be written locally
+to the directory specified by the \code{out} argument of
+\code{gen\_hw\_set}. This directory will include files with some basic
+information about the climate models and the closest grid points of each
+climate model to each location, as well as a directory with files of
+identified and classified extreme events for each ensemble member,
+including all characteristics in Table \ref{tab:hwcharacteristics}.
 
@@
 
-The default heat wave definition for this package is:
+By default, the package identifies extreme events in climate model
+output data using a specific definition for heat waves:
 
@@
 A \dfn{heat wave} is two or more days at or above a city-specific
-threshold temperature, with the threshold determined as the \(98^{th}\)
-percentile of year-round temperature in the city during some reference
-period (by default, 1990--1999).
+threshold temperature, with the threshold determined as the
+98\textsuperscript{th} percentile of year-round temperature in the city
+during some reference period (by default, 1990--1999).
 \end{quote}
 
-Many different definitions of heat waves exist (e.g.,
-\citet{smith2013heat, kent2014heat}), so researchers will often want to
-use alternative ways to define heat waves. Researchers might want to use
-a specific definition, for example, because it matches the definition
-used by local health officials to declare heat wave warnings or, in the
-case of health impact assessments, to match with a definition used in an
-epidemiological study. Three components of the heat wave definition can
-be easily customized in the \texttt{gen\_hw\_set} function call, without
-creating a new R function to use to identify heat waves. The
-customization of the heat wave definition is even more extensive as one
-has the option of writing a custom R function.
-
-First, the percentile used to identify a heat wave can be changed using
-the \texttt{probThreshold} option in \texttt{gen\_hw\_set}. This option
-can take values between 0 and 1. The default value is 0.98, or a
-definition with a threshold of the \(98^{th}\) percentile of the
-location's temperature. For example, to identify heat waves as two or
-more days at or above the city's \(99^{th}\) percentile of year-round
-temperature, the user could run:
-
-\begin{Schunk}
-\begin{Sinput}
-gen_hw_set(out = "example_results",
-           dataFolder = projection_dir_location ,
-           dataDirectories = list("historical" = c(1990, 1999),
-                                        "rcp85" = c(2060, 2079)),
-           citycsv = city_file_location,
-           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
-           tasFilenames = "tas_NorthAmerica_12mo.csv",
-           timeFilenames = "time_NorthAmerica_12mo.csv",
-           probThreshold = 0.99)
-\end{Sinput}
-\end{Schunk}
-
-\noindent Second, the user can change the number of days used in the
-heat wave definition using the \texttt{numDays} argument in the
-\texttt{gen\_hw\_set} function. Combined, these two customization
-choices can allow the user to identify heat waves using many of the heat
-wave definitions used in previous climate and health research-- for
-example, 9 of the 16 heat wave definitions used in \citet{kent2014heat}
-could be fit using different combinations of these two options for
-specifying threshold percentile and number of days.
-
-Third, it is possible to specify the range of years that should be used
-when determining this threshold. For example, if you wanted to base the
-threshold for each location on its current climate, you could leave this
-option as its default, while if you wanted to use a different set of
-years (i.e.~a threshold relative to future projected temperatures), you
-could set the start and end year bounds for this reference period using
-the \texttt{thresholdBoundaries} argument in the function
-\texttt{gen\_hw\_set}. For example, to use temperatures in each city
-from 2070 to 2079 to determine threshold temperatures, you would run:
-
-\begin{Schunk}
-\begin{Sinput}
-gen_hw_set(out = "example_results",
-           dataFolder = projection_dir_location ,
-           dataDirectories = list("historical" = c(1990, 1999),
-                                        "rcp85" = c(2060, 2079)),
-           citycsv = city_file_location,
-           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
-           tasFilenames = "tas_NorthAmerica_12mo.csv",
-           timeFilenames = "time_NorthAmerica_12mo.csv",
-           thresholdBoundaries = c(2070, 2079))
-\end{Sinput}
-\end{Schunk}
-
-It is also possible to use a completely customized heat wave definition.
-This functionality allows the user to use heat wave definitions that
-either require a number of days above an absolute threshold (e.g.,
-maximum temperature of \(\ge95^oF\) for \(\ge1\) day;
-\citet{kent2014heat, tan2007heat}) or that require a combination of
-thresholds to be met (e.g., maximum daily temperature above a lower
-threshold every day of the heat wave and above a higher threshold for a
-certain number of days; \citet{kent2014heat, peng2011toward}).
-
-To use a customized heat wave definition, the user can write and load an
-R function that implements the definition. To work correctly, this
-custom function must allow only specific inputs and generate only
-specific outputs. Details about this required structure are provided in
-the \pkg{futureheatwaves} package vignette.
-
-To help in developing custom heat wave identification functions, the
-package includes an example of the required input data frame,
-\texttt{datafr}, which can be loaded with \texttt{data(datafr)}. If a
-custom heat wave function is properly set up to use in this package, it
-can take as input this example data frame and a threshold value and will
-return the original data frame with the columns for whether the day was
-part of a heat wave (\texttt{hw}) and, if so, the sequential number of
-the heat wave out of all heat waves in that location
-(\texttt{hw.number}).
-
-The custom function can then be referenced using the
-\texttt{IDheatwavesFunction} option in \texttt{gen\_hw\_set}, and it
-will be used to identify heat waves in the data set. As an example, we
-have included an alternative ID definition function in this package,
-called \texttt{IDHeatwavesAlternative}, which identifies heat waves as
-two or more days above the higher of either the location's \(98^{th}\)
-percentile temperature or \(80^{o}F\), with the idea that this
-definition prevents the identification of heat waves at mild
-temperatures in locations with mild or cool climates. The following code
-processes the example climate projecting files using this function as
-the heat wave definition:
-
-\begin{Schunk}
-\begin{Sinput}
-gen_hw_set(out = "example_results",
-           dataFolder = projection_dir_location ,
-           dataDirectories = list("historical" = c(1990, 1999),
-                                        "rcp85" = c(2060, 2079)),
-           citycsv = city_file_location,
-           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
-           tasFilenames = "tas_NorthAmerica_12mo.csv",
-           timeFilenames = "time_NorthAmerica_12mo.csv",
-           IDheatwavesFunction = "IDHeatwavesAlternative")
-\end{Sinput}
-\end{Schunk}
-
-With this ability to write and use a custom heat wave identification
-function, most heat wave definitions can be applied with this package.
-The exception is definitions that require two separate temperature
-metrics (e.g., maximum temperature above a certain threshold or minimum
-temperature above a different threshold;
-\citet{kent2014heat, robinson2001definition}), because the
-\texttt{gen\_hw\_set} function cannot take as input projections of more
-than one temperature metric. Definitions that use a heat metric that
-combines multiple inputs (e.g., heat index, which is a combined measure
-of temperature and air moisture) can be used, but require that the
-calculation of the metric is done when creating the input files to
-include in the ``Climate projections" directory (Figure
-\ref{fig:overview}, left), rather than as part of the heat wave
-definition function.
-
-To increase processing speed when identifying heat waves, we coded parts
-of the default heat wave identification function in C++ and synced it
-with R using the \CRANpkg{Rcpp} package \citep{Rcpp}. Users should
-consider a similar strategy for custom heat wave definitions, especially
-if processing a large number of climate projection files.
+\noindent However, the user can extensively customize the definition
+used to identify extreme events. Users can write a custom R function
+with either a different heat wave definition (see \citet{smith2013heat}
+and \citet{kent2014heat} for listings of definitions used in scientific
+studies) or with a definition appropriate for a different type of
+extreme event (e.g., one or more days at or below 32\textsuperscript{o}F
+/ 0\textsuperscript{o}C for frost day spells). For heat wave
+identification, researchers might want to use a specific definition, for
+example, because it matches the definition used by local health
+officials to declare heat wave warnings or, in the case of health impact
+assessments, to match with a definition used in an epidemiological
+study. For other extreme events, a heat wave definition likely will not
+be applicable and so a customized definition is necessary.
+
+Three components of the extreme event definition can be easily
+customized in the \code{gen\_hw\_set} function call, without creating a
+new R function to use to identify heat waves. First, often users wish to
+define an extreme event based on conditions that are rare in the study
+location \citep{IPCCch1}. In this case, the percentile of the variable
+of interest required for an extreme event can be changed using the
+\code{probThreshold} option in \code{gen\_hw\_set}. For example, this
+option could be used to identify heat waves based on the
+99\textsuperscript{th} percentile, rather than the
+98\textsuperscript{th} percentile, of a city's year-round temperature.
+Second, the user can change the number of days used in the heat wave
+definition using the \code{numDays} argument in the \code{gen\_hw\_set}
+function (the default is \samp{numDays = 2}, corresponding to two or
+more days). Combined, these two customization choices allow the user to
+identify heat waves using many of the heat wave definitions used in
+previous climate and health research-- for example, 9 of the 16 heat
+wave definitions used in \citet{kent2014heat} (e.g., {[}some examples of
+the 9 that could be fit{]}) could be fit using different combinations of
+these two options for specifying threshold percentile and number of
+days. Third, some extreme events like cold waves and frost day spells
+are defined as a certain number of days below, rather than above, a
+threshold. While the default is to identify events by searching for days
+above a threshold, this behavior can be changed with the
+\code{above\_threshold = FALSE} argument in the
+\code{gen\_hw\_set function}.
+
+Beyond these simpler options, the customization of the event definition
+is even more extensive as one has the option of writing a custom R
+function. This functionality allows the user to use heat wave
+definitions that either require a number of days above an absolute
+threshold (e.g., maximum temperature of \(\ge\) 95\textsuperscript{o}F
+for \(\ge1\) day \citet{kent2014heat, tan2007heat}; minimum temperature
+\(\le\) 0\textsuperscript{o}C for \(\ge1\) day for frost day spells) or
+that require a combination of thresholds to be met (e.g., maximum daily
+temperature one threshold every day of the heat wave and above a higher
+threshold for a certain number of days;
+\citet{kent2014heat, peng2011toward}). To use a customized event
+definition, the user can write and load an R function that implements
+the definition. The custom function is passed to the \code{gen\_hw\_set}
+function using the \code{IDheatwavesFunction} argument. To work
+correctly, this custom function must allow only specific inputs and
+generate only specific outputs. Details about this required structure
+are provided in the \pkg{futureheatwaves} package vignette. To increase
+processing speed when identifying extreme events, we coded parts of the
+default event definition function in C++ and synced it with R using the
+\CRANpkg{Rcpp} package \citep{Rcpp}. Users should consider a similar
+strategy for custom heat wave definitions, especially if processing a
+large number of climate projection files.
+
+Finally, extreme events are often based on conditions that are rare for
+a specific location using location-specific relative thresholds. These
+thresholds are often defined for climate impact studies based on a
+variable's distribution at that location in present-day or historical
+data. However, it can be interesting to explore trends in extreme events
+under climate change if extreme events are identified based on variable
+distributions during the projection period, or another future period. It
+is possible to specify the time period to use when determining a
+relative threshold for an event definition using the
+\code{thresholdBoundaries} argument in the function \code{gen\_hw\_set}.
 
@@
 
-Once you have created a directory of files with characterized heat waves
-for each ensemble member, the results can be explored using the
-\texttt{apply\_all\_models} function. This function allows the user to
-apply custom R functions across all heat wave data frames created by the
-\texttt{gen\_hw\_sets} call. The user can apply any R function that
+Once the user has created a directory of characterized event files for
+each ensemble member (``Extreme event data sets'', Figure
+\ref{fig:overview}), he or she can explore the results using the
+\code{apply\_all\_models} function. This function allows the user to
+apply custom R functions across all extreme event data frames created by
+the \code{gen\_hw\_sets} call. The user can apply any R function that
 follows certain standards in accepting input and returning output. Full
@@
 
-As an example, if the user wanted to to get the average temperature of
-the heat waves identified within each ensemble member, he or she could
+As an example, if the user wanted to calculate the average temperature
+of the heat waves identified for each ensemble member, he or she could
 write a simple function:
@@
 
-\noindent The \texttt{apply\_all\_models} function can then apply this
-\texttt{average\_mean\_temp} function across the heat wave data frames
-for all ensemble members in all climate models:
+\noindent This function could then be applied across all extreme event
+data sets output by \code{gen\_hw\_set} using the
+\code{apply\_all\_models} function. For example, to apply this function
+to all the example output results that come with the package, the user
+could run:
 
@@
 
-Note that the location of the directory with the heat wave data frames
-must be specified using the \texttt{out} argument when calling
+This output gives the results (\code{value} column) of running the
+custom function for each ensemble member of each climate model. Note
+that the location of the directory with the heat wave data frames must
+be specified using the \code{out} argument when calling
 \code{apply\_all\_models}. Typically, this will be the directory path
-for the directory specified with the \texttt{out} argument in
-\code{gen\_hw\_set}. Location-specific results can be generated using
-the \code{city\_specific} argument in \code{apply\_all\_models}:
+for the directory specified with the \code{out} argument in
+\code{gen\_hw\_set}.
+
+Location-specific results can be generated using the
+\code{city\_specific} argument in \code{apply\_all\_models}:
 
@@
 
-The \texttt{apply\_all\_models} function can also be used to project the
-health impacts of heat waves. As a very simplistic example,
-\citet{anderson2009weather} estimated that heat waves, defined as two or
-more days at or above a community's \(98^{th}\) percentile temperature,
-were associated with an added relative risk of 1.032 for
-cardiorespiratory mortality risk in 107 U.S. communities. A simple
-estimate of excess deaths associated with this added heat wave risk in a
-community can be calculated as \citep{peng2011toward}:
+The \code{apply\_all\_models} function can also be used to project the
+health impacts of heat waves. A simple estimate of excess deaths
+associated with heat wave risk in a community can be calculated as
+\citep{peng2011toward}:
 
 \begin{equation}
-E_c = N_c * (RR - 1) * L_c 
+E_c = B_c * (RR - 1) * L_c 
 \label{eq:healthimpacts}
@@
 \begin{itemize}
-\tightlist
-\item
-  \(E_c\) is the total number of excess deaths in community \(c\);
-\item
-  \(N_c\) is the baseline average daily mortality in community \(c\);
-\item
-  \(RR\) is the relative risk of cardiorespiratory mortality per day
-  associated with a heat wave; and
-\item
-  \(L_c\) is the total number of heat wave days in community \(c\) over
-  the study period.
+\item $E_c$ is the total number of excess deaths in community $c$;
+\item $B_c$ is the baseline average daily mortality in community $c$; 
+\item $RR$ is the relative risk of mortality associated with a heat wave day; and 
+\item $L_c$ is the total number of heat wave days in community $c$ over the study period.
 \end{itemize}
@@
 equation \ref{eq:healthimpacts} for each heat wave, and then sums up the
-total excess deaths across all heat waves:
+total excess deaths across all heat waves. In this example, we have set
+the default relative risk for a heat wave as 1.032, based on the
+estimated added heat wave relative risk in 107 U.S. communities for heat
+waves defined as two or more days at or above a community's
+98\textsuperscript{th} percentile temperature
+\citep{anderson2009weather}.
 
@@
                                       by = "city") %>%
-                dplyr::mutate(excess_deaths = base_mort * length * RR)
+                dplyr::mutate(excess_deaths = base_mort * (RR - 1) * length)
         out <- sum(hw_datafr$excess_deaths)
+        out <- round(out)
         return(out)
@@
 heat waves from all climate models' ensemble members, provided that you
-have a data frame called \texttt{city\_mortality} with columns with each
-community's identifier (\texttt{city}) and baseline mortality rate
-(\texttt{base\_mort}), using the call:
+have a data frame called \code{city\_mortality} with columns with each
+community's identifier (\code{city}) and baseline mortality rate
+(\code{base\_mort}), using the call (this example uses hypothetical
+baseline mortality rates for the cities and requires the \CRANpkg{dplyr}
+package \citep{dplyr}):
 
@@
 \begin{Sinput}
+library(dplyr)
+city_mortality <- data.frame(city = c("balt", "nwk", "ny", "phil", "prov"),
+                             base_mort = c(3, 2.5, 35, 6, 2.5))
 apply_all_models(out = out, FUN = excess_deaths, base_mortality = city_mortality)
 \end{Sinput}
+\begin{Soutput}
+#>   model ensemble value
+#> 1  bcc1        1   930
+#> 2  ccsm        1   968
+#> 3  ccsm        2   967
+\end{Soutput}
 \end{Schunk}
 
-As examples of the type of functions that can be used with
-\texttt{apply\_all\_models}, we have included several simple functions
-for exploring heat waves (Table \ref{tab:applyfunctions}). We included
-an example data frame representative of the heat wave data frames that
-\texttt{gen\_hw\_set} outputs, to use to test new functions. This data
-can be loaded with \texttt{data(hw\_datafr)}. The process for creating a
-new function to use to explore heatwaves should be to:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Load the example data frame with \texttt{load(hw\_datafr)};
-\item
-  Build a function that works with \texttt{hw\_datafr} as an input; and
-\item
-  Apply the function with \texttt{apply\_all\_models} to process all of
-  the heat wave files.
-\end{enumerate}
-
-\begin{table}
-\centering
-\begin{tabular}{ll}
-\toprule
-Function & Description \\
-\midrule
-\code{number\_of\_heatwaves} & Determines the number of extreme events\\
-\code{heatwave\_days} & Sums up the total number of extreme event days across all heat waves\\
-\code{average\_length} & Calculates the average length of extreme events\\
-\code{average\_mean\_temp} & Calculates the average mean temperature across all extreme events\\
-\bottomrule
-\end{tabular}
-\caption{Example functions included in the \pkg{futureheatwaves} package that can be used to explore extreme events using the \code{apply\_all\_models} function.}
-\label{tab:applyfunctions}
-\end{table}
+\noindent While this example impact analysis is simplistic, it provides
+an example of how the \code{apply\_all\_models} function can be used in
+conjunction with epidemiologic results to estimate the potential health
+impacts of extreme events with climate change.
 
@@
 by taking the absolute temperature of the heat wave (e.g., average
-temperature during the heat wave is \(90^oF\)) and comparing it to the
-heat wave location's typical temperature distribution. This process
-generates relative measures of how intense the heat wave is compared to
-what is normal in that location (e.g., \(90^oF\) is in the \(99^{th}\)
-percentile of year-round temperatures in the location).
+temperature during the heat wave is 90\textsuperscript{o}F) and
+comparing it to the heat wave location's typical temperature
+distribution. This process generates relative measures of how intense
+the heat wave is compared to what is normal in that location (e.g.,
+90\textsuperscript{o}F is in the 99\textsuperscript{th} percentile of
+year-round temperatures in the location).
 
@@
 day for that location versus if the temperature is compared to typical
-temperatures much later in the \(21^{st}\) century for the same
-location.
+temperatures much later in the 21\textsuperscript{st} century for the
+same location.
 
@@
 The lines on the map connect each climate model grid point to the study location(s) for which that grid point was used.
-The interactive maps include pop-ups with city identiers; one is shown open in this snapshot as an example.}
+The interactive maps include pop-ups with city identiers; one is shown open in this snapshot as an example.
+From this map, you can see that the climate model grid point closest to New York City for this climate model is over the Atlantic Ocean.}
 \label{fig:gridmap}
@@
 calendar year, average of May--September temperatures, days above
-\(90^{o}F\)) might not be meaningful for studies of other types of
-events. Further, because event periods are usually defined as a string
-of multiple days exceeding some threshold, the functions in this package
-may miss the first and last event of the time period. For example, if
-the first day of the time series is the last day of an event, this
-function would not identify that event because it lacks data from the
-earlier days that allow this day to meet the event definition. This
-issue would lead to, at most, missing two events out of each projection,
-but should be considered if studying events that might occur near the
-start and end of projection data. If there is adequate interest from
-researchers, in the future we may adapt the package to make these
+90\textsuperscript{o}F) might not be meaningful for studies of other
+types of events. Further, because event periods are usually defined as a
+string of multiple days exceeding some threshold, the functions in this
+package may miss the first and last event of the time period. For
+example, if the first day of the time series is the last day of an
+event, this function would not identify that event because it lacks data
+from the earlier days that allow this day to meet the event definition.
+This issue would lead to, at most, missing two events out of each
+projection, but should be considered if studying events that might occur
+near the start and end of projection data. If there is adequate interest
+from researchers, in the future we may adapt the package to make these
 secondary applications part of the package.

### 2016-12-29 paper_drafts/RJournalRevisions/anderson-eason-barnes.Rmd
Update article draft

@@
 
-## Customizing dates for projections and reference temperatures
+# Customizing dates for projections and reference temperatures
 
-By default, the `gen_hw_set` function will generate heat wave projections for the years 2070 to 2079, to align with the example data.
-This projection date range can be changed by specifying alternative starting and ending years in the `projectionBoundaries` argument of the `gen_hw_set` function.
-For example, to create projections for 2060 to 2079, the user would call:
+Depending on the climate model output data used and the research question, different projection periods may be of interest.
+For example, one research may want to explore extreme events in the 2050--2075 period, while another may be interested in 2090--2100.
+The projection date range can be changed by specifying starting and ending years in the \code{projectionBoundaries} argument of the \code{gen\_hw\_set} function.
 
-```{r, eval = FALSE}
-gen_hw_set(out = "example_results",
-           dataFolder = projection_dir_location ,
-           dataDirectories = list("historical" = c(1990, 1999),
-                                        "rcp85" = c(2060, 2079)),
-           citycsv = city_file_location,
-           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
-           tasFilenames = "tas_NorthAmerica_12mo.csv",
-           timeFilenames = "time_NorthAmerica_12mo.csv",
-           projectionBoundaries = c(2060, 2079))
-```
+Further, some of the event characteristics (e.g., \code{mean.temp.quantile}, Table \ref{tab:hwcharacteristics}) are calculated based on relative temperature, which is a measure of how the value of the variable of interest during an extreme event compares to the typical distribution of that variable at that location.
+These characteristics are measured for each of the extreme events identified by the \code{gen\_hw\_set} function by taking the absolute value of the variable during the event (e.g., average temperature during the heat wave is 90\textsuperscript{o}F \ 32.2\textsuperscript{o}C) and comparing it to the location's typical variable distribution.
+This process generates relative measures of how intense the event is compared to what is normal in that location (e.g., 90\textsuperscript{o}F \ 32.2\textsuperscript{o}C is in the 99\textsuperscript{th} percentile of year-round temperatures in the location).
 
-The heat wave data sets characterize heat wave in several ways that are based on relative temperature (for example, the `mean.temp.quantile`, `max.temp.quantile`, and `min.temp.quantile` columns described in Table \ref{tab:hwcharacteristics}).
-These characteristics are measured for each of the heat waves identified by the `gen_hw_set` function by taking the absolute temperature of the heat wave (e.g., average temperature during the heat wave is 90\textsuperscript{o}F) and comparing it to the heat wave location's typical temperature distribution.
-This process generates relative measures of how intense the heat wave is compared to what is normal in that location (e.g., 90\textsuperscript{o}F is in the 99\textsuperscript{th} percentile of year-round temperatures in the location).
-
-The values of these relative heat wave characteristics will depend on what time period is used to determine the location's typical temperature distribution.
-For example, the same heat wave would likely have much higher values for these relative characteristics if the temperature during the heat wave is compared to typical temperatures in the present day for that location versus if the temperature is compared to typical temperatures much later in the 21\textsuperscript{st} century for the same location.
-
-The default in the `gen_hw_set` function is to use temperatures for the period 2070 to 2079 for these reference temperatures to calculate all relative heat wave characteristics.
-However, the user can change this specification using the `referenceBoundaries` option of `gen_hw_set`.
-This functionality can be useful in exploring the role of adaptation in future heat waves.
-For example, to use temperature projections for 1990 to 1999 when calculating relative characteristics of heat waves, to explore the assumption that cities remain adapted to their present-day climate, rather than changing in adaptation as climate change increases temperatures, a user could run:
-
-```{r, eval = FALSE}
-gen_hw_set(out = "example_results",
-           dataFolder = projection_dir_location ,
-           dataDirectories = list("historical" = c(1990, 1999),
-                                        "rcp85" = c(2060, 2079)),
-           citycsv = city_file_location,
-           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
-           tasFilenames = "tas_NorthAmerica_12mo.csv",
-           timeFilenames = "time_NorthAmerica_12mo.csv",
-           referenceBoundaries = c(1990, 1999))
-```
-
-For these date range specifications, there are some restrictions on which year ranges can be selected.
-The starting year cannot be earlier than the first year in the first subdirectory, the ending year cannot be later than the last year of the second subdirectory, and the custom date boundaries cannot span the two subdirectories.
+These relative event characteristics will vary depending on whether you calculate them based on a location's present-day variable distribution or on the location's variable distribution in the future, since the distributions of many relevant variables (e.g., temperature, precipitation) are expected to change in many locations with climate change.
+The package therefore allows the user to specify date ranges of the temperature distributions to be used in calculating these relative temperature metrics in each location, which can be done using the \code{referenceBoundaries} option of \code{gen\_hw\_set}.
 
 ## Mapping grid points
-
-It can be useful to explore the location of the climate model grid point used to pull climate model output for each study location with a given climate model.
-Therefore, the package has a function called `map_grid_leaflet` that plots the locations of grid points used for each location from each climate model.
-This function is built using the htmlWidget \pkg{leaflet} package \citep{leaflet}.
-The following code illustrates the use of this function with the example data to create Figure \ref{fig:gridmap}, which plots the grid points used in the example data from the BCC climate model in the example data:
-
-```{r eval = FALSE}
-out <- system.file("extdata/example_results", package = "futureheatwaves")
-map_grid_leaflet(plot_model = "bcc1", out = out)
-```
 
@@
 
+It can be useful to explore the location of the climate model grid point used to pull climate model output for each study location with a given climate model.
+Therefore, the package has a function called `map_grid_leaflet` that plots the locations of grid points used for each location from each climate model.
+This function is built using the htmlWidget \pkg{leaflet} package \citep{leaflet}.
+The following code illustrates the use of this function with the example data to create Figure \ref{fig:gridmap}, which plots the grid points used in the example data from the BCC climate model in the example data:
+
+```{r eval = FALSE}
+out <- system.file("extdata/example_results", package = "futureheatwaves")
+map_grid_leaflet(plot_model = "bcc1", out = out)
+```
+
 ## Extensions
 
-The functionality of this package can be easily expanded by loops.
-For example, to explore the role of the heat wave definition on projections, the user could create a loop to run `gen_hw_set` and `apply_all_models` to the same directory of climate projections but with a variety of different functions used to identify the heat waves in the projections.
-
-Also, while this package was created to be used for research on heat waves in climate change projections, with some modifications it can be used more broadly.
-For example, there are other episodes like wildfires and air pollution where it may be interesting to identify extended periods of high exposures in projection time series, and this package could be applied to explore these exposures.
-The directory of projection data would need to be set up in the same structure as for exploring heat waves, and the `input_metric` should be set as `fahrenheit`, to pass the exposure values through to the characterized data sets without performing a conversion.
-A user could also use this package to explore events that must be lower than some minimum threshold (e.g., cold waves), but it would take some extra coding and conversions, since the functions in this package are written to identify periods above a threshold (for example, the user could multiple all projected temperatures by -1, and then the coldest temperatures would register as being the highest).
-
-```{r eval = FALSE}
-## Cold wave example
-gen_hw_set(out = "example_results",
-           dataFolder = projection_dir_location ,
-           dataDirectories = list("historical" = c(1990, 1999),
-                                  "rcp85" = c(2060, 2079)),
-           citycsv = city_file_location,
-           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
-           tasFilenames = "tas_NorthAmerica_12mo.csv",
-           timeFilenames = "time_NorthAmerica_12mo.csv", 
-           probThreshold = 0.10, 
-           above_threshold = FALSE, 
-           absolute_thresholds = c(266, 263, 260, 258), 
-           seasonal_months = c(12, 1, 2))
-```
-
->  Extreme statistics and spell-lengths: "A 'frost day' is defined as one during which the minimum temperature falls below freezing point (0 degC).
-This is described as a climatological statistic, in which the minimum temperature is first calculated within each day, and then the number of days or spell lengths meeting the specified condition are evaluated." (from one example in http://cfconventions.org/Data/cf-conventions/cf-conventions-1.6/build/cf-conventions.html#calendar)
-
-Users doing these kinds of extensions will need to pay attention to a few points.
-First, some of the event characteristics (first in the calendar year, average of May--September temperatures, days above 90\textsuperscript{o}F) might not be meaningful for studies of other types of events.
-Further, because event periods are usually defined as a string of multiple days exceeding some threshold, the functions in this package may miss the first and last event of the time period.
-For example, if the first day of the time series is the last day of an event, this function would not identify that event because it lacks data from the earlier days that allow this day to meet the event definition.
-This issue would lead to, at most, missing two events out of each projection, but should be considered if studying events that might occur near the start and end of projection data.
-If there is adequate interest from researchers, in the future we may adapt the package to make these secondary applications part of the package.
+While this package was created to be used for research on extreme events in climate change projections, it can be used more broadly.
+For example, there are other episodes like wildfires and air pollution where it may be interesting to identify extended periods of high exposures in projection time series, and this package could be applied to gridded air pollution model output to explore these exposures.
 
@@
 This work was supported by grants from the National Institute of Environmental Health Sciences (R00ES022631), the National Science Foundation (1331399), and the Colorado State University Vice President for Research.
+Thank you to early testers of the package:  Julia Bromberek, Wande Benka-Coker, Josh Ferreri, Ryan Gan, Molly Gutilla, Mike Lyons, Casey Quinn, Rachel Severson, and Meilin Yan.
 

### 2016-12-29 paper_drafts/RJournalRevisions/anderson-eason-barnes.tex
Update article draft

@@
 
-\subsection{Customizing dates for projections and reference
+\section{Customizing dates for projections and reference
 temperatures}\label{customizing-dates-for-projections-and-reference-temperatures}
 
-By default, the \texttt{gen\_hw\_set} function will generate heat wave
-projections for the years 2070 to 2079, to align with the example data.
-This projection date range can be changed by specifying alternative
-starting and ending years in the \texttt{projectionBoundaries} argument
-of the \texttt{gen\_hw\_set} function. For example, to create
-projections for 2060 to 2079, the user would call:
+Depending on the climate model output data used and the research
+question, different projection periods may be of interest. For example,
+one research may want to explore extreme events in the 2050--2075
+period, while another may be interested in 2090--2100. The projection
+date range can be changed by specifying starting and ending years in the
+\code{projectionBoundaries} argument of the \code{gen\_hw\_set}
+function.
 
-\begin{Schunk}
-\begin{Sinput}
-gen_hw_set(out = "example_results",
-           dataFolder = projection_dir_location ,
-           dataDirectories = list("historical" = c(1990, 1999),
-                                        "rcp85" = c(2060, 2079)),
-           citycsv = city_file_location,
-           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
-           tasFilenames = "tas_NorthAmerica_12mo.csv",
-           timeFilenames = "time_NorthAmerica_12mo.csv",
-           projectionBoundaries = c(2060, 2079))
-\end{Sinput}
-\end{Schunk}
-
-The heat wave data sets characterize heat wave in several ways that are
-based on relative temperature (for example, the
-\texttt{mean.temp.quantile}, \texttt{max.temp.quantile}, and
-\texttt{min.temp.quantile} columns described in Table
-\ref{tab:hwcharacteristics}). These characteristics are measured for
-each of the heat waves identified by the \texttt{gen\_hw\_set} function
-by taking the absolute temperature of the heat wave (e.g., average
-temperature during the heat wave is 90\textsuperscript{o}F) and
-comparing it to the heat wave location's typical temperature
-distribution. This process generates relative measures of how intense
-the heat wave is compared to what is normal in that location (e.g.,
-90\textsuperscript{o}F is in the 99\textsuperscript{th} percentile of
+Further, some of the event characteristics (e.g.,
+\code{mean.temp.quantile}, Table \ref{tab:hwcharacteristics}) are
+calculated based on relative temperature, which is a measure of how the
+value of the variable of interest during an extreme event compares to
+the typical distribution of that variable at that location. These
+characteristics are measured for each of the extreme events identified
+by the \code{gen\_hw\_set} function by taking the absolute value of the
+variable during the event (e.g., average temperature during the heat
+wave is 90\textsuperscript{o}F ~32.2\textsuperscript{o}C) and comparing
+it to the location's typical variable distribution. This process
+generates relative measures of how intense the event is compared to what
+is normal in that location (e.g., 90\textsuperscript{o}F
+~32.2\textsuperscript{o}C is in the 99\textsuperscript{th} percentile of
 year-round temperatures in the location).
 
-The values of these relative heat wave characteristics will depend on
-what time period is used to determine the location's typical temperature
-distribution. For example, the same heat wave would likely have much
-higher values for these relative characteristics if the temperature
-during the heat wave is compared to typical temperatures in the present
-day for that location versus if the temperature is compared to typical
-temperatures much later in the 21\textsuperscript{st} century for the
-same location.
-
-The default in the \texttt{gen\_hw\_set} function is to use temperatures
-for the period 2070 to 2079 for these reference temperatures to
-calculate all relative heat wave characteristics. However, the user can
-change this specification using the \texttt{referenceBoundaries} option
-of \texttt{gen\_hw\_set}. This functionality can be useful in exploring
-the role of adaptation in future heat waves. For example, to use
-temperature projections for 1990 to 1999 when calculating relative
-characteristics of heat waves, to explore the assumption that cities
-remain adapted to their present-day climate, rather than changing in
-adaptation as climate change increases temperatures, a user could run:
-
-\begin{Schunk}
-\begin{Sinput}
-gen_hw_set(out = "example_results",
-           dataFolder = projection_dir_location ,
-           dataDirectories = list("historical" = c(1990, 1999),
-                                        "rcp85" = c(2060, 2079)),
-           citycsv = city_file_location,
-           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
-           tasFilenames = "tas_NorthAmerica_12mo.csv",
-           timeFilenames = "time_NorthAmerica_12mo.csv",
-           referenceBoundaries = c(1990, 1999))
-\end{Sinput}
-\end{Schunk}
-
-For these date range specifications, there are some restrictions on
-which year ranges can be selected. The starting year cannot be earlier
-than the first year in the first subdirectory, the ending year cannot be
-later than the last year of the second subdirectory, and the custom date
-boundaries cannot span the two subdirectories.
+These relative event characteristics will vary depending on whether you
+calculate them based on a location's present-day variable distribution
+or on the location's variable distribution in the future, since the
+distributions of many relevant variables (e.g., temperature,
+precipitation) are expected to change in many locations with climate
+change. The package therefore allows the user to specify date ranges of
+the temperature distributions to be used in calculating these relative
+temperature metrics in each location, which can be done using the
+\code{referenceBoundaries} option of \code{gen\_hw\_set}.
 
 \subsection{Mapping grid points}\label{mapping-grid-points}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width = 0.9\textwidth]{ExampleLeaflet}
+\end{center}
+\caption{Snapshot of an example interactive map created using the \code{map\_grid\_leaflet} showing the locations of study cities and their matching climate model grid points for the BCC climate model example data included with \pkg{futureheatwaves}.
+The lines on the map connect each climate model grid point to the study location(s) for which that grid point was used.
+The interactive maps include pop-ups with city identiers; one is shown open in this snapshot as an example.
+From this map, you can see that the climate model grid point closest to New York City for this climate model is over the Atlantic Ocean.}
+\label{fig:gridmap}
+\end{figure}
 
@@
 
-\begin{figure}
-\begin{center}
-\includegraphics[width = 0.9\textwidth]{ExampleLeaflet}
-\end{center}
-\caption{Snapshot of an example interactive map created using the \code{map\_grid\_leaflet} showing the locations of study cities and their matching climate model grid points for the BCC climate model example data included with \pkg{futureheatwaves}.
-The lines on the map connect each climate model grid point to the study location(s) for which that grid point was used.
-The interactive maps include pop-ups with city identiers; one is shown open in this snapshot as an example.
-From this map, you can see that the climate model grid point closest to New York City for this climate model is over the Atlantic Ocean.}
-\label{fig:gridmap}
-\end{figure}
-
 \subsection{Extensions}\label{extensions}
 
-The functionality of this package can be easily expanded by loops. For
-example, to explore the role of the heat wave definition on projections,
-the user could create a loop to run \texttt{gen\_hw\_set} and
-\texttt{apply\_all\_models} to the same directory of climate projections
-but with a variety of different functions used to identify the heat
-waves in the projections.
-
-Also, while this package was created to be used for research on heat
-waves in climate change projections, with some modifications it can be
-used more broadly. For example, there are other episodes like wildfires
-and air pollution where it may be interesting to identify extended
-periods of high exposures in projection time series, and this package
-could be applied to explore these exposures. The directory of projection
-data would need to be set up in the same structure as for exploring heat
-waves, and the \texttt{input\_metric} should be set as
-\texttt{fahrenheit}, to pass the exposure values through to the
-characterized data sets without performing a conversion. A user could
-also use this package to explore events that must be lower than some
-minimum threshold (e.g., cold waves), but it would take some extra
-coding and conversions, since the functions in this package are written
-to identify periods above a threshold (for example, the user could
-multiple all projected temperatures by -1, and then the coldest
-temperatures would register as being the highest).
-
-\begin{Schunk}
-\begin{Sinput}
-## Cold wave example
-gen_hw_set(out = "example_results",
-           dataFolder = projection_dir_location ,
-           dataDirectories = list("historical" = c(1990, 1999),
-                                  "rcp85" = c(2060, 2079)),
-           citycsv = city_file_location,
-           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
-           tasFilenames = "tas_NorthAmerica_12mo.csv",
-           timeFilenames = "time_NorthAmerica_12mo.csv", 
-           probThreshold = 0.10, 
-           above_threshold = FALSE, 
-           absolute_thresholds = c(266, 263, 260, 258), 
-           seasonal_months = c(12, 1, 2))
-\end{Sinput}
-\end{Schunk}
-
-\begin{quote}
-Extreme statistics and spell-lengths: ``A `frost day' is defined as one
-during which the minimum temperature falls below freezing point (0
-degC). This is described as a climatological statistic, in which the
-minimum temperature is first calculated within each day, and then the
-number of days or spell lengths meeting the specified condition are
-evaluated.'' (from one example in
-\url{http://cfconventions.org/Data/cf-conventions/cf-conventions-1.6/build/cf-conventions.html\#calendar})
-\end{quote}
-
-Users doing these kinds of extensions will need to pay attention to a
-few points. First, some of the event characteristics (first in the
-calendar year, average of May--September temperatures, days above
-90\textsuperscript{o}F) might not be meaningful for studies of other
-types of events. Further, because event periods are usually defined as a
-string of multiple days exceeding some threshold, the functions in this
-package may miss the first and last event of the time period. For
-example, if the first day of the time series is the last day of an
-event, this function would not identify that event because it lacks data
-from the earlier days that allow this day to meet the event definition.
-This issue would lead to, at most, missing two events out of each
-projection, but should be considered if studying events that might occur
-near the start and end of projection data. If there is adequate interest
-from researchers, in the future we may adapt the package to make these
-secondary applications part of the package.
+While this package was created to be used for research on extreme events
+in climate change projections, it can be used more broadly. For example,
+there are other episodes like wildfires and air pollution where it may
+be interesting to identify extended periods of high exposures in
+projection time series, and this package could be applied to gridded air
+pollution model output to explore these exposures.
 
@@
 Foundation (1331399), and the Colorado State University Vice President
-for Research.
+for Research. Thank you to early testers of the package: Julia
+Bromberek, Wande Benka-Coker, Josh Ferreri, Ryan Gan, Molly Gutilla,
+Mike Lyons, Casey Quinn, Rachel Severson, and Meilin Yan.
 

### 2016-12-29 paper_drafts/RJournalRevisions/anderson-eason-barnes.Rmd
Work on R paper draft

@@
 
-\begin{figure}
+\begin{widefigure}
 \begin{center}
-\includegraphics[width = 0.9\textwidth]{netcdf_structure_example}
+\includegraphics[width = \textwidth]{netcdf_figure}
 \end{center}
@@
 \label{fig:netcdfexample}
-\end{figure}
+\end{widefigure}
 
@@
 In this case, the \code{varid} parameter should be set using the appropriate CMIP5 variable name (e.g., "tas" for near-surface air temperature); these variable names can be found in the CMIP Requested Output tables \citep{taylor2010cmip5}.
-In practice, you can use the dimensional time and location data to identify the location of the variable data you need in the netCDF array and use indexing to read only that data into memory, without needing to read in the full file \citep{michna2013rnetcdf}.
+You can use the dimensional time and location data to identify the location of the variable data you need in the netCDF array and use indexing to read only that data into memory, without needing to read in the full file (for example, with the \code{nc.get.var.subset.by.axes} function in \pkg{ncdf4.helpers}).
 Once the user is done reading in data from the file, the connection can be closed (e.g., with the \code{nc\_close} function from \pkg{ncdf4}).
@@
 The following code gives an example of using a CMIP5 file and some of the R packages and functions discussed.
-Results include a map of near-surface air temperatures from a single climate model on a specific date (July 1, 2075) for China (Figure [x]) and a time series of daily near-surface air temperature simulations for the climate model grid point closest to Beijing, China (Figure [x]).
-To try this example, you will need to download the file ...
-from ...
-and save it in the "tmp" subdirectory of your home directory.
+Results include a map of near-surface air temperatures from a single climate model on a specific day in July 2075 (Figure \ref{fig:worldmap}) and a time series of daily near-surface air temperature simulations for the climate model grid point closest to Beijing, China (Figure \ref{fig:timeseries}).
+To try this example, you will need to download the file "tas_day_GFDL-ESM2G_rcp85_r1i1p1_20710101-20751231.nc" from https://pcmdi.llnl.gov/search/cmip5/ and save it in the "tmp" subdirectory of your home directory.
 
-```{r}
+```{r eval = FALSE, echo = FALSE}
+# To try this example,you will need to download the file 
+# “tas_day_GFDL-ESM2G_rcp85_r1i1p1_20710101-20751231.nc”
+# from https://pcmdi.llnl.gov/search/cmip5/and save it in the “tmp” 
+# subdirectory of your home directory
 
+# Load required packages
+# Packages for working with CMIP5 files
+library(ncdf4)
+library(ncdf4.helpers)
+library(PCICt)
+# Some general purpose packages
+library(readr)
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(ggmap)
+library(viridis)
+library(weathermetrics)
+
+# Open connection to CMIP5 netCDF file
+climate_filepath <- paste0("~/tmp/tas_day_GFDL-ESM2G_rcp85",
+                           "_r1i1p1_20710101-20751231.nc")
+climate_output <- nc_open(climate_filepath)
+
+# Load dimension variables (longitude, latitude, time) and climate output
+# variable ("tas", near-surface air temperature)
+lon <- ncvar_get(climate_output, varid = "lon")
+lat <- ncvar_get(climate_output, varid = "lat")
+tas_time <- nc.get.time.series(climate_output, v = "tas",
+                               time.dim.name = "time")
+
+# Subset to a single date (July 1, 2075) to create map
+pdf("worldmapexample.pdf", height = 3, width = 6)
+time_index <- which(format(tas_time, "%Y-%m-%d") == "2075-07-15")
+tas <- nc.get.var.subset.by.axes(climate_output, "tas",
+                                 axis.indices = list(T = time_index))
+expand.grid(lon, lat) %>%
+        rename(lon = Var1, lat = Var2) %>%
+        mutate(lon = ifelse(lon > 180, -(360 - lon), lon),
+               tas = as.vector(tas)) %>% 
+        mutate(tas = convert_temperature(tas, "k", "c")) %>%
+        ggplot() + 
+        geom_point(aes(x = lon, y = lat, color = tas),
+                   size = 0.8) + 
+        borders("world", colour="black", fill=NA) + 
+        scale_color_viridis(name = "Temperature (C)") + 
+        theme_void() + 
+        coord_quickmap() + 
+        ggtitle("Modeled temperature on a day in July 2075",
+                subtitle = "GFDL-ESM2G model, RCP8.5 experiment, r1i1p1 ensemble member") 
+dev.off()
+
+# Subset the time series for the grid point closest to Beijing, China, to
+# create figure
+pdf("timeseriesexample.pdf", height = 3, width = 6)
+lon_index <- which.min(abs(lon - 116.4))
+lat_index <- which.min(abs(lat - 39.9))
+tas <- nc.get.var.subset.by.axes(climate_output, "tas",
+                                 axis.indices = list(X = lon_index,
+                                                     Y = lat_index))
+data_frame(time = tas_time, 
+           tas = as.vector(tas)) %>%
+        mutate(time = as.Date(format(time, "%Y-%m-%d"))) %>%
+        ggplot(aes(x = time, y = tas)) + 
+        geom_line() + 
+        xlab("Date") + ylab("Temperature (K)") + 
+        ggtitle("Daily modeled near-surface air temprature, 2071-2075",
+                subtitle = "At model grid point nearest Beijing, China") + 
+        theme_classic()
+dev.off()
+
+# Close file connection to netCDF file
+nc_close(climate_output)
 ```
+
+\begin{figure}
+\begin{center}
+\includegraphics[width = \textwidth]{worldmapexample}
+\end{center}
+\caption{Example of mapping near-surface air temperature data worldwide for a single day of climate model output data.
+This map uses data from the Geophysical Fluid Dynamics Laboratory's Earth System Model 2G, r1i1p1 ensemble member, on a single day in the summer of 2075.}
+\label{fig:worldmap}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width = 0.8\textwidth]{timeseriesexample}
+\end{center}
+\caption{Example of plotting a time series of temperature simulations between 2071 and 2075 from CMIP5 daily climate model output data for the model grid cell point closest to Beijing, China.
+This plot uses data from the Geophysical Fluid Dynamics Laboratory's Earth System Model 2G, r1i1p1 ensemble member.}
+\label{fig:timeseries}
+\end{figure}
 

### 2016-12-29 paper_drafts/RJournalRevisions/anderson-eason-barnes.tex
Work on R paper draft

@@
 near-surface air temperature); these variable names can be found in the
-CMIP Requested Output tables \citep{taylor2010cmip5}. In practice, you
-can use the dimensional time and location data to identify the location
-of the variable data you need in the netCDF array and use indexing to
-read only that data into memory, without needing to read in the full
-file \citep{michna2013rnetcdf}. Once the user is done reading in data
-from the file, the connection can be closed (e.g., with the
-\code{nc\_close} function from \pkg{ncdf4}).
+CMIP Requested Output tables \citep{taylor2010cmip5}. You can use the
+dimensional time and location data to identify the location of the
+variable data you need in the netCDF array and use indexing to read only
+that data into memory, without needing to read in the full file (for
+example, with the \code{nc.get.var.subset.by.axes} function in
+\pkg{ncdf4.helpers}). Once the user is done reading in data from the
+file, the connection can be closed (e.g., with the \code{nc\_close}
+function from \pkg{ncdf4}).
 
@@
 near-surface air temperatures from a single climate model on a specific
-date (July 1, 2075) for China (Figure {[}x{]}) and a time series of
-daily near-surface air temperature simulations for the climate model
-grid point closest to Beijing, China (Figure {[}x{]}). To try this
-example, you will need to download the file \ldots{} from \ldots{} and
-save it in the ``tmp'' subdirectory of your home directory.
+day in July 2075 (Figure \ref{fig:worldmap}) and a time series of daily
+near-surface air temperature simulations for the climate model grid
+point closest to Beijing, China (Figure \ref{fig:timeseries}). To try
+this example, you will need to download the file
+``tas\_day\_GFDL-ESM2G\_rcp85\_r1i1p1\_20710101-20751231.nc'' from
+\url{https://pcmdi.llnl.gov/search/cmip5/} and save it in the ``tmp''
+subdirectory of your home directory.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width = \textwidth]{worldmapexample}
+\end{center}
+\caption{Example of mapping near-surface air temperature data worldwide for a single day of climate model output data.
+This map uses data from the Geophysical Fluid Dynamics Laboratory's Earth System Model 2G, r1i1p1 ensemble member, on a single day in the summer of 2075.}
+\label{fig:worldmap}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width = 0.8\textwidth]{timeseriesexample}
+\end{center}
+\caption{Example of plotting a time series of temperature simulations between 2071 and 2075 from CMIP5 daily climate model output data for the model grid cell point closest to Beijing, China.
+This plot uses data from the Geophysical Fluid Dynamics Laboratory's Earth System Model 2G, r1i1p1 ensemble member.}
+\label{fig:timeseries}
+\end{figure}
 

### 2016-12-29 paper_drafts/RJournalRevisions/anderson-eason-barnes.Rmd
Update paper draft

@@
 
-The following code gives an example of using a CMIP5 file and some of the R packages and functions discussed.
-Results include a map of near-surface air temperatures from a single climate model on a specific day in July 2075 (Figure \ref{fig:worldmap}) and a time series of daily near-surface air temperature simulations for the climate model grid point closest to Beijing, China (Figure \ref{fig:timeseries}).
-To try this example, you will need to download the file "tas_day_GFDL-ESM2G_rcp85_r1i1p1_20710101-20751231.nc" from https://pcmdi.llnl.gov/search/cmip5/ and save it in the "tmp" subdirectory of your home directory.
+The functions and packages just described can be used to work with CMIP5 netCDF files in R to do things like map near-surface air temperatures from a single climate model on a specific day in July 2075 (Figure \ref{fig:worldmap}) and pull a time series of daily near-surface air temperature simulations for the climate model grid point closest to Beijing, China (Figure \ref{fig:timeseries}).
+The "starting_from_netcdf" vignette that comes with the "futureheatwaves" package provides all code required to create these figures, as well as more details and code examples on working with CMIP5 netCDF files in R.
 

### 2016-12-30 paper_drafts/RJournalRevisions/anderson-eason-barnes.Rmd
Add Libby's direct revisions

@@
 abstract: >
-  Research on climate change impacts can require extensive processing, to incorporate output from ensembles of different climate models and different simulations of each model.
+  Research on climate change impacts can require extensive processing to incorporate output from ensembles of different climate models and different simulations of each model.
 This processing can be particularly extensive when identifying and characterizing extreme events like heat waves and frost day spells, as these must be processed from model output with daily time steps and may use identification threholds based on local distributions (e.g., 98\textsuperscript{th} percentile temperature for that location).
@@
 Further, we present the \pkg{futureheatwaves} package, which we developed to ease the process of identifying, characterizing, and exploring multi-day extreme events in climate model output.
-This package takes as input a directory of projection files from one or many climate models, identifies all of a specific type of extreme event (e.g., heat waves) for multiple cities using a user-specified event definition, and writes an output directory of files with these extreme events and their characteristics (e.g., length, absolute and relative temperatures).
+This package takes as input a directory of projection files from one or many climate models, identifies all occurrences of a specific type of extreme event (e.g., heat waves) for multiple cities using a user-specified event definition, and writes an output directory of files with these extreme events and their characteristics (e.g., length, absolute and relative temperatures).
 The \pkg{futureheatwaves} package then allows the user to create custom functions and apply them across all event files; this functionality can be used either to explore extreme event patterns (e.g., average length and intensity across different climate models) or to apply epidemiological estimates to project health impacts under different scenarios.
@@
 Further, it is often of interest to create summaries of multiple characteristics of these extreme events.
-One interest, for example, in interpreting climate change scenarios output is whether frequency of heat waves or warm spells will change, as well as how certain characteristics of these extreme events (e.g., length, intensity) will change under different scenarios \citep{IPCCch1}.
+For example, one may be interested in determining whether the frequency or characteristics (e.g., length, intensity) of heat waves or warm spells will change under certain climate change scenarios \citep{IPCCch1}.
 
@@
 
-After this overview, we present the \pkg{futureheatwaves} package.
-which we created to aid in identifying and characterizing any type of multi-day extreme event from daily climate model output (Table \ref{tab:goals}).
+After this overview, we present the \pkg{futureheatwaves} package which we created to aid in identifying and characterizing any type of multi-day extreme event from daily climate model output (Table \ref{tab:goals}).
 This package can be used to identify and characterize a variety of multi-day extreme events, but provides some functionality particularly useful in identifying and characterizing heat waves specifically.
-Quantifying the impacts of heat waves on human health suffer from additional sources of uncertainty beyond those inherent in projections of regional changes in surface temperature.
+Quantification of the impacts of heat waves on human health suffer from additional sources of uncertainty beyond those inherent in projections of regional changes in surface temperature.
 These include: (1) uncertainty in the ability of communities to adapt to changing temperatures (the adaptation scenario) and (2) uncertainty in the definition of a heat wave itself.
@@
 Further, other types of extreme events will be defined differently than heat waves (for example, frost day spells may be defined as one or more days with temperature at or below 32\textsuperscript{o}F / 0\textsuperscript{o}C).
-This package therefore allows the user to create and use a custom the extreme event definition used to identify events in the climate model output.
+This package therefore allows the user to create and use a custom extreme event definition to identify events in the climate model output.
 Finally, this package allows the user to explore the extreme events identified in the climate model output with a function that can take a user-defined R function and apply it across all extreme event files generated for the separate ensemble members.
@@
 
-For climate impact studies, a top source for climate model output files is the Coupled Model Intercomparison Project, which is currently in its fifth phase (CMIP5).
+For climate impact studies, a main source for climate model output is the Coupled Model Intercomparison Project, which is currently in its fifth phase (CMIP5).
 Over 20 climate modeling groups have created one or more climate models which, for this project, are run using standardized scenarios \citep{taylor2012overview}.
 The resulting output is uniform across modeling groups and has a consistent structure, which allows comparison of simulations from different models \citep{IPCCch9}.
-For CMIP5, each group created simulations under several experiments, with experiments varying in terms of the radiative forcing.
+For CMIP5, each group ran simulations under several experiments, with experiments varying in terms of the radiative forcing.
 This radiative forcing depends on time-varying model inputs (greenhouse gas emissions or concentrations, land use changes, etc.), which are specified for each experiment \citep{taylor2012overview, IPCCch9}.
@@
 Some modeling groups created ensembles of output for a specific model and experiment, in which they ran the experiment multiple times with the model with very small changes to the initial conditions, resulting in multiple ensemble members for a single climate model and experiment.
-CMIP5 climate model output is created at a number of different time steps (e.g., daily, monthly, seasonal, yearly) \citep{taylor2010cmip5}, and some variables are reported at multiple depths in the ocean or atmosphere (e.g., ocean temperature).
-Here, we will focus on data with a daily time step for variables reported at a single depth (e.g., near-surface air temperature).
+CMIP5 climate model output is archived at a number of different time steps (e.g., daily, monthly, seasonal, yearly) \citep{taylor2010cmip5}, and some variables are reported at multiple levels in the ocean or atmosphere (e.g., ocean temperature).
+Here, we will focus on data with a daily time step for variables reported at a single level (e.g., near-surface air temperature).
 
@@
 These climate model output files can be as large as several gigabytes [?].
-For climate model output at a single depth (e.g., near-surface air temperature), the data is a 3-dimensional array, with dimensions representing time and two metrics of location (e.g., latitude and longitude) (Figure \ref{fig:netcdfexample}).
+For climate model output at a single level (e.g., near-surface air temperature), the data is a 3-dimensional array, with dimensions representing time and two coordinates of location (e.g., latitude and longitude) (Figure \ref{fig:netcdfexample}).
 Global climate models generate output at regularly-spaced time steps, typically at regularly-spaced grid points around the world.
@@
 For daily output files, the time unit is in days since a specified origin date-time (e.g., days since 1850-01-01 00:00:00) \citep{taylor2010cmip5}.
-All CMIP5 output files are requred to include certain metadata  \citep{taylor2010cmip5}.
+All CMIP5 output files are required to include certain metadata  \citep{taylor2010cmip5}.
 This required metadata includes the experiment, forcing agents input to the model to create the simulation, time step, institution and institutional contact information, climate model, and modeling realm \citep{taylor2010cmip5}.
@@
 More recent packages, including \pkg{ncdf4} \citep{ncdf4} and \pkg{RNetCDF} \citep{michna2013rnetcdf, RNetCDF}, work with both version 4 and netCDF's older version 3.
-Climate model output data for CMIP5 is required to conform with the earlier version (version 3) \citep{taylor2010cmip5} and so should work with any of these packages, although it is safer to write code using functions that can be used with version for in case future phases of CMIP do not require files to conform with netCDF version 3.
+Climate model output data for CMIP5 is required to conform with the earlier version (version 3) \citep{taylor2010cmip5} and so should work with any of these packages, although it is safer to write code using functions that can be used with either version in case future phases of CMIP do not require files to conform with netCDF version 3.
 
@@
 Since the late 1500s, Western dates have been set using the Gregorian calendar, which has 365.2425-day years.
-Some climate models, however, output to different calendars, including the Julian calendar (365.25-day years), a calendar where there are no leap years (365-day years), a calendar where every year is a leap year (366-day years), and a calendar of twelve 30-day months (360-day years) \citep{cfconventions}.
+Some climate models, however, are run using different calendars, including the Julian calendar (365.25-day years), a calendar where there are no leap years (365-day years), a calendar where every year is a leap year (366-day years), and a calendar of twelve 30-day months (360-day years) \citep{cfconventions}.
 With these non-Gregorian calendars, R's base functions for converting a vector to a Date class based on the number of days since an origin date (\code{as.Date}, \code{as.POSIXct}) do not return the desired values.
@@
 
-The functions and packages just described can be used to work with CMIP5 netCDF files in R to do things like map near-surface air temperatures from a single climate model on a specific day in July 2075 (Figure \ref{fig:worldmap}) and pull a time series of daily near-surface air temperature simulations for the climate model grid point closest to Beijing, China (Figure \ref{fig:timeseries}).
+The functions and packages described can be used to work with CMIP5 netCDF files in R to do things like map near-surface air temperatures from a single climate model on a specific day in July 2075 (Figure \ref{fig:worldmap}) and pull a time series of daily near-surface air temperature simulations for the climate model grid point closest to Beijing, China (Figure \ref{fig:timeseries}).
 The "starting_from_netcdf" vignette that comes with the "futureheatwaves" package provides all code required to create these figures, as well as more details and code examples on working with CMIP5 netCDF files in R.
@@
 
-We have included data files in the package to serve as example files, so users can try this package before applying it to their own directory of climate projection files.
+We have included data files in the package to serve as example files so that users can try this package before applying it to their own directory of climate projection files.
 This example data comes from two climate models that are a part of CMIP5: (1) the model of the Beijing Climate Center, China Meteorological Administration (BCC) \citep{xin2013introduction} and (2) the National Center for Atmospheric Research's (NCAR's) Community Climate System Model, version 4 (CCSM4) \citep{gent2011community}.
@@
 
-\noindent While this example impact analysis is simplistic, it provides an example of how the \code{apply\_all\_models} function can be used in conjunction with epidemiologic results to estimate the potential health impacts of extreme events with climate change.
+\noindent While the above impact analysis is simplistic, it provides an example of how the \code{apply\_all\_models} function can be used in conjunction with epidemiologic results to estimate the potential health impacts of extreme events with climate change.
 
@@
 Depending on the climate model output data used and the research question, different projection periods may be of interest.
-For example, one research may want to explore extreme events in the 2050--2075 period, while another may be interested in 2090--2100.
+For example, one researcher may wish to explore extreme events in the 2050--2075 period, while another may be interested in 2090--2100.
 The projection date range can be changed by specifying starting and ending years in the \code{projectionBoundaries} argument of the \code{gen\_hw\_set} function.

### 2016-12-30 paper_drafts/RJournalRevisions/anderson-eason-barnes.tex
Add Libby's direct revisions

@@
 \abstract{%
-Research on climate change impacts can require extensive processing, to
+Research on climate change impacts can require extensive processing to
 incorporate output from ensembles of different climate models and
@@
 climate model output. This package takes as input a directory of
-projection files from one or many climate models, identifies all of a
-specific type of extreme event (e.g., heat waves) for multiple cities
-using a user-specified event definition, and writes an output directory
-of files with these extreme events and their characteristics (e.g.,
-length, absolute and relative temperatures). The \pkg{futureheatwaves}
-package then allows the user to create custom functions and apply them
-across all event files; this functionality can be used either to explore
-extreme event patterns (e.g., average length and intensity across
-different climate models) or to apply epidemiological estimates to
-project health impacts under different scenarios.
+projection files from one or many climate models, identifies all
+occurrences of a specific type of extreme event (e.g., heat waves) for
+multiple cities using a user-specified event definition, and writes an
+output directory of files with these extreme events and their
+characteristics (e.g., length, absolute and relative temperatures). The
+\pkg{futureheatwaves} package then allows the user to create custom
+functions and apply them across all event files; this functionality can
+be used either to explore extreme event patterns (e.g., average length
+and intensity across different climate models) or to apply
+epidemiological estimates to project health impacts under different
+scenarios.
 }
@@
 climate model output. Further, it is often of interest to create
-summaries of multiple characteristics of these extreme events. One
-interest, for example, in interpreting climate change scenarios output
-is whether frequency of heat waves or warm spells will change, as well
-as how certain characteristics of these extreme events (e.g., length,
-intensity) will change under different scenarios \citep{IPCCch1}.
+summaries of multiple characteristics of these extreme events. For
+example, one may be interested in determining whether the frequency or
+characteristics (e.g., length, intensity) of heat waves or warm spells
+will change under certain climate change scenarios \citep{IPCCch1}.
 
@@
 
-After this overview, we present the \pkg{futureheatwaves} package. which
+After this overview, we present the \pkg{futureheatwaves} package which
 we created to aid in identifying and characterizing any type of
@@
 particularly useful in identifying and characterizing heat waves
-specifically. Quantifying the impacts of heat waves on human health
-suffer from additional sources of uncertainty beyond those inherent in
-projections of regional changes in surface temperature. These include:
-(1) uncertainty in the ability of communities to adapt to changing
-temperatures (the adaptation scenario) and (2) uncertainty in the
-definition of a heat wave itself. Thus, identifying and characterizing
-the impacts of future heat waves in state-of-the-art climate models
-requires analyzing hundreds of projections based on combinations of
-anthropogenic activity, climate model and ensemble member, adaptation
-scenario, and heat wave definition. Such an analysis is non-trivial. For
-example, to estimate the possible impacts of heat waves using 32
-ensemble members of CMIP5, two forcing scenarios, two different heat
-wave definitions, and five progressive assumptions about adaptation to
-heat, would require that one identify and characterize heat waves in 640
-separate temperature time series for each region of interest.
+specifically. Quantification of the impacts of heat waves on human
+health suffer from additional sources of uncertainty beyond those
+inherent in projections of regional changes in surface temperature.
+These include: (1) uncertainty in the ability of communities to adapt to
+changing temperatures (the adaptation scenario) and (2) uncertainty in
+the definition of a heat wave itself. Thus, identifying and
+characterizing the impacts of future heat waves in state-of-the-art
+climate models requires analyzing hundreds of projections based on
+combinations of anthropogenic activity, climate model and ensemble
+member, adaptation scenario, and heat wave definition. Such an analysis
+is non-trivial. For example, to estimate the possible impacts of heat
+waves using 32 ensemble members of CMIP5, two forcing scenarios, two
+different heat wave definitions, and five progressive assumptions about
+adaptation to heat, would require that one identify and characterize
+heat waves in 640 separate temperature time series for each region of
+interest.
 
@@
 32\textsuperscript{o}F / 0\textsuperscript{o}C). This package therefore
-allows the user to create and use a custom the extreme event definition
-used to identify events in the climate model output. Finally, this
-package allows the user to explore the extreme events identified in the
-climate model output with a function that can take a user-defined R
-function and apply it across all extreme event files generated for the
-separate ensemble members.
+allows the user to create and use a custom extreme event definition to
+identify events in the climate model output. Finally, this package
+allows the user to explore the extreme events identified in the climate
+model output with a function that can take a user-defined R function and
+apply it across all extreme event files generated for the separate
+ensemble members.
 
@@
 
-For climate impact studies, a top source for climate model output files
-is the Coupled Model Intercomparison Project, which is currently in its
+For climate impact studies, a main source for climate model output is
+the Coupled Model Intercomparison Project, which is currently in its
 fifth phase (CMIP5). Over 20 climate modeling groups have created one or
@@
 comparison of simulations from different models \citep{IPCCch9}. For
-CMIP5, each group created simulations under several experiments, with
+CMIP5, each group ran simulations under several experiments, with
 experiments varying in terms of the radiative forcing. This radiative
@@
 multiple ensemble members for a single climate model and experiment.
-CMIP5 climate model output is created at a number of different time
+CMIP5 climate model output is archived at a number of different time
 steps (e.g., daily, monthly, seasonal, yearly) \citep{taylor2010cmip5},
-and some variables are reported at multiple depths in the ocean or
+and some variables are reported at multiple levels in the ocean or
 atmosphere (e.g., ocean temperature). Here, we will focus on data with a
-daily time step for variables reported at a single depth (e.g.,
+daily time step for variables reported at a single level (e.g.,
 near-surface air temperature).
@@
 array. These climate model output files can be as large as several
-gigabytes {[}?{]}. For climate model output at a single depth (e.g.,
+gigabytes {[}?{]}. For climate model output at a single level (e.g.,
 near-surface air temperature), the data is a 3-dimensional array, with
-dimensions representing time and two metrics of location (e.g., latitude
-and longitude) (Figure \ref{fig:netcdfexample}). Global climate models
-generate output at regularly-spaced time steps, typically at
+dimensions representing time and two coordinates of location (e.g.,
+latitude and longitude) (Figure \ref{fig:netcdfexample}). Global climate
+models generate output at regularly-spaced time steps, typically at
 regularly-spaced grid points around the world. The latitude and
@@
 specified origin date-time (e.g., days since 1850-01-01 00:00:00)
-\citep{taylor2010cmip5}. All CMIP5 output files are requred to include
+\citep{taylor2010cmip5}. All CMIP5 output files are required to include
 certain metadata \citep{taylor2010cmip5}. This required metadata
@@
 although it is safer to write code using functions that can be used with
-version for in case future phases of CMIP do not require files to
+either version in case future phases of CMIP do not require files to
 conform with netCDF version 3.
@@
 calendar, which has 365.2425-day years. Some climate models, however,
-output to different calendars, including the Julian calendar (365.25-day
-years), a calendar where there are no leap years (365-day years), a
-calendar where every year is a leap year (366-day years), and a calendar
-of twelve 30-day months (360-day years) \citep{cfconventions}. With
-these non-Gregorian calendars, R's base functions for converting a
-vector to a Date class based on the number of days since an origin date
-(\code{as.Date}, \code{as.POSIXct}) do not return the desired values.
-The \pkg{PCICt} \citep{PCICt} and \pkg{ncdf4.helpers}
-\citep{ncdf4.helpers} packages provide further functionality with netCDF
-files that can be useful when working with climate model output data and
-provide particular help in working with different calendars. CMIP5
-netCDF files include information on the calendar used, which the
-\code{nc.get.time.series} function in \pkg{ncdf4.helpers} use to convert
-the ``time'' variable in the file to an object of the PCICt class, which
-provides Date-like functionality for 360- and 365-day calendars
-\citep{PCICt}. As a note, while these functions will help with handling
-most CMIP5 files, the CMIP5 standards allows use of other calendars
-which may not be succesfully handled by these functions, so it is
-important to assess whether the time variable range in the PCICt object
-correctly matches the expected date ranges for a file as you process
-CMIP5 data in R.
+are run using different calendars, including the Julian calendar
+(365.25-day years), a calendar where there are no leap years (365-day
+years), a calendar where every year is a leap year (366-day years), and
+a calendar of twelve 30-day months (360-day years)
+\citep{cfconventions}. With these non-Gregorian calendars, R's base
+functions for converting a vector to a Date class based on the number of
+days since an origin date (\code{as.Date}, \code{as.POSIXct}) do not
+return the desired values. The \pkg{PCICt} \citep{PCICt} and
+\pkg{ncdf4.helpers} \citep{ncdf4.helpers} packages provide further
+functionality with netCDF files that can be useful when working with
+climate model output data and provide particular help in working with
+different calendars. CMIP5 netCDF files include information on the
+calendar used, which the \code{nc.get.time.series} function in
+\pkg{ncdf4.helpers} use to convert the ``time'' variable in the file to
+an object of the PCICt class, which provides Date-like functionality for
+360- and 365-day calendars \citep{PCICt}. As a note, while these
+functions will help with handling most CMIP5 files, the CMIP5 standards
+allows use of other calendars which may not be succesfully handled by
+these functions, so it is important to assess whether the time variable
+range in the PCICt object correctly matches the expected date ranges for
+a file as you process CMIP5 data in R.
 
@@
 
-The following code gives an example of using a CMIP5 file and some of
-the R packages and functions discussed. Results include a map of
-near-surface air temperatures from a single climate model on a specific
-day in July 2075 (Figure \ref{fig:worldmap}) and a time series of daily
-near-surface air temperature simulations for the climate model grid
-point closest to Beijing, China (Figure \ref{fig:timeseries}). To try
-this example, you will need to download the file
-``tas\_day\_GFDL-ESM2G\_rcp85\_r1i1p1\_20710101-20751231.nc'' from
-\url{https://pcmdi.llnl.gov/search/cmip5/} and save it in the ``tmp''
-subdirectory of your home directory.
+The functions and packages described can be used to work with CMIP5
+netCDF files in R to do things like map near-surface air temperatures
+from a single climate model on a specific day in July 2075 (Figure
+\ref{fig:worldmap}) and pull a time series of daily near-surface air
+temperature simulations for the climate model grid point closest to
+Beijing, China (Figure \ref{fig:timeseries}). The
+``starting\_from\_netcdf'' vignette that comes with the
+``futureheatwaves'' package provides all code required to create these
+figures, as well as more details and code examples on working with CMIP5
+netCDF files in R.
 
@@
 
-We have included data files in the package to serve as example files, so
-users can try this package before applying it to their own directory of
-climate projection files. This example data comes from two climate
-models that are a part of CMIP5: (1) the model of the Beijing Climate
-Center, China Meteorological Administration (BCC)
+We have included data files in the package to serve as example files so
+that users can try this package before applying it to their own
+directory of climate projection files. This example data comes from two
+climate models that are a part of CMIP5: (1) the model of the Beijing
+Climate Center, China Meteorological Administration (BCC)
 \citep{xin2013introduction} and (2) the National Center for Atmospheric
@@
 
-\noindent While this example impact analysis is simplistic, it provides
-an example of how the \code{apply\_all\_models} function can be used in
+\noindent While the above impact analysis is simplistic, it provides an
+example of how the \code{apply\_all\_models} function can be used in
 conjunction with epidemiologic results to estimate the potential health
@@
 question, different projection periods may be of interest. For example,
-one research may want to explore extreme events in the 2050--2075
+one researcher may wish to explore extreme events in the 2050--2075
 period, while another may be interested in 2090--2100. The projection

### 2016-12-30 README.Rmd
Add Travis status to README
Update vignette

@@
 <!-- README.md is generated from README.Rmd. Please edit that file -->
+
+[![Build Status](https://travis-ci.org/geanders/futureheatwaves.svg?branch=master)](https://travis-ci.org/geanders/futureheatwaves)
 
@@
 
-The `futureheatwaves` package takes a directory of climate projection files and, for each, identifies and characterizes all heat waves.
-The definition used to identify heat waves can be customized.
-Characterizations include several metrics of heat wave length, intensity, and timing in the year.
-Heat waves can be explored by applying custom functions across all generated heat wave files.
+The `futureheatwaves` package takes a directory of climate projection files and, for each, identifies and characterizes all of a certain type of multi-day extreme events (e.g., heat waves).
+The definition used to identify extreme events can be customized.
+Characterizations include several metrics of event length, intensity, and timing in the year.
+The identified extreme events can be explored by applying custom functions across all generated heat wave files.
 

### 2016-12-31 paper_drafts/RJournalRevisions/anderson-eason-barnes.tex
Update paper draft

--- 2016-12-29 paper_drafts/RJournalRevisions/anderson-eason-barnes.tex
@@
 
-The following code gives an example of using a CMIP5 file and some of
-the R packages and functions discussed. Results include a map of
-near-surface air temperatures from a single climate model on a specific
-day in July 2075 (Figure \ref{fig:worldmap}) and a time series of daily
-near-surface air temperature simulations for the climate model grid
-point closest to Beijing, China (Figure \ref{fig:timeseries}). To try
-this example, you will need to download the file
-``tas\_day\_GFDL-ESM2G\_rcp85\_r1i1p1\_20710101-20751231.nc'' from
-\url{https://pcmdi.llnl.gov/search/cmip5/} and save it in the ``tmp''
-subdirectory of your home directory.
+The functions and packages just described can be used to work with CMIP5
+netCDF files in R to do things like map near-surface air temperatures
+from a single climate model on a specific day in July 2075 (Figure
+\ref{fig:worldmap}) and pull a time series of daily near-surface air
+temperature simulations for the climate model grid point closest to
+Beijing, China (Figure \ref{fig:timeseries}). The
+``starting\_from\_netcdf'' vignette that comes with the
+``futureheatwaves'' package provides all code required to create these
+figures, as well as more details and code examples on working with CMIP5
+netCDF files in R.
 

### 2016-12-31 README.Rmd

@@
 
-A version of this vignette is also available through CRAN [here](https://cran.r-project.org/web/packages/futureheatwaves/vignettes/futureheatwaves.html).
+A version of this vignette is also available through CRAN [here](https://CRAN.R-project.org/package=futureheatwaves/vignettes/futureheatwaves.html).
 

### 2017-01-04 paper_drafts/RJournalRevisions/anderson-eason-barnes.Rmd
More work on article draft

--- 2016-12-30 paper_drafts/RJournalRevisions/anderson-eason-barnes.Rmd
@@
 abstract: >
-  Research on climate change impacts can require extensive processing to incorporate output from ensembles of different climate models and different simulations of each model.
-This processing can be particularly extensive when identifying and characterizing extreme events like heat waves and frost day spells, as these must be processed from model output with daily time steps and may use identification threholds based on local distributions (e.g., 98\textsuperscript{th} percentile temperature for that location).
+  Research on climate change impacts can require extensive processing, especially when using ensemble techniques to incorporate output from different climate models and different simulations of each model.
+This processing can be particularly extensive when identifying and characterizing multi-day extreme events like heat waves and frost day spells, as these must be processed from model output with daily time steps.
+Further, climate model output is in a format and follows standards that may be unfamiliar to most R users.
 Here, we provide an overview of working with daily climate model output data in R.
-Further, we present the \pkg{futureheatwaves} package, which we developed to ease the process of identifying, characterizing, and exploring multi-day extreme events in climate model output.
-This package takes as input a directory of projection files from one or many climate models, identifies all occurrences of a specific type of extreme event (e.g., heat waves) for multiple cities using a user-specified event definition, and writes an output directory of files with these extreme events and their characteristics (e.g., length, absolute and relative temperatures).
-The \pkg{futureheatwaves} package then allows the user to create custom functions and apply them across all event files; this functionality can be used either to explore extreme event patterns (e.g., average length and intensity across different climate models) or to apply epidemiological estimates to project health impacts under different scenarios.
+We then present the \pkg{futureheatwaves} package, which we developed to ease the process of identifying, characterizing, and exploring multi-day extreme events in climate model output.
 preamble: >
@@
 
-Research on climate change impacts can require extensive processing.
-This is not only because output files for a single climate model can be large, but also because of the rising popularity of ensemble techniques \citep{IPCCch1}, in which, to better characterize uncertainty in projections, impacts are assessed for multiple climate models, multiple simulations of each climate model, and multiple climate experiments.
-Projections of regional climate change over the next century are subject to high uncertainty due to three distinct sources: (1) internal climate variability, i.e.
+Research on climate change impacts can require extensive processing of climate model output data.
+This is not only because output files for a single climate model can be large, but also because of the rising popularity of ensemble techniques \citep{IPCCch1} in which, to better characterize uncertainty in projections, impacts are assessed for multiple climate models, multiple simulations of each climate model, and multiple climate experiments.
+Such ensemble techniques help characterize uncertainty in projections of regional climate change over the next century due to three distinct sources: (1) internal climate variability, i.e.
 climate noise, (2) climate model uncertainty, i.e.
@@
 \citet{hawkins2009potential}).
-One approach for ensuring that these sources of uncertainty are characterized is to simulate the future climate many times with multiple models and for multiple future scenarios.
 
-For example, the Coupled Model Intercomparison Project, phase 5 (CMIP5; \citet{taylor2012overview}) brought together dozens of major climate modeling groups around the world to simulate the same future radiative forcing scenarios, but with their own models.
+A key source of data for ensemble techniques is the Coupled Model Intercomparison Project, phase 5 (CMIP5; \citet{taylor2012overview}).
+This project brought together dozens of major climate modeling groups around the world to simulate the same future radiative forcing scenarios, but with their own models.
 This created an ensemble of state-of-the-art climate model projections that allow researchers to study projections and their uncertainties.
@@
 
-Processing climate output data is particularly intensive for data in smaller time steps, like daily output.
-While some climate impacts can be assessed using climate model output at monthly, seasonal, or yearly time steps, the impacts of multi-day extreme events must be assessed using output in daily time steps.
-Such multi-day extreme events include heat waves, cold spells, frost day spells, and droughts.
-The assessment is further complicated if extreme events are identified based on conditions that are rare for a certain location (e.g., 98th percentile of local temperature distribution for identifying heat waves) \citep{IPCCch1}, as this event definition must be determined at each study location from climate model output.
-Further, it is often of interest to create summaries of multiple characteristics of these extreme events.
-For example, one may be interested in determining whether the frequency or characteristics (e.g., length, intensity) of heat waves or warm spells will change under certain climate change scenarios \citep{IPCCch1}.
-
-We begin this article with an overview of climate model output data, particularly daily data, for R users.
+We begin this article with an overview of CMIP5 climate model output data for R users, focusing on output with a daily time step.
 We outline where data from CMIP5 can be obtained as well as how to work with the file format (netCDF) from R.
@@
 
-After this overview, we present the \pkg{futureheatwaves} package which we created to aid in identifying and characterizing any type of multi-day extreme event from daily climate model output (Table \ref{tab:goals}).
-This package can be used to identify and characterize a variety of multi-day extreme events, but provides some functionality particularly useful in identifying and characterizing heat waves specifically.
+After this overview, we present the \pkg{futureheatwaves} package, which we created to aid in identifying and characterizing any type of multi-day extreme event from daily climate model output (Table \ref{tab:goals}).
+The impacts of multi-day extreme events must be assessed using output in daily time step, unlike other climate impacts that can be assessed using climate model output at monthly, seasonal, or yearly time steps.
+Further, extreme events are identified based on conditions that are rare for a certain location (e.g., 98th percentile of local temperature distribution for identifying heat waves) \citep{IPCCch1}.
+In this case, assessment is further complicated, as the event definition must be determined at each study location from climate model output.
+Finally, it is often of interest to create summaries of multiple characteristics of these extreme events.
+For example, one may be interested in determining whether the frequency or characteristics (e.g., length, intensity) of heat waves or warm spells will change under certain climate change scenarios \citep{IPCCch1}.
+
+The \pkg{futureheatwaves} package handles these challenges and can be used to identify and characterize a variety of multi-day extreme events across different ensemble members of one or more climate models.
+It also provides some functionality particularly useful in identifying and characterizing heat waves specifically.
 Quantification of the impacts of heat waves on human health suffer from additional sources of uncertainty beyond those inherent in projections of regional changes in surface temperature.
-These include: (1) uncertainty in the ability of communities to adapt to changing temperatures (the adaptation scenario) and (2) uncertainty in the definition of a heat wave itself.
-Thus, identifying and characterizing the impacts of future heat waves in state-of-the-art climate models requires analyzing hundreds of projections based on combinations of anthropogenic activity, climate model and ensemble member, adaptation scenario, and heat wave definition.
-Such an analysis is non-trivial.
-For example, to estimate the possible impacts of heat waves using 32 ensemble members of CMIP5, two forcing scenarios, two different heat wave definitions, and five progressive assumptions about adaptation to heat, would require that one identify and characterize heat waves in 640 separate temperature time series for each region of interest.
+These include: (1) uncertainty in the definition of a heat wave itself and (2) uncertainty in the ability of communities to adapt to changing temperatures (the adaptation scenario).
+This package therefore allows the user to create and use a custom extreme event definition to identify events in the climate model output, as well as options to explore different scenarios of adaptation to heat.
 
@@
 \midrule
-1 & Make processing of large sets of climate projections more practical for researchers exploring the potential impacts of heat waves and other multi-day extreme events.\\
+1 & Make processing of large ensembles of climate simulations more practical for researchers exploring the potential impacts of heat waves and other multi-day extreme events.\\
 2 & Speed up processing time by incorporating C++ in event identification.\\
-3 & Keep track of the names of climate models, and number of ensemble members processed for each.\\
-4 & Not only identify, but also characterize, all extreme events within each climate projection, to allow the exploration of patterns in these characteristics across different projections and also to allow the use of more complex impacts models, including models that include effect modification by event characteristics (e.g., event length, event intensity).
+3 & Keep track of the names of climate models and number of ensemble members processed for each.\\
+4 & Not only identify, but also characterize, all extreme events within each climate simulation, to allow the exploration of patterns in these characteristics across different simulations and also to allow the use of more complex impacts models, including models that incorporate event characteristics (e.g., event length, event intensity).
 For example, this package allows the user to apply a health effects model where risk of mortality is not the same for every heat wave, but rather is modified by heat wave length, intensity, or other measured characteristics.\\
-5 & Give users extensive power in customizing the process, including allowing custom event definitions.\\
-6 & Allow users to easily explore the extreme events identified within all climate projections by applying custom functions across heat wave data sets from all projections at once.\\
+5 & Give users extensive power in customizing the process, including allowing custom extreme event definitions.\\
+6 & Allow users to easily explore the extreme events identified within all climate simulations.\\
 7 & Create output that is in a ``tidy" data format, allowing it to work well with \pkg{ggplot2} for visualization.\\
@@
 
-The \pkg{futureheatwaves} package automates the process of identifying and characterizing multi-day extreme events across different ensemble members of one or more climate models.
-A variety of different heat wave definitions have been used to identify heat waves in a time series of temperature data \citep{smith2013heat}, and the choice of heat wave definitions can influence both health effect estimates \citep{chen2015influence, kent2014heat} and projected heat wave trends \citep{smith2013heat}.
-Further, other types of extreme events will be defined differently than heat waves (for example, frost day spells may be defined as one or more days with temperature at or below 32\textsuperscript{o}F / 0\textsuperscript{o}C).
-This package therefore allows the user to create and use a custom extreme event definition to identify events in the climate model output.
-Finally, this package allows the user to explore the extreme events identified in the climate model output with a function that can take a user-defined R function and apply it across all extreme event files generated for the separate ensemble members.
-
 # An overview of climate model output for R users
@@
 For climate impact studies, a main source for climate model output is the Coupled Model Intercomparison Project, which is currently in its fifth phase (CMIP5).
-Over 20 climate modeling groups have created one or more climate models which, for this project, are run using standardized scenarios \citep{taylor2012overview}.
+Over 20 climate modeling groups created one or more climate models which, for this project, were run using standardized scenarios \citep{taylor2012overview}.
 The resulting output is uniform across modeling groups and has a consistent structure, which allows comparison of simulations from different models \citep{IPCCch9}.
-For CMIP5, each group ran simulations under several experiments, with experiments varying in terms of the radiative forcing.
-This radiative forcing depends on time-varying model inputs (greenhouse gas emissions or concentrations, land use changes, etc.), which are specified for each experiment \citep{taylor2012overview, IPCCch9}.
-Experiments include historical experiments (run using forcings consistent with observed and reconstructed data for 1850--2005), pre-Industrial control experiments, and experiments of future scenarios of radiative forcing over the 21st century or longer (e.g., RCP4.5, RCP8.5) \citep{taylor2012overview}.
-Some modeling groups created ensembles of output for a specific model and experiment, in which they ran the experiment multiple times with the model with very small changes to the initial conditions, resulting in multiple ensemble members for a single climate model and experiment.
 CMIP5 climate model output is archived at a number of different time steps (e.g., daily, monthly, seasonal, yearly) \citep{taylor2010cmip5}, and some variables are reported at multiple levels in the ocean or atmosphere (e.g., ocean temperature).
 Here, we will focus on data with a daily time step for variables reported at a single level (e.g., near-surface air temperature).
+
+Each modeling group ran simulations under several experiments, with experiments varying in terms of radiative forcing by using different scenarios of time-varying model inputs (greenhouse gas emissions or concentrations, land use changes, etc.) \citep{taylor2012overview, IPCCch9}.
+Experiments include historical experiments (run using radiative forcing consistent with observed and reconstructed data for 1850--2005), pre-Industrial control experiments, and experiments of future scenarios of radiative forcing over the 21st century or longer (e.g., RCP4.5, RCP8.5) \citep{taylor2012overview}.
+Some modeling groups created ensembles of output for a specific model and experiment, in which they ran the experiment multiple times with the model with very small changes to the initial conditions.
 
@@
 There is a separate file for each combination of climate model, experiment, modeling realm (e.g., atmosphere, ocean), variable, time step, and ensemble member \citep{taylor2012overview, taylor2010cmip5}.
-For finer time scales, this output is further split across multiple files for specific year ranges (e.g., 5 years of output for each file) \citep{taylor2010cmip5}.
-Filenames for CMIP5 files can be parsed to generate information about the output variable, climate model, experiment, and ensemble member for the simulation \citep{taylor2010cmip5}.
-Files can be searched and downloaded through a point-and-click web interface.
-The data portal also allows you to generate a specific wget script, which you can use to download many files at once.
-A wget script can also be generated through Earth System Grid Federation's Search RESTful API.
+For finer time scales, the output is further split across multiple files for specific year ranges (e.g., 5 years of output for each file) \citep{taylor2010cmip5}.
+Each file's name includes the output variable, climate model, experiment, and ensemble member for the simulation \citep{taylor2010cmip5}.
+
+CMIP5 files can be searched and downloaded through a point-and-click web interface.
+They can also be downloaded in bulk to computers with Unix or Mac operating systems using the \code{wget} file downloading utility.
+Appropriate \code{wget} scripts can be created either through the World Climate Research Programme CMIP5 data portal or through Earth System Grid Federation's Search RESTful API.
 Tips on efficiently searching and downloading the data, including through use of wget scripts and the search API, are available as user tutorials through the website of the University of Colorado Boulder's Earth System CoG (e.g., https://www.earthsystemcog.org/projects/cog/doc/wget for a tutorial on downloading files using wget).
 
-CMIP5 files are saved in Network Common Data Format (netCDF).
-NetCDF is a binary file format that allows storage of data representing a regular array.
-These climate model output files can be as large as several gigabytes [?].
-For climate model output at a single level (e.g., near-surface air temperature), the data is a 3-dimensional array, with dimensions representing time and two coordinates of location (e.g., latitude and longitude) (Figure \ref{fig:netcdfexample}).
+CMIP5 files are saved in Network Common Data Format (netCDF), a binary file format that allows storage of data representing a regular array.
+For climate model output at a single level (e.g., near-surface air temperature), the data is a 3-dimensional array, with dimensions representing time and two coordinates of location (e.g., latitude and longitude).
+Figure \ref{fig:netcdfexample} provides a sketch of the structure of netCDF files for single-level climate model output.
 Global climate models generate output at regularly-spaced time steps, typically at regularly-spaced grid points around the world.
@@
 For daily output files, the time unit is in days since a specified origin date-time (e.g., days since 1850-01-01 00:00:00) \citep{taylor2010cmip5}.
-All CMIP5 output files are required to include certain metadata  \citep{taylor2010cmip5}.
-This required metadata includes the experiment, forcing agents input to the model to create the simulation, time step, institution and institutional contact information, climate model, and modeling realm \citep{taylor2010cmip5}.
+All CMIP5 output files are required to include certain metadata  \citep{taylor2010cmip5}, including the experiment, forcing agents input to the model to create the simulation, time step, institution and institutional contact information, climate model, and modeling realm \citep{taylor2010cmip5}.
 The metadata also must include units for all of the coordinate variables (e.g., longitude, latitude, time).
+The netCDF format allows you to access metadata and variables describing the dimensions of the data without reading the full file into memory.
 
-\begin{widefigure}
+\begin{figure}
 \begin{center}
@@
 \label{fig:netcdfexample}
-\end{widefigure}
+\end{figure}
 
-\citet{taylor2012overview} and \citet{meehl2007wcrp} are excellent resources for finding out more about the CMIP climate model output data.
+To find out more about the CMIP climate model output data, \citet{taylor2012overview} and \citet{meehl2007wcrp} are excellent resources.
 
@@
 
-A few R packages can be used to work with the netCDF file format used for CMIP5 files.
-Earlier packages to work with netCDF files included \pkg{ncdf} and \pkg{ncvar}, but these do not work with the newer netCDF version 4 released in 2008 and are no longer available through CRAN.
+When working with daily climate model output data, challenges to R users include: (1) the file format; (2) use of non-Gregorian calendars; and (3) large file sizes.
+This section explains these challenges and offers some strategies for dealing with them.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width = \textwidth]{worldmapexample}
+\end{center}
+\caption{Example of mapping near-surface air temperature data worldwide for a single day of climate model output data.
+This map uses data from the Geophysical Fluid Dynamics Laboratory's Earth System Model 2G, r1i1p1 ensemble member, on a single day in the summer of 2075.
+Full code for recreating the map is available in the "starting\_from\_netcdf" vignette of the \pkg{futureheatwaves} package.}
+\label{fig:worldmap}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width = 0.8\textwidth]{timeseriesexample}
+\end{center}
+\caption{Example of plotting a time series of temperature simulations between 2071 and 2075 from CMIP5 daily climate model output data for the model grid cell point closest to Beijing, China.
+This plot uses data from the Geophysical Fluid Dynamics Laboratory's Earth System Model 2G, r1i1p1 ensemble member.
+Full code for recreating the map is available in the "starting\_from\_netcdf" vignette of the \pkg{futureheatwaves} package.}
+\label{fig:timeseries}
+\end{figure}
+
+CMIP5 data are available as netCDF files, for which base R import functions do not exist.
+There are, however, a few R packages that can be used to work with the netCDF file format used for CMIP5 files.
+Older packages include \pkg{ncdf} and \pkg{ncvar}, but these do not work with the newer netCDF version 4 released in 2008 and are no longer available through CRAN.
 More recent packages, including \pkg{ncdf4} \citep{ncdf4} and \pkg{RNetCDF} \citep{michna2013rnetcdf, RNetCDF}, work with both version 4 and netCDF's older version 3.
-Climate model output data for CMIP5 is required to conform with the earlier version (version 3) \citep{taylor2010cmip5} and so should work with any of these packages, although it is safer to write code using functions that can be used with either version in case future phases of CMIP do not require files to conform with netCDF version 3.
+While climate model output data for CMIP5 is required to conform with the earlier version (version 3) \citep{taylor2010cmip5}, it is safer to write code using functions that can be used with either version, in case future phases of CMIP do not require files to conform with netCDF version 3.
 
-The netCDF format allows you to access metadata and variables describing the dimensions of the data without reading the full file into memory.
-The metadata describes the dimensions of the array and the units, etc., of each variable and is printed by the print method of the R object returned by the \code{nc\_open} function of \pkg{ncdf4}.
-For CMIP5 files, the dimension variables include the times and locations corresponding to each array in the data.
-Before reading in a variable from a netCDF file, a connection to the file must be opened, for example with the \code{nc\_open} function from \pkg{ncdf4}.
-Variables can then be read in using the \code{ncvar\_get} function from \pkg{ncdf4}, with the \code{varid} parameter set to "lat", "lon", or "time" (as a caveat, many climate models output to non-Gregorian calendars, in which case the time variable should be read in using a different function, as discussed later in this section).
+You can do a number of things with netCDF files in R using these functions.
+For example, \pkg{ncdf4}'s \code{nc\_open} function can be used to open a connection to a netCDF file, and the object returned by the function includes the file's attribute data.
+Once a file connection is open, variables can be read in using the \code{ncvar\_get} function.
+For example, the variables defining the dimensions of the sketched netCDF file in Figure \ref{fig:netcdfexample} could be read into R with \code{ncvar\_get} with the \code{varid} parameter set to "lat", "lon", or "time".
 The climate output variable (e.g., near-surface air temperature) can similarly be read in using \code{ncvar\_get}.
 In this case, the \code{varid} parameter should be set using the appropriate CMIP5 variable name (e.g., "tas" for near-surface air temperature); these variable names can be found in the CMIP Requested Output tables \citep{taylor2010cmip5}.
-You can use the dimensional time and location data to identify the location of the variable data you need in the netCDF array and use indexing to read only that data into memory, without needing to read in the full file (for example, with the \code{nc.get.var.subset.by.axes} function in \pkg{ncdf4.helpers}).
+If you only need data from a subset of the full file, you can use the dimensional time and location data to identify the location of the variable data you need in the netCDF array and use indexing to read only that data into memory, without needing to read in the full file (for example, with the \code{nc.get.var.subset.by.axes} function in \pkg{ncdf4.helpers}).
 Once the user is done reading in data from the file, the connection can be closed (e.g., with the \code{nc\_close} function from \pkg{ncdf4}).
 
+A second challenge when working with climate model output data in R is that some climate models output to non-Gregorian calendars.
 Since the late 1500s, Western dates have been set using the Gregorian calendar, which has 365.2425-day years.
@@
 With these non-Gregorian calendars, R's base functions for converting a vector to a Date class based on the number of days since an origin date (\code{as.Date}, \code{as.POSIXct}) do not return the desired values.
-The \pkg{PCICt} \citep{PCICt} and \pkg{ncdf4.helpers} \citep{ncdf4.helpers} packages provide further functionality with netCDF files that can be useful when working with climate model output data and provide particular help in working with different calendars.
-CMIP5 netCDF files include information on the calendar used, which the \code{nc.get.time.series} function in \pkg{ncdf4.helpers} use to convert the "time" variable in the file to an object of the PCICt class, which provides Date-like functionality for 360- and 365-day calendars \citep{PCICt}.
-As a note, while these functions will help with handling most CMIP5 files, the CMIP5 standards allows use of other calendars which may not be succesfully handled by these functions, so it is important to assess whether the time variable range in the PCICt object correctly matches the expected date ranges for a file as you process CMIP5 data in R.
 
-The size of the climate model output files can be large enough that it may make more sense to work with smaller chunks of the data, rather than reading all data into memory and working with the data all at once \citep{RCMIP5}.
+Two R packages provide help with non-Gregorian calendars: \pkg{PCICt} \citep{PCICt} and \pkg{ncdf4.helpers} \citep{ncdf4.helpers}.
+The \code{nc.get.time.series} function in \pkg{ncdf4.helpers} pulls and uses metadata on the calendar stored in the CMIP5 netCDF file's attributes to convert the "time" variable in the file to an object of the \code{PCICt} class.
+This class is defined in the \pkg{PCICt} package and provides Date-like functionality for 360- and 365-day calendars \citep{PCICt}.
+However, while these functions will help with handling most CMIP5 files, the CMIP5 standards allows use of other calendars which may not be succesfully handled by these functions, so it is important to assess whether the time variable range in the \code{PCICt} object correctly matches the expected date ranges for a file when processing CMIP5 data in R.
+
+Finally, the size of CMIP5 files can make them difficult to work with in R.
+CMIP5 climate model output files can be as large as several gigabytes.
+The size of the files can therefore be large enough that it may make more sense to work with smaller chunks of the data in R, rather than reading all data into memory and working with the data all at once \citep{RCMIP5}.
 This problem aggregates when working with multiple climate models and more than one ensemble member for each of those climate models.
 
-The functions and packages described can be used to work with CMIP5 netCDF files in R to do things like map near-surface air temperatures from a single climate model on a specific day in July 2075 (Figure \ref{fig:worldmap}) and pull a time series of daily near-surface air temperature simulations for the climate model grid point closest to Beijing, China (Figure \ref{fig:timeseries}).
+In addition to these general packages for working with netCDF files, there are several R packages specifically for working with climate model output data, including \pkg{RCMIP5} \citep{RCMIP5} and \pkg{wux} \citep{wux}.
+However, these packages are more useful for working with data output at time steps of a month or higher and have limited utility with the daily climate model output data required for studies of multi-day extreme events.
+
+The \pkg{RCMIP5} package includes functions to read in CMIP5 data from netCDF files, scan a directory of CMIP5 files and determine models with continuous available data, create objects of a special \code{cmip5data} class to work with CMIP5 data within R, and parse the filenames for all files in a directory to extract information within the filename.
+For this package, most functions only work with monthly or less frequent data \citep{RCMIP5}.
+While the \code{loadCMIP5} function does successfully load daily data as a \code{cmip5data} object, most of the methods for this object type do not do anything meaningful for daily data.
+The package's \code{getFileInfo} function, however, will work with CMIP5 files of any time step; this function identifies all CMIP5 files in a directory and creates a dataframe with information parsed from the file name.
+As a note, the \code{get.split.filename.cmip5} function in the \pkg{ncdf4.helper} package similarly can be used to parse information contained in CMIP5 file names \citep{ncdf4.helpers}.
+
+The \pkg{wux} package \citep{wux} includes functions that allow the user to download CMIP5 monthly-aggregated output directly from within R with the \code{CMIP5fromESGF} function.
+The package then uses the \code{models2wux} function to read in climate model output netCDF files and convert them to "WUX" dataframes, which can be used by other functions in the package.
+While this function can input climate model output with daily time steps (the "what.timesteps" element of the \code{modelinput} list input must be set to "daily"), the function aggregates this data to a monthly or less frequent (e.g., seasonal) aggregation when creating the WUX dataframe.
+Therefore, while this package provides very useful functionality for working with averaged output of daily climate model output data, it cannot easily be used to identify and characterize multi-day extreme events like heat waves.
+
+The functions and packages described in this section can be used with CMIP5 netCDF files to do things in R like map near-surface air temperatures from a single climate model on a specific day (Figure \ref{fig:worldmap}) or pull a time series of daily near-surface air temperature simulations at a specific climate model grid point (Figure \ref{fig:timeseries}).
 The "starting_from_netcdf" vignette that comes with the "futureheatwaves" package provides all code required to create these figures, as well as more details and code examples on working with CMIP5 netCDF files in R.
@@
 
-\begin{figure}
-\begin{center}
-\includegraphics[width = \textwidth]{worldmapexample}
-\end{center}
-\caption{Example of mapping near-surface air temperature data worldwide for a single day of climate model output data.
-This map uses data from the Geophysical Fluid Dynamics Laboratory's Earth System Model 2G, r1i1p1 ensemble member, on a single day in the summer of 2075.}
-\label{fig:worldmap}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width = 0.8\textwidth]{timeseriesexample}
-\end{center}
-\caption{Example of plotting a time series of temperature simulations between 2071 and 2075 from CMIP5 daily climate model output data for the model grid cell point closest to Beijing, China.
-This plot uses data from the Geophysical Fluid Dynamics Laboratory's Earth System Model 2G, r1i1p1 ensemble member.}
-\label{fig:timeseries}
-\end{figure}
-
-In addition to these general packages for working with netCDF files, there are several R packages specifically for working with climate model output data, including \pkg{RCMIP5} \citep{RCMIP5} and \pkg{wux} \citep{wux}.
-However, these packages are more useful for working with data output at time steps of a month or higher and have limited utility with the daily climate model output data required for studies of multi-day extreme events.
-
-The \pkg{RCMIP5} package includes functions to read in CMIP5 data from netCDF files, scan a directory of CMIP5 files and determine models with continuous available data, create objects of a special "cmip5data" class to work with CMIP5 data within R, and parse the filenames for all files in a directory to extract information within the filename.
-For this package, most functions only work with monthly or less frequent data (e.g., the functions \code{checkTimePeriod} and \code{cmip5data}, as well as functions that work with \code{cmip5data} objects including \code{filterDimensions} and \code{getProjectionMatrix}) \citep{RCMIP5}.
-While the \code{loadCMIP5} function does successfully load the daily data as a \code{cmip5data} object, most of the methods for this object type do not do anything meaningful for daily data.
-The package's \code{getFileInfo} function, however, will work with CMIP5 files of any time step; this function identifies all CMIP5 files in a directory and creates a dataframe that parses the information contained in the file name.
-As a note, the \code{get.split.filename.cmip5} function in the \pkg{ncdf4.helper} package similarly can be used to parse information contained in CMIP5 file names \citep{ncdf4.helpers}.
-
-The \pkg{wux} package \citep{wux} includes functions that allow the user to download CMIP5 monthly-aggregated output directly from within R with the \code{CMIP5fromESGF} function.
-However, this function does not allow downloading of climate model output with finer time steps, like daily data.
-This package uses the \code{models2wux} function to read in climate model output netCDF files and convert to a WUX dataframe that can be used by other functions in the package.
-While this function can input climate model output with daily time steps, if the element "what.timesteps" of the \code{modelinput} list input is set to "daily", the function aggregates this data to a monthly or less frequent (e.g., seasonal) aggregation when creating the WUX dataframe.
-Therefore, while this package provides very useful functionality for working with averaged output of daily climate model output data, it cannot easily be used to identify and characterize multi-day extreme events like heat waves.
-
-More tips on working with CMIP5 files in R are provided in the "starting_from_netcdf" vignette of the \pkg{futureheatwaves} package.
-
 # The \pkg{futureheatwaves} package
@@
 First, the \code{gen\_hw\_set} function processes a directory of climate projection files that are stored locally on the user's computer (Figure \ref{fig:overview}, "Climate projections"), to generate a list of all extreme events in each projection, as well as over a dozen characteristics of each identified extreme event (Table \ref{tab:hwcharacteristics}).
-The extreme events are identified and characterized at one or more study locations (e.g., cities), with study locations specified by the user in an input file.
+This package start from files rather than R objects to avoid loading data from all climate model ensembles at once; instead, the function loads, processes, and saves output for a single climate model ensemble member at a time.
+The extreme events are identified and characterized at one or more study locations (e.g., cities), which the user specifies in an input file.
 The extreme events identified for each ensemble member are output as separate files in a directory specified by the user (Figure \ref{fig:overview}, ``Extreme events datasets").
 
-Once the user creates these data frames of location-specific heat waves, the \code{apply\_all\_models} function allows the user to apply custom functions across all the extreme event data frames generated by \code{gen\_hw\_set} to summaries of extreme events across all climate models and ensemble members (Figure \ref{fig:overview}, right).
-This functionality can be used to generate summary statistics (e.g., determine average heat wave length or total frost days), or can be used to apply more complex functions (e.g., apply epidemiologic effect estimates across the heat waves to generate health impact estimates).
+Once the user creates these data frames of location-specific heat waves, the \code{apply\_all\_models} function can be used to apply custom functions across all the extreme event data frames generated by \code{gen\_hw\_set}.
+This functionality allows users to create summaries of extreme events across all climate models and ensemble members (Figure \ref{fig:overview}, right).
+The function can be used to generate summary statistics (e.g., determine average heat wave length or total frost days) or to apply more complex functions (e.g., apply epidemiologic effect estimates across the heat waves to generate health impact estimates).
+
+When using this package, CMIP5 climate model output data requires some pre-processing.
+Data will need to be saved in a specific format, with files stored in a specific directory structure.
+Full details of the required file and directory structure are provided in the package's "futureheatwaves" vignette, while tips and an R script for conducting this processing starting from CMIP5 netCDF files are given in the "starting_from_netcdf" vignette.
+
+This package can be used for study locations worldwide.
+The "starting_from_netcdf"" package vignette provides an example of using this package to identify and explore future heat waves in several Chinese cities.
 
@@
 \code{start.month} & Month in which the event started (1 = January) \\
-\code{days.above.abs.thresh.1} & Number of days in the event above a specified absolute threshold (default is the number of days in the event above 80\textsuperscript{o}F / 26.7\textsuperscript{o}C, but this can be changed with the \code{absolute\_thresholds} argument in \code{gen\_hw\_set}) \\
+\code{days.above.abs.thresh.1} & Number of days in the event above a specified absolute threshold (default is the number of days in the event above 80\textsuperscript{o}F / 26.7\textsuperscript{o}C, but this and the following three absolute thresholds can be changed with the \code{absolute\_thresholds} argument in \code{gen\_hw\_set}) \\
 \code{days.above.abs.thresh.2} & Number of days in the event above a specified absolute threshold (default is the number of days in the event above 85\textsuperscript{o}F / 29.4\textsuperscript{o}C) \\
@@
 We include one ensemble member from BCC (r1i1p1) and two from CCSM (r1i1p1 and r2i1p1).
+Once the \pkg{futureheatwaves} package is installed and loaded, the user can find the local location of these files using R's \code{system.file} function.
 
@@
 
-Once the \pkg{futureheatwaves} package is installed and loaded, the user can find the local location of these files using R's \code{system.file} function.
-In the later sections of this article, we show how to use the package functions with these example files as inputs.
-
-When using this package with data other than this example data, the climate model output data will require some pre-processing.
-Data will need to be saved in a specific format, with files stored in a specific directory structure.
-This preprocessing can be done using some of the R packages described earlier for working with CMIP5 data in R.
-Full details of the required file and directory structure are provided in the package's "futureheatwaves" vignette, while tips on conducting this processing starting from CMIP5 netCDF files are given in the "starting_from_netcdf" vignette.
-
-## Basic example of identifying and characterizing extreme events
+## Basic example of using \pkg{futureheatwaves}
 
@@
 This code first identifies and saves as objects the path names on the user's computer of the example climate projections directory (\code{projection\_dir\_location}) and the file of study locations (\code{city\_file\_location}).
-The \code{gen\_hw\_set} function processes this example input and creates a new directory, \file{example\_results}, with files of identified and characterized heat waves, in the user's current working directory.
+The \code{gen\_hw\_set} function processes the example input and creates a new directory, \file{example\_results}, with files of identified and characterized heat waves, in the user's current working directory.
 In this example code, the processing is done using  default values for the event definition, years for which to generate the heat wave data sets, etc.
@@
 This directory will include files with some basic information about the climate models and the closest grid points of each climate model to each location, as well as a directory with files of identified and classified extreme events for each ensemble member, including all characteristics in Table \ref{tab:hwcharacteristics}.
-
-## Customizing the extreme event definition
-
-By default, the package identifies extreme events in climate model output data using a specific definition for heat waves:
-
-> A \dfn{heat wave} is two or more days at or above a city-specific threshold temperature, with the threshold determined as the 98\textsuperscript{th} percentile of year-round temperature in the city during some reference period (by default, 1990--1999).
-
-\noindent However, the user can extensively customize the definition used to identify extreme events.
-Users can write a custom R function with either a different heat wave definition (see \citet{smith2013heat} and \citet{kent2014heat} for listings of definitions used in scientific studies) or with a definition appropriate for a different type of extreme event (e.g., one or more days at or below 32\textsuperscript{o}F / 0\textsuperscript{o}C for frost day spells).
-For heat wave identification, researchers might want to use a specific definition, for example, because it matches the definition used by local health officials to declare heat wave warnings or, in the case of health impact assessments, to match with a definition used in an epidemiological study.
-For other extreme events, a heat wave definition likely will not be applicable and so a customized definition is necessary.
-
-Three components of the extreme event definition can be easily customized in the \code{gen\_hw\_set} function call, without creating a new R function to use to identify heat waves.
-First, often users wish to define an extreme event based on conditions that are rare in the study location \citep{IPCCch1}.
-In this case, the percentile of the variable of interest required for an extreme event can be changed using the \code{probThreshold} option in \code{gen\_hw\_set}.
-For example, this option could be used to identify heat waves based on the 99\textsuperscript{th} percentile, rather than the 98\textsuperscript{th} percentile, of a city's year-round temperature.
-Second, the user can change the number of days used in the heat wave definition using the \code{numDays} argument in the \code{gen\_hw\_set} function (the default is \samp{numDays = 2}, corresponding to two or more days).
-Combined, these two customization choices allow the user to identify heat waves using many of the heat wave definitions used in previous climate and health research-- for example, 9 of the 16 heat wave definitions used in \citet{kent2014heat} (e.g., [some examples of the 9 that could be fit]) could be fit using different combinations of these two options for specifying threshold percentile and number of days.
-Third, some extreme events like cold waves and frost day spells are defined as a certain number of days below, rather than above, a threshold.
-While the default is to identify events by searching for days above a threshold, this behavior can be changed with the \code{above\_threshold = FALSE} argument in the \code{gen\_hw\_set function}.
-
-Beyond these simpler options, the customization of the event definition is even more extensive as one has the option of writing a custom R function.
-This functionality allows the user to use heat wave definitions that either require a number of days above an absolute threshold (e.g., maximum temperature of $\ge$ 95\textsuperscript{o}F for $\ge1$ day \citet{kent2014heat, tan2007heat}; minimum temperature $\le$ 0\textsuperscript{o}C for $\ge1$ day for frost day spells) or that require a combination of thresholds to be met (e.g., maximum daily temperature one threshold every day of the heat wave and above a higher threshold for a certain number of days; \citet{kent2014heat, peng2011toward}).
-To use a customized event definition, the user can write and load an R function that implements the definition.
-The custom function is passed to the \code{gen\_hw\_set} function using the \code{IDheatwavesFunction} argument.
-To work correctly, this custom function must allow only specific inputs and generate only specific outputs.
-Details about this required structure are provided in the \pkg{futureheatwaves} package vignette.
-To increase processing speed when identifying extreme events, we coded parts of the default event definition function in C++ and synced it with R using the \CRANpkg{Rcpp} package \citep{Rcpp}.
-Users should consider a similar strategy for custom heat wave definitions, especially if processing a large number of climate projection files.
-
-Finally, extreme events are often based on conditions that are rare for a specific location using location-specific relative thresholds.
-These thresholds are often defined for climate impact studies based on a variable's distribution at that location in present-day or historical data.
-However, it can be interesting to explore trends in extreme events under climate change if extreme events are identified based on variable distributions during the projection period, or another future period.
-It is possible to specify the time period to use when determining a relative threshold for an event definition using the \code{thresholdBoundaries} argument in the function \code{gen\_hw\_set}.
-
-## Exploring extreme events
+See the package's vignettes for more details on the content and structure of this output.
 
@@
 This output is structured as ``tidy" data \citep{wickham2014tidy}, allowing it to be used easily with the graphing package \CRANpkg{ggplot2} \citep{ggplot2}.
+This functionality can be used to create a number of other summaries of the identified extreme events.
+For example, it could be used to determine average heat wave length or estimate how much earlier in the year events are expected to start across an ensemble of climate model simulations.
+The functionality can also be used for more complex analysis of extreme event files.
+For example, it can be used to apply epidemiological models of heat wave to estimate excess heat-related mortality under different future scenarios; an example of this application is provided in the "futureheatwaves" vignette that comes with the package.
 
-The \code{apply\_all\_models} function can also be used to project the health impacts of heat waves.
-A simple estimate of excess deaths associated with heat wave risk in a community can be calculated as \citep{peng2011toward}:
+## Customizing the extreme event definition
 
-\begin{equation}
-E_c = B_c * (RR - 1) * L_c 
-\label{eq:healthimpacts}
-\end{equation}
+By default, the package identifies extreme events in climate model output data using a specific definition for heat waves that has been used in some epidemiological and climate impact research (e.g., \citet{anderson2009weather}):
 
-where:
+> A \dfn{heat wave} is two or more days at or above a city-specific threshold temperature, with the threshold determined as the 98\textsuperscript{th} percentile of year-round temperature in the city during some reference period (by default, 1990--1999).
 
-\begin{itemize}
-\item $E_c$ is the total number of excess deaths in community $c$;
-\item $B_c$ is the baseline average daily mortality in community $c$; 
-\item $RR$ is the relative risk of mortality associated with a heat wave day; and 
-\item $L_c$ is the total number of heat wave days in community $c$ over the study period.
-\end{itemize}
+However, this is not the only accepted heat wave definition in the scientific literature.
+A variety of different heat wave definitions have been used to identify heat waves in a time series of temperature data \citep{smith2013heat}, and the choice of heat wave definitions can influence both projected heat wave trends \citep{smith2013heat} and estimates of health risks during events \citep{chen2015influence, kent2014heat}.
+Further, other types of extreme events will be defined differently than heat waves (for example, frost day spells may be defined as one or more days with temperature at or below 32\textsuperscript{o}F / 0\textsuperscript{o}C).
 
-This impact assessment equation can be translated into a function that merges each projection's heat wave data frame with a data frame of community-specific baseline mortality rates ($B_c$), calculates equation \ref{eq:healthimpacts} for each heat wave, and then sums up the total excess deaths across all heat waves.
-In this example, we have set the default relative risk for a heat wave as 1.032, based on the estimated added heat wave relative risk in 107 U.S.
-communities for heat waves defined as two or more days at or above a community's 98\textsuperscript{th} percentile temperature \citep{anderson2009weather}.
+Therefore, this package allows the user to extensively customize the definition used to identify extreme events.
+Users can write a custom R function with either a different heat wave definition (see \citet{smith2013heat} and \citet{kent2014heat} for listings of definitions used in scientific studies) or with a definition appropriate for a different type of extreme event (e.g., one or more days at or below 32\textsuperscript{o}F / 0\textsuperscript{o}C for frost day spells).
+For heat wave identification, researchers might want to use a specific definition, for example, because it matches the definition used by local health officials to declare heat wave warnings or, in the case of health impact assessments, to match with a definition used in an epidemiological study.
+For other extreme events, a heat wave definition likely will not be applicable and so a customized definition is necessary.
 
-```
-excess_deaths <- function(hw_datafr, base_mortality, RR = 1.032){
-        hw_datafr <- dplyr::left_join(hw_datafr, base_mortality,
-                                      by = "city") %>%
-                dplyr::mutate(excess_deaths = base_mort * (RR - 1) * length)
-        out <- sum(hw_datafr$excess_deaths)
-        out <- round(out)
-        return(out)
-}
-```
+Three components of the extreme event definition can be easily customized in the \code{gen\_hw\_set} function call, without creating a new R function to use to identify heat waves.
+First, often users wish to define an extreme event based on conditions that are rare in the study location \citep{IPCCch1}.
+In this case, the percentile of the variable of interest required for an extreme event can be changed using the \code{probThreshold} option in \code{gen\_hw\_set}.
+For example, this option could be used to identify heat waves based on the 99\textsuperscript{th} percentile, rather than the 98\textsuperscript{th} percentile, of a city's year-round temperature.
+Second, the user can change the number of days used in the heat wave definition using the \code{numDays} argument in the \code{gen\_hw\_set} function (the default is \samp{numDays = 2}, corresponding to two or more days).
+Combined, these two customization choices allow the user to identify heat waves using many of the heat wave definitions used in previous climate and health research-- for example, 9 of the 16 heat wave definitions used in \citet{kent2014heat} (e.g., [some examples of the 9 that could be fit]) could be fit using different combinations of these two options for specifying threshold percentile and number of days.
+Third, some extreme events like cold waves and frost day spells are defined as a certain number of days below, rather than above, a threshold.
+While the default is to identify events by searching for days above a threshold, this behavior can be changed with the \code{above\_threshold = FALSE} argument in the \code{gen\_hw\_set function}.
 
-```{r echo = FALSE}
-excess_deaths <- function(hw_datafr, base_mortality, RR = 1.032){
-        hw_datafr <- dplyr::left_join(hw_datafr, base_mortality,
-                                      by = "city") %>%
-                dplyr::mutate(excess_deaths = base_mort * (RR - 1) * length)
-        out <- sum(hw_datafr$excess_deaths)
-        out <- round(out)
-        return(out)
-}
-```
+Beyond these simpler options, the customization of the event definition is even more extensive as one has the option of writing a custom R function.
+This functionality allows the user to use heat wave definitions that either require a number of days above an absolute threshold (e.g., maximum temperature of $\ge$ 95\textsuperscript{o}F for $\ge1$ day \citet{kent2014heat, tan2007heat}; minimum temperature $\le$ 0\textsuperscript{o}C for $\ge1$ day for frost day spells) or that require a combination of thresholds to be met (e.g., maximum daily temperature one threshold every day of the heat wave and above a higher threshold for a certain number of days; \citet{kent2014heat, peng2011toward}).
+To use a customized event definition, the user can write and load an R function that implements the definition.
+The custom function is passed to the \code{gen\_hw\_set} function using the \code{IDheatwavesFunction} argument.
+To work correctly, this custom function must allow only specific inputs and generate only specific outputs.
+Details about this required structure are provided in the \pkg{futureheatwaves} package vignette.
+To increase processing speed when identifying extreme events, we coded parts of the default event definition function in C++ and synced it with R using the \CRANpkg{Rcpp} package \citep{Rcpp}.
+Users should consider a similar strategy for custom heat wave definitions, especially if processing a large number of climate projection files.
 
-\noindent Once defined in R, this function can be applied across all heat waves from all climate models' ensemble members, provided that you have a data frame called \code{city\_mortality} with columns with each community's identifier (\code{city}) and baseline mortality rate (\code{base\_mort}), using the call (this example uses hypothetical baseline mortality rates for the cities and requires the \CRANpkg{dplyr} package \citep{dplyr}):
+Finally, extreme events are often based on conditions that are rare for a specific location using location-specific relative thresholds.
+These thresholds are often defined for climate impact studies based on a variable's distribution at that location in present-day or historical data.
+However, it can be interesting to explore trends in extreme events under climate change if extreme events are identified based on variable distributions during the projection period, or another future period.
+It is possible to specify the time period to use when determining a relative threshold for an event definition using the \code{thresholdBoundaries} argument in the function \code{gen\_hw\_set}.
 
-```{r message = FALSE}
-library(dplyr)
-city_mortality <- data.frame(city = c("balt", "nwk", "ny", "phil", "prov"),
-                             base_mort = c(3, 2.5, 35, 6, 2.5))
-apply_all_models(out = out, FUN = excess_deaths, base_mortality = city_mortality)
-```
-
-\noindent While the above impact analysis is simplistic, it provides an example of how the \code{apply\_all\_models} function can be used in conjunction with epidemiologic results to estimate the potential health impacts of extreme events with climate change.
-
-# Customizing dates for projections and reference temperatures
+## Customizing dates for projections and reference temperatures
 

### 2017-01-04 paper_drafts/RJournalRevisions/anderson-eason-barnes.tex
More work on article draft

--- 2016-12-30 paper_drafts/RJournalRevisions/anderson-eason-barnes.tex
@@
 \abstract{%
-Research on climate change impacts can require extensive processing to
-incorporate output from ensembles of different climate models and
-different simulations of each model. This processing can be particularly
-extensive when identifying and characterizing extreme events like heat
-waves and frost day spells, as these must be processed from model output
-with daily time steps and may use identification threholds based on
-local distributions (e.g., 98\textsuperscript{th} percentile temperature
-for that location). Here, we provide an overview of working with daily
-climate model output data in R. Further, we present the
-\pkg{futureheatwaves} package, which we developed to ease the process of
-identifying, characterizing, and exploring multi-day extreme events in
-climate model output. This package takes as input a directory of
-projection files from one or many climate models, identifies all
-occurrences of a specific type of extreme event (e.g., heat waves) for
-multiple cities using a user-specified event definition, and writes an
-output directory of files with these extreme events and their
-characteristics (e.g., length, absolute and relative temperatures). The
-\pkg{futureheatwaves} package then allows the user to create custom
-functions and apply them across all event files; this functionality can
-be used either to explore extreme event patterns (e.g., average length
-and intensity across different climate models) or to apply
-epidemiological estimates to project health impacts under different
-scenarios.
+Research on climate change impacts can require extensive processing,
+especially when using ensemble techniques to incorporate output from
+different climate models and different simulations of each model. This
+processing can be particularly extensive when identifying and
+characterizing multi-day extreme events like heat waves and frost day
+spells, as these must be processed from model output with daily time
+steps. Further, climate model output is in a format and follows
+standards that may be unfamiliar to most R users. Here, we provide an
+overview of working with daily climate model output data in R. We then
+present the \pkg{futureheatwaves} package, which we developed to ease
+the process of identifying, characterizing, and exploring multi-day
+extreme events in climate model output.
 }
@@
 
-Research on climate change impacts can require extensive processing.
-This is not only because output files for a single climate model can be
-large, but also because of the rising popularity of ensemble techniques
-\citep{IPCCch1}, in which, to better characterize uncertainty in
-projections, impacts are assessed for multiple climate models, multiple
-simulations of each climate model, and multiple climate experiments.
-Projections of regional climate change over the next century are subject
-to high uncertainty due to three distinct sources: (1) internal climate
-variability, i.e.~climate noise, (2) climate model uncertainty, i.e.~the
-same forcing can produce a different response in different models and
-(3) scenario uncertainty, i.e.~uncertainty in future climate forcings
-(e.g. \citet{hawkins2009potential}). One approach for ensuring that
-these sources of uncertainty are characterized is to simulate the future
-climate many times with multiple models and for multiple future
-scenarios.
+Research on climate change impacts can require extensive processing of
+climate model output data. This is not only because output files for a
+single climate model can be large, but also because of the rising
+popularity of ensemble techniques \citep{IPCCch1} in which, to better
+characterize uncertainty in projections, impacts are assessed for
+multiple climate models, multiple simulations of each climate model, and
+multiple climate experiments. Such ensemble techniques help characterize
+uncertainty in projections of regional climate change over the next
+century due to three distinct sources: (1) internal climate variability,
+i.e.~climate noise, (2) climate model uncertainty, i.e.~the same forcing
+can produce a different response in different models and (3) scenario
+uncertainty, i.e.~uncertainty in future climate forcings (e.g.
+\citet{hawkins2009potential}).
 
-For example, the Coupled Model Intercomparison Project, phase 5 (CMIP5;
-\citet{taylor2012overview}) brought together dozens of major climate
-modeling groups around the world to simulate the same future radiative
-forcing scenarios, but with their own models. This created an ensemble
-of state-of-the-art climate model projections that allow researchers to
+A key source of data for ensemble techniques is the Coupled Model
+Intercomparison Project, phase 5 (CMIP5; \citet{taylor2012overview}).
+This project brought together dozens of major climate modeling groups
+around the world to simulate the same future radiative forcing
+scenarios, but with their own models. This created an ensemble of
+state-of-the-art climate model projections that allow researchers to
 study projections and their uncertainties. Most of these modeling groups
@@
 
-Processing climate output data is particularly intensive for data in
-smaller time steps, like daily output. While some climate impacts can be
-assessed using climate model output at monthly, seasonal, or yearly time
-steps, the impacts of multi-day extreme events must be assessed using
-output in daily time steps. Such multi-day extreme events include heat
-waves, cold spells, frost day spells, and droughts. The assessment is
-further complicated if extreme events are identified based on conditions
-that are rare for a certain location (e.g., 98th percentile of local
-temperature distribution for identifying heat waves) \citep{IPCCch1}, as
-this event definition must be determined at each study location from
-climate model output. Further, it is often of interest to create
-summaries of multiple characteristics of these extreme events. For
-example, one may be interested in determining whether the frequency or
-characteristics (e.g., length, intensity) of heat waves or warm spells
-will change under certain climate change scenarios \citep{IPCCch1}.
+We begin this article with an overview of CMIP5 climate model output
+data for R users, focusing on output with a daily time step. We outline
+where data from CMIP5 can be obtained as well as how to work with the
+file format (netCDF) from R. We overview some R packages that can be
+useful when working with this data, as well as aspects of the data
+(e.g., non-standard calendars) of which users should be aware when
+working with daily climate model output in R.
 
-We begin this article with an overview of climate model output data,
-particularly daily data, for R users. We outline where data from CMIP5
-can be obtained as well as how to work with the file format (netCDF)
-from R. We overview some R packages that can be useful when working with
-this data, as well as aspects of the data (e.g., non-standard calendars)
-of which users should be aware when working with daily climate model
-output in R.
-
-After this overview, we present the \pkg{futureheatwaves} package which
+After this overview, we present the \pkg{futureheatwaves} package, which
 we created to aid in identifying and characterizing any type of
 multi-day extreme event from daily climate model output (Table
-\ref{tab:goals}). This package can be used to identify and characterize
-a variety of multi-day extreme events, but provides some functionality
-particularly useful in identifying and characterizing heat waves
-specifically. Quantification of the impacts of heat waves on human
-health suffer from additional sources of uncertainty beyond those
-inherent in projections of regional changes in surface temperature.
-These include: (1) uncertainty in the ability of communities to adapt to
-changing temperatures (the adaptation scenario) and (2) uncertainty in
-the definition of a heat wave itself. Thus, identifying and
-characterizing the impacts of future heat waves in state-of-the-art
-climate models requires analyzing hundreds of projections based on
-combinations of anthropogenic activity, climate model and ensemble
-member, adaptation scenario, and heat wave definition. Such an analysis
-is non-trivial. For example, to estimate the possible impacts of heat
-waves using 32 ensemble members of CMIP5, two forcing scenarios, two
-different heat wave definitions, and five progressive assumptions about
-adaptation to heat, would require that one identify and characterize
-heat waves in 640 separate temperature time series for each region of
-interest.
+\ref{tab:goals}). The impacts of multi-day extreme events must be
+assessed using output in daily time step, unlike other climate impacts
+that can be assessed using climate model output at monthly, seasonal, or
+yearly time steps. Further, extreme events are identified based on
+conditions that are rare for a certain location (e.g., 98th percentile
+of local temperature distribution for identifying heat waves)
+\citep{IPCCch1}. In this case, assessment is further complicated, as the
+event definition must be determined at each study location from climate
+model output. Finally, it is often of interest to create summaries of
+multiple characteristics of these extreme events. For example, one may
+be interested in determining whether the frequency or characteristics
+(e.g., length, intensity) of heat waves or warm spells will change under
+certain climate change scenarios \citep{IPCCch1}.
+
+The \pkg{futureheatwaves} package handles these challenges and can be
+used to identify and characterize a variety of multi-day extreme events
+across different ensemble members of one or more climate models. It also
+provides some functionality particularly useful in identifying and
+characterizing heat waves specifically. Quantification of the impacts of
+heat waves on human health suffer from additional sources of uncertainty
+beyond those inherent in projections of regional changes in surface
+temperature. These include: (1) uncertainty in the definition of a heat
+wave itself and (2) uncertainty in the ability of communities to adapt
+to changing temperatures (the adaptation scenario). This package
+therefore allows the user to create and use a custom extreme event
+definition to identify events in the climate model output, as well as
+options to explore different scenarios of adaptation to heat.
 
@@
 \end{table}
-
-The \pkg{futureheatwaves} package automates the process of identifying
-and characterizing multi-day extreme events across different ensemble
-members of one or more climate models. A variety of different heat wave
-definitions have been used to identify heat waves in a time series of
-temperature data \citep{smith2013heat}, and the choice of heat wave
-definitions can influence both health effect estimates
-\citep{chen2015influence, kent2014heat} and projected heat wave trends
-\citep{smith2013heat}. Further, other types of extreme events will be
-defined differently than heat waves (for example, frost day spells may
-be defined as one or more days with temperature at or below
-32\textsuperscript{o}F / 0\textsuperscript{o}C). This package therefore
-allows the user to create and use a custom extreme event definition to
-identify events in the climate model output. Finally, this package
-allows the user to explore the extreme events identified in the climate
-model output with a function that can take a user-defined R function and
-apply it across all extreme event files generated for the separate
-ensemble members.
 
@@
 the Coupled Model Intercomparison Project, which is currently in its
-fifth phase (CMIP5). Over 20 climate modeling groups have created one or
-more climate models which, for this project, are run using standardized
+fifth phase (CMIP5). Over 20 climate modeling groups created one or more
+climate models which, for this project, were run using standardized
 scenarios \citep{taylor2012overview}. The resulting output is uniform
 across modeling groups and has a consistent structure, which allows
-comparison of simulations from different models \citep{IPCCch9}. For
-CMIP5, each group ran simulations under several experiments, with
-experiments varying in terms of the radiative forcing. This radiative
-forcing depends on time-varying model inputs (greenhouse gas emissions
-or concentrations, land use changes, etc.), which are specified for each
-experiment \citep{taylor2012overview, IPCCch9}. Experiments include
-historical experiments (run using forcings consistent with observed and
+comparison of simulations from different models \citep{IPCCch9}. CMIP5
+climate model output is archived at a number of different time steps
+(e.g., daily, monthly, seasonal, yearly) \citep{taylor2010cmip5}, and
+some variables are reported at multiple levels in the ocean or
+atmosphere (e.g., ocean temperature). Here, we will focus on data with a
+daily time step for variables reported at a single level (e.g.,
+near-surface air temperature).
+
+Each modeling group ran simulations under several experiments, with
+experiments varying in terms of radiative forcing by using different
+scenarios of time-varying model inputs (greenhouse gas emissions or
+concentrations, land use changes, etc.)
+\citep{taylor2012overview, IPCCch9}. Experiments include historical
+experiments (run using radiative forcing consistent with observed and
 reconstructed data for 1850--2005), pre-Industrial control experiments,
@@
 and experiment, in which they ran the experiment multiple times with the
-model with very small changes to the initial conditions, resulting in
-multiple ensemble members for a single climate model and experiment.
-CMIP5 climate model output is archived at a number of different time
-steps (e.g., daily, monthly, seasonal, yearly) \citep{taylor2010cmip5},
-and some variables are reported at multiple levels in the ocean or
-atmosphere (e.g., ocean temperature). Here, we will focus on data with a
-daily time step for variables reported at a single level (e.g.,
-near-surface air temperature).
+model with very small changes to the initial conditions.
 
@@
 variable, time step, and ensemble member
-\citep{taylor2012overview, taylor2010cmip5}. For finer time scales, this
+\citep{taylor2012overview, taylor2010cmip5}. For finer time scales, the
 output is further split across multiple files for specific year ranges
-(e.g., 5 years of output for each file) \citep{taylor2010cmip5}.
-Filenames for CMIP5 files can be parsed to generate information about
-the output variable, climate model, experiment, and ensemble member for
-the simulation \citep{taylor2010cmip5}. Files can be searched and
-downloaded through a point-and-click web interface. The data portal also
-allows you to generate a specific wget script, which you can use to
-download many files at once. A wget script can also be generated through
-Earth System Grid Federation's Search RESTful API. Tips on efficiently
-searching and downloading the data, including through use of wget
-scripts and the search API, are available as user tutorials through the
-website of the University of Colorado Boulder's Earth System CoG (e.g.,
+(e.g., 5 years of output for each file) \citep{taylor2010cmip5}. Each
+file's name includes the output variable, climate model, experiment, and
+ensemble member for the simulation \citep{taylor2010cmip5}.
+
+CMIP5 files can be searched and downloaded through a point-and-click web
+interface. They can also be downloaded in bulk to computers with Unix or
+Mac operating systems using the \code{wget} file downloading utility.
+Appropriate \code{wget} scripts can be created either through the World
+Climate Research Programme CMIP5 data portal or through Earth System
+Grid Federation's Search RESTful API. Tips on efficiently searching and
+downloading the data, including through use of wget scripts and the
+search API, are available as user tutorials through the website of the
+University of Colorado Boulder's Earth System CoG (e.g.,
 \url{https://www.earthsystemcog.org/projects/cog/doc/wget} for a
@@
 
-CMIP5 files are saved in Network Common Data Format (netCDF). NetCDF is
-a binary file format that allows storage of data representing a regular
-array. These climate model output files can be as large as several
-gigabytes {[}?{]}. For climate model output at a single level (e.g.,
-near-surface air temperature), the data is a 3-dimensional array, with
-dimensions representing time and two coordinates of location (e.g.,
-latitude and longitude) (Figure \ref{fig:netcdfexample}). Global climate
-models generate output at regularly-spaced time steps, typically at
-regularly-spaced grid points around the world. The latitude and
+CMIP5 files are saved in Network Common Data Format (netCDF), a binary
+file format that allows storage of data representing a regular array.
+For climate model output at a single level (e.g., near-surface air
+temperature), the data is a 3-dimensional array, with dimensions
+representing time and two coordinates of location (e.g., latitude and
+longitude). Figure \ref{fig:netcdfexample} provides a sketch of the
+structure of netCDF files for single-level climate model output. Global
+climate models generate output at regularly-spaced time steps, typically
+at regularly-spaced grid points around the world. The latitude and
 longitude spacing of grid points vary by climate model, but are
@@
 \citep{taylor2010cmip5}. All CMIP5 output files are required to include
-certain metadata \citep{taylor2010cmip5}. This required metadata
-includes the experiment, forcing agents input to the model to create the
-simulation, time step, institution and institutional contact
-information, climate model, and modeling realm \citep{taylor2010cmip5}.
-The metadata also must include units for all of the coordinate variables
-(e.g., longitude, latitude, time).
+certain metadata \citep{taylor2010cmip5}, including the experiment,
+forcing agents input to the model to create the simulation, time step,
+institution and institutional contact information, climate model, and
+modeling realm \citep{taylor2010cmip5}. The metadata also must include
+units for all of the coordinate variables (e.g., longitude, latitude,
+time). The netCDF format allows you to access metadata and variables
+describing the dimensions of the data without reading the full file into
+memory.
 
-\begin{widefigure}
+\begin{figure}
 \begin{center}
@@
 \label{fig:netcdfexample}
-\end{widefigure}
+\end{figure}
 
+To find out more about the CMIP climate model output data,
 \citet{taylor2012overview} and \citet{meehl2007wcrp} are excellent
-resources for finding out more about the CMIP climate model output data.
+resources.
 
@@
 
-A few R packages can be used to work with the netCDF file format used
-for CMIP5 files. Earlier packages to work with netCDF files included
-\pkg{ncdf} and \pkg{ncvar}, but these do not work with the newer netCDF
-version 4 released in 2008 and are no longer available through CRAN.
-More recent packages, including \pkg{ncdf4} \citep{ncdf4} and
-\pkg{RNetCDF} \citep{michna2013rnetcdf, RNetCDF}, work with both version
-4 and netCDF's older version 3. Climate model output data for CMIP5 is
-required to conform with the earlier version (version 3)
-\citep{taylor2010cmip5} and so should work with any of these packages,
-although it is safer to write code using functions that can be used with
-either version in case future phases of CMIP do not require files to
-conform with netCDF version 3.
-
-The netCDF format allows you to access metadata and variables describing
-the dimensions of the data without reading the full file into memory.
-The metadata describes the dimensions of the array and the units, etc.,
-of each variable and is printed by the print method of the R object
-returned by the \code{nc\_open} function of \pkg{ncdf4}. For CMIP5
-files, the dimension variables include the times and locations
-corresponding to each array in the data. Before reading in a variable
-from a netCDF file, a connection to the file must be opened, for example
-with the \code{nc\_open} function from \pkg{ncdf4}. Variables can then
-be read in using the \code{ncvar\_get} function from \pkg{ncdf4}, with
-the \code{varid} parameter set to ``lat'', ``lon'', or ``time'' (as a
-caveat, many climate models output to non-Gregorian calendars, in which
-case the time variable should be read in using a different function, as
-discussed later in this section). The climate output variable (e.g.,
-near-surface air temperature) can similarly be read in using
-\code{ncvar\_get}. In this case, the \code{varid} parameter should be
-set using the appropriate CMIP5 variable name (e.g., ``tas'' for
-near-surface air temperature); these variable names can be found in the
-CMIP Requested Output tables \citep{taylor2010cmip5}. You can use the
-dimensional time and location data to identify the location of the
-variable data you need in the netCDF array and use indexing to read only
-that data into memory, without needing to read in the full file (for
-example, with the \code{nc.get.var.subset.by.axes} function in
-\pkg{ncdf4.helpers}). Once the user is done reading in data from the
-file, the connection can be closed (e.g., with the \code{nc\_close}
-function from \pkg{ncdf4}).
-
-Since the late 1500s, Western dates have been set using the Gregorian
-calendar, which has 365.2425-day years. Some climate models, however,
-are run using different calendars, including the Julian calendar
-(365.25-day years), a calendar where there are no leap years (365-day
-years), a calendar where every year is a leap year (366-day years), and
-a calendar of twelve 30-day months (360-day years)
-\citep{cfconventions}. With these non-Gregorian calendars, R's base
-functions for converting a vector to a Date class based on the number of
-days since an origin date (\code{as.Date}, \code{as.POSIXct}) do not
-return the desired values. The \pkg{PCICt} \citep{PCICt} and
-\pkg{ncdf4.helpers} \citep{ncdf4.helpers} packages provide further
-functionality with netCDF files that can be useful when working with
-climate model output data and provide particular help in working with
-different calendars. CMIP5 netCDF files include information on the
-calendar used, which the \code{nc.get.time.series} function in
-\pkg{ncdf4.helpers} use to convert the ``time'' variable in the file to
-an object of the PCICt class, which provides Date-like functionality for
-360- and 365-day calendars \citep{PCICt}. As a note, while these
-functions will help with handling most CMIP5 files, the CMIP5 standards
-allows use of other calendars which may not be succesfully handled by
-these functions, so it is important to assess whether the time variable
-range in the PCICt object correctly matches the expected date ranges for
-a file as you process CMIP5 data in R.
-
-The size of the climate model output files can be large enough that it
-may make more sense to work with smaller chunks of the data, rather than
-reading all data into memory and working with the data all at once
-\citep{RCMIP5}. This problem aggregates when working with multiple
-climate models and more than one ensemble member for each of those
-climate models.
-
-The functions and packages described can be used to work with CMIP5
-netCDF files in R to do things like map near-surface air temperatures
-from a single climate model on a specific day in July 2075 (Figure
-\ref{fig:worldmap}) and pull a time series of daily near-surface air
-temperature simulations for the climate model grid point closest to
-Beijing, China (Figure \ref{fig:timeseries}). The
-``starting\_from\_netcdf'' vignette that comes with the
-``futureheatwaves'' package provides all code required to create these
-figures, as well as more details and code examples on working with CMIP5
-netCDF files in R.
+When working with daily climate model output data, challenges to R users
+include: (1) the file format; (2) use of non-Gregorian calendars; and
+(3) large file sizes. This section explains these challenges and offers
+some strategies for dealing with them.
 
@@
 
+CMIP5 data are available as netCDF files, for which base R import
+functions do not exist. There are, however, a few R packages that can be
+used to work with the netCDF file format used for CMIP5 files. Older
+packages include \pkg{ncdf} and \pkg{ncvar}, but these do not work with
+the newer netCDF version 4 released in 2008 and are no longer available
+through CRAN. More recent packages, including \pkg{ncdf4} \citep{ncdf4}
+and \pkg{RNetCDF} \citep{michna2013rnetcdf, RNetCDF}, work with both
+version 4 and netCDF's older version 3. While climate model output data
+for CMIP5 is required to conform with the earlier version (version 3)
+\citep{taylor2010cmip5}, it is safer to write code using functions that
+can be used with either version, in case future phases of CMIP do not
+require files to conform with netCDF version 3.
+
+You can do a number of things with netCDF files in R using these
+functions. For example, \pkg{ncdf4}'s \code{nc\_open} function can be
+used to open a connection to a netCDF file, and the object returned by
+the function includes the file's attribute data. Once a file connection
+is open, variables can be read in using the \code{ncvar\_get} function.
+For example, the variables defining the dimensions of the sketched
+netCDF file in Figure \ref{fig:netcdfexample} could be read into R with
+\code{ncvar\_get} with the \code{varid} parameter set to ``lat'',
+``lon'', or ``time''. The climate output variable (e.g., near-surface
+air temperature) can similarly be read in using \code{ncvar\_get}. In
+this case, the \code{varid} parameter should be set using the
+appropriate CMIP5 variable name (e.g., ``tas'' for near-surface air
+temperature); these variable names can be found in the CMIP Requested
+Output tables \citep{taylor2010cmip5}. If you only need data from a
+subset of the full file, you can use the dimensional time and location
+data to identify the location of the variable data you need in the
+netCDF array and use indexing to read only that data into memory,
+without needing to read in the full file (for example, with the
+\code{nc.get.var.subset.by.axes} function in \pkg{ncdf4.helpers}). Once
+the user is done reading in data from the file, the connection can be
+closed (e.g., with the \code{nc\_close} function from \pkg{ncdf4}).
+
+A second challenge when working with climate model output data in R is
+that some climate models output to non-Gregorian calendars. Since the
+late 1500s, Western dates have been set using the Gregorian calendar,
+which has 365.2425-day years. Some climate models, however, are run
+using different calendars, including the Julian calendar (365.25-day
+years), a calendar where there are no leap years (365-day years), a
+calendar where every year is a leap year (366-day years), and a calendar
+of twelve 30-day months (360-day years) \citep{cfconventions}. With
+these non-Gregorian calendars, R's base functions for converting a
+vector to a Date class based on the number of days since an origin date
+(\code{as.Date}, \code{as.POSIXct}) do not return the desired values.
+
+Two R packages provide help with non-Gregorian calendars: \pkg{PCICt}
+\citep{PCICt} and \pkg{ncdf4.helpers} \citep{ncdf4.helpers}. The
+\code{nc.get.time.series} function in \pkg{ncdf4.helpers} pulls and uses
+metadata on the calendar stored in the CMIP5 netCDF file's attributes to
+convert the ``time'' variable in the file to an object of the
+\code{PCICt} class. This class is defined in the \pkg{PCICt} package and
+provides Date-like functionality for 360- and 365-day calendars
+\citep{PCICt}. However, while these functions will help with handling
+most CMIP5 files, the CMIP5 standards allows use of other calendars
+which may not be succesfully handled by these functions, so it is
+important to assess whether the time variable range in the \code{PCICt}
+object correctly matches the expected date ranges for a file when
+processing CMIP5 data in R.
+
+Finally, the size of CMIP5 files can make them difficult to work with in
+R. CMIP5 climate model output files can be as large as several
+gigabytes. The size of the files can therefore be large enough that it
+may make more sense to work with smaller chunks of the data in R, rather
+than reading all data into memory and working with the data all at once
+\citep{RCMIP5}. This problem aggregates when working with multiple
+climate models and more than one ensemble member for each of those
+climate models.
+
 In addition to these general packages for working with netCDF files,
@@
 netCDF files, scan a directory of CMIP5 files and determine models with
-continuous available data, create objects of a special ``cmip5data''
+continuous available data, create objects of a special \code{cmip5data}
 class to work with CMIP5 data within R, and parse the filenames for all
@@
 this package, most functions only work with monthly or less frequent
-data (e.g., the functions \code{checkTimePeriod} and \code{cmip5data},
-as well as functions that work with \code{cmip5data} objects including
-\code{filterDimensions} and \code{getProjectionMatrix}) \citep{RCMIP5}.
-While the \code{loadCMIP5} function does successfully load the daily
-data as a \code{cmip5data} object, most of the methods for this object
-type do not do anything meaningful for daily data. The package's
-\code{getFileInfo} function, however, will work with CMIP5 files of any
-time step; this function identifies all CMIP5 files in a directory and
-creates a dataframe that parses the information contained in the file
-name. As a note, the \code{get.split.filename.cmip5} function in the
-\pkg{ncdf4.helper} package similarly can be used to parse information
-contained in CMIP5 file names \citep{ncdf4.helpers}.
+data \citep{RCMIP5}. While the \code{loadCMIP5} function does
+successfully load daily data as a \code{cmip5data} object, most of the
+methods for this object type do not do anything meaningful for daily
+data. The package's \code{getFileInfo} function, however, will work with
+CMIP5 files of any time step; this function identifies all CMIP5 files
+in a directory and creates a dataframe with information parsed from the
+file name. As a note, the \code{get.split.filename.cmip5} function in
+the \pkg{ncdf4.helper} package similarly can be used to parse
+information contained in CMIP5 file names \citep{ncdf4.helpers}.
 
@@
 to download CMIP5 monthly-aggregated output directly from within R with
-the \code{CMIP5fromESGF} function. However, this function does not allow
-downloading of climate model output with finer time steps, like daily
-data. This package uses the \code{models2wux} function to read in
-climate model output netCDF files and convert to a WUX dataframe that
-can be used by other functions in the package. While this function can
-input climate model output with daily time steps, if the element
-``what.timesteps'' of the \code{modelinput} list input is set to
-``daily'', the function aggregates this data to a monthly or less
-frequent (e.g., seasonal) aggregation when creating the WUX dataframe.
-Therefore, while this package provides very useful functionality for
-working with averaged output of daily climate model output data, it
-cannot easily be used to identify and characterize multi-day extreme
-events like heat waves.
+the \code{CMIP5fromESGF} function. The package then uses the
+\code{models2wux} function to read in climate model output netCDF files
+and convert them to ``WUX'' dataframes, which can be used by other
+functions in the package. While this function can input climate model
+output with daily time steps (the ``what.timesteps'' element of the
+\code{modelinput} list input must be set to ``daily''), the function
+aggregates this data to a monthly or less frequent (e.g., seasonal)
+aggregation when creating the WUX dataframe. Therefore, while this
+package provides very useful functionality for working with averaged
+output of daily climate model output data, it cannot easily be used to
+identify and characterize multi-day extreme events like heat waves.
 
-More tips on working with CMIP5 files in R are provided in the
-``starting\_from\_netcdf'' vignette of the \pkg{futureheatwaves}
-package.
+The functions and packages described in this section can be used with
+CMIP5 netCDF files to do things in R like map near-surface air
+temperatures from a single climate model on a specific day (Figure
+\ref{fig:worldmap}) or pull a time series of daily near-surface air
+temperature simulations at a specific climate model grid point (Figure
+\ref{fig:timeseries}). The ``starting\_from\_netcdf'' vignette that
+comes with the ``futureheatwaves'' package provides all code required to
+create these figures, as well as more details and code examples on
+working with CMIP5 netCDF files in R.
 
@@
 characteristics of each identified extreme event (Table
-\ref{tab:hwcharacteristics}). The extreme events are identified and
-characterized at one or more study locations (e.g., cities), with study
-locations specified by the user in an input file. The extreme events
+\ref{tab:hwcharacteristics}). This package start from files rather than
+R objects to avoid loading data from all climate model ensembles at
+once; instead, the function loads, processes, and saves output for a
+single climate model ensemble member at a time. The extreme events are
+identified and characterized at one or more study locations (e.g.,
+cities), which the user specifies in an input file. The extreme events
 identified for each ensemble member are output as separate files in a
@@
 Once the user creates these data frames of location-specific heat waves,
-the \code{apply\_all\_models} function allows the user to apply custom
+the \code{apply\_all\_models} function can be used to apply custom
 functions across all the extreme event data frames generated by
-\code{gen\_hw\_set} to summaries of extreme events across all climate
-models and ensemble members (Figure \ref{fig:overview}, right). This
-functionality can be used to generate summary statistics (e.g.,
-determine average heat wave length or total frost days), or can be used
-to apply more complex functions (e.g., apply epidemiologic effect
-estimates across the heat waves to generate health impact estimates).
+\code{gen\_hw\_set}. This functionality allows users to create summaries
+of extreme events across all climate models and ensemble members (Figure
+\ref{fig:overview}, right). The function can be used to generate summary
+statistics (e.g., determine average heat wave length or total frost
+days) or to apply more complex functions (e.g., apply epidemiologic
+effect estimates across the heat waves to generate health impact
+estimates).
+
+When using this package, CMIP5 climate model output data requires some
+pre-processing. Data will need to be saved in a specific format, with
+files stored in a specific directory structure. Full details of the
+required file and directory structure are provided in the package's
+``futureheatwaves'' vignette, while tips and an R script for conducting
+this processing starting from CMIP5 netCDF files are given in the
+``starting\_from\_netcdf'' vignette.
+
+This package can be used for study locations worldwide. The
+``starting\_from\_netcdf''" package vignette provides an example of
+using this package to identify and explore future heat waves in several
+Chinese cities.
 
@@
 \citep{gent2011community}. We include one ensemble member from BCC
-(r1i1p1) and two from CCSM (r1i1p1 and r2i1p1).
+(r1i1p1) and two from CCSM (r1i1p1 and r2i1p1). Once the
+\pkg{futureheatwaves} package is installed and loaded, the user can find
+the local location of these files using R's \code{system.file} function.
 
@@
 
-Once the \pkg{futureheatwaves} package is installed and loaded, the user
-can find the local location of these files using R's \code{system.file}
-function. In the later sections of this article, we show how to use the
-package functions with these example files as inputs.
-
-When using this package with data other than this example data, the
-climate model output data will require some pre-processing. Data will
-need to be saved in a specific format, with files stored in a specific
-directory structure. This preprocessing can be done using some of the R
-packages described earlier for working with CMIP5 data in R. Full
-details of the required file and directory structure are provided in the
-package's ``futureheatwaves'' vignette, while tips on conducting this
-processing starting from CMIP5 netCDF files are given in the
-``starting\_from\_netcdf'' vignette.
-
-\subsection{Basic example of identifying and characterizing extreme
-events}\label{basic-example-of-identifying-and-characterizing-extreme-events}
+\subsection{\texorpdfstring{Basic example of using
+\pkg{futureheatwaves}}{Basic example of using }}\label{basic-example-of-using}
 
@@
 (\code{city\_file\_location}). The \code{gen\_hw\_set} function
-processes this example input and creates a new directory,
+processes the example input and creates a new directory,
 \file{example\_results}, with files of identified and characterized heat
@@
 identified and classified extreme events for each ensemble member,
-including all characteristics in Table \ref{tab:hwcharacteristics}.
-
-\subsection{Customizing the extreme event
-definition}\label{customizing-the-extreme-event-definition}
-
-By default, the package identifies extreme events in climate model
-output data using a specific definition for heat waves:
-
-\begin{quote}
-A \dfn{heat wave} is two or more days at or above a city-specific
-threshold temperature, with the threshold determined as the
-98\textsuperscript{th} percentile of year-round temperature in the city
-during some reference period (by default, 1990--1999).
-\end{quote}
-
-\noindent However, the user can extensively customize the definition
-used to identify extreme events. Users can write a custom R function
-with either a different heat wave definition (see \citet{smith2013heat}
-and \citet{kent2014heat} for listings of definitions used in scientific
-studies) or with a definition appropriate for a different type of
-extreme event (e.g., one or more days at or below 32\textsuperscript{o}F
-/ 0\textsuperscript{o}C for frost day spells). For heat wave
-identification, researchers might want to use a specific definition, for
-example, because it matches the definition used by local health
-officials to declare heat wave warnings or, in the case of health impact
-assessments, to match with a definition used in an epidemiological
-study. For other extreme events, a heat wave definition likely will not
-be applicable and so a customized definition is necessary.
-
-Three components of the extreme event definition can be easily
-customized in the \code{gen\_hw\_set} function call, without creating a
-new R function to use to identify heat waves. First, often users wish to
-define an extreme event based on conditions that are rare in the study
-location \citep{IPCCch1}. In this case, the percentile of the variable
-of interest required for an extreme event can be changed using the
-\code{probThreshold} option in \code{gen\_hw\_set}. For example, this
-option could be used to identify heat waves based on the
-99\textsuperscript{th} percentile, rather than the
-98\textsuperscript{th} percentile, of a city's year-round temperature.
-Second, the user can change the number of days used in the heat wave
-definition using the \code{numDays} argument in the \code{gen\_hw\_set}
-function (the default is \samp{numDays = 2}, corresponding to two or
-more days). Combined, these two customization choices allow the user to
-identify heat waves using many of the heat wave definitions used in
-previous climate and health research-- for example, 9 of the 16 heat
-wave definitions used in \citet{kent2014heat} (e.g., {[}some examples of
-the 9 that could be fit{]}) could be fit using different combinations of
-these two options for specifying threshold percentile and number of
-days. Third, some extreme events like cold waves and frost day spells
-are defined as a certain number of days below, rather than above, a
-threshold. While the default is to identify events by searching for days
-above a threshold, this behavior can be changed with the
-\code{above\_threshold = FALSE} argument in the
-\code{gen\_hw\_set function}.
-
-Beyond these simpler options, the customization of the event definition
-is even more extensive as one has the option of writing a custom R
-function. This functionality allows the user to use heat wave
-definitions that either require a number of days above an absolute
-threshold (e.g., maximum temperature of \(\ge\) 95\textsuperscript{o}F
-for \(\ge1\) day \citet{kent2014heat, tan2007heat}; minimum temperature
-\(\le\) 0\textsuperscript{o}C for \(\ge1\) day for frost day spells) or
-that require a combination of thresholds to be met (e.g., maximum daily
-temperature one threshold every day of the heat wave and above a higher
-threshold for a certain number of days;
-\citet{kent2014heat, peng2011toward}). To use a customized event
-definition, the user can write and load an R function that implements
-the definition. The custom function is passed to the \code{gen\_hw\_set}
-function using the \code{IDheatwavesFunction} argument. To work
-correctly, this custom function must allow only specific inputs and
-generate only specific outputs. Details about this required structure
-are provided in the \pkg{futureheatwaves} package vignette. To increase
-processing speed when identifying extreme events, we coded parts of the
-default event definition function in C++ and synced it with R using the
-\CRANpkg{Rcpp} package \citep{Rcpp}. Users should consider a similar
-strategy for custom heat wave definitions, especially if processing a
-large number of climate projection files.
-
-Finally, extreme events are often based on conditions that are rare for
-a specific location using location-specific relative thresholds. These
-thresholds are often defined for climate impact studies based on a
-variable's distribution at that location in present-day or historical
-data. However, it can be interesting to explore trends in extreme events
-under climate change if extreme events are identified based on variable
-distributions during the projection period, or another future period. It
-is possible to specify the time period to use when determining a
-relative threshold for an event definition using the
-\code{thresholdBoundaries} argument in the function \code{gen\_hw\_set}.
-
-\subsection{Exploring extreme events}\label{exploring-extreme-events}
+including all characteristics in Table \ref{tab:hwcharacteristics}. See
+the package's vignettes for more details on the content and structure of
+this output.
 
@@
 allowing it to be used easily with the graphing package
-\CRANpkg{ggplot2} \citep{ggplot2}.
+\CRANpkg{ggplot2} \citep{ggplot2}. This functionality can be used to
+create a number of other summaries of the identified extreme events. For
+example, it could be used to determine average heat wave length or
+estimate how much earlier in the year events are expected to start
+across an ensemble of climate model simulations. The functionality can
+also be used for more complex analysis of extreme event files. For
+example, it can be used to apply epidemiological models of heat wave to
+estimate excess heat-related mortality under different future scenarios;
+an example of this application is provided in the ``futureheatwaves''
+vignette that comes with the package.
 
-The \code{apply\_all\_models} function can also be used to project the
-health impacts of heat waves. A simple estimate of excess deaths
-associated with heat wave risk in a community can be calculated as
-\citep{peng2011toward}:
+\subsection{Customizing the extreme event
+definition}\label{customizing-the-extreme-event-definition}
 
-\begin{equation}
-E_c = B_c * (RR - 1) * L_c 
-\label{eq:healthimpacts}
-\end{equation}
+By default, the package identifies extreme events in climate model
+output data using a specific definition for heat waves that has been
+used in some epidemiological and climate impact research (e.g.,
+\citet{anderson2009weather}):
 
-where:
+\begin{quote}
+A \dfn{heat wave} is two or more days at or above a city-specific
+threshold temperature, with the threshold determined as the
+98\textsuperscript{th} percentile of year-round temperature in the city
+during some reference period (by default, 1990--1999).
+\end{quote}
 
-\begin{itemize}
-\item $E_c$ is the total number of excess deaths in community $c$;
-\item $B_c$ is the baseline average daily mortality in community $c$; 
-\item $RR$ is the relative risk of mortality associated with a heat wave day; and 
-\item $L_c$ is the total number of heat wave days in community $c$ over the study period.
-\end{itemize}
+However, this is not the only accepted heat wave definition in the
+scientific literature. A variety of different heat wave definitions have
+been used to identify heat waves in a time series of temperature data
+\citep{smith2013heat}, and the choice of heat wave definitions can
+influence both projected heat wave trends \citep{smith2013heat} and
+estimates of health risks during events
+\citep{chen2015influence, kent2014heat}. Further, other types of extreme
+events will be defined differently than heat waves (for example, frost
+day spells may be defined as one or more days with temperature at or
+below 32\textsuperscript{o}F / 0\textsuperscript{o}C).
 
-This impact assessment equation can be translated into a function that
-merges each projection's heat wave data frame with a data frame of
-community-specific baseline mortality rates (\(B_c\)), calculates
-equation \ref{eq:healthimpacts} for each heat wave, and then sums up the
-total excess deaths across all heat waves. In this example, we have set
-the default relative risk for a heat wave as 1.032, based on the
-estimated added heat wave relative risk in 107 U.S. communities for heat
-waves defined as two or more days at or above a community's
-98\textsuperscript{th} percentile temperature
-\citep{anderson2009weather}.
+Therefore, this package allows the user to extensively customize the
+definition used to identify extreme events. Users can write a custom R
+function with either a different heat wave definition (see
+\citet{smith2013heat} and \citet{kent2014heat} for listings of
+definitions used in scientific studies) or with a definition appropriate
+for a different type of extreme event (e.g., one or more days at or
+below 32\textsuperscript{o}F / 0\textsuperscript{o}C for frost day
+spells). For heat wave identification, researchers might want to use a
+specific definition, for example, because it matches the definition used
+by local health officials to declare heat wave warnings or, in the case
+of health impact assessments, to match with a definition used in an
+epidemiological study. For other extreme events, a heat wave definition
+likely will not be applicable and so a customized definition is
+necessary.
 
-\begin{verbatim}
-excess_deaths <- function(hw_datafr, base_mortality, RR = 1.032){
-        hw_datafr <- dplyr::left_join(hw_datafr, base_mortality,
-                                      by = "city") %>%
-                dplyr::mutate(excess_deaths = base_mort * (RR - 1) * length)
-        out <- sum(hw_datafr$excess_deaths)
-        out <- round(out)
-        return(out)
-}
-\end{verbatim}
+Three components of the extreme event definition can be easily
+customized in the \code{gen\_hw\_set} function call, without creating a
+new R function to use to identify heat waves. First, often users wish to
+define an extreme event based on conditions that are rare in the study
+location \citep{IPCCch1}. In this case, the percentile of the variable
+of interest required for an extreme event can be changed using the
+\code{probThreshold} option in \code{gen\_hw\_set}. For example, this
+option could be used to identify heat waves based on the
+99\textsuperscript{th} percentile, rather than the
+98\textsuperscript{th} percentile, of a city's year-round temperature.
+Second, the user can change the number of days used in the heat wave
+definition using the \code{numDays} argument in the \code{gen\_hw\_set}
+function (the default is \samp{numDays = 2}, corresponding to two or
+more days). Combined, these two customization choices allow the user to
+identify heat waves using many of the heat wave definitions used in
+previous climate and health research-- for example, 9 of the 16 heat
+wave definitions used in \citet{kent2014heat} (e.g., {[}some examples of
+the 9 that could be fit{]}) could be fit using different combinations of
+these two options for specifying threshold percentile and number of
+days. Third, some extreme events like cold waves and frost day spells
+are defined as a certain number of days below, rather than above, a
+threshold. While the default is to identify events by searching for days
+above a threshold, this behavior can be changed with the
+\code{above\_threshold = FALSE} argument in the
+\code{gen\_hw\_set function}.
 
-\noindent Once defined in R, this function can be applied across all
-heat waves from all climate models' ensemble members, provided that you
-have a data frame called \code{city\_mortality} with columns with each
-community's identifier (\code{city}) and baseline mortality rate
-(\code{base\_mort}), using the call (this example uses hypothetical
-baseline mortality rates for the cities and requires the \CRANpkg{dplyr}
-package \citep{dplyr}):
+Beyond these simpler options, the customization of the event definition
+is even more extensive as one has the option of writing a custom R
+function. This functionality allows the user to use heat wave
+definitions that either require a number of days above an absolute
+threshold (e.g., maximum temperature of \(\ge\) 95\textsuperscript{o}F
+for \(\ge1\) day \citet{kent2014heat, tan2007heat}; minimum temperature
+\(\le\) 0\textsuperscript{o}C for \(\ge1\) day for frost day spells) or
+that require a combination of thresholds to be met (e.g., maximum daily
+temperature one threshold every day of the heat wave and above a higher
+threshold for a certain number of days;
+\citet{kent2014heat, peng2011toward}). To use a customized event
+definition, the user can write and load an R function that implements
+the definition. The custom function is passed to the \code{gen\_hw\_set}
+function using the \code{IDheatwavesFunction} argument. To work
+correctly, this custom function must allow only specific inputs and
+generate only specific outputs. Details about this required structure
+are provided in the \pkg{futureheatwaves} package vignette. To increase
+processing speed when identifying extreme events, we coded parts of the
+default event definition function in C++ and synced it with R using the
+\CRANpkg{Rcpp} package \citep{Rcpp}. Users should consider a similar
+strategy for custom heat wave definitions, especially if processing a
+large number of climate projection files.
 
-\begin{Schunk}
-\begin{Sinput}
-library(dplyr)
-city_mortality <- data.frame(city = c("balt", "nwk", "ny", "phil", "prov"),
-                             base_mort = c(3, 2.5, 35, 6, 2.5))
-apply_all_models(out = out, FUN = excess_deaths, base_mortality = city_mortality)
-\end{Sinput}
-\begin{Soutput}
-#>   model ensemble value
-#> 1  bcc1        1   930
-#> 2  ccsm        1   968
-#> 3  ccsm        2   967
-\end{Soutput}
-\end{Schunk}
+Finally, extreme events are often based on conditions that are rare for
+a specific location using location-specific relative thresholds. These
+thresholds are often defined for climate impact studies based on a
+variable's distribution at that location in present-day or historical
+data. However, it can be interesting to explore trends in extreme events
+under climate change if extreme events are identified based on variable
+distributions during the projection period, or another future period. It
+is possible to specify the time period to use when determining a
+relative threshold for an event definition using the
+\code{thresholdBoundaries} argument in the function \code{gen\_hw\_set}.
 
-\noindent While the above impact analysis is simplistic, it provides an
-example of how the \code{apply\_all\_models} function can be used in
-conjunction with epidemiologic results to estimate the potential health
-impacts of extreme events with climate change.
-
-\section{Customizing dates for projections and reference
+\subsection{Customizing dates for projections and reference
 temperatures}\label{customizing-dates-for-projections-and-reference-temperatures}

### 2017-01-04 paper_drafts/RJournalRevisions/anderson-eason-barnes.Rmd
Update article draft

@@
 
-The CMIP5 climate model output data is distributed across data nodes at different climate modeling centers \citep{taylor2012overview}, but can be accessed centrally at the World Climate Research Programme CMIP5 data portal at https://pcmdi.llnl.gov/search/cmip5/.
+The CMIP5 climate model output data are distributed across data nodes at different climate modeling centers \citep{taylor2012overview}, but can be accessed centrally at the World Climate Research Programme CMIP5 data portal at https://pcmdi.llnl.gov/search/cmip5/.
 Users must register before downloading data, and some data are restricted to non-commercial use.
@@
 \end{center}
-\caption{Example of structure of a NetCDF climate model output file for a variable reported at a single depth, like surface air temperature.
-Data are stored in a three-dimensional array, with measurements at each time step and grid location.
-Surface temperature data are typically indexed in climate model output files by longitude, latitude, and time, in that order.
-For example, if the air surface temperature ("tas") is read into an R object called \code{tas}, you can access the value for the first day at the fourth longitude and third latitude with \code{tas[4, 3, 1]}.
+\caption{Example of structure of a NetCDF climate model output file for a variable reported at a single level, like near-surface air temperature.
+Data are stored in a three-dimensional array, with measurements at each time step and climate grid location.
+These data are typically indexed in the netCDF file by longitude, latitude, and time, in that order.
+For example, if the near-surface air temperature is read into an R object called \code{tas}, you can access the value for the first day at the fourth longitude and third latitude with \code{tas[4, 3, 1]}.
 In addition to the output variable (temperature in this example), vectors with the ordered values of each dimension (longitude, latitude, and time) can also be read in from the netCDF file, as well as attribute data (e.g., units for variables, the calendar used for time).}
@@
 This section explains these challenges and offers some strategies for dealing with them.
+
+CMIP5 data are available as netCDF files, for which base R import functions do not exist.
+There are, however, a few R packages that can be used to work with the netCDF file format used for CMIP5 files.
+Older packages include \pkg{ncdf} and \pkg{ncvar}, but these do not work with the newer netCDF version 4 released in 2008 and are no longer available through CRAN.
+More recent packages, including \pkg{ncdf4} \citep{ncdf4} and \pkg{RNetCDF} \citep{michna2013rnetcdf, RNetCDF}, work with both version 4 and netCDF's older version 3.
+While climate model output data for CMIP5 are required to conform with the earlier version (version 3) \citep{taylor2010cmip5}, it is safer to write code using functions that can be used with either version, in case future phases of CMIP do not require files to conform with netCDF version 3.
+
+You can do a number of things with netCDF files in R using these functions.
+For example, \pkg{ncdf4}'s \code{nc\_open} function can be used to open a connection to a netCDF file, and the object returned by the function includes the file's attribute data.
+Once a file connection is open, variables can be read in using the \code{ncvar\_get} function.
+For example, the variables defining the dimensions of the sketched netCDF file in Figure \ref{fig:netcdfexample} could be read into R with \code{ncvar\_get} with the \code{varid} parameter set to "lat", "lon", or "time".
+The climate output variable (e.g., near-surface air temperature) can similarly be read in using \code{ncvar\_get}.
+In this case, the \code{varid} parameter should be set using the appropriate CMIP5 variable name (e.g., "tas" for near-surface air temperature); these variable names can be found in the CMIP Requested Output tables \citep{taylor2010cmip5}.
+If you only need data from a subset of the full file, you can use the dimensional time and location data to identify the location of the variable data you need in the netCDF array and use indexing to read only that data into memory, without needing to read in the full file (for example, with the \code{nc.get.var.subset.by.axes} function in \pkg{ncdf4.helpers}).
+Once the user is done reading in data from the file, the connection can be closed (e.g., with the \code{nc\_close} function from \pkg{ncdf4}).
 
@@
 \end{figure}
-
-CMIP5 data are available as netCDF files, for which base R import functions do not exist.
-There are, however, a few R packages that can be used to work with the netCDF file format used for CMIP5 files.
-Older packages include \pkg{ncdf} and \pkg{ncvar}, but these do not work with the newer netCDF version 4 released in 2008 and are no longer available through CRAN.
-More recent packages, including \pkg{ncdf4} \citep{ncdf4} and \pkg{RNetCDF} \citep{michna2013rnetcdf, RNetCDF}, work with both version 4 and netCDF's older version 3.
-While climate model output data for CMIP5 is required to conform with the earlier version (version 3) \citep{taylor2010cmip5}, it is safer to write code using functions that can be used with either version, in case future phases of CMIP do not require files to conform with netCDF version 3.
-
-You can do a number of things with netCDF files in R using these functions.
-For example, \pkg{ncdf4}'s \code{nc\_open} function can be used to open a connection to a netCDF file, and the object returned by the function includes the file's attribute data.
-Once a file connection is open, variables can be read in using the \code{ncvar\_get} function.
-For example, the variables defining the dimensions of the sketched netCDF file in Figure \ref{fig:netcdfexample} could be read into R with \code{ncvar\_get} with the \code{varid} parameter set to "lat", "lon", or "time".
-The climate output variable (e.g., near-surface air temperature) can similarly be read in using \code{ncvar\_get}.
-In this case, the \code{varid} parameter should be set using the appropriate CMIP5 variable name (e.g., "tas" for near-surface air temperature); these variable names can be found in the CMIP Requested Output tables \citep{taylor2010cmip5}.
-If you only need data from a subset of the full file, you can use the dimensional time and location data to identify the location of the variable data you need in the netCDF array and use indexing to read only that data into memory, without needing to read in the full file (for example, with the \code{nc.get.var.subset.by.axes} function in \pkg{ncdf4.helpers}).
-Once the user is done reading in data from the file, the connection can be closed (e.g., with the \code{nc\_close} function from \pkg{ncdf4}).
 
@@
 While the \code{loadCMIP5} function does successfully load daily data as a \code{cmip5data} object, most of the methods for this object type do not do anything meaningful for daily data.
-The package's \code{getFileInfo} function, however, will work with CMIP5 files of any time step; this function identifies all CMIP5 files in a directory and creates a dataframe with information parsed from the file name.
+The package's \code{getFileInfo} function, however, will work with CMIP5 files of any time step; this function identifies all CMIP5 files in a directory and creates a data frame with information parsed from the file name.
 As a note, the \code{get.split.filename.cmip5} function in the \pkg{ncdf4.helper} package similarly can be used to parse information contained in CMIP5 file names \citep{ncdf4.helpers}.
@@
 The \pkg{wux} package \citep{wux} includes functions that allow the user to download CMIP5 monthly-aggregated output directly from within R with the \code{CMIP5fromESGF} function.
-The package then uses the \code{models2wux} function to read in climate model output netCDF files and convert them to "WUX" dataframes, which can be used by other functions in the package.
-While this function can input climate model output with daily time steps (the "what.timesteps" element of the \code{modelinput} list input must be set to "daily"), the function aggregates this data to a monthly or less frequent (e.g., seasonal) aggregation when creating the WUX dataframe.
+The package then uses the \code{models2wux} function to read in climate model output netCDF files and convert them to "WUX" data frames, which can be used by other functions in the package.
+While this function can input climate model output with daily time steps (the "what.timesteps" element of the \code{modelinput} list input must be set to "daily"), the function aggregates this data to a monthly or less frequent (e.g., seasonal) aggregation when creating the WUX data frame.
 Therefore, while this package provides very useful functionality for working with averaged output of daily climate model output data, it cannot easily be used to identify and characterize multi-day extreme events like heat waves.
@@
 The package takes a directory with climate projection files (left), for one or more climate models, with one or more ensemble members for each climate model (this example figure shows four climate models with one or two ensemble members each).
-The gen\_hw\_set function processes these files to create a data frame for each ensemble member, identifying and characterizing all multi-day extreme events (e.g., heat waves) in the time series projection for that ensemble member.
-The apply\_all\_models function allows users to explore these extreme events by applying user-created functions across all the extreme event data frames, creating a summary data frame with results.}
+The \code{gen\_hw\_set} function processes these files to create a data frame for each ensemble member, identifying and characterizing all multi-day extreme events (e.g., heat waves) in the time series projection for that ensemble member.
+The \code{apply\_all\_models} function allows users to explore these extreme events by applying user-created functions across all the extreme event data frames, creating a summary data frame with results.}
 \label{fig:overview}
@@
 
-When using this package, CMIP5 climate model output data requires some pre-processing.
+When using this package, CMIP5 climate model output data require some pre-processing.
 Data will need to be saved in a specific format, with files stored in a specific directory structure.
@@
 We have included data files in the package to serve as example files so that users can try this package before applying it to their own directory of climate projection files.
-This example data comes from two climate models that are a part of CMIP5: (1) the model of the Beijing Climate Center, China Meteorological Administration (BCC) \citep{xin2013introduction} and (2) the National Center for Atmospheric Research's (NCAR's) Community Climate System Model, version 4 (CCSM4) \citep{gent2011community}.
+These example data come from two climate models that are a part of CMIP5: (1) the model of the Beijing Climate Center, China Meteorological Administration (BCC) \citep{xin2013introduction} and (2) the National Center for Atmospheric Research's (NCAR's) Community Climate System Model, version 4 (CCSM4) \citep{gent2011community}.
 We include one ensemble member from BCC (r1i1p1) and two from CCSM (r1i1p1 and r2i1p1).
@@
 
-However, this is not the only accepted heat wave definition in the scientific literature.
-A variety of different heat wave definitions have been used to identify heat waves in a time series of temperature data \citep{smith2013heat}, and the choice of heat wave definitions can influence both projected heat wave trends \citep{smith2013heat} and estimates of health risks during events \citep{chen2015influence, kent2014heat}.
+\noindent However, this is not the only accepted heat wave definition in the scientific literature.
+A variety of different heat wave definitions have been used to identify heat waves in a time series of temperature data \citep{smith2013heat, kent2014heat, chen2015influence, anderson2009weather}, and the choice of heat wave definitions can influence both projected heat wave trends \citep{smith2013heat} and estimates of health risks during events \citep{chen2015influence, kent2014heat, anderson2009weather}.
 Further, other types of extreme events will be defined differently than heat waves (for example, frost day spells may be defined as one or more days with temperature at or below 32\textsuperscript{o}F / 0\textsuperscript{o}C).
@@
 Therefore, this package allows the user to extensively customize the definition used to identify extreme events.
-Users can write a custom R function with either a different heat wave definition (see \citet{smith2013heat} and \citet{kent2014heat} for listings of definitions used in scientific studies) or with a definition appropriate for a different type of extreme event (e.g., one or more days at or below 32\textsuperscript{o}F / 0\textsuperscript{o}C for frost day spells).
+Users can write a custom R function with either a different heat wave definition (see \citet{smith2013heat} and \citet{kent2014heat} for listings of some of the definitions used in scientific studies) or with a definition appropriate for a different type of extreme event (e.g., one or more days at or below 32\textsuperscript{o}F / 0\textsuperscript{o}C for frost day spells).
 For heat wave identification, researchers might want to use a specific definition, for example, because it matches the definition used by local health officials to declare heat wave warnings or, in the case of health impact assessments, to match with a definition used in an epidemiological study.
-For other extreme events, a heat wave definition likely will not be applicable and so a customized definition is necessary.
+For studies of other extreme events, a heat wave definition likely will not be applicable and so a customized definition is necessary.
 
 Three components of the extreme event definition can be easily customized in the \code{gen\_hw\_set} function call, without creating a new R function to use to identify heat waves.
-First, often users wish to define an extreme event based on conditions that are rare in the study location \citep{IPCCch1}.
-In this case, the percentile of the variable of interest required for an extreme event can be changed using the \code{probThreshold} option in \code{gen\_hw\_set}.
-For example, this option could be used to identify heat waves based on the 99\textsuperscript{th} percentile, rather than the 98\textsuperscript{th} percentile, of a city's year-round temperature.
-Second, the user can change the number of days used in the heat wave definition using the \code{numDays} argument in the \code{gen\_hw\_set} function (the default is \samp{numDays = 2}, corresponding to two or more days).
-Combined, these two customization choices allow the user to identify heat waves using many of the heat wave definitions used in previous climate and health research-- for example, 9 of the 16 heat wave definitions used in \citet{kent2014heat} (e.g., [some examples of the 9 that could be fit]) could be fit using different combinations of these two options for specifying threshold percentile and number of days.
+First, many extreme event definitions are based on conditions that are rare in the study location \citep{IPCCch1}, but may vary in how rare conditions must be for a period to qualify as an extreme event.
+For example, some of the different definitions used to identify heat waves vary only in the percentile temperature used for a threshold (e.g., one definition is $\ge2$ days at or above the 98\textsuperscript{th} percentile temperature at a location while another is $\ge2$ days at or above the 99\textsuperscript{th} percentile temperature; \citet{kent2014heat, smith2013heat}).
+Therefore, the \pkg{futureheatwaves} package allows users to change the percentile of the variable of interest required for an extreme event using the \code{probThreshold} option in \code{gen\_hw\_set}.
+Other heat wave definitions vary only in the number of consecutive days that must be over the threshold for a period to quality as an extreme event (e.g., one definition is $\ge2$ days at or above the 98\textsuperscript{th} percentile temperature at a location while another is $\ge4$ days at or above the 98\textsuperscript{th} percentile temperature; \citet{anderson2009weather}).
+Therefore, the package allows the user to change the number of days used in the heat wave definition using the \code{numDays} argument in the \code{gen\_hw\_set} function.
+Combined, these two customization choices allow the user to identify heat waves using many of the heat wave definitions used in previous climate and health research-- for example, 9 of the 16 heat wave definitions used in \citet{kent2014heat} could be fit using different combinations of these two options for specifying threshold percentile and number of days.
 Third, some extreme events like cold waves and frost day spells are defined as a certain number of days below, rather than above, a threshold.
@@
 
-Beyond these simpler options, the customization of the event definition is even more extensive as one has the option of writing a custom R function.
-This functionality allows the user to use heat wave definitions that either require a number of days above an absolute threshold (e.g., maximum temperature of $\ge$ 95\textsuperscript{o}F for $\ge1$ day \citet{kent2014heat, tan2007heat}; minimum temperature $\le$ 0\textsuperscript{o}C for $\ge1$ day for frost day spells) or that require a combination of thresholds to be met (e.g., maximum daily temperature one threshold every day of the heat wave and above a higher threshold for a certain number of days; \citet{kent2014heat, peng2011toward}).
-To use a customized event definition, the user can write and load an R function that implements the definition.
-The custom function is passed to the \code{gen\_hw\_set} function using the \code{IDheatwavesFunction} argument.
-To work correctly, this custom function must allow only specific inputs and generate only specific outputs.
-Details about this required structure are provided in the \pkg{futureheatwaves} package vignette.
+Beyond these simpler options, the customization of the event definition is even more extensive as one has the option of writing and using a custom R function to identify extreme events.
+This functionality allows the user to use definitions that either require a number of days above an absolute threshold (e.g., maximum temperature of $\ge$ 95\textsuperscript{o}F for $\ge1$ day \citet{kent2014heat, tan2007heat}; minimum temperature $\le$ 0\textsuperscript{o}C for $\ge1$ day for frost day spells) or that require a combination of thresholds to be met (e.g., maximum daily temperature above a lower threshold every day of the heat wave and above a higher threshold for a certain number of days within the heat wave; \citet{kent2014heat, peng2011toward}).
+To use a customized event definition, the user must write and load an R function that implements the definition.
+This custom function is passed to the \code{gen\_hw\_set} function using the \code{IDheatwavesFunction} argument.
+To work correctly, this custom function must allow only specific inputs and generate only specific outputs; details about the required structure are provided in the \pkg{futureheatwaves} package vignette.
 To increase processing speed when identifying extreme events, we coded parts of the default event definition function in C++ and synced it with R using the \CRANpkg{Rcpp} package \citep{Rcpp}.
@@
 
-Finally, extreme events are often based on conditions that are rare for a specific location using location-specific relative thresholds.
+## Exploring adaptation scenarios
+
+Extreme events are often based on conditions that are rare for a specific location using location-specific relative thresholds.
 These thresholds are often defined for climate impact studies based on a variable's distribution at that location in present-day or historical data.
 However, it can be interesting to explore trends in extreme events under climate change if extreme events are identified based on variable distributions during the projection period, or another future period.
-It is possible to specify the time period to use when determining a relative threshold for an event definition using the \code{thresholdBoundaries} argument in the function \code{gen\_hw\_set}.
+For some extreme events, impacts are associated with how rare the conditions during the event are compared to norms in the location \citep{anderson2009weather}.
+This relationship raises the question of whether extreme events should be defined using a percentile threshold based on present-day variable distributions or based on distributions in the time period being projected.
+The \pkg{futureheatwaves} package allows users to specify the time period to use when determining a relative threshold for an event definition using the \code{thresholdBoundaries} argument in the function \code{gen\_hw\_set}, allowing users to explore how sensitive projections of impacts are to this choice of the time period to use when determining relative variable measures, including thresholds used for percentile-based event definitions.
 
-## Customizing dates for projections and reference temperatures
-
-Depending on the climate model output data used and the research question, different projection periods may be of interest.
-For example, one researcher may wish to explore extreme events in the 2050--2075 period, while another may be interested in 2090--2100.
-The projection date range can be changed by specifying starting and ending years in the \code{projectionBoundaries} argument of the \code{gen\_hw\_set} function.
-
-Further, some of the event characteristics (e.g., \code{mean.temp.quantile}, Table \ref{tab:hwcharacteristics}) are calculated based on relative temperature, which is a measure of how the value of the variable of interest during an extreme event compares to the typical distribution of that variable at that location.
+Similarly, some of the event characteristics (e.g., \code{mean.temp.quantile}, Table \ref{tab:hwcharacteristics}) are also calculated by the package based on relative temperature, providing measures of how the value of the variable of interest during an extreme event compares to the typical distribution of that variable at that location (e.g., the "mean.var.quantile", "min.var.quantile", and "max.var.quantile" characteristics, Table \ref{tab:hwcharacteristics}) or how long conditions of a certain rarity persisted during the event (e.g., the "days.above.99th" and "days.above.99.5th" characteristics, Table \ref{tab:hwcharacteristics}).
 These characteristics are measured for each of the extreme events identified by the \code{gen\_hw\_set} function by taking the absolute value of the variable during the event (e.g., average temperature during the heat wave is 90\textsuperscript{o}F \ 32.2\textsuperscript{o}C) and comparing it to the location's typical variable distribution.
@@
 \end{center}
-\caption{Snapshot of an example interactive map created using the \code{map\_grid\_leaflet} showing the locations of study cities and their matching climate model grid points for the BCC climate model example data included with \pkg{futureheatwaves}.
+\caption{Snapshot of an interactive map created using the \code{map\_grid\_leaflet} showing the locations of study cities and their matching climate model grid points for the BCC climate model example data included with \pkg{futureheatwaves}.
 The lines on the map connect each climate model grid point to the study location(s) for which that grid point was used.
@@
 
+\noindent This interactive map can be panned and zoomed to explore the locations of climate model grid points used to represent each study location.
+This mapping function works for study locations worldwide.
+
 ## Extensions

### 2017-01-04 paper_drafts/RJournalRevisions/anderson-eason-barnes.tex
Update article draft

@@
 
-The CMIP5 climate model output data is distributed across data nodes at
+The CMIP5 climate model output data are distributed across data nodes at
 different climate modeling centers \citep{taylor2012overview}, but can
@@
 
-\begin{figure}
-\begin{center}
-\includegraphics[width = \textwidth]{worldmapexample}
-\end{center}
-\caption{Example of mapping near-surface air temperature data worldwide for a single day of climate model output data.
-This map uses data from the Geophysical Fluid Dynamics Laboratory's Earth System Model 2G, r1i1p1 ensemble member, on a single day in the summer of 2075.}
-\label{fig:worldmap}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width = 0.8\textwidth]{timeseriesexample}
-\end{center}
-\caption{Example of plotting a time series of temperature simulations between 2071 and 2075 from CMIP5 daily climate model output data for the model grid cell point closest to Beijing, China.
-This plot uses data from the Geophysical Fluid Dynamics Laboratory's Earth System Model 2G, r1i1p1 ensemble member.}
-\label{fig:timeseries}
-\end{figure}
-
 CMIP5 data are available as netCDF files, for which base R import
@@
 version 4 and netCDF's older version 3. While climate model output data
-for CMIP5 is required to conform with the earlier version (version 3)
+for CMIP5 are required to conform with the earlier version (version 3)
 \citep{taylor2010cmip5}, it is safer to write code using functions that
@@
 closed (e.g., with the \code{nc\_close} function from \pkg{ncdf4}).
+
+\begin{figure}
+\begin{center}
+\includegraphics[width = \textwidth]{worldmapexample}
+\end{center}
+\caption{Example of mapping near-surface air temperature data worldwide for a single day of climate model output data.
+This map uses data from the Geophysical Fluid Dynamics Laboratory's Earth System Model 2G, r1i1p1 ensemble member, on a single day in the summer of 2075.
+Full code for recreating the map is available in the "starting\_from\_netcdf" vignette of the \pkg{futureheatwaves} package.}
+\label{fig:worldmap}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width = 0.8\textwidth]{timeseriesexample}
+\end{center}
+\caption{Example of plotting a time series of temperature simulations between 2071 and 2075 from CMIP5 daily climate model output data for the model grid cell point closest to Beijing, China.
+This plot uses data from the Geophysical Fluid Dynamics Laboratory's Earth System Model 2G, r1i1p1 ensemble member.
+Full code for recreating the map is available in the "starting\_from\_netcdf" vignette of the \pkg{futureheatwaves} package.}
+\label{fig:timeseries}
+\end{figure}
 
@@
 CMIP5 files of any time step; this function identifies all CMIP5 files
-in a directory and creates a dataframe with information parsed from the
+in a directory and creates a data frame with information parsed from the
 file name. As a note, the \code{get.split.filename.cmip5} function in
@@
 \code{models2wux} function to read in climate model output netCDF files
-and convert them to ``WUX'' dataframes, which can be used by other
+and convert them to ``WUX'' data frames, which can be used by other
 functions in the package. While this function can input climate model
@@
 aggregates this data to a monthly or less frequent (e.g., seasonal)
-aggregation when creating the WUX dataframe. Therefore, while this
+aggregation when creating the WUX data frame. Therefore, while this
 package provides very useful functionality for working with averaged
@@
 
-When using this package, CMIP5 climate model output data requires some
+When using this package, CMIP5 climate model output data require some
 pre-processing. Data will need to be saved in a specific format, with
@@
 that users can try this package before applying it to their own
-directory of climate projection files. This example data comes from two
+directory of climate projection files. These example data come from two
 climate models that are a part of CMIP5: (1) the model of the Beijing
@@
 
-However, this is not the only accepted heat wave definition in the
-scientific literature. A variety of different heat wave definitions have
-been used to identify heat waves in a time series of temperature data
-\citep{smith2013heat}, and the choice of heat wave definitions can
-influence both projected heat wave trends \citep{smith2013heat} and
-estimates of health risks during events
-\citep{chen2015influence, kent2014heat}. Further, other types of extreme
-events will be defined differently than heat waves (for example, frost
-day spells may be defined as one or more days with temperature at or
-below 32\textsuperscript{o}F / 0\textsuperscript{o}C).
+\noindent However, this is not the only accepted heat wave definition in
+the scientific literature. A variety of different heat wave definitions
+have been used to identify heat waves in a time series of temperature
+data
+\citep{smith2013heat, kent2014heat, chen2015influence, anderson2009weather},
+and the choice of heat wave definitions can influence both projected
+heat wave trends \citep{smith2013heat} and estimates of health risks
+during events
+\citep{chen2015influence, kent2014heat, anderson2009weather}. Further,
+other types of extreme events will be defined differently than heat
+waves (for example, frost day spells may be defined as one or more days
+with temperature at or below 32\textsuperscript{o}F /
+0\textsuperscript{o}C).
 
@@
 function with either a different heat wave definition (see
-\citet{smith2013heat} and \citet{kent2014heat} for listings of
-definitions used in scientific studies) or with a definition appropriate
-for a different type of extreme event (e.g., one or more days at or
-below 32\textsuperscript{o}F / 0\textsuperscript{o}C for frost day
-spells). For heat wave identification, researchers might want to use a
-specific definition, for example, because it matches the definition used
-by local health officials to declare heat wave warnings or, in the case
-of health impact assessments, to match with a definition used in an
-epidemiological study. For other extreme events, a heat wave definition
-likely will not be applicable and so a customized definition is
-necessary.
+\citet{smith2013heat} and \citet{kent2014heat} for listings of some of
+the definitions used in scientific studies) or with a definition
+appropriate for a different type of extreme event (e.g., one or more
+days at or below 32\textsuperscript{o}F / 0\textsuperscript{o}C for
+frost day spells). For heat wave identification, researchers might want
+to use a specific definition, for example, because it matches the
+definition used by local health officials to declare heat wave warnings
+or, in the case of health impact assessments, to match with a definition
+used in an epidemiological study. For studies of other extreme events, a
+heat wave definition likely will not be applicable and so a customized
+definition is necessary.
 
@@
 customized in the \code{gen\_hw\_set} function call, without creating a
-new R function to use to identify heat waves. First, often users wish to
-define an extreme event based on conditions that are rare in the study
-location \citep{IPCCch1}. In this case, the percentile of the variable
-of interest required for an extreme event can be changed using the
-\code{probThreshold} option in \code{gen\_hw\_set}. For example, this
-option could be used to identify heat waves based on the
-99\textsuperscript{th} percentile, rather than the
-98\textsuperscript{th} percentile, of a city's year-round temperature.
-Second, the user can change the number of days used in the heat wave
-definition using the \code{numDays} argument in the \code{gen\_hw\_set}
-function (the default is \samp{numDays = 2}, corresponding to two or
-more days). Combined, these two customization choices allow the user to
-identify heat waves using many of the heat wave definitions used in
-previous climate and health research-- for example, 9 of the 16 heat
-wave definitions used in \citet{kent2014heat} (e.g., {[}some examples of
-the 9 that could be fit{]}) could be fit using different combinations of
-these two options for specifying threshold percentile and number of
-days. Third, some extreme events like cold waves and frost day spells
-are defined as a certain number of days below, rather than above, a
+new R function to use to identify heat waves. First, many extreme event
+definitions are based on conditions that are rare in the study location
+\citep{IPCCch1}, but may vary in how rare conditions must be for a
+period to qualify as an extreme event. For example, some of the
+different definitions used to identify heat waves vary only in the
+percentile temperature used for a threshold (e.g., one definition is
+\(\ge2\) days at or above the 98\textsuperscript{th} percentile
+temperature at a location while another is \(\ge2\) days at or above the
+99\textsuperscript{th} percentile temperature;
+\citet{kent2014heat, smith2013heat}). Therefore, the
+\pkg{futureheatwaves} package allows users to change the percentile of
+the variable of interest required for an extreme event using the
+\code{probThreshold} option in \code{gen\_hw\_set}. Other heat wave
+definitions vary only in the number of consecutive days that must be
+over the threshold for a period to quality as an extreme event (e.g.,
+one definition is \(\ge2\) days at or above the 98\textsuperscript{th}
+percentile temperature at a location while another is \(\ge4\) days at
+or above the 98\textsuperscript{th} percentile temperature;
+\citet{anderson2009weather}). Therefore, the package allows the user to
+change the number of days used in the heat wave definition using the
+\code{numDays} argument in the \code{gen\_hw\_set} function. Combined,
+these two customization choices allow the user to identify heat waves
+using many of the heat wave definitions used in previous climate and
+health research-- for example, 9 of the 16 heat wave definitions used in
+\citet{kent2014heat} could be fit using different combinations of these
+two options for specifying threshold percentile and number of days.
+Third, some extreme events like cold waves and frost day spells are
+defined as a certain number of days below, rather than above, a
 threshold. While the default is to identify events by searching for days
@@
 Beyond these simpler options, the customization of the event definition
-is even more extensive as one has the option of writing a custom R
-function. This functionality allows the user to use heat wave
-definitions that either require a number of days above an absolute
-threshold (e.g., maximum temperature of \(\ge\) 95\textsuperscript{o}F
-for \(\ge1\) day \citet{kent2014heat, tan2007heat}; minimum temperature
-\(\le\) 0\textsuperscript{o}C for \(\ge1\) day for frost day spells) or
-that require a combination of thresholds to be met (e.g., maximum daily
-temperature one threshold every day of the heat wave and above a higher
-threshold for a certain number of days;
+is even more extensive as one has the option of writing and using a
+custom R function to identify extreme events. This functionality allows
+the user to use definitions that either require a number of days above
+an absolute threshold (e.g., maximum temperature of \(\ge\)
+95\textsuperscript{o}F for \(\ge1\) day
+\citet{kent2014heat, tan2007heat}; minimum temperature \(\le\)
+0\textsuperscript{o}C for \(\ge1\) day for frost day spells) or that
+require a combination of thresholds to be met (e.g., maximum daily
+temperature above a lower threshold every day of the heat wave and above
+a higher threshold for a certain number of days within the heat wave;
 \citet{kent2014heat, peng2011toward}). To use a customized event
-definition, the user can write and load an R function that implements
-the definition. The custom function is passed to the \code{gen\_hw\_set}
-function using the \code{IDheatwavesFunction} argument. To work
-correctly, this custom function must allow only specific inputs and
-generate only specific outputs. Details about this required structure
-are provided in the \pkg{futureheatwaves} package vignette. To increase
-processing speed when identifying extreme events, we coded parts of the
-default event definition function in C++ and synced it with R using the
-\CRANpkg{Rcpp} package \citep{Rcpp}. Users should consider a similar
-strategy for custom heat wave definitions, especially if processing a
-large number of climate projection files.
+definition, the user must write and load an R function that implements
+the definition. This custom function is passed to the
+\code{gen\_hw\_set} function using the \code{IDheatwavesFunction}
+argument. To work correctly, this custom function must allow only
+specific inputs and generate only specific outputs; details about the
+required structure are provided in the \pkg{futureheatwaves} package
+vignette. To increase processing speed when identifying extreme events,
+we coded parts of the default event definition function in C++ and
+synced it with R using the \CRANpkg{Rcpp} package \citep{Rcpp}. Users
+should consider a similar strategy for custom heat wave definitions,
+especially if processing a large number of climate projection files.
 
-Finally, extreme events are often based on conditions that are rare for
-a specific location using location-specific relative thresholds. These
+\subsection{Exploring adaptation
+scenarios}\label{exploring-adaptation-scenarios}
+
+Extreme events are often based on conditions that are rare for a
+specific location using location-specific relative thresholds. These
 thresholds are often defined for climate impact studies based on a
@@
 under climate change if extreme events are identified based on variable
-distributions during the projection period, or another future period. It
-is possible to specify the time period to use when determining a
+distributions during the projection period, or another future period.
+For some extreme events, impacts are associated with how rare the
+conditions during the event are compared to norms in the location
+\citep{anderson2009weather}. This relationship raises the question of
+whether extreme events should be defined using a percentile threshold
+based on present-day variable distributions or based on distributions in
+the time period being projected. The \pkg{futureheatwaves} package
+allows users to specify the time period to use when determining a
 relative threshold for an event definition using the
-\code{thresholdBoundaries} argument in the function \code{gen\_hw\_set}.
+\code{thresholdBoundaries} argument in the function \code{gen\_hw\_set},
+allowing users to explore how sensitive projections of impacts are to
+this choice of the time period to use when determining relative variable
+measures, including thresholds used for percentile-based event
+definitions.
 
-\subsection{Customizing dates for projections and reference
-temperatures}\label{customizing-dates-for-projections-and-reference-temperatures}
-
-Depending on the climate model output data used and the research
-question, different projection periods may be of interest. For example,
-one researcher may wish to explore extreme events in the 2050--2075
-period, while another may be interested in 2090--2100. The projection
-date range can be changed by specifying starting and ending years in the
-\code{projectionBoundaries} argument of the \code{gen\_hw\_set}
-function.
-
-Further, some of the event characteristics (e.g.,
-\code{mean.temp.quantile}, Table \ref{tab:hwcharacteristics}) are
-calculated based on relative temperature, which is a measure of how the
-value of the variable of interest during an extreme event compares to
-the typical distribution of that variable at that location. These
+Similarly, some of the event characteristics (e.g.,
+\code{mean.temp.quantile}, Table \ref{tab:hwcharacteristics}) are also
+calculated by the package based on relative temperature, providing
+measures of how the value of the variable of interest during an extreme
+event compares to the typical distribution of that variable at that
+location (e.g., the ``mean.var.quantile'', ``min.var.quantile'', and
+``max.var.quantile'' characteristics, Table \ref{tab:hwcharacteristics})
+or how long conditions of a certain rarity persisted during the event
+(e.g., the ``days.above.99th'' and ``days.above.99.5th''
+characteristics, Table \ref{tab:hwcharacteristics}). These
 characteristics are measured for each of the extreme events identified
@@
 \end{Schunk}
+
+\noindent This interactive map can be panned and zoomed to explore the
+locations of climate model grid points used to represent each study
+location. This mapping function works for study locations worldwide.
 

### 2017-01-05 paper_drafts/RJournalRevisions/ToSubmit/anderson-eason-barnes.tex
Update paper draft

--- 2017-01-04 paper_drafts/RJournalRevisions/anderson-eason-barnes.tex
@@
 % !TeX root = RJwrapper.tex
-\title{Working with daily climate model output data in R and the
-\pkg{futureheatwaves} package}
+\title{Working with Daily Climate Model Output Data in R and the
+\pkg{futureheatwaves} Package}
 \author{by G. Brooke Anderson, Colin Eason, Elizabeth A. Barnes}
@@
 \abstract{%
-Research on climate change impacts can require extensive processing,
-especially when using ensemble techniques to incorporate output from
-different climate models and different simulations of each model. This
-processing can be particularly extensive when identifying and
-characterizing multi-day extreme events like heat waves and frost day
-spells, as these must be processed from model output with daily time
-steps. Further, climate model output is in a format and follows
-standards that may be unfamiliar to most R users. Here, we provide an
-overview of working with daily climate model output data in R. We then
-present the \pkg{futureheatwaves} package, which we developed to ease
-the process of identifying, characterizing, and exploring multi-day
-extreme events in climate model output.
+Research on climate change impacts can require extensive processing of
+climate model output, especially when using ensemble techniques to
+incorporate output from multiple climate models and multiple simulations
+of each model. This processing can be particularly extensive when
+identifying and characterizing multi-day extreme events like heat waves
+and frost day spells, as these must be processed from model output with
+daily time steps. Further, climate model output is in a format and
+follows standards that may be unfamiliar to most R users. Here, we
+provide an overview of working with daily climate model output data in
+R. We then present the \CRANpkg{futureheatwaves} package, which we
+developed to ease the process of identifying, characterizing, and
+exploring multi-day extreme events in climate model output. This package
+can input a directory of climate model output files, identify all
+extreme events using customizable event definitions, and summarize the
+output using user-specified functions.
 }
@@
 Intercomparison Project, phase 5 (CMIP5; \citet{taylor2012overview}).
-This project brought together dozens of major climate modeling groups
-around the world to simulate the same future radiative forcing
-scenarios, but with their own models. This created an ensemble of
-state-of-the-art climate model projections that allow researchers to
-study projections and their uncertainties. Most of these modeling groups
-additionally performed more than one simulation for each scenario and
-model (i.e.~multiple ensemble members), perturbing the initial
-conditions by a very tiny amount to quantify uncertainties due to
-internal climate variability.
+This project brought together major climate modeling groups around the
+world to simulate the same future radiative forcing scenarios, but with
+their own models. This created an ensemble of state-of-the-art climate
+model projections that allow researchers to study projections and their
+uncertainties. Most of these modeling groups additionally performed more
+than one simulation for each scenario and model (i.e.~multiple ensemble
+members), perturbing the initial conditions by a very tiny amount to
+quantify uncertainties due to internal climate variability.
 
@@
 
-After this overview, we present the \pkg{futureheatwaves} package, which
-we created to aid in identifying and characterizing any type of
+After this overview, we present the \CRANpkg{futureheatwaves} package,
+which we created to aid in identifying and characterizing any type of
 multi-day extreme event from daily climate model output (Table
@@
 of local temperature distribution for identifying heat waves)
-\citep{IPCCch1}. In this case, assessment is further complicated, as the
-event definition must be determined at each study location from climate
-model output. Finally, it is often of interest to create summaries of
+\citep{IPCCch1}. In this case, the event definition must be determined
+at each study location from climate model output before events can be
+identified. Finally, it is often of interest to create summaries of
 multiple characteristics of these extreme events. For example, one may
@@
 definition to identify events in the climate model output, as well as
-options to explore different scenarios of adaptation to heat.
+provides options to explore different scenarios of adaptation to heat.
 
@@
 6 & Allow users to easily explore the extreme events identified within all climate simulations.\\
-7 & Create output that is in a ``tidy" data format, allowing it to work well with \pkg{ggplot2} for visualization.\\
+7 & Create output that is in a ``tidy" data format, allowing it to work well with \CRANpkg{ggplot2} for visualization.\\
 \bottomrule
@@
 
-For climate impact studies, a main source for climate model output is
-the Coupled Model Intercomparison Project, which is currently in its
-fifth phase (CMIP5). Over 20 climate modeling groups created one or more
-climate models which, for this project, were run using standardized
-scenarios \citep{taylor2012overview}. The resulting output is uniform
-across modeling groups and has a consistent structure, which allows
-comparison of simulations from different models \citep{IPCCch9}. CMIP5
-climate model output is archived at a number of different time steps
-(e.g., daily, monthly, seasonal, yearly) \citep{taylor2010cmip5}, and
-some variables are reported at multiple levels in the ocean or
-atmosphere (e.g., ocean temperature). Here, we will focus on data with a
-daily time step for variables reported at a single level (e.g.,
-near-surface air temperature).
+For climate impact studies, a main source of data is the Coupled Model
+Intercomparison Project, which is currently in its fifth phase (CMIP5).
+Over 20 climate modeling groups created one or more climate models
+which, for this project, were run using standardized scenarios
+\citep{taylor2012overview}. The resulting output is uniform across
+modeling groups and has a consistent structure, which allows comparison
+of simulations from different models \citep{IPCCch9}. CMIP5 climate
+model output is archived at a number of different time steps (e.g.,
+daily, monthly, seasonal, yearly) \citep{taylor2010cmip5}, and some
+variables are reported at multiple levels in the ocean or atmosphere
+(e.g., ocean temperature is reported at different ocean depths). Here,
+we will focus on data with a daily time step for variables reported at a
+single level (e.g., near-surface air temperature).
 
-Each modeling group ran simulations under several experiments, with
-experiments varying in terms of radiative forcing by using different
-scenarios of time-varying model inputs (greenhouse gas emissions or
-concentrations, land use changes, etc.)
+Each modeling group ran simulations for CMIP5 under several experiments,
+with experiments varying in terms of radiative forcing scenarios through
+the use of different scenarios of time-varying model inputs (greenhouse
+gas emissions or concentrations, land use changes, etc.)
 \citep{taylor2012overview, IPCCch9}. Experiments include historical
@@
 longitude). Figure \ref{fig:netcdfexample} provides a sketch of the
-structure of netCDF files for single-level climate model output. Global
-climate models generate output at regularly-spaced time steps, typically
-at regularly-spaced grid points around the world. The latitude and
-longitude spacing of grid points vary by climate model, but are
-typically 1--2 degrees for atmospheric variables in CMIP5 models
-\citep{IPCCch9}. Each data point in the netCDF array gives the modeled
-value of the variable (e.g., surface temperature) for a single time
-point and location. For CMIP5 climate model output, the location units
-are in degrees east and degrees north for longitude and latitude,
+structure of netCDF files for single-level climate model output.
+
+Each data point in the netCDF array gives the modeled value of the
+variable (e.g., surface temperature) for a single time point and
+location. Global climate models generate output at regularly-spaced time
+steps, typically at regularly-spaced grid points around the world. The
+latitude and longitude spacing of grid points vary by climate model, but
+are typically 1--2 degrees for atmospheric variables in CMIP5 models
+\citep{IPCCch9}. For CMIP5 climate model output, the location units are
+in degrees east and degrees north for longitude and latitude,
 respectively. For daily output files, the time unit is in days since a
 specified origin date-time (e.g., days since 1850-01-01 00:00:00)
-\citep{taylor2010cmip5}. All CMIP5 output files are required to include
-certain metadata \citep{taylor2010cmip5}, including the experiment,
+\citep{taylor2010cmip5}.
+
+All CMIP5 output files are required to include certain metadata (or
+``attributes'') \citep{taylor2010cmip5}, including the experiment,
 forcing agents input to the model to create the simulation, time step,
@@
 modeling realm \citep{taylor2010cmip5}. The metadata also must include
-units for all of the coordinate variables (e.g., longitude, latitude,
+units for all of the dimension variables (e.g., longitude, latitude,
 time). The netCDF format allows you to access metadata and variables
@@
 \end{center}
-\caption{Example of structure of a NetCDF climate model output file for a variable reported at a single level, like near-surface air temperature.
+\caption{Example of structure of a netCDF climate model output file for a variable reported at a single level, like near-surface air temperature.
 Data are stored in a three-dimensional array, with measurements at each time step and climate grid location.
 These data are typically indexed in the netCDF file by longitude, latitude, and time, in that order.
-For example, if the near-surface air temperature is read into an R object called \code{tas}, you can access the value for the first day at the fourth longitude and third latitude with \code{tas[4, 3, 1]}.
+For example, if the near-surface air temperature in the example netCDF shown here is read into an R object called \code{tas}, you can access the value for the first day at the fourth longitude and third latitude with \code{tas[4, 3, 1]}.
 In addition to the output variable (temperature in this example), vectors with the ordered values of each dimension (longitude, latitude, and time) can also be read in from the netCDF file, as well as attribute data (e.g., units for variables, the calendar used for time).}
@@
 the newer netCDF version 4 released in 2008 and are no longer available
-through CRAN. More recent packages, including \pkg{ncdf4} \citep{ncdf4}
-and \pkg{RNetCDF} \citep{michna2013rnetcdf, RNetCDF}, work with both
-version 4 and netCDF's older version 3. While climate model output data
-for CMIP5 are required to conform with the earlier version (version 3)
+through CRAN. More recent packages, including \CRANpkg{ncdf4}
+\citep{CRANncdf4} and \pkg{CRANRNetCDF}
+\citep{michna2013rnetcdf, RNetCDF}, work with both version 4 and
+netCDF's older version 3. While climate model output data for CMIP5 are
+required to conform with the earlier version (version 3)
 \citep{taylor2010cmip5}, it is safer to write code using functions that
@@
 You can do a number of things with netCDF files in R using these
-functions. For example, \pkg{ncdf4}'s \code{nc\_open} function can be
+packages. For example, \pkg{ncdf4}'s \code{nc\_open} function can be
 used to open a connection to a netCDF file, and the object returned by
@@
 \code{ncvar\_get} with the \code{varid} parameter set to ``lat'',
-``lon'', or ``time''. The climate output variable (e.g., near-surface
-air temperature) can similarly be read in using \code{ncvar\_get}. In
-this case, the \code{varid} parameter should be set using the
-appropriate CMIP5 variable name (e.g., ``tas'' for near-surface air
-temperature); these variable names can be found in the CMIP Requested
-Output tables \citep{taylor2010cmip5}. If you only need data from a
-subset of the full file, you can use the dimensional time and location
-data to identify the location of the variable data you need in the
-netCDF array and use indexing to read only that data into memory,
-without needing to read in the full file (for example, with the
-\code{nc.get.var.subset.by.axes} function in \pkg{ncdf4.helpers}). Once
-the user is done reading in data from the file, the connection can be
-closed (e.g., with the \code{nc\_close} function from \pkg{ncdf4}).
+``lon'', or ``time''.
+
+The climate output variable (e.g., near-surface air temperature) can
+similarly be read in using \code{ncvar\_get}. In this case, the
+\code{varid} parameter should be set using the appropriate CMIP5
+variable name (e.g., ``tas'' for near-surface air temperature); these
+variable names can be found in the CMIP Requested Output tables
+\citep{taylor2010cmip5}. If only a subset of the full file is needed,
+the dimensional time and location data can be used to identify the
+location of the needed data in the netCDF array and this information can
+then be used to read a portion of data into memory (for example, with
+the \code{nc.get.var.subset.by.axes} function in
+\CRANpkg{ncdf4.helpers}). Once the user is done reading in data from the
+file, the connection to the netCDF should be closed (e.g., with the
+\code{nc\_close} function from \pkg{ncdf4}).
 
@@
 
-Two R packages provide help with non-Gregorian calendars: \pkg{PCICt}
-\citep{PCICt} and \pkg{ncdf4.helpers} \citep{ncdf4.helpers}. The
-\code{nc.get.time.series} function in \pkg{ncdf4.helpers} pulls and uses
-metadata on the calendar stored in the CMIP5 netCDF file's attributes to
-convert the ``time'' variable in the file to an object of the
-\code{PCICt} class. This class is defined in the \pkg{PCICt} package and
-provides Date-like functionality for 360- and 365-day calendars
-\citep{PCICt}. However, while these functions will help with handling
-most CMIP5 files, the CMIP5 standards allows use of other calendars
-which may not be succesfully handled by these functions, so it is
-important to assess whether the time variable range in the \code{PCICt}
-object correctly matches the expected date ranges for a file when
-processing CMIP5 data in R.
+Two R packages provide help with non-Gregorian calendars:
+\CRANpkg{PCICt} \citep{PCICt} and \CRANpkg{ncdf4.helpers}
+\citep{ncdf4.helpers}. The \code{nc.get.time.series} function in
+\pkg{ncdf4.helpers} pulls and uses metadata on the calendar stored in
+the CMIP5 netCDF file's attributes to convert the ``time'' variable in
+the file to an object of the \code{PCICt} class. This class is defined
+in the \pkg{PCICt} package and provides date-like functionality for 360-
+and 365-day calendars \citep{PCICt}. However, while these functions will
+help with handling most CMIP5 files, the CMIP5 standards allows use of
+other calendars which may not be succesfully handled by these functions,
+so it is important to assess whether the time variable range in the
+\code{PCICt} object correctly matches the expected date ranges for a
+file when processing CMIP5 data in R.
 
@@
 there are several R packages specifically for working with climate model
-output data, including \pkg{RCMIP5} \citep{RCMIP5} and \pkg{wux}
+output data, including \CRANpkg{RCMIP5} \citep{RCMIP5} and \CRANpkg{wux}
 \citep{wux}. However, these packages are more useful for working with
@@
 in a directory and creates a data frame with information parsed from the
-file name. As a note, the \code{get.split.filename.cmip5} function in
-the \pkg{ncdf4.helper} package similarly can be used to parse
-information contained in CMIP5 file names \citep{ncdf4.helpers}.
+file name. The \code{get.split.filename.cmip5} function in the
+\pkg{ncdf4.helper} package similarly can be used to parse information
+contained in CMIP5 file names \citep{ncdf4.helpers}.
 
 The \pkg{wux} package \citep{wux} includes functions that allow the user
-to download CMIP5 monthly-aggregated output directly from within R with
-the \code{CMIP5fromESGF} function. The package then uses the
-\code{models2wux} function to read in climate model output netCDF files
-and convert them to ``WUX'' data frames, which can be used by other
-functions in the package. While this function can input climate model
-output with daily time steps (the ``what.timesteps'' element of the
+to download CMIP5 output at a monthly time step directly from R with the
+\code{CMIP5fromESGF} function. The package then uses the
+\code{models2wux} function to read climate model output netCDF files and
+convert it to ``WUX'' data frames, which can be used by other functions
+in the package. While this function can input climate model output with
+daily time steps (the ``what.timesteps'' element of the
 \code{modelinput} list input must be set to ``daily''), the function
@@
 package provides very useful functionality for working with averaged
-output of daily climate model output data, it cannot easily be used to
-identify and characterize multi-day extreme events like heat waves.
+output of daily climate model output data or with output data at a
+larger time step, it cannot easily be used to identify and characterize
+multi-day extreme events like heat waves.
 
@@
 temperature simulations at a specific climate model grid point (Figure
-\ref{fig:timeseries}). The ``starting\_from\_netcdf'' vignette that
-comes with the ``futureheatwaves'' package provides all code required to
-create these figures, as well as more details and code examples on
-working with CMIP5 netCDF files in R.
+\ref{fig:timeseries}). The \pkg{futureheatwaves}' ``Starting from netCDF
+files'' vignette
+(\url{https://cran.r-project.org/web/packages/futureheatwaves/vignettes/starting_from_netcdf.html})
+provides all code required to create these figures, as well as more
+details and code examples on working with CMIP5 netCDF files in R.
 
@@
 
-Once the user creates these data frames of location-specific heat waves,
-the \code{apply\_all\_models} function can be used to apply custom
-functions across all the extreme event data frames generated by
-\code{gen\_hw\_set}. This functionality allows users to create summaries
-of extreme events across all climate models and ensemble members (Figure
-\ref{fig:overview}, right). The function can be used to generate summary
-statistics (e.g., determine average heat wave length or total frost
-days) or to apply more complex functions (e.g., apply epidemiologic
-effect estimates across the heat waves to generate health impact
-estimates).
+Once the user creates these data frames of location-specific extreme
+events, the \code{apply\_all\_models} function can be used to apply
+custom functions across all the extreme event data frames. This
+functionality allows users to create summaries of extreme events across
+all climate models and ensemble members (Figure \ref{fig:overview},
+right). The function can be used to generate summary statistics (e.g.,
+determine average heat wave length or total frost days) or to apply more
+complex functions (e.g., apply epidemiologic effect estimates across the
+heat waves to generate health impact estimates).
 
@@
 files stored in a specific directory structure. Full details of the
-required file and directory structure are provided in the package's
-``futureheatwaves'' vignette, while tips and an R script for conducting
-this processing starting from CMIP5 netCDF files are given in the
-``starting\_from\_netcdf'' vignette.
+required file and directory structure are provided in the package's main
+vignette
+(\url{https://cran.r-project.org/web/packages/futureheatwaves/vignettes/futureheatwaves.html}),
+while tips and an R script for conducting this processing starting from
+CMIP5 netCDF files are given in the ``Starting from netCDF files''
+vignette
+(\url{https://cran.r-project.org/web/packages/futureheatwaves/vignettes/starting_from_netcdf.html}).
 
-This package can be used for study locations worldwide. The
-``starting\_from\_netcdf''" package vignette provides an example of
-using this package to identify and explore future heat waves in several
-Chinese cities.
+This package can be used for study locations worldwide. The ``Starting
+from netCDF files'' package vignette provides an example of using this
+package to identify and explore future heat waves in several Chinese
+cities.
 
@@
 \pkg{futureheatwaves} package is installed and loaded, the user can find
-the local location of these files using R's \code{system.file} function.
+the location of these files on his or her computer using R's
+\code{system.file} function.
 
@@
 the processing is done using default values for the event definition,
-years for which to generate the heat wave data sets, etc. How and why to
-customize these choices are explained later in the text. Function
-arguments (e.g., \code{dataDirectories}, \code{tasFilenames}) are used
-to specify the format of the data and the directory structure.
+adaptation scenario, etc. How and why to customize these choices are
+explained later in the text. Function arguments (e.g.,
+\code{dataDirectories}, \code{tasFilenames}) are used to specify the
+format of the data and the directory structure.
 
@@
 follows certain standards in accepting input and returning output. Full
-details on these standards are given in the \pkg{futureheatwaves}
-package vignette.
+details on these standards are given in the main package vignette.
 
 As an example, if the user wanted to calculate the average temperature
-of the heat waves identified for each ensemble member, he or she could
-write a simple function:
+of the heat waves identified for each ensemble member in the output
+generated by the code above, he or she could write a simple function:
 
@@
 
-\noindent This function could then be applied across all extreme event
+\noindent This function can then be applied across all extreme event
 data sets output by \code{gen\_hw\_set} using the
@@
 
-This output is structured as ``tidy" data \citep{wickham2014tidy},
-allowing it to be used easily with the graphing package
-\CRANpkg{ggplot2} \citep{ggplot2}. This functionality can be used to
-create a number of other summaries of the identified extreme events. For
-example, it could be used to determine average heat wave length or
-estimate how much earlier in the year events are expected to start
-across an ensemble of climate model simulations. The functionality can
-also be used for more complex analysis of extreme event files. For
-example, it can be used to apply epidemiological models of heat wave to
-estimate excess heat-related mortality under different future scenarios;
-an example of this application is provided in the ``futureheatwaves''
-vignette that comes with the package.
+The same process can be used to create a number of other summaries of
+the identified extreme events. For example, it could be used to
+determine average length of extreme events or estimate how much earlier
+in the year events are expected to start across an ensemble of climate
+model simulations. The functionality can also be used for more complex
+analysis of extreme event files. For example, it can be used to apply
+epidemiological models of heat wave to estimate excess heat-related
+mortality under different future scenarios; an example of this
+application is provided in the main \pkg{futureheatwaves} vignette. The
+output from \code{apply\_all\_models} is structured as ``tidy" data
+\citep{wickham2014tidy}, allowing it to be used easily with the graphing
+package \CRANpkg{ggplot2} \citep{ggplot2} and other packages in the
+tidyverse.
 
@@
 frost day spells). For heat wave identification, researchers might want
-to use a specific definition, for example, because it matches the
+to use a different event definition because, for example, it matches the
 definition used by local health officials to declare heat wave warnings
@@
 definitions are based on conditions that are rare in the study location
-\citep{IPCCch1}, but may vary in how rare conditions must be for a
-period to qualify as an extreme event. For example, some of the
-different definitions used to identify heat waves vary only in the
-percentile temperature used for a threshold (e.g., one definition is
-\(\ge2\) days at or above the 98\textsuperscript{th} percentile
-temperature at a location while another is \(\ge2\) days at or above the
-99\textsuperscript{th} percentile temperature;
-\citet{kent2014heat, smith2013heat}). Therefore, the
-\pkg{futureheatwaves} package allows users to change the percentile of
-the variable of interest required for an extreme event using the
+\citep{IPCCch1}, but definitions may vary in how rare conditions must
+be. For example, some of the different definitions used to identify heat
+waves vary only in the percentile temperature used for a threshold
+(e.g., one definition is \(\ge2\) days at or above the
+98\textsuperscript{th} percentile temperature at a location while
+another is \(\ge2\) days at or above the 99\textsuperscript{th}
+percentile temperature; \citet{kent2014heat, smith2013heat}). Therefore,
+the \pkg{futureheatwaves} package allows users to change the percentile
+of the variable of interest required for an extreme event using the
 \code{probThreshold} option in \code{gen\_hw\_set}. Other heat wave
@@
 using many of the heat wave definitions used in previous climate and
-health research-- for example, 9 of the 16 heat wave definitions used in
+health research-- for example, 9 of 16 heat wave definitions outlined in
 \citet{kent2014heat} could be fit using different combinations of these
@@
 the user to use definitions that either require a number of days above
-an absolute threshold (e.g., maximum temperature of \(\ge\)
+or below an absolute threshold (e.g., maximum temperature of \(\ge\)
 95\textsuperscript{o}F for \(\ge1\) day
@@
 specific inputs and generate only specific outputs; details about the
-required structure are provided in the \pkg{futureheatwaves} package
-vignette. To increase processing speed when identifying extreme events,
-we coded parts of the default event definition function in C++ and
-synced it with R using the \CRANpkg{Rcpp} package \citep{Rcpp}. Users
-should consider a similar strategy for custom heat wave definitions,
-especially if processing a large number of climate projection files.
+required structure are provided in the main \pkg{futureheatwaves}
+package vignette. To increase processing speed when identifying extreme
+events, we coded parts of the default event definition function in C++
+and synced it with R using the \CRANpkg{Rcpp} package \citep{Rcpp}.
+Users should consider a similar strategy for custom heat wave
+definitions, especially if processing a large number of climate
+projection files.
 
-\subsection{Exploring adaptation
-scenarios}\label{exploring-adaptation-scenarios}
+\subsection{Exploring sensitivity of results to
+adaptation}\label{exploring-sensitivity-of-results-to-adaptation}
 
-Extreme events are often based on conditions that are rare for a
-specific location using location-specific relative thresholds. These
-thresholds are often defined for climate impact studies based on a
+Extreme events tend to be identified based on conditions that are rare
+for a specific location using location-specific relative thresholds.
+These thresholds are often defined for climate impact studies based on a
 variable's distribution at that location in present-day or historical
-data. However, it can be interesting to explore trends in extreme events
-under climate change if extreme events are identified based on variable
-distributions during the projection period, or another future period.
-For some extreme events, impacts are associated with how rare the
-conditions during the event are compared to norms in the location
-\citep{anderson2009weather}. This relationship raises the question of
-whether extreme events should be defined using a percentile threshold
-based on present-day variable distributions or based on distributions in
-the time period being projected. The \pkg{futureheatwaves} package
-allows users to specify the time period to use when determining a
-relative threshold for an event definition using the
-\code{thresholdBoundaries} argument in the function \code{gen\_hw\_set},
-allowing users to explore how sensitive projections of impacts are to
-this choice of the time period to use when determining relative variable
-measures, including thresholds used for percentile-based event
-definitions.
+data. However, for some extreme events, impacts are associated with how
+rare the conditions during the event are compared to current norms in
+the location \citep{anderson2009weather}, which suggests some capacity
+for adaptation to heat and raises the question of whether extreme events
+should be defined using a percentile threshold based on present-day
+variable distributions or based on distributions in the time period
+being projected. Therefore, it can be interesting to explore trends in
+extreme events under climate change if extreme events are identified
+based on variable distributions during the projection period or another
+future period. The \pkg{futureheatwaves} package allows users to specify
+the time period to use when determining a location-specific relative
+threshold for an event definition using the \code{thresholdBoundaries}
+argument in the function \code{gen\_hw\_set}. This feature allows users
+to explore how sensitive projections of impacts are to this choice of
+the time period to use when determining relative variable measures,
+including thresholds used for percentile-based event definitions.
 
@@
 
-It can be useful to explore the location of the climate model grid point
-used to pull climate model output for each study location with a given
-climate model. Therefore, the package has a function called
+Finally, it can be useful to explore the location of the climate model
+grid point used to pull climate model output for each study location
+with a given climate model. Therefore, the package has a function called
 \texttt{map\_grid\_leaflet} that plots the locations of grid points used
 for each location from each climate model. This function is built using
-the htmlWidget \pkg{leaflet} package \citep{leaflet}. The following code
-illustrates the use of this function with the example data to create
-Figure \ref{fig:gridmap}, which plots the grid points used in the
+the htmlWidget \CRANpkg{leaflet} package \citep{leaflet}. The following
+code illustrates the use of this function with the example data to
+create Figure \ref{fig:gridmap}, which plots the grid points used in the
 example data from the BCC climate model in the example data:
@@
 be interesting to identify extended periods of high exposures in
-projection time series, and this package could be applied to gridded air
-pollution model output to explore these exposures.
+projection time series. The \pkg{futureheatwaves} package is not
+exclusive to CMIP5 model output data, and so could be applied to gridded
+air pollution model output to explore these exposures.
 

### 2017-01-05 paper_drafts/RJournalRevisions/anderson-eason-barnes.Rmd
Update paper draft

@@
 title: >
-    Working with daily climate model output data in R and the \pkg{futureheatwaves} package
+    Working with Daily Climate Model Output Data in R and the \pkg{futureheatwaves} Package
 author:
@@
 abstract: >
-  Research on climate change impacts can require extensive processing, especially when using ensemble techniques to incorporate output from different climate models and different simulations of each model.
+  Research on climate change impacts can require extensive processing of climate model output, especially when using ensemble techniques to incorporate output from multiple climate models and multiple simulations of each model.
 This processing can be particularly extensive when identifying and characterizing multi-day extreme events like heat waves and frost day spells, as these must be processed from model output with daily time steps.
@@
 Here, we provide an overview of working with daily climate model output data in R.
-We then present the \pkg{futureheatwaves} package, which we developed to ease the process of identifying, characterizing, and exploring multi-day extreme events in climate model output.
+We then present the \CRANpkg{futureheatwaves} package, which we developed to ease the process of identifying, characterizing, and exploring multi-day extreme events in climate model output.
+This package can input a directory of climate model output files, identify all extreme events using customizable event definitions, and summarize the output using user-specified functions.
 preamble: >
@@
 A key source of data for ensemble techniques is the Coupled Model Intercomparison Project, phase 5 (CMIP5; \citet{taylor2012overview}).
-This project brought together dozens of major climate modeling groups around the world to simulate the same future radiative forcing scenarios, but with their own models.
+This project brought together major climate modeling groups around the world to simulate the same future radiative forcing scenarios, but with their own models.
 This created an ensemble of state-of-the-art climate model projections that allow researchers to study projections and their uncertainties.
@@
 
-After this overview, we present the \pkg{futureheatwaves} package, which we created to aid in identifying and characterizing any type of multi-day extreme event from daily climate model output (Table \ref{tab:goals}).
+After this overview, we present the \CRANpkg{futureheatwaves} package, which we created to aid in identifying and characterizing any type of multi-day extreme event from daily climate model output (Table \ref{tab:goals}).
 The impacts of multi-day extreme events must be assessed using output in daily time step, unlike other climate impacts that can be assessed using climate model output at monthly, seasonal, or yearly time steps.
 Further, extreme events are identified based on conditions that are rare for a certain location (e.g., 98th percentile of local temperature distribution for identifying heat waves) \citep{IPCCch1}.
-In this case, assessment is further complicated, as the event definition must be determined at each study location from climate model output.
+In this case, the event definition must be determined at each study location from climate model output before events can be identified.
 Finally, it is often of interest to create summaries of multiple characteristics of these extreme events.
@@
 These include: (1) uncertainty in the definition of a heat wave itself and (2) uncertainty in the ability of communities to adapt to changing temperatures (the adaptation scenario).
-This package therefore allows the user to create and use a custom extreme event definition to identify events in the climate model output, as well as options to explore different scenarios of adaptation to heat.
+This package therefore allows the user to create and use a custom extreme event definition to identify events in the climate model output, as well as provides options to explore different scenarios of adaptation to heat.
 
@@
 
-For climate impact studies, a main source for climate model output is the Coupled Model Intercomparison Project, which is currently in its fifth phase (CMIP5).
+For climate impact studies, a main source of data is the Coupled Model Intercomparison Project, which is currently in its fifth phase (CMIP5).
 Over 20 climate modeling groups created one or more climate models which, for this project, were run using standardized scenarios \citep{taylor2012overview}.
 The resulting output is uniform across modeling groups and has a consistent structure, which allows comparison of simulations from different models \citep{IPCCch9}.
-CMIP5 climate model output is archived at a number of different time steps (e.g., daily, monthly, seasonal, yearly) \citep{taylor2010cmip5}, and some variables are reported at multiple levels in the ocean or atmosphere (e.g., ocean temperature).
+CMIP5 climate model output is archived at a number of different time steps (e.g., daily, monthly, seasonal, yearly) \citep{taylor2010cmip5}, and some variables are reported at multiple levels in the ocean or atmosphere (e.g., ocean temperature is reported at different ocean depths).
 Here, we will focus on data with a daily time step for variables reported at a single level (e.g., near-surface air temperature).
 
-Each modeling group ran simulations under several experiments, with experiments varying in terms of radiative forcing by using different scenarios of time-varying model inputs (greenhouse gas emissions or concentrations, land use changes, etc.) \citep{taylor2012overview, IPCCch9}.
+Each modeling group ran simulations for CMIP5 under several experiments, with experiments varying in terms of radiative forcing scenarios through the use of different scenarios of time-varying model inputs (greenhouse gas emissions or concentrations, land use changes, etc.) \citep{taylor2012overview, IPCCch9}.
 Experiments include historical experiments (run using radiative forcing consistent with observed and reconstructed data for 1850--2005), pre-Industrial control experiments, and experiments of future scenarios of radiative forcing over the 21st century or longer (e.g., RCP4.5, RCP8.5) \citep{taylor2012overview}.
@@
 
-The CMIP5 climate model output data are distributed across data nodes at different climate modeling centers \citep{taylor2012overview}, but can be accessed centrally at the World Climate Research Programme CMIP5 data portal at https://pcmdi.llnl.gov/search/cmip5/.
+The CMIP5 climate model output data are distributed across data nodes at different climate modeling centers \citep{taylor2012overview}, but can be accessed centrally at the World Climate Research Programme CMIP5 data portal at \url{https://pcmdi.llnl.gov/search/cmip5/}.
 Users must register before downloading data, and some data are restricted to non-commercial use.
@@
 Appropriate \code{wget} scripts can be created either through the World Climate Research Programme CMIP5 data portal or through Earth System Grid Federation's Search RESTful API.
-Tips on efficiently searching and downloading the data, including through use of wget scripts and the search API, are available as user tutorials through the website of the University of Colorado Boulder's Earth System CoG (e.g., https://www.earthsystemcog.org/projects/cog/doc/wget for a tutorial on downloading files using wget).
+Tips on efficiently searching and downloading the data, including through use of wget scripts and the search API, are available as user tutorials through the website of the University of Colorado Boulder's Earth System CoG (e.g., \url{https://www.earthsystemcog.org/projects/cog/doc/wget} for a tutorial on downloading files using wget).
 
@@
 Figure \ref{fig:netcdfexample} provides a sketch of the structure of netCDF files for single-level climate model output.
+
+Each data point in the netCDF array gives the modeled value of the variable (e.g., surface temperature) for a single time point and location.
 Global climate models generate output at regularly-spaced time steps, typically at regularly-spaced grid points around the world.
 The latitude and longitude spacing of grid points vary by climate model, but are typically 1--2 degrees for atmospheric variables in CMIP5 models \citep{IPCCch9}.
-Each data point in the netCDF array gives the modeled value of the variable (e.g., surface temperature) for a single time point and location.
 For CMIP5 climate model output, the location units are in degrees east and degrees north for longitude and latitude, respectively.
 For daily output files, the time unit is in days since a specified origin date-time (e.g., days since 1850-01-01 00:00:00) \citep{taylor2010cmip5}.
-All CMIP5 output files are required to include certain metadata  \citep{taylor2010cmip5}, including the experiment, forcing agents input to the model to create the simulation, time step, institution and institutional contact information, climate model, and modeling realm \citep{taylor2010cmip5}.
-The metadata also must include units for all of the coordinate variables (e.g., longitude, latitude, time).
+
+All CMIP5 output files are required to include certain metadata (or "attributes")  \citep{taylor2010cmip5}, including the experiment, forcing agents input to the model to create the simulation, time step, institution and institutional contact information, climate model, and modeling realm \citep{taylor2010cmip5}.
+The metadata also must include units for all of the dimension variables (e.g., longitude, latitude, time).
 The netCDF format allows you to access metadata and variables describing the dimensions of the data without reading the full file into memory.
@@
 Older packages include \pkg{ncdf} and \pkg{ncvar}, but these do not work with the newer netCDF version 4 released in 2008 and are no longer available through CRAN.
-More recent packages, including \pkg{ncdf4} \citep{ncdf4} and \pkg{RNetCDF} \citep{michna2013rnetcdf, RNetCDF}, work with both version 4 and netCDF's older version 3.
+More recent packages, including \CRANpkg{ncdf4} \citep{CRANncdf4} and \pkg{CRANRNetCDF} \citep{michna2013rnetcdf, RNetCDF}, work with both version 4 and netCDF's older version 3.
 While climate model output data for CMIP5 are required to conform with the earlier version (version 3) \citep{taylor2010cmip5}, it is safer to write code using functions that can be used with either version, in case future phases of CMIP do not require files to conform with netCDF version 3.
 
-You can do a number of things with netCDF files in R using these functions.
+You can do a number of things with netCDF files in R using these packages.
 For example, \pkg{ncdf4}'s \code{nc\_open} function can be used to open a connection to a netCDF file, and the object returned by the function includes the file's attribute data.
@@
 For example, the variables defining the dimensions of the sketched netCDF file in Figure \ref{fig:netcdfexample} could be read into R with \code{ncvar\_get} with the \code{varid} parameter set to "lat", "lon", or "time".
+
 The climate output variable (e.g., near-surface air temperature) can similarly be read in using \code{ncvar\_get}.
 In this case, the \code{varid} parameter should be set using the appropriate CMIP5 variable name (e.g., "tas" for near-surface air temperature); these variable names can be found in the CMIP Requested Output tables \citep{taylor2010cmip5}.
-If you only need data from a subset of the full file, you can use the dimensional time and location data to identify the location of the variable data you need in the netCDF array and use indexing to read only that data into memory, without needing to read in the full file (for example, with the \code{nc.get.var.subset.by.axes} function in \pkg{ncdf4.helpers}).
-Once the user is done reading in data from the file, the connection can be closed (e.g., with the \code{nc\_close} function from \pkg{ncdf4}).
+If only a subset of the full file is needed, the dimensional time and location data can be used to identify the location of the needed data in the netCDF array and this information can then be used to read a portion of data into memory (for example, with the \code{nc.get.var.subset.by.axes} function in \CRANpkg{ncdf4.helpers}).
+Once the user is done reading in data from the file, the connection to the netCDF should be closed (e.g., with the \code{nc\_close} function from \pkg{ncdf4}).
 
@@
 
-Two R packages provide help with non-Gregorian calendars: \pkg{PCICt} \citep{PCICt} and \pkg{ncdf4.helpers} \citep{ncdf4.helpers}.
+Two R packages provide help with non-Gregorian calendars: \CRANpkg{PCICt} \citep{PCICt} and \CRANpkg{ncdf4.helpers} \citep{ncdf4.helpers}.
 The \code{nc.get.time.series} function in \pkg{ncdf4.helpers} pulls and uses metadata on the calendar stored in the CMIP5 netCDF file's attributes to convert the "time" variable in the file to an object of the \code{PCICt} class.
-This class is defined in the \pkg{PCICt} package and provides Date-like functionality for 360- and 365-day calendars \citep{PCICt}.
+This class is defined in the \pkg{PCICt} package and provides date-like functionality for 360- and 365-day calendars \citep{PCICt}.
 However, while these functions will help with handling most CMIP5 files, the CMIP5 standards allows use of other calendars which may not be succesfully handled by these functions, so it is important to assess whether the time variable range in the \code{PCICt} object correctly matches the expected date ranges for a file when processing CMIP5 data in R.
@@
 
-In addition to these general packages for working with netCDF files, there are several R packages specifically for working with climate model output data, including \pkg{RCMIP5} \citep{RCMIP5} and \pkg{wux} \citep{wux}.
+In addition to these general packages for working with netCDF files, there are several R packages specifically for working with climate model output data, including \CRANpkg{RCMIP5} \citep{RCMIP5} and \CRANpkg{wux} \citep{wux}.
 However, these packages are more useful for working with data output at time steps of a month or higher and have limited utility with the daily climate model output data required for studies of multi-day extreme events.
@@
 The package's \code{getFileInfo} function, however, will work with CMIP5 files of any time step; this function identifies all CMIP5 files in a directory and creates a data frame with information parsed from the file name.
-As a note, the \code{get.split.filename.cmip5} function in the \pkg{ncdf4.helper} package similarly can be used to parse information contained in CMIP5 file names \citep{ncdf4.helpers}.
+The \code{get.split.filename.cmip5} function in the \pkg{ncdf4.helper} package similarly can be used to parse information contained in CMIP5 file names \citep{ncdf4.helpers}.
 
-The \pkg{wux} package \citep{wux} includes functions that allow the user to download CMIP5 monthly-aggregated output directly from within R with the \code{CMIP5fromESGF} function.
-The package then uses the \code{models2wux} function to read in climate model output netCDF files and convert them to "WUX" data frames, which can be used by other functions in the package.
+The \pkg{wux} package \citep{wux} includes functions that allow the user to download CMIP5 output at a monthly time step directly from R with the \code{CMIP5fromESGF} function.
+The package then uses the \code{models2wux} function to read climate model output netCDF files and convert it to "WUX" data frames, which can be used by other functions in the package.
 While this function can input climate model output with daily time steps (the "what.timesteps" element of the \code{modelinput} list input must be set to "daily"), the function aggregates this data to a monthly or less frequent (e.g., seasonal) aggregation when creating the WUX data frame.
-Therefore, while this package provides very useful functionality for working with averaged output of daily climate model output data, it cannot easily be used to identify and characterize multi-day extreme events like heat waves.
+Therefore, while this package provides very useful functionality for working with averaged output of daily climate model output data or with output data at a larger time step, it cannot easily be used to identify and characterize multi-day extreme events like heat waves.
 
 The functions and packages described in this section can be used with CMIP5 netCDF files to do things in R like map near-surface air temperatures from a single climate model on a specific day (Figure \ref{fig:worldmap}) or pull a time series of daily near-surface air temperature simulations at a specific climate model grid point (Figure \ref{fig:timeseries}).
-The "starting_from_netcdf" vignette that comes with the "futureheatwaves" package provides all code required to create these figures, as well as more details and code examples on working with CMIP5 netCDF files in R.
+The \pkg{futureheatwaves}' "Starting from netCDF files" vignette (\url{https://cran.r-project.org/web/packages/futureheatwaves/vignettes/starting_from_netcdf.html}) provides all code required to create these figures, as well as more details and code examples on working with CMIP5 netCDF files in R.
 
@@
 
-Once the user creates these data frames of location-specific heat waves, the \code{apply\_all\_models} function can be used to apply custom functions across all the extreme event data frames generated by \code{gen\_hw\_set}.
+Once the user creates these data frames of location-specific extreme events, the \code{apply\_all\_models} function can be used to apply custom functions across all the extreme event data frames.
 This functionality allows users to create summaries of extreme events across all climate models and ensemble members (Figure \ref{fig:overview}, right).
@@
 Data will need to be saved in a specific format, with files stored in a specific directory structure.
-Full details of the required file and directory structure are provided in the package's "futureheatwaves" vignette, while tips and an R script for conducting this processing starting from CMIP5 netCDF files are given in the "starting_from_netcdf" vignette.
+Full details of the required file and directory structure are provided in the package's main vignette (\url{https://cran.r-project.org/web/packages/futureheatwaves/vignettes/futureheatwaves.html}), while tips and an R script for conducting this processing starting from CMIP5 netCDF files are given in the "Starting from netCDF files" vignette (\url{https://cran.r-project.org/web/packages/futureheatwaves/vignettes/starting_from_netcdf.html}).
 
 This package can be used for study locations worldwide.
-The "starting_from_netcdf"" package vignette provides an example of using this package to identify and explore future heat waves in several Chinese cities.
+The "Starting from netCDF files" package vignette provides an example of using this package to identify and explore future heat waves in several Chinese cities.
 
@@
 We include one ensemble member from BCC (r1i1p1) and two from CCSM (r1i1p1 and r2i1p1).
-Once the \pkg{futureheatwaves} package is installed and loaded, the user can find the local location of these files using R's \code{system.file} function.
+Once the \pkg{futureheatwaves} package is installed and loaded, the user can find the location of these files on his or her computer using R's \code{system.file} function.
 
@@
 ```{r eval = FALSE}
+library(futureheatwaves)
 projection_dir_location <- system.file("extdata/cmip5",
@@
 The \code{gen\_hw\_set} function processes the example input and creates a new directory, \file{example\_results}, with files of identified and characterized heat waves, in the user's current working directory.
-In this example code, the processing is done using  default values for the event definition, years for which to generate the heat wave data sets, etc.
+In this example code, the processing is done using  default values for the event definition, adaptation scenario, etc.
 How and why to customize these choices are explained later in the text.
@@
 The user can apply any R function that follows certain standards in accepting input and returning output.
-Full details on these standards are given in the \pkg{futureheatwaves} package vignette.
+Full details on these standards are given in the main package vignette.
 
-As an example, if the user wanted to calculate the average temperature of the heat waves identified for each ensemble member, he or she could write a simple function:
+As an example, if the user wanted to calculate the average temperature of the heat waves identified for each ensemble member in the output generated by the code above, he or she could write a simple function:
 
@@
 
-\noindent This function could then be applied across all extreme event data sets output by \code{gen\_hw\_set} using the \code{apply\_all\_models} function.
+\noindent This function can then be applied across all extreme event data sets output by \code{gen\_hw\_set} using the \code{apply\_all\_models} function.
 For example, to apply this function to all the example output results that come with the package, the user could run:
@@
 
-This output is structured as ``tidy" data \citep{wickham2014tidy}, allowing it to be used easily with the graphing package \CRANpkg{ggplot2} \citep{ggplot2}.
-This functionality can be used to create a number of other summaries of the identified extreme events.
-For example, it could be used to determine average heat wave length or estimate how much earlier in the year events are expected to start across an ensemble of climate model simulations.
+The same process can be used to create a number of other summaries of the identified extreme events.
+For example, it could be used to determine average length of extreme events or estimate how much earlier in the year events are expected to start across an ensemble of climate model simulations.
 The functionality can also be used for more complex analysis of extreme event files.
-For example, it can be used to apply epidemiological models of heat wave to estimate excess heat-related mortality under different future scenarios; an example of this application is provided in the "futureheatwaves" vignette that comes with the package.
+For example, it can be used to apply epidemiological models of heat wave to estimate excess heat-related mortality under different future scenarios; an example of this application is provided in the main \pkg{futureheatwaves} vignette.
+The output from \code{apply\_all\_models} is structured as ``tidy" data \citep{wickham2014tidy}, allowing it to be used easily with the graphing package \CRANpkg{ggplot2} \citep{ggplot2} and other packages in the tidyverse.
 
@@
 Users can write a custom R function with either a different heat wave definition (see \citet{smith2013heat} and \citet{kent2014heat} for listings of some of the definitions used in scientific studies) or with a definition appropriate for a different type of extreme event (e.g., one or more days at or below 32\textsuperscript{o}F / 0\textsuperscript{o}C for frost day spells).
-For heat wave identification, researchers might want to use a specific definition, for example, because it matches the definition used by local health officials to declare heat wave warnings or, in the case of health impact assessments, to match with a definition used in an epidemiological study.
+For heat wave identification, researchers might want to use a different event definition because, for example, it matches the definition used by local health officials to declare heat wave warnings or, in the case of health impact assessments, to match with a definition used in an epidemiological study.
 For studies of other extreme events, a heat wave definition likely will not be applicable and so a customized definition is necessary.
@@
 Three components of the extreme event definition can be easily customized in the \code{gen\_hw\_set} function call, without creating a new R function to use to identify heat waves.
-First, many extreme event definitions are based on conditions that are rare in the study location \citep{IPCCch1}, but may vary in how rare conditions must be for a period to qualify as an extreme event.
+First, many extreme event definitions are based on conditions that are rare in the study location \citep{IPCCch1}, but definitions may vary in how rare conditions must be.
 For example, some of the different definitions used to identify heat waves vary only in the percentile temperature used for a threshold (e.g., one definition is $\ge2$ days at or above the 98\textsuperscript{th} percentile temperature at a location while another is $\ge2$ days at or above the 99\textsuperscript{th} percentile temperature; \citet{kent2014heat, smith2013heat}).
@@
 Therefore, the package allows the user to change the number of days used in the heat wave definition using the \code{numDays} argument in the \code{gen\_hw\_set} function.
-Combined, these two customization choices allow the user to identify heat waves using many of the heat wave definitions used in previous climate and health research-- for example, 9 of the 16 heat wave definitions used in \citet{kent2014heat} could be fit using different combinations of these two options for specifying threshold percentile and number of days.
+Combined, these two customization choices allow the user to identify heat waves using many of the heat wave definitions used in previous climate and health research-- for example, 9 of 16 heat wave definitions outlined in \citet{kent2014heat} could be fit using different combinations of these two options for specifying threshold percentile and number of days.
 Third, some extreme events like cold waves and frost day spells are defined as a certain number of days below, rather than above, a threshold.
@@
 Beyond these simpler options, the customization of the event definition is even more extensive as one has the option of writing and using a custom R function to identify extreme events.
-This functionality allows the user to use definitions that either require a number of days above an absolute threshold (e.g., maximum temperature of $\ge$ 95\textsuperscript{o}F for $\ge1$ day \citet{kent2014heat, tan2007heat}; minimum temperature $\le$ 0\textsuperscript{o}C for $\ge1$ day for frost day spells) or that require a combination of thresholds to be met (e.g., maximum daily temperature above a lower threshold every day of the heat wave and above a higher threshold for a certain number of days within the heat wave; \citet{kent2014heat, peng2011toward}).
+This functionality allows the user to use definitions that either require a number of days above or below an absolute threshold (e.g., maximum temperature of $\ge$ 95\textsuperscript{o}F for $\ge1$ day \citet{kent2014heat, tan2007heat}; minimum temperature $\le$ 0\textsuperscript{o}C for $\ge1$ day for frost day spells) or that require a combination of thresholds to be met (e.g., maximum daily temperature above a lower threshold every day of the heat wave and above a higher threshold for a certain number of days within the heat wave; \citet{kent2014heat, peng2011toward}).
 To use a customized event definition, the user must write and load an R function that implements the definition.
 This custom function is passed to the \code{gen\_hw\_set} function using the \code{IDheatwavesFunction} argument.
-To work correctly, this custom function must allow only specific inputs and generate only specific outputs; details about the required structure are provided in the \pkg{futureheatwaves} package vignette.
+To work correctly, this custom function must allow only specific inputs and generate only specific outputs; details about the required structure are provided in the main \pkg{futureheatwaves} package vignette.
 To increase processing speed when identifying extreme events, we coded parts of the default event definition function in C++ and synced it with R using the \CRANpkg{Rcpp} package \citep{Rcpp}.
@@
 
-## Exploring adaptation scenarios
+## Exploring sensitivity of results to adaptation
 
-Extreme events are often based on conditions that are rare for a specific location using location-specific relative thresholds.
+Extreme events tend to be identified based on conditions that are rare for a specific location using location-specific relative thresholds.
 These thresholds are often defined for climate impact studies based on a variable's distribution at that location in present-day or historical data.
-However, it can be interesting to explore trends in extreme events under climate change if extreme events are identified based on variable distributions during the projection period, or another future period.
-For some extreme events, impacts are associated with how rare the conditions during the event are compared to norms in the location \citep{anderson2009weather}.
-This relationship raises the question of whether extreme events should be defined using a percentile threshold based on present-day variable distributions or based on distributions in the time period being projected.
-The \pkg{futureheatwaves} package allows users to specify the time period to use when determining a relative threshold for an event definition using the \code{thresholdBoundaries} argument in the function \code{gen\_hw\_set}, allowing users to explore how sensitive projections of impacts are to this choice of the time period to use when determining relative variable measures, including thresholds used for percentile-based event definitions.
+However, for some extreme events, impacts are associated with how rare the conditions during the event are compared to current norms in the location \citep{anderson2009weather}, which suggests some capacity for adaptation to heat and raises the question of whether extreme events should be defined using a percentile threshold based on present-day variable distributions or based on distributions in the time period being projected.
+Therefore, it can be interesting to explore trends in extreme events under climate change if extreme events are identified based on variable distributions during the projection period or another future period.
+The \pkg{futureheatwaves} package allows users to specify the time period to use when determining a location-specific relative threshold for an event definition using the \code{thresholdBoundaries} argument in the function \code{gen\_hw\_set}.
+This feature allows users to explore how sensitive projections of impacts are to this choice of the time period to use when determining relative variable measures, including thresholds used for percentile-based event definitions.
 
@@
 
-It can be useful to explore the location of the climate model grid point used to pull climate model output for each study location with a given climate model.
+Finally, it can be useful to explore the location of the climate model grid point used to pull climate model output for each study location with a given climate model.
 Therefore, the package has a function called `map_grid_leaflet` that plots the locations of grid points used for each location from each climate model.
-This function is built using the htmlWidget \pkg{leaflet} package \citep{leaflet}.
+This function is built using the htmlWidget \CRANpkg{leaflet} package \citep{leaflet}.
 The following code illustrates the use of this function with the example data to create Figure \ref{fig:gridmap}, which plots the grid points used in the example data from the BCC climate model in the example data:
@@
 While this package was created to be used for research on extreme events in climate change projections, it can be used more broadly.
-For example, there are other episodes like wildfires and air pollution where it may be interesting to identify extended periods of high exposures in projection time series, and this package could be applied to gridded air pollution model output to explore these exposures.
+For example, there are other episodes like wildfires and air pollution where it may be interesting to identify extended periods of high exposures in projection time series.
+The \pkg{futureheatwaves} package is not exclusive to CMIP5 model output data, and so could be applied to gridded air pollution model output to explore these exposures.
 

### 2017-01-09 README.Rmd
Add CRAN badge
Try to get badges on same line

@@
 [![Build Status](https://travis-ci.org/geanders/futureheatwaves.svg?branch=master)](https://travis-ci.org/geanders/futureheatwaves)
+[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/futureheatwaves)](https://cran.r-project.org/package=futureheatwaves)
 

### 2017-05-22 R/processHeatwaves.R
Update `createHwDataframe` to use `lubridate`

@@ -290,33 +290,30 @@ createHwDataframe <- function(city, threshold, heatwaves,
                 dplyr::summarize_(mean.var = ~ mean(tmpd),
                                  max.var = ~ max(tmpd),
                                  min.var = ~ min(tmpd),
-                                 length = ~ length(unique(date)),
-                                 end.date = ~ date[length(date)],
-                                 start.date.year = ~ ifelse(is.na(date[1]), # Feb. 30 problem in some climate data
-                                                            as.POSIXlt(date[length(date)])$year +
-                                                                    1900,
-                                                          NA),
+                                 length = ~ n(),
+                                 end.date = ~ last(date),
+                                 start.date.year = ~ ifelse(is.na(first(date)), # Feb. 30 problem in some climate data
+                                                            as.POSIXlt(last(date))$year + 1900,
+                                                            NA),
                                  start.date.2 = ~ as.Date(paste(start.date.year,
-                                                           "03", "01", sep = "-"),
+                                                                "03", "01", sep = "-"),
                                                           format = "%Y-%m-%d"),
-                                 start.date = ~ as.Date(ifelse(is.na(date[1]),
-                                                       start.date.2,
-                                                       date[1]),
-                                                       origin = c("1970-01-01")),
-                                 start.doy = ~ as.POSIXlt(as.Date(start.date,
-                                                          origin = "1970-01-01"))$yday,
-                                 start.month = ~ as.POSIXlt(as.Date(start.date,
-                                                            origin = "1970-01-01"))$mon + 1,
-                                 days.above.abs.thresh.1 = ~ length(date[tmpd > custom$absolute_thresholds[1]]),
-                                 days.above.abs.thresh.2 = ~ length(date[tmpd > custom$absolute_thresholds[2]]),
-                                 days.above.abs.thresh.3 = ~ length(date[tmpd > custom$absolute_thresholds[3]]),
-                                 days.above.abs.thresh.4 = ~ length(date[tmpd > custom$absolute_thresholds[4]]),
-                                 days.above.99th = ~ length(date[tmpd >
-                                                stats::quantile(ref_temps, .99,
-                                                         na.rm = TRUE)]),
-                                 days.above.99.5th = ~ length(date[tmpd >
-                                                stats::quantile(ref_temps, .995,
[…]

### 2017-05-22 src/RcppExports.cpp
Update build

@@ -124,3 +124,16 @@ RcppExport SEXP futureheatwaves_RcppExport_registerCCallable() {
     R_RegisterCCallable("futureheatwaves", "futureheatwaves_RcppExport_validate", (DL_FUNC)futureheatwaves_RcppExport_validate);
     return R_NilValue;
 }
+
+static const R_CallMethodDef CallEntries[] = {
+    {"futureheatwaves_IDHeatwavesCPP", (DL_FUNC) &futureheatwaves_IDHeatwavesCPP, 2},
+    {"futureheatwaves_storeHeatwaveEntry", (DL_FUNC) &futureheatwaves_storeHeatwaveEntry, 5},
+    {"futureheatwaves_storeZeroes", (DL_FUNC) &futureheatwaves_storeZeroes, 4},
+    {"futureheatwaves_RcppExport_registerCCallable", (DL_FUNC) &futureheatwaves_RcppExport_registerCCallable, 0},
+    {NULL, NULL, 0}
+};
+
+RcppExport void R_init_futureheatwaves(DllInfo *dll) {
+    R_registerRoutines(dll, NULL, CallEntries, NULL, NULL);
+    R_useDynamicSymbols(dll, FALSE);
+}

### 2017-05-23 paper_drafts/RJournalRevisions2/response_to_reviewers.txt
Work on revisions for R Journal paper

Thank you very much for the opportunity to revise and resubmit this manuscript. The
reviewers made some excellent suggestions, and we have modified the manuscript to
respond to these. In cases where suggested changes are beyond the scope of the described
package, we have added text to the manuscript to explain how such proposed approaches could
contribute as R programmers continue to evolve software solutions to working with climate model
output in R, and we have tried to make some connections to how R-based approaches to working with
large data continue to evolve in another topical areas with large, array-based data (flow
cytometry). In addition, we have revised and updated the package that is described by this
manuscript, both in response to reviewer suggestions and also to ensure functionality with recent
updates to `dplyr` and other key imported packages that have undergone major updates since the
original submission.

Comments from Reviewer #1

Comment: The authors answered quite satisfactory to the questions and also greatly
improved the manuscript and the package. I recommend that the article
should be published "as is" after solving a minor issue of the package: the
function 'map_grid_leaflet' does not work correctly because it requires a
local copy of the leaflet library.  Is it possible to put the http link of
leaflet instead of a local one so that function works for any connected
machine?

Response: The current version of the package imports the latest version of
`leaflet` from CRAN (version 1.1.0) and works with this version of leaflet,
rather than a local copy of the leaflet library. The function will now
work for any connected machine.

... https://github.com/rstudio/leaflet/issues/192#issuecomment-286883019
Also, "a network proxy which would block https requests" can cause problems?
"Do you have a network proxy or VPN active on your machine or your network?"
RStudio advice: https://support.rstudio.com/hc/en-us/articles/200488488-Configuring-R-to-Use-an-HTTP-or-HTTPS-Proxy

Comments from Reviewer #2

Comment: Climate science is fundamental to informing key societal decisions in the 21st
century. The Climate Intercomparison Project (currently in Phase 5) is a valuable
suite of Early system model outputs that continues to produce novel science. As
such, R packages to ingest and analyze these outputs could be quite valuable to the
community, particularly to field scientists and experimentalists who are not familiar
with the more idiosyncratic pipelines traditionally used by Earth system modelers. That
being said, this package does not address the most pressing barriers to data access and
interpretation, and I'm unclear that this manuscript describes a package that is
sufficiently novel to warrent publication.

Response: We appreciate the reviewer's point that the described package does not
resolve all the challenges of working with climate model output data in R.
However, we think that the package described offers an important step forward in making the
ensemble techniques that have become popular in climate science research more accessible to
researchers who are conducting climate impact studies but are not themselves climate scientists.
We have added a "Motivation" subsection to the manuscript section describing the `futureheatwaves`
package to clarify this point:

Added text: "We created the `futureheatwaves` package to aid in identifying, characterizing, and
exploring multi-day extreme events in daily climate model output data. While most of the discrete
tasks involved in identifying and characterizing multi-day extreme events are fairly
straightforward, the full process can be code-intensive, especially for multi-city studies and
studies that test sensitivity to how an event is defined or that incorporate different scenarios of
adaptation in the case of events defined using a threshold relative to community climate. Our aim
in developing this package was therefore to make the full process of identifying and characterizing
these extreme events much more convenient and so facilitate the use of multi-model, multi-ensemble
member analyses in climate impact studies conducted by non-climate scientists."

Further, we have also added a section to the end of the paper ("Future directions for working
with climate model output in R") that describes how future software development in R could
tackle the challenges provided by climate data, to help put the package described in the current
manuscript in context within the evolving use of R for research involving climate model output
data.

Comment: This article correctly identifies memory use and data format as a key issue for the
usability of CMIP5 data, neither of which are addressed by this package. The user is required
to use `ncdf4` to preprocess the CMIP5 outputs and script their own sub-setting routines. This
sidesteps both the memory issue and data formatting. The comments on calendars being non-standard
in the CMIP5 models is, again, useful and accurately identified but not addressed systematically
in the package and left as an issue for the user to solve, albeit, with guidance. If parts of this
process were hard coded (for example, Model A has a 365-day calendar starting from 1800 and model
B has a leap-year calendar starting from 1900) this could be very useful to the community.

Response: We appreciate the reviewer's point that having information at hand on which calendar
a climate models uses might be useful to those working with climate model output in R. Such a
table, however, would be fairly complex if it included both the calendar and the reference date
from which days are counted for all CMIP5 data. This is because, for some
of the CMIP5 climate models, different calendars are used for different
experiments, and difference reference dates are also used for different experiments within climate
models.

To provide some guidance to R users, however, we have added a link to the manuscript for a table
provided in the Frequently Asked Questions section of the Earth System CoG website. This table
gives the calendar (although not all starting reference dates) for each of the CMIP5 models,
including notes of exceptions for certain experiments within some models. Here is the sentence
we added to the manuscript with this link:

Added text: "While most CMIP5 climate models use the same calendar for all experiments, a few do not; a full table of the calendars used for each climate model and experiment, pulled from netCDF metadata, is available at https://www.earthsystemcog.org/projects/cog/faq_data."

In terms of the challenges presented by the large file sizes of climate
model output, we do recommend some pre-processing of data files to subset files down
to the geographic area and time range required for a study (e.g., North America for a national
study of the United States) which would help reduce data size prior to processing climate model
output through the described package. As the reviewer notes, we offer suggestions and guidance for
conducting this pre-processing in R, both through text in the manuscript. We offer additional
guidance through a vignette provided with the package described in the manuscript
(https://cran.r-project.org/web/packages/futureheatwaves/vignettes/starting_from_netcdf.html).
Even after pre-processing the files prior to using the `futureheatwaves` package, many research
applications will still involve working with very large files. For example, a national study of the
United States may still use files with daily temperatures for many decades at a large number of
grid points, and an analysis that uses multiple climate models, some with multiple ensemble
members, would still need to process a number of very large files. The `futureheatwaves` package
addresses these large file sizes by inputting and processing data from each ensemble member of each
climate model separately, avoiding the memory problems created by trying to load all data from all
ensemble members into memory at once and processing the data using `apply`, `map`, `plyr`, or
similar tools applied across all data at once.

We acknowledge, however, that while this approach is successful in terms of allowing the
identification and characterization of sustained extreme events in R, future software development
could continue refining and optimizing the handling of climate model data in R. We have added a
section to the end of the manuscript on future directions for climate model output data in R and
have included comments on how such future work might be able to successfully draw on CDF-backed
approaches recently created for handling flow cytometry data in R or on "data vault"-style
approaches geared to improve ...

[Added text]

Comment: Some acknowledgement of NCL or CDO for merging, sub-setting, averaging, and otherwise
manipulating netCDF files is warranted in this manuscript. These processes are MUCH faster in
these specialty languages and then done via R and not to mention this when dealing with CMIP5
data is remiss. One idea, although this completely changes the focus of the package, would be
to create a wrapper for these speciality languages in R to leverage that speed.

Response: Thank you for this suggestion. In the revised version of the manuscript, we have
added references to both software tools, as well as text describing the availability of
NCL and CDO for initial processing of climate model output files, including both the computation speed advantages and the disadvantage of needing to learn the syntax of a new language:

"CMIP5 data are available in the netCDF file format. Free specialty software exists to work with
climate model output files in this format, including a collection of command line tools developed
by the Max Planck Institute called Climate Model Operators (CDO) (Schulzweida, 2017) and an
interpreted language developed by the National Center for Atmospheric Research (NCAR) called the
NCAR Command Language (NCL) (UCAR/NCAR Computational and Information SystemsLaboratory, 2017).
Although such software can be used to quickly process netCDF climate model output files, they
require learning a new language syntax, and so for R users may not be worth the computational speed
gain compared to alternative solutions that can be scripted in the R language."

Further, we have added a section to the end of the paper, in response to this and other comments
from this reviewer, called "Future directions for working with climate model output in R". In this
section, we have added some text outlining the reviewer's idea that wrappers for these two
speciality languages could be an interesting next step to pursue as R software for working with
climate model output data evolves, including references to existing libraries with wrappers in
other languages (Python and Ruby):

Added text: "One important step in future development of R software to work with climate model output could be the development of R wrappers for some of the existing command line tools available
through the Climate Data Operators (CDO) software (Schulzweida, 2017). Libraries already exist for
Python and Ruby that allow the functionality of CDO tools to be used within these scripting
languages (available from the Max Planck Institute at
https://code.zmaw.de/projects/cdo/wiki/Cdo%7Brbpy%7D). While one R package (`ncdf4.helpers`;
Bronaugh, 2014) already provides R wrappers for a few CDO operators, such functionality could be
extended through future R software to capture more of the full functionality of the CDO toolkit."

Comment: The analysis that this package does is fairly straightforward using `plyr` or other
common R packages / functions. The main function described in this article is the identifying
and summing ranges of events. If a user made it far enough to do the pre-processing of the CMIP5
output data I doubt this would be a computational barrier to further analysis. If authors could
show notable speed or memory requirement improvement then this could be interesting.

Response:

Comment: Readability of the article is excellent.

Response: Thank you. We have tried to maintain the same level of readability through our revisions
in response to reviewers' comments.

### 2017-05-23 paper_drafts/RJournalRevisions2/anderson-eason-barnes.Rmd
Work on revisions for R Journal paper

--- 2017-01-05 paper_drafts/RJournalRevisions/anderson-eason-barnes.Rmd
@@
 
-CMIP5 data are available as netCDF files, for which base R import functions do not exist.
-There are, however, a few R packages that can be used to work with the netCDF file format used for CMIP5 files.
+CMIP5 data are available in the netCDF file format.
+Free specialty software exists to work with climate model output files in this format, including a collection of command line tools developed by the Max Planck Institute called Climate Model Operators (CDO) \citep{schulzweida2006cdo} and an interpreted language developed by the National Center for Atmospheric Research (NCAR) called the NCAR Command Language (NCL) \citep{ncl}.
+Although such software can be used to quickly process netCDF climate model output files, they require learning a new language syntax, and so for R users may not be worth the computational speed gain compared to alternative solutions that can be scripted in the R language.
+While base R import functions do not exist for netCDF files, there are a few R package extensions that allow R users to work with the netCDF file format used for CMIP5 files directly from R.
 Older packages include \pkg{ncdf} and \pkg{ncvar}, but these do not work with the newer netCDF version 4 released in 2008 and are no longer available through CRAN.
@@
 This class is defined in the \pkg{PCICt} package and provides date-like functionality for 360- and 365-day calendars \citep{PCICt}.
-However, while these functions will help with handling most CMIP5 files, the CMIP5 standards allows use of other calendars which may not be succesfully handled by these functions, so it is important to assess whether the time variable range in the \code{PCICt} object correctly matches the expected date ranges for a file when processing CMIP5 data in R.
+However, while these functions will help with handling most CMIP5 files, the CMIP5 standards allows use of other calendars which may not be successfully handled by these functions, so it is important to assess whether the time variable range in the \code{PCICt} object correctly matches the expected date ranges for a file when processing CMIP5 data in R.
+While most CMIP5 climate models use the same calendar for all experiments, a few do not; a full table of the calendars used for each climate model and experiment, pulled from netCDF metadata, is available at \url{https://www.earthsystemcog.org/projects/cog/faq_data}.
 
@@
 
-The \pkg{RCMIP5} package includes functions to read in CMIP5 data from netCDF files, scan a directory of CMIP5 files and determine models with continuous available data, create objects of a special \code{cmip5data} class to work with CMIP5 data within R, and parse the filenames for all files in a directory to extract information within the filename.
+The \pkg{RCMIP5} package includes functions to read in CMIP5 data from netCDF files, scan a directory of CMIP5 files and determine models with continuous available data, create objects of a special \code{cmip5data} class to work with CMIP5 data within R, and parse the file names for all files in a directory to extract information within the file name.
 For this package, most functions only work with monthly or less frequent data \citep{RCMIP5}.
@@
 
+## Motivation
+
+We created the \pkg{futureheatwaves} package to aid in identifying, characterizing, and exploring multi-day extreme events in daily climate model output data.
+While most of the discrete tasks involved in identifying and characterizing multi-day extreme events are fairly straightforward, the full process can be code-intensive, especially for multi-city studies and studies that test sensitivity to how an event is defined or that incorporate different scenarios of adaptation in the case of events defined using a threshold relative to community climate.
+Our aim in developing this package was therefore to make the full process of identifying and characterizing these extreme events much more convenient and so facilitate the use of multi-model, multi-ensemble member analyses in climate impact studies conducted by non-climate scientists.
+
 ## How the package works
@@
 
-We created the \pkg{futureheatwaves} package to aid in identifying, characterizing, and exploring multi-day extreme events in daily climate model output data.
 Figure \ref{fig:overview} gives an overview of the two primary functions of the \pkg{futureheatwaves} package.
@@
 The lines on the map connect each climate model grid point to the study location(s) for which that grid point was used.
-The interactive maps include pop-ups with city identiers; one is shown open in this snapshot as an example.
+The interactive maps include pop-ups with city identifiers; one is shown open in this snapshot as an example.
 From this map, you can see that the climate model grid point closest to New York City for this climate model is over the Atlantic Ocean.}
@@
 
+# Future directions for working with climate model output in R
+
+Research that assesses the potential impacts of climate change is critical in informing current policy choices, and R is an important tool for many researchers performing such assessments.
+While the `futureheatwaves` package described here takes steps to facilitate the assessment of impacts related to sustained, multi-day events, a number of challenges remain in working with climate model output data in R, and future R software development offers the potential to further address the challenges of working with this data.
+
+One important step in future development of R software to work with climate model output could be the development of R wrappers for some of the existing command line tools available through the Climate Data Operators (CDO) software \citep{schulzweida2006cdo}.
+Libraries already exist for Python and Ruby that allow the functionality of CDO tools to be used within these scripting languages (available from the Max Planck Institute at \url{https://code.zmaw.de/projects/cdo/wiki/Cdo%7Brbpy%7D}).
+While one R package (\CRANpkg{ncdf4.helpers}; \citet{ncdf.4helpers}) already provides R wrappers for a few CDO operators, such functionality could be extended through future R software to capture more of the full functionality of the CDO toolkit.
+
+Another important path for development could be through approaches that allow researchers to take advantage of the statistical tools offered by R while maintaining large climate model output files in a netCDF format.
+For example, Goncalves and coauthors recently described a ``round table" approach of connecting as-needed data access from netCDF climate data files through to functionality available in R and CDO through the intermediary of a MonetDB database system \citep{goncalves2015round}.
+In other topical areas, R programmers are also improving the efficiency of working with data in large netCDF files through approaches that avoid loading all data in-memory.
+For example, the Bioconductor package \pkg{ncdfFlow} enables R users to conduct analyses of hundreds of large flow cytometry output files through the creation and use of an \code{ncdfFlowSet} class that stores the data in an HDF5 format rather than in-memory \citep{}.
+Such an approach could be promising for future R software development for working with large climate model files.
+
 # Acknowledgements

### 2017-05-23 paper_drafts/RJournalRevisions2/anderson-eason-barnes.tex
Work on revisions for R Journal paper

--- 2017-01-05 paper_drafts/RJournalRevisions/ToSubmit/anderson-eason-barnes.tex
@@
 
-CMIP5 data are available as netCDF files, for which base R import
-functions do not exist. There are, however, a few R packages that can be
-used to work with the netCDF file format used for CMIP5 files. Older
-packages include \pkg{ncdf} and \pkg{ncvar}, but these do not work with
-the newer netCDF version 4 released in 2008 and are no longer available
-through CRAN. More recent packages, including \CRANpkg{ncdf4}
-\citep{CRANncdf4} and \pkg{CRANRNetCDF}
+CMIP5 data are available in the netCDF file format. Free specialty
+software exists to work with climate model output files in this format,
+including a collection of command line tools developed by the Max Planck
+Institute called Climate Model Operators (CDO)
+\citep{schulzweida2006cdo} and an interpreted language developed by the
+National Center for Atmospheric Research (NCAR) called the NCAR Command
+Language (NCL) \citep{ncl}. Although such software can be used to
+quickly process netCDF climate model output files, they require learning
+a new language syntax, and so for R users may not be worth the
+computational speed gain compared to alternative solutions that can be
+scripted in the R language. While base R import functions do not exist
+for netCDF files, there are a few R package extensions that allow R
+users to work with the netCDF file format used for CMIP5 files directly
+from R. Older packages include \pkg{ncdf} and \pkg{ncvar}, but these do
+not work with the newer netCDF version 4 released in 2008 and are no
+longer available through CRAN. More recent packages, including
+\CRANpkg{ncdf4} \citep{CRANncdf4} and \pkg{CRANRNetCDF}
 \citep{michna2013rnetcdf, RNetCDF}, work with both version 4 and
@@
 help with handling most CMIP5 files, the CMIP5 standards allows use of
-other calendars which may not be succesfully handled by these functions,
-so it is important to assess whether the time variable range in the
-\code{PCICt} object correctly matches the expected date ranges for a
-file when processing CMIP5 data in R.
+other calendars which may not be successfully handled by these
+functions, so it is important to assess whether the time variable range
+in the \code{PCICt} object correctly matches the expected date ranges
+for a file when processing CMIP5 data in R. While most CMIP5 climate
+models use the same calendar for all experiments, a few do not; a full
+table of the calendars used for each climate model and experiment,
+pulled from netCDF metadata, is available at
+\url{https://www.earthsystemcog.org/projects/cog/faq_data}.
 
@@
 continuous available data, create objects of a special \code{cmip5data}
-class to work with CMIP5 data within R, and parse the filenames for all
-files in a directory to extract information within the filename. For
+class to work with CMIP5 data within R, and parse the file names for all
+files in a directory to extract information within the file name. For
 this package, most functions only work with monthly or less frequent
@@
 
+\subsection{Motivation}\label{motivation}
+
+We created the \pkg{futureheatwaves} package to aid in identifying,
+characterizing, and exploring multi-day extreme events in daily climate
+model output data. While most of the discrete tasks involved in
+identifying and characterizing multi-day extreme events are fairly
+straightforward, the full process can be code-intensive, especially for
+multi-city studies and studies that test sensitivity to how an event is
+defined or that incorporate different scenarios of adaptation in the
+case of events defined using a threshold relative to community climate.
+Our aim in developing this package was therefore to make the full
+process of identifying and characterizing these extreme events much more
+convenient and so facilitate the use of multi-model, multi-ensemble
+member analyses in climate impact studies conducted by non-climate
+scientists.
+
 \subsection{How the package works}\label{how-the-package-works}
@@
 
-We created the \pkg{futureheatwaves} package to aid in identifying,
-characterizing, and exploring multi-day extreme events in daily climate
-model output data. Figure \ref{fig:overview} gives an overview of the
-two primary functions of the \pkg{futureheatwaves} package. First, the
-\code{gen\_hw\_set} function processes a directory of climate projection
-files that are stored locally on the user's computer (Figure
-\ref{fig:overview}, ``Climate projections''), to generate a list of all
-extreme events in each projection, as well as over a dozen
-characteristics of each identified extreme event (Table
-\ref{tab:hwcharacteristics}). This package start from files rather than
-R objects to avoid loading data from all climate model ensembles at
-once; instead, the function loads, processes, and saves output for a
-single climate model ensemble member at a time. The extreme events are
-identified and characterized at one or more study locations (e.g.,
-cities), which the user specifies in an input file. The extreme events
-identified for each ensemble member are output as separate files in a
-directory specified by the user (Figure \ref{fig:overview}, ``Extreme
-events datasets``).
+Figure \ref{fig:overview} gives an overview of the two primary functions
+of the \pkg{futureheatwaves} package. First, the \code{gen\_hw\_set}
+function processes a directory of climate projection files that are
+stored locally on the user's computer (Figure \ref{fig:overview},
+``Climate projections''), to generate a list of all extreme events in
+each projection, as well as over a dozen characteristics of each
+identified extreme event (Table \ref{tab:hwcharacteristics}). This
+package start from files rather than R objects to avoid loading data
+from all climate model ensembles at once; instead, the function loads,
+processes, and saves output for a single climate model ensemble member
+at a time. The extreme events are identified and characterized at one or
+more study locations (e.g., cities), which the user specifies in an
+input file. The extreme events identified for each ensemble member are
+output as separate files in a directory specified by the user (Figure
+\ref{fig:overview}, ``Extreme events datasets``).
 
@@
 \begin{Sinput}
+library(futureheatwaves)
 projection_dir_location <- system.file("extdata/cmip5",
@@
 
+\section{Future directions for working with climate model output in
+R}\label{future-directions-for-working-with-climate-model-output-in-r}
+
+Research that assesses the potential impacts of climate change is
+critical in informing current policy choices, and R is an important tool
+for many researchers performing such assessments. While the
+\texttt{futureheatwaves} package described here takes steps to
+facilitate the assessment of impacts related to sustained, multi-day
+events, a number of challenges remain in working with climate model
+output data in R, and future R software development offers the potential
+to further address the challenges of working with this data.
+
+One important step in future development of R software to work with
+climate model output could be the development of R wrappers for some of
+the existing command line tools available through the Climate Data
+Operators (CDO) software \citep{schulzweida2006cdo}. Libraries already
+exist for Python and Ruby that allow the functionality of CDO tools to
+be used within these scripting languages (available from the Max Planck
+Institute at
+\url{https://code.zmaw.de/projects/cdo/wiki/Cdo%7Brbpy%7D}). While one R
+package (\CRANpkg{ncdf4.helpers}; \citet{ncdf.4helpers}) already
+provides R wrappers for a few CDO operators, such functionality could be
+extended through future R software to capture more of the full
+functionality of the CDO toolkit.
+
+Another important path for development could be through approaches that
+allow researchers to take advantage of the statistical tools offered by
+R while maintaining large climate model output files in a netCDF format.
+For example, Goncalves and coauthors recently described a ``round table"
+approach of connecting as-needed data access from netCDF climate data
+files through to functionality available in R and CDO through the
+intermediary of a MonetDB database system \citep{goncalves2015round}.
+
+In other topical areas, R programmers are also improving the efficiency
+of working with data in large netCDF files through approaches that work
+with the data on-disc {[}?{]}. For example, the recent Bioconductor
+package \pkg{openCyto} {[}?{]} allows scientists to work with the large
+files generated by flow cytometry that characterize cell populations in
+a sample. Similar approaches have been taken in metabolomics (e.g.,
+{[}packages{]}) and other 'omics fields. The developers of future R
+packages to work with climate model output may find it useful to borrow
+ideas from these fields to create solutions with improved memory
+management {[}?{]} for large climate output files.
+
 \section{Acknowledgements}\label{acknowledgements}

### 2017-05-24 paper_drafts/RJournalRevisions2/response_to_reviewers.txt
Continue work on revisions to R Journal manuscript

@@
 work for any connected machine.
-
-... https://github.com/rstudio/leaflet/issues/192#issuecomment-286883019
-Also, "a network proxy which would block https requests" can cause problems?
-"Do you have a network proxy or VPN active on your machine or your network?"
-RStudio advice: https://support.rstudio.com/hc/en-us/articles/200488488-Configuring-R-to-Use-an-HTTP-or-HTTPS-Proxy
 
@@
 section to the end of the manuscript on future directions for climate model output data in R and
-have included comments on how such future work might be able to successfully draw on CDF-backed
-approaches recently created for handling flow cytometry data in R or on "data vault"-style
-approaches geared to improve ...
+have included comments on how such future work could continue to tackle the challenges introduced
+by the large file sizes of climate model output data:
 
-[Added text]
+Added text: "Another important path for development could be through approaches that allow
+researchers to take advantage of the statistical tools offered by R while maintaining large climate
+model output files in a netCDF format. For example, Goncalves and coauthors recently described a
+"round table" approach of connecting as-needed data access from netCDF climate data files through
+to functionality available in R and CDO through the intermediary of a MonetDB database system
+(Goncalves et al., 2015). In other topical areas, R programmers are also improving the efficiency
+of working with data in large netCDF files through approaches that avoid loading all data
+in-memory. For example, the Bioconductor package `ncdfFlow` enables R users to conduct analyses of
+hundreds of large flow cytometry output files through the creation and use of an `ncdfFlowSet`
+class that stores the data in an HDF5 format rather than in-memory (Jiang et al., 2017;
+Finak et al., 2014). Such an approach could be promising for future R software development for
+working with large climate model files."
 

### 2017-05-24 paper_drafts/RJournalRevisions2/anderson-eason-barnes.Rmd
Continue work on revisions to R Journal manuscript

@@
 Libraries already exist for Python and Ruby that allow the functionality of CDO tools to be used within these scripting languages (available from the Max Planck Institute at \url{https://code.zmaw.de/projects/cdo/wiki/Cdo%7Brbpy%7D}).
-While one R package (\CRANpkg{ncdf4.helpers}; \citet{ncdf.4helpers}) already provides R wrappers for a few CDO operators, such functionality could be extended through future R software to capture more of the full functionality of the CDO toolkit.
+While one R package (\CRANpkg{ncdf4.helpers}; \citet{ncdf4helpers}) already provides R wrappers for a few CDO operators, such functionality could be extended through future R software to capture more of the full functionality of the CDO toolkit.
 
@@
 In other topical areas, R programmers are also improving the efficiency of working with data in large netCDF files through approaches that avoid loading all data in-memory.
-For example, the Bioconductor package \pkg{ncdfFlow} enables R users to conduct analyses of hundreds of large flow cytometry output files through the creation and use of an \code{ncdfFlowSet} class that stores the data in an HDF5 format rather than in-memory \citep{}.
+For example, the Bioconductor package \pkg{ncdfFlow} enables R users to conduct analyses of hundreds of large flow cytometry output files through the creation and use of an \code{ncdfFlowSet} class that stores the data in an HDF5 format rather than in-memory \citep{ncdfFlow, finak2014opencyto}.
 Such an approach could be promising for future R software development for working with large climate model files.

### 2017-05-24 paper_drafts/RJournalRevisions2/anderson-eason-barnes.tex
Continue work on revisions to R Journal manuscript

@@
 \url{https://code.zmaw.de/projects/cdo/wiki/Cdo%7Brbpy%7D}). While one R
-package (\CRANpkg{ncdf4.helpers}; \citet{ncdf.4helpers}) already
-provides R wrappers for a few CDO operators, such functionality could be
-extended through future R software to capture more of the full
-functionality of the CDO toolkit.
+package (\CRANpkg{ncdf4.helpers}; \citet{ncdf4helpers}) already provides
+R wrappers for a few CDO operators, such functionality could be extended
+through future R software to capture more of the full functionality of
+the CDO toolkit.
 
@@
 files through to functionality available in R and CDO through the
-intermediary of a MonetDB database system \citep{goncalves2015round}.
-
-In other topical areas, R programmers are also improving the efficiency
-of working with data in large netCDF files through approaches that work
-with the data on-disc {[}?{]}. For example, the recent Bioconductor
-package \pkg{openCyto} {[}?{]} allows scientists to work with the large
-files generated by flow cytometry that characterize cell populations in
-a sample. Similar approaches have been taken in metabolomics (e.g.,
-{[}packages{]}) and other 'omics fields. The developers of future R
-packages to work with climate model output may find it useful to borrow
-ideas from these fields to create solutions with improved memory
-management {[}?{]} for large climate output files.
+intermediary of a MonetDB database system \citep{goncalves2015round}. In
+other topical areas, R programmers are also improving the efficiency of
+working with data in large netCDF files through approaches that avoid
+loading all data in-memory. For example, the Bioconductor package
+\pkg{ncdfFlow} enables R users to conduct analyses of hundreds of large
+flow cytometry output files through the creation and use of an
+\code{ncdfFlowSet} class that stores the data in an HDF5 format rather
+than in-memory \citep{ncdfFlow, finak2014opencyto}. Such an approach
+could be promising for future R software development for working with
+large climate model files.
 

### 2017-05-24 paper_drafts/RJournalRevisions2/response_to_reviewers.txt
Latest updates of revisions

@@
 
-Response:
+Response: As noted in a previous response, the advantage of this package is to improve
+convenience for researchers identifying and characterizing sustained, multi-day events in
+daily climate model output, especially for analyses that incorporate multiple climate
+models and ensemble members. We have added a paragraph to the manuscript making this
+motivation for the described package clearer:
+
+Added text: "We created the `futureheatwaves` package to aid in identifying,
+characterizing, and exploring multi-day extreme events in daily climate model output data.
+While most of the discrete tasks involved in identifying and characterizing multi-day extreme
+events are fairly straightforward, the full process can be code-intensive, especially for
+multi-city studies and studies that test sensitivity to how an event is defined or that incorporate
+different scenarios of adaptation in the case of events defined using a threshold relative to
+community climate. Our aim in developing this package was therefore to make the full process of
+identifying and characterizing these extreme events much more convenient and so facilitate the use
+of multi-model, multi-ensemble member analyses in climate impact studies conducted by non-climate
+scientists."
 

### 2017-06-01 paper_drafts/RJournalRevisions2/response_to_reviewers.txt
Create submission for R Journal

@@
 output in R, and we have tried to make some connections to how R-based approaches to working with
-large data continue to evolve in another topical areas with large, array-based data (flow
+large data continue to evolve in other topical areas with large, array-based data (e.g., flow
 cytometry). In addition, we have revised and updated the package that is described by this
@@
 updates to `dplyr` and other key imported packages that have undergone major updates since the
-original submission.
+original submission. This updated version of the package is available through GitHub
+(https://www.github.com/geanders/futureheatwaves), and we will submit this updated package
+version to CRAN as soon as we have satisfactorily responded to reviewers.
 
@@
 with climate model output in R") that describes how future software development in R could
-tackle the challenges provided by climate data, to help put the package described in the current
-manuscript in context within the evolving use of R for research involving climate model output
-data.
+continue to tackle the challenges provided by climate data, to help put the package described in
+the current manuscript in context within the evolving use of R for research involving climate
+model output data.
 

### 2017-06-01 paper_drafts/RJournalRevisions2/to_submit/AndersonRScript.R

@@ -1,0 +1,33 @@
+# Identify and characterize heat waves in the example data provided
+# with the `futureheatwaves` package
+library(futureheatwaves)
+projection_dir_location <- system.file("extdata/cmip5",
+                                       package = "futureheatwaves")
+city_file_location <- system.file("extdata/cities.csv",
+                                  package = "futureheatwaves")
+
+gen_hw_set(out = "example_results",
+           dataFolder = projection_dir_location ,
+           dataDirectories = list("historical" = c(1990, 1999),
+                                  "rcp85" = c(2060, 2079)),
+           citycsv = city_file_location,
+           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
+           tasFilenames = "tas_NorthAmerica_12mo.csv",
+           timeFilenames = "time_NorthAmerica_12mo.csv")
+
+# Define and apply a function to measure mean temperature during
+# heat waves in all the climate model ensemble members of the example
+# data.
+average_mean_temp <- function(hw_datafr){
+        out <- mean(hw_datafr$mean.var)
+        return(out)
+}
+
+out <- system.file("extdata/example_results", package = "futureheatwaves")
+
+apply_all_models(out = out, FUN = average_mean_temp)
+apply_all_models(out = out, FUN = average_mean_temp, city_specific = TRUE)
+
+# Create an interactive leaflet map with study locations and matching
+# climate model grid points for example data
+map_grid_leaflet(plot_model = "bcc1", out = out)
